### 2020-07-15 manuscript/v1_medrxiv/data_methods.Rmd
init git repository

# Data

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 23 February 2020 to 8 July 2020 ($T=137$), since we used a lag of 8 days for all confounders. We did not exclude any states or districts ($K=401$). We analyzed the daily reported number of new cases as outcome ($K\cdot T=54{\ }937$ observations). The set of possible predictors was derived from our causal DAG (see Table \@ref(tab:model-vars) and Figure \@ref(fig:dag-covid-19)). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal analysis, we computed adjustment sets in three different scenarios for separate exposures within the DAG: i) mobility of population, ii) awareness of COVID-19 (i.e. Google searches for "corona"), iii) weather (i.e. temperature).

## Variables

We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, [@casenumbers_rki], acquired on 12 July 2020) and aggregated them by date. The number of daily active cases for day $d$ was derived by subtracting the total number of reported cases on day $d$ and day $d-14$ (14 days as a conservative estimate for the infectious period, which corresponds here to the required quarantine time in Germany).

To assess the mobility of the German population, we used data publicly available on German state level from Google [@google_mobility]. Measurements are daily relative changes of mobility in percent compared to the period of 3 January 2020 to 6 February 2020. Missing values ($25$ out of $13{\ }488$) were imputed with value $0$ and the state level measurements were passed onto districts within the corresponding state. Google mobility data was available for six different sectors of daily life ("retail and recreation", "grocery and pharmacy", "parks", "transit stations", "workplaces", "residential") which means that "mobility" is a construct consisting of several variables. All variables but "residential" mobility are relative changes of daily visitor numbers to the corresponding sectors compared to the reference period. "Residential" mobility is the relative change of daily time spent at residential areas.

The notion of awareness in the population of COVID-19 describes the general state of alertness about the new infectious disease. As such, it was hard to measure directly. As a proxy, we used the relative interest in the topic term "corona" as indicated by Google searches. The daily data was available on state level [@google_trends] and passed onto district level. As a second proxy for awareness, we used the daily reported number of COVID-19 cases on the day of the exposure: Since media reported case numbers prominently, we assumed that this could reflect individual awareness, too.

We constructed daily weather from four variables ("temperature", "rainfall", "humidity", "wind"). Weather data was downloaded from Deutscher Wetterdienst (DWD, [@dwd_weather]) for all weather stations in Germany below 1000 meters altitude with daily records in 2020 until 8 July 2020. District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district). Missing values were imputed with mean values ($n=59$ for wind).

The reported number of COVID-19 cases varied strongly by day of the week. Thus, we included "weekday" as a categorical variable. Similarly, the reported cases and the exposure to the virus were affected by official holidays. Within the observation period, this included among others Good Friday, Easter Monday, and Labor Day. To correct for effects of these days, we included two variables in the model, "Holiday (report)" (indicates if the day of the report was a holiday, because governmental health departments were less likely to be on full duty) and "Holiday (exposure)" (indicates if the day of exposure to the virus was a holiday, because the population behaves differently on holidays).

For different official and political measures we used one-hot encoded daily variables, i.e. ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.

We included several social, economic, and demographic factors on the district level with direct or indirect influence on the risk of exposure to SARS-CoV-2 in our analysis. All are readily available from INKAR database [@inkar]. We used the share of population that is 65 years or older and the share of population that is younger than 18 years (Age), the share of females in population (Gender), the population density, the share of foreign citizenships and the share of the population seeking refuge (Foreign citizenship), the share of low-income households (Socio-economic status), voter turnout, share of right-wing populist party votes, and the number of nursing (retirement) homes.

<!-- next lines are for proper display of sources/references in table -->

(ref:googlemobility) Google [@google_mobility]

(ref:dwdweather) DWD [@dwd_weather]

(ref:postfacemasks) IZA [@mitze2020face]

(ref:inkar) INKAR [@inkar]

(ref:googletrends) Google [@google_trends]

(ref:rki) RKI [@casenumbers_rki]

```{r model-vars, echo=FALSE, message=FALSE, warning=FALSE}
library("dplyr")
library("readr")
library("knitr")
library("kableExtra")
my_variables <- read_delim("tables/t_variables.csv", delim=";") %>%
  filter(observed!="unobs") %>%
  dplyr::select(-contains("missing"), -contains("observ"))
options(knitr.kable.NA = '-')
kable(my_variables, caption = "Observed model variables", # landscape(
                booktabs = TRUE,  format = "latex") %>% # 
  kable_styling(font_size = 6, full_width = FALSE) %>%
  pack_rows("Mobility", 4, 9) %>%
  pack_rows("Weather", 10, 13) %>%
  pack_rows("Policies", 14, 17) %>%
  pack_rows("Socio-demographic", 18, 25) %>%
  pack_rows("Awareness", 26, 27) %>%
  pack_rows("Case numbers", 28, 29) %>%
  # pack_rows("Unobserved", 31, 36) %>%
  column_spec(5, width = "10em") # 
# , margin="1cm")
```

All variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered for numerical stability. We did not scale variables to unit variance to maintain interpretability of effects on the original scale of variables. Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 8 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 8 days.

# Methods

## Causal analysis with DAG and adjustment sets

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. Every node $v_i$ in the graph is the graphical representation of an observed or unobserved variable $x_i$, a directed edge $e_{ij}$ is an arrow from node $v_i$ to $v_j$ that implies a direct causal relationship from variable $x_i$ onto variable $x_j$. The set of all nodes is denoted by $V$, the set of all edges by $E$, as such, the complete DAG is the tuple $G=(V,E)$. The seminal works of Spirtes and Pearl [@spirtes2000causation;@pearl2009causality] introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable $x_i$ onto a variable $x_j$. The direct effect is the effect associated with the edge $e_{ij}$ only (if it exists), while the total effect takes indirect effects via other paths from $v_i$ to $v_j$ into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the *reported* number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect [@pearl2009causality]. The "minimal adjustment set" [@greenland1999causal] is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest asymptotic variance in the resulting causal effect estimates.

We analyzed the DAG from Section \@ref(causal-model) with the R Software [@rsoftware] and the R packages `dagitty` (formal representation of the graph and minimal adjustment sets [@textor_robust_2017]) and `pcalg` (for finding an optimal adjustment set [@pcalg]). For the defined exposures and the outcome "Reported new cases of COVID-19", we computed the minimal and optimal adjustment sets. Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the highest pseudo-$R^2$ (see next section) to estimate the final total causal effect from exposure to outcome.

## Regression with negative binomial model

We can estimate the causal effect from exposure to outcome by regression [@pearl2009causality]. Since the outcome "Reported new cases of COVID-19" is a count variable, one should not employ a linear regression model with Gaussian errors, but instead we assumed a log-linear relationship between the expected value of the outcome $Y$ (new cases) and regressors $x$, as well as a Poisson or negative binomial distribution for $Y$:
\begin{equation}
\log(\mathbb{E}[Y|x]) = \alpha + \sum_{i\in S}\beta_i\cdot x_{i}, (\#eq:regmodel)
\end{equation}
where $\alpha$ is the regression intercept, $S$ is the set of adjustment variables for the exposure $i^{\ast}$ including the exposure variable itself, $\beta_i$ are the regression coefficients corresponding to the variables $x_i$. As such $\beta_{i^{\ast}}$ is the total causal effect from exposure variable $x_{i^{\ast}}$ on the outcome Y.

The Poisson regression assumes equality of mean and variance. If this is not the case one observes so-called overdispersion (the variance is higher than the mean), this indicates one should use regression with a negative binomial distribution instead to estimate the variance parameter separately from the mean.

We needed to account for the fact that our outcome is not counted per time unit (one day) only, but depends on the number of active COVID-19 cases: Holding all other variables fixed, the number of new cases $Y$ is a constant proportion of the number of active cases $A$. This was modeled by including an offset $\log(A+1)$ in the regression model \@ref(eq:regmodel):
\begin{align}
\log(\mathbb{E}[Y|x]) &= \alpha + \log(A+1) + \sum_{i\in S}\beta_i\cdot x_{i} \nonumber \\
\Leftrightarrow \log(\frac{\mathbb{E}[Y|x]}{A+1}) &= \alpha + \sum\beta_i\cdot x_{i} (\#eq:logratemodel) \\
\Leftrightarrow \frac{\mathbb{E}[Y|x]}{A+1} &= \exp(\alpha)\cdot \prod \exp(\beta_i)^{x_i}. (\#eq:ratechange)
\end{align}
Here we added a pseudocount "$+1$" to ensure a finite logarithm and avoid division by $0$.

One can interpret the model as approximating the log-ratio of new cases and active cases by a linear combination of the regressor variables \@ref(eq:logratemodel). If all variables $x_i$ are centered in \@ref(eq:ratechange), we have for the baseline $\forall i\ x_i=0 \Rightarrow \mathbb{E}[Y|x=0]=\exp(\alpha)\cdot (A+1)$. In other words, the exponentiated intercept is the baseline daily infection rate (how many people does one infected individual infect in one day). If we hold all variables $x_i$ fixed (e.g. at baseline 0) in \@ref(eq:ratechange) but now increase the exposure variable $x_{i^{\ast}}=0$ by one unit to $x_{i^{\ast}}+1=0+1$, we have $\mathbb{E}[Y|x']=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}})$, which means the exponentiated coefficient $\beta_{i^{\ast}}$ describes the rate change of the outcome by one unit increase of the exposure.

In practice, given observations of $Y$ and $x$ we estimate the regression coefficients $\alpha$ and $\beta_i$  by *maximum likelihood* [@maxlikelihood]. Our observational measurements are $y_{kt}$ and $x_{ikt}$, where $k$ indicates the corresponding district and $t$ the date of measurement.

When we analyzed different adjustment sets given by analysis of the causal DAG (i.e. the minimal and optimal adjustment sets), we first checked if the set included unobserved variables. If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function `gac` in package `pcalg` [@perkovi2015complete]). If a minimal adjustment set contained unobserved variables, we discarded the whole set. We conducted a log-linear regression (function `glm` with `family=poisson()` for Poisson regression, and `glm.nb` from the `MASS` package for the negative binomial regression [@mass]) for every remaining valid adjustment set as regressors and calculated a Pseudo-$R^2$ given by $1-V_m/V_0$, where $V_m$ is the sum of squared prediction errors of the current model and $V_0$ is the sum of squared prediction errors of the null model (intercept and offset only). That is, our Pseudo-$R^2$ is $1$ minus the fraction of variance unexplained. Finally, we decided for the model/adjustment set with the highest pseudo-$R^2$. We report the exponentiated estimated coefficients along with 99 percent confidence intervals of the estimates.

### 2020-07-15 manuscript/v1_medrxiv/discussion.Rmd
init git repository

## Main findings

Our objective was to identify causal effects for COVID-19 cases. We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers). We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population. The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.

Furthermore, we made a strong case for the use of causal DAGs in epidemiology and a pandemic like COVID-19: DAGs allow to choose confounders for the analysis in a principled and statistically correct way while reducing possible causes for bias. Also, the DAG formalization allows for discussion about the underlying causal structure.

## Comparison with previous research

Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects [@fowler_effect_2020;@li_retrospective_2020;@shi_impact_2020;@Wang2020temperature]. To reliably identify causal drivers, one needs to adjust for confounders. To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph. A causal approach is used in another current COVID-19 analysis [@gencoglu2020causal]. There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.

Several studies assessing the impact of public health measures on mobility have each observed a downward trend accompanied by a decrease in the number of newly reported cases [@chang_modeling_2020;@cowling2020impact;@fowler_effect_2020;@kraemer_effect_2020;@lai_effect_2020;@linka_outbreak_2020].

Our findings regarding awareness/Google Trends analysis are in good agreement with the correlations found by @effenberger_association_2020, @higgins_correlations_2020, and @yuan_trends_2020, who conclude that alertness to COVID-19 rises several days before the highest number of cases are reported. At this point it should be noted, that awareness is substantially influenced by public media coverage, which should be considered, if possible, in future studies [@higgins_correlations_2020]. As such, awareness is difficult to measure and here the number of Google searches for "corona" could only be a proxy for this concept.

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases [@bannister-tyrrell_preliminary_2020;@demongeot_temperature_2020;@liu_impact_2020;@qi_covid-19_2020;@shi_impact_2020;@sobral_association_2020;@tosepu_correlation_2020;@Wang2020temperature;@wu_effects_2020]. It is however controversial to other scientific literature describing no effects [@briz-redon_spatio-temporal_2020;@iqbal_nexus_2020;@jahangiri_sensitivity_2020;@juni_impact_2020;@yao_no_2020] or even converse correlations [@auler_evidence_2020;@xie_association_2020]. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. In regards of humidity, our findings (higher humidity leading to increased case numbers) are in agreement with previous research [@qi_covid-19_2020;@wu_effects_2020;@juni_impact_2020]. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].

## Limitations and strengths

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.

As we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984]. The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions. Due to limited available data for the different variables, there is currently no way to overcome this.

Our observation period was restricted to succession from late winter to spring and summer (February to July). Nevertheless, this transition with increasing temperature was a natural experiment that allowed clues on weather effects.

We could not include data on health care utilization during the pandemic into our models due to the lack of available resources. This is planned for a later follow up to this paper since we rank health care utilization and mobility within health care facilities among the strong factors for COVID-19 progression: personnel in hospitals and private practices is particularly exposed to infection, while the lack of adequate care for other diseases has severe effects on general health of the population. At the same time, health care facilities are key for testing and surveillance of COVID-19 patients.

While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.

The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository^[The repository is located here: https://github.com/zidatalab/causalcovid19] of the article.

## Discussion of causal effects

In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy make a strong case for the successful implementation of policies like contact restrictions which limit the number of individual interactions that lead to infections. This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.

The causal effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret. We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.

Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections. In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

## Conclusions

To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available). Causal reasoning with a DAG allows us to estimate the causal effects reliably.

Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.

### 2020-07-15 manuscript/v1_medrxiv/manuscript.Rmd
init git repository

---
title: Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
author:
  - name: Edgar Steiger
    email: esteiger@zi.de
    affiliation: Zi
  - name: Tobias Mußgnug
    email: tmussgnug@zi.de
    affiliation: Zi
  - name: Lars Eric Kroll
    email: lkroll@zi.de
    affiliation: Zi
    corresponding: lkroll@zi.de
address:
  - code: Zi
    address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g. Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature and of increased public mobility for essential shopping (grocery and pharmacy) as well as awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

author_summary: |
  to be written

journal: "medRxiv"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
    base_format: rticles::elsevier_article 
    citation_package: natbib
---

# Introduction

As the COVID-19 pandemic progresses, research on mechanisms behind the transmission of SARS-CoV-2 shows conflicting evidence [@who2020report;@Chinazzi395;@guan2020clinical]. While effects of mobility have been extensively discussed, less is known on other factors such as changing awareness in the population  [@higgins_correlations_2020;@li_retrospective_2020;@yuan_trends_2020] or the effects of temperature [@bannister-tyrrell_preliminary_2020;@demongeot_temperature_2020;@liu_impact_2020]. A limiting factor in many studies is the lack of a causal approach to assess the causal contributions of various factors [@Greenland1999]. This can lead to distorted estimates of the causal factors with observational data [@Greenland1999;@schipf_directed_2011;@textor_robust_2017].

With COVID-19, we find ourselves in a situation in which information on the causal contribution of various influencing factors in the population is urgently needed to inform politicians and health authories. On the other hand, trials cannot be carried out for obvious ethical and legal reasons. Therefore, when assessing the effects of determinants of SARS-CoV-2 spread, special attention must be paid to strategies for the selection of confounding factors.

Another problem with assessing the effects of various determinants of SARS-CoV-2 spread is the heterogeneity of the countries and regions examined for example in the Johns Hopkins University (JHU) COVID-19 database [@jhucovid19db2020]. The comparison of time series of case numbers from different countries and observational periods can be strongly distorted by different factors like testing capacities and regional variations.

Our objective is to provide estimates of the causal effects of the main drivers of the pandemic with reduced bias. We conducted a scoping review of the available studies regarding signaling pathways and determinants of the spread of SARS-CoV-2 infections and the reported new COVID-19 cases. Then we integrated the current findings into a directed acyclic graph for the progress of the pandemic at the regional level. Using the resulting model and the do-calculus we found identifiable effects without blocked causal paths whose effects can be analyzed with observational data [@Pearl_2014]. We used regional time series data of all German districts ($401$) from various publicly available sources to analyze these questions on a regional level. Germany is a good choice in this regard, because it has ample data on contributing factors on the regional level and has had high testing and treatment capacities from early on in the pandemic.

```{r child = 'causal_model.Rmd'}
```

```{r child = 'data_methods.Rmd'}
```

```{r child = 'results.Rmd'}
```

# Discussion

```{r child = 'discussion.Rmd'}
```

# References

### 2020-07-15 manuscript/v1_medrxiv/manuscript.tex
init git repository

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{medRxiv} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header
\usepackage{geometry}
% \usepackage[numbers,sort&compress]{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Causal analysis of COVID-19 observational data in German districts
reveals effects of mobility, awareness, and temperature}
    \author[Zi]{Edgar Steiger}
   \ead{esteiger@zi.de} 
    \author[Zi]{Tobias Mußgnug}
   \ead{tmussgnug@zi.de} 
    \author[Zi]{Lars Eric Kroll}
   \ead{lkroll@zi.de} 
      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi),
Salzufer 8, D-10587 Berlin, Germany}

  \begin{abstract}
  Mobility, awareness, and weather are suspected to be causal drivers for
  new cases of COVID-19 infection. Correcting for possible confounders, we
  estimated their causal effects on reported case numbers. To this end, we
  used a directed acyclic graph (DAG) as a graphical representation of the
  hypothesized causal effects of the aforementioned determinants on new
  reported cases of COVID-19. Based on this, we computed valid adjustment
  sets of the possible confounding factors. We collected data for Germany
  from publicly available sources (e.g.~Robert Koch Institute, Germany's
  National Meteorological Service, Google) for 401 German districts over
  the period of 15 February to 8 July 2020, and estimated total causal
  effects based on our DAG analysis by negative binomial regression. Our
  analysis revealed favorable causal effects of increasing temperature and
  of increased public mobility for essential shopping (grocery and
  pharmacy) as well as awareness measured by COVID-19 burden, all of them
  reducing the outcome of newly reported COVID-19 cases. Conversely, we
  saw adverse effects of public mobility in retail and recreational areas,
  awareness measured by searches for ``corona'' in Google and higher
  rainfall, leading to an increase in new COVID-19 cases. This
  comprehensive causal analysis of a variety of determinants affecting
  COVID-19 progression gives strong evidence for the driving forces of
  mobility, public awareness, and temperature, whose implications need to
  be taken into account for future decisions regarding pandemic
  management.
  \end{abstract}

 \end{frontmatter}

\section{Introduction}\label{introduction}

As the COVID-19 pandemic progresses, research on mechanisms behind the
transmission of SARS-CoV-2 shows conflicting evidence
\citep{who2020report, Chinazzi395, guan2020clinical}. While effects of
mobility have been extensively discussed, less is known on other factors
such as changing awareness in the population
\citep{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
or the effects of temperature
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}.
A limiting factor in many studies is the lack of a causal approach to
assess the causal contributions of various factors
\citep{Greenland1999}. This can lead to distorted estimates of the
causal factors with observational data
\citep{Greenland1999, schipf_directed_2011, textor_robust_2017}.

With COVID-19, we find ourselves in a situation in which information on
the causal contribution of various influencing factors in the population
is urgently needed to inform politicians and health authories. On the
other hand, trials cannot be carried out for obvious ethical and legal
reasons. Therefore, when assessing the effects of determinants of
SARS-CoV-2 spread, special attention must be paid to strategies for the
selection of confounding factors.

Another problem with assessing the effects of various determinants of
SARS-CoV-2 spread is the heterogeneity of the countries and regions
examined for example in the Johns Hopkins University (JHU) COVID-19
database \citep{jhucovid19db2020}. The comparison of time series of case
numbers from different countries and observational periods can be
strongly distorted by different factors like testing capacities and
regional variations.

Our objective is to provide estimates of the causal effects of the main
drivers of the pandemic with reduced bias. We conducted a scoping review
of the available studies regarding signaling pathways and determinants
of the spread of SARS-CoV-2 infections and the reported new COVID-19
cases. Then we integrated the current findings into a directed acyclic
graph for the progress of the pandemic at the regional level. Using the
resulting model and the do-calculus we found identifiable effects
without blocked causal paths whose effects can be analyzed with
observational data \citep{Pearl_2014}. We used regional time series data
of all German districts (\(401\)) from various publicly available
sources to analyze these questions on a regional level. Germany is a
good choice in this regard, because it has ample data on contributing
factors on the regional level and has had high testing and treatment
capacities from early on in the pandemic.

\section{Causal Model}\label{causal-model}

We used a directed acyclic graph (DAG)
\citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze
the causal relationships between several exposures and SARS-CoV-2
spread. To get an overview on published associations, a scoping review
was conducted from 20th to 22nd of May 2020 within Pubmed and Google
scholar. Restrictions were applied to English and German language and
the publication date in the last one year. The following search terms
were applied to abstracts and title in Pubmed (``COVID-19'' OR
``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and
connected separately in each case with the exposure variables
(``mobility'', ``public awareness'', ``awareness'', ``google
trends'',``ambient temperature'', ``temperature''). For ``mobility'', we
analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together
with the first ten pages (100 results) in Google scholar
(``awareness''/``public awareness''/``google trends'' \(n=9\),
\(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
We integrated these findings where possible into the construction of our
DAG, which can be seen in Figure \ref{fig:dag-covid-19}.

\begin{figure}
\centering
\includegraphics{figures/f_full_covid19_dag.eps}
\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
cases on the district level. Unobserved variables are light gray,
variables marked with an asterisk (*) are confounded by
weekday/holiday.}
\end{figure}

A number of studies report a strong association of \textbf{mobility}
restrictions on the number of new COVID-19 cases: Restrictive measures
(e.g. ``stay-at-home'' orders, travel bans, or school closures) are
shown to possibly reduce the COVID-19 incidence
\citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
However, some studies point out the combination of various
non-pharmaceutical interventions (NPIs) is decisive to prevent new
infections \citep{juni_impact_2020, lai_effect_2020}.

Google Trends \citep{google_trends} data can be used as a tool to get
insights into public interest (\textbf{awareness}) in the coronavirus
disease. Several recent studies imply a connection of relative search
volumes (RSV) indices and reported new COVID-19 cases
\citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
infected cases/total number of cases by roughly 7 to 14 days for
different countries
\citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
Additionally, we acknowledged that individual risk-aware behavior might
be a reaction to the current COVID-19 burden (measured as reported cases
at the day of exposure).

Mixed evidence is available regarding the effect of
\textbf{temperature}: On the one hand several papers report an
association between increase in temperature and decrease in newly
infected COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
On the other hand, also the opposite has been found
\citep{auler_evidence_2020, xie_association_2020}. Some studies found no
association at all
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
It should be noted that few studies considered other confounding
variables than meteorological ones (especially age and population
density among others
\citep{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
In addition, the transferability of results between different climate
zones is questionable. To avoid possible bias caused by weather
variables other than temperature, we included rain, wind, and humidity
in our model.

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We include some unobserved variables in our DAG
(e.g. ``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG (mobility, awareness, temperature).

\section{Data}\label{data}

We collected and aggregated data on reported COVID-19 cases, regional
socio-demographic factors, weather, and general mobility on district and
state level in Germany for the period of 15 February 2020 to 8 July
2020. Our observation period for the outcome consisted of all dates from
23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
days for all confounders. We did not exclude any states or districts
(\(K=401\)). We analyzed the daily reported number of new cases as
outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
predictors was derived from our causal DAG (see Table
\ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to
modelling and data limitations, some of the predictors were unobserved
or were modelled as a construct consisting of several variables. For our
causal analysis, we computed adjustment sets in three different
scenarios for separate exposures within the DAG: i) mobility of
population, ii) awareness of COVID-19 (i.e.~Google searches for
``corona''), iii) weather (i.e.~temperature).

\subsection{Variables}\label{variables}

We downloaded German daily case numbers on district level reported by
Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July
2020) and aggregated them by date. The number of daily active cases for
day \(d\) was derived by subtracting the total number of reported cases
on day \(d\) and day \(d-14\) (14 days as a conservative estimate for
the infectious period, which corresponds here to the required quarantine
time in Germany).

To assess the mobility of the German population, we used data publicly
available on German state level from Google \citep{google_mobility}.
Measurements are daily relative changes of mobility in percent compared
to the period of 3 January 2020 to 6 February 2020. Missing values
(\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the
state level measurements were passed onto districts within the
corresponding state. Google mobility data was available for six
different sectors of daily life (``retail and recreation'', ``grocery
and pharmacy'', ``parks'', ``transit stations'', ``workplaces'',
``residential'') which means that ``mobility'' is a construct consisting
of several variables. All variables but ``residential'' mobility are
relative changes of daily visitor numbers to the corresponding sectors
compared to the reference period. ``Residential'' mobility is the
relative change of daily time spent at residential areas.

The notion of awareness in the population of COVID-19 describes the
general state of alertness about the new infectious disease. As such, it
was hard to measure directly. As a proxy, we used the relative interest
in the topic term ``corona'' as indicated by Google searches. The daily
data was available on state level \citep{google_trends} and passed onto
district level. As a second proxy for awareness, we used the daily
reported number of COVID-19 cases on the day of the exposure: Since
media reported case numbers prominently, we assumed that this could
reflect individual awareness, too.

We constructed daily weather from four variables (``temperature'',
``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
stations in Germany below 1000 meters altitude with daily records in
2020 until 8 July 2020. District level daily weather data was aggregated
per district by averaging the data from the three nearest weather
stations (which includes weather stations inside the district). Missing
values were imputed with mean values (\(n=59\) for wind).

The reported number of COVID-19 cases varied strongly by day of the
week. Thus, we included ``weekday'' as a categorical variable.
Similarly, the reported cases and the exposure to the virus were
affected by official holidays. Within the observation period, this
included among others Good Friday, Easter Monday, and Labor Day. To
correct for effects of these days, we included two variables in the
model, ``Holiday (report)'' (indicates if the day of the report was a
holiday, because governmental health departments were less likely to be
on full duty) and ``Holiday (exposure)'' (indicates if the day of
exposure to the virus was a holiday, because the population behaves
differently on holidays).

For different official and political measures we used one-hot encoded
daily variables, i.e.~ban of mass gatherings, school and kindergarten
closures and their gradual reopening, contact restrictions, and
mandatory face masks for shopping and public transport.

We included several social, economic, and demographic factors on the
district level with direct or indirect influence on the risk of exposure
to SARS-CoV-2 in our analysis. All are readily available from INKAR
database \citep{inkar}. We used the share of population that is 65 years
or older and the share of population that is younger than 18 years
(Age), the share of females in population (Gender), the population
density, the share of foreign citizenships and the share of the
population seeking refuge (Foreign citizenship), the share of low-income
households (Socio-economic status), voter turnout, share of right-wing
populist party votes, and the number of nursing (retirement) homes.

\begin{table}

\caption{\label{tab:model-vars}Observed model variables}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{llll>{\raggedright\arraybackslash}p{10em}l}
\toprule
variable & dynamics & level & type & unit/comment & source\\
\midrule
Weekday & daily & national & categorical & Sat through Thu as six binary variables, Fri as baseline & -\\
Holiday (report) & daily & national & binary & - & -\\
Holiday (exposure) & daily & national & binary & - & -\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}grocery and pharmacy & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}parks & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}workplaces & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}residential & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}transit stations & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & daily & district & numeric & mm (l/sqm) & DWD \citep{dwd_weather}\\
\hspace{1em}Temperature & daily & district & numeric & °C & DWD \citep{dwd_weather}\\
\hspace{1em}Humidity & daily & district & numeric & relative humidity (\%) & DWD \citep{dwd_weather}\\
\hspace{1em}Wind & daily & district & numeric & m/s & DWD \citep{dwd_weather}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & daily & national & binary & - & -\\
\hspace{1em}School and kindergarten closures & daily & state & numeric & 0 for no closure, 1 for full closure, 0.5 for partial reopening & -\\
\hspace{1em}Contact restrictions & daily & national & binary & - & -\\
\hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \citep{inkar}\\
\hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \citep{inkar}\\
\hspace{1em}Population density & constant & district & numeric & population per sqkm & INKAR \citep{inkar}\\
\hspace{1em}Foreign citizens & constant & district & numeric & 2 variables: share of foreign citizens \& of population seeking refuge & INKAR \citep{inkar}\\
\hspace{1em}Socio-economic status & constant & district & numeric & share of households with low income & INKAR \citep{inkar}\\
\hspace{1em}Turnout & constant & district & numeric & voter turnout in last election & INKAR \citep{inkar}\\
\hspace{1em}Right-wing populist party votes & constant & district & numeric & share of votes for AfD in last election & INKAR \citep{inkar}\\
\hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \citep{inkar}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Case numbers}}\\
\hspace{1em}Reported new cases of COVID-19 & daily & district & numeric & - & RKI \citep{casenumbers_rki}\\
\hspace{1em}Active cases & daily & district & numeric & active cases on day of report & RKI \citep{casenumbers_rki}\\
\bottomrule
\end{tabular}
\end{table}

All variables but the outcome ``Reported new cases of COVID-19'' and the
offset ``Active cases'' were centered for numerical stability. We did
not scale variables to unit variance to maintain interpretability of
effects on the original scale of variables. Additionally, we lagged the
effect of all variables (but outcome, offset, and the non-dynamic
socio-demographic variables) by 8 days (see Section \ref{results}) which
means that we assumed that their effects on the outcome will be visible
after 8 days.

\section{Methods}\label{methods}

\subsection{Causal analysis with DAG and adjustment
sets}\label{causal-analysis-with-dag-and-adjustment-sets}

We used a directed acyclic graph as a graphical representation of the
hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
virus, onset of COVID-19, and finally reports of COVID-19 cases. Every
node \(v_i\) in the graph is the graphical representation of an observed
or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
from node \(v_i\) to \(v_j\) that implies a direct causal relationship
from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
denoted by \(V\), the set of all edges by \(E\), as such, the complete
DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
\citep{spirtes2000causation, pearl2009causality} introduce the theory of
causal analysis, do-calculus, and how to analyze a DAG to estimate the
total or direct causal effect from a variable \(x_i\) onto a variable
\(x_j\). The direct effect is the effect associated with the edge
\(e_{ij}\) only (if it exists), while the total effect takes indirect
effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
we estimated total effects only, since most of our variables were not
hypothesized to have a direct effect on the \emph{reported} number of
new COVID-19 cases. In contrast to prediction tasks, where one would
include all variables available, it is actually ill-advised to use all
available variables to estimate causal effects, due to introducing bias
by adjusting for unnecessary variables within the causal DAG. This is
why we need to identify a valid set of necessary variables (an
adjustment set) to estimate the proper causal effect
\citep{pearl2009causality}. The ``minimal adjustment set''
\citep{greenland1999causal} is a valid adjustment set of variables that
does not contain another valid adjustment set as a subset. However,
identifying a minimal adjustment set might not be enough to reliably
estimate the causal effect. Thus, we identified the ``optimal adjustment
set'' \citep{henckel2019graphical} as the set of variables which is a
valid adjustment set while having the lowest asymptotic variance in the
resulting causal effect estimates.

We analyzed the DAG from Section \ref{causal-model} with the R Software
\citep{rsoftware} and the R packages \texttt{dagitty} (formal
representation of the graph and minimal adjustment sets
\citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
adjustment set \citep{pcalg}). For the defined exposures and the outcome
``Reported new cases of COVID-19'', we computed the minimal and optimal
adjustment sets. Since it was possible that these sets contained
unobserved variables that needed to be left out of the regression model,
we chose the valid set with the highest pseudo-\(R^2\) (see next
section) to estimate the final total causal effect from exposure to
outcome.

\subsection{Regression with negative binomial
model}\label{regression-with-negative-binomial-model}

We can estimate the causal effect from exposure to outcome by regression
\citep{pearl2009causality}. Since the outcome ``Reported new cases of
COVID-19'' is a count variable, one should not employ a linear
regression model with Gaussian errors, but instead we assumed a
log-linear relationship between the expected value of the outcome \(Y\)
(new cases) and regressors \(x\), as well as a Poisson or negative
binomial distribution for \(Y\):

\begin{equation}
\log(\mathbb{E}[Y|x]) = \alpha + \sum_{i\in S}\beta_i\cdot x_{i}, \label{eq:regmodel}
\end{equation}

where \(\alpha\) is the regression intercept, \(S\) is the set of
adjustment variables for the exposure \(i^{\ast}\) including the
exposure variable itself, \(\beta_i\) are the regression coefficients
corresponding to the variables \(x_i\). As such \(\beta_{i^{\ast}}\) is
the total causal effect from exposure variable \(x_{i^{\ast}}\) on the
outcome Y.

The Poisson regression assumes equality of mean and variance. If this is
not the case one observes so-called overdispersion (the variance is
higher than the mean), this indicates one should use regression with a
negative binomial distribution instead to estimate the variance
parameter separately from the mean.

We needed to account for the fact that our outcome is not counted per
time unit (one day) only, but depends on the number of active COVID-19
cases: Holding all other variables fixed, the number of new cases \(Y\)
is a constant proportion of the number of active cases \(A\). This was
modeled by including an offset \(\log(A+1)\) in the regression model
\eqref{eq:regmodel}:

\begin{align}
\log(\mathbb{E}[Y|x]) &= \alpha + \log(A+1) + \sum_{i\in S}\beta_i\cdot x_{i} \nonumber \\
\Leftrightarrow \log(\frac{\mathbb{E}[Y|x]}{A+1}) &= \alpha + \sum\beta_i\cdot x_{i} \label{eq:logratemodel} \\
\Leftrightarrow \frac{\mathbb{E}[Y|x]}{A+1} &= \exp(\alpha)\cdot \prod \exp(\beta_i)^{x_i}. \label{eq:ratechange}
\end{align}

Here we added a pseudocount ``\(+1\)'' to ensure a finite logarithm and
avoid division by \(0\).

One can interpret the model as approximating the log-ratio of new cases
and active cases by a linear combination of the regressor variables
\eqref{eq:logratemodel}. If all variables \(x_i\) are centered in
\eqref{eq:ratechange}, we have for the baseline
\(\forall i\ x_i=0 \Rightarrow \mathbb{E}[Y|x=0]=\exp(\alpha)\cdot (A+1)\).
In other words, the exponentiated intercept is the baseline daily
infection rate (how many people does one infected individual infect in
one day). If we hold all variables \(x_i\) fixed (e.g.~at baseline 0) in
\eqref{eq:ratechange} but now increase the exposure variable
\(x_{i^{\ast}}=0\) by one unit to \(x_{i^{\ast}}+1=0+1\), we have
\(\mathbb{E}[Y|x']=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}})\),
which means the exponentiated coefficient \(\beta_{i^{\ast}}\) describes
the rate change of the outcome by one unit increase of the exposure.

In practice, given observations of \(Y\) and \(x\) we estimate the
regression coefficients \(\alpha\) and \(\beta_i\) by \emph{maximum
likelihood} \citep{maxlikelihood}. Our observational measurements are
\(y_{kt}\) and \(x_{ikt}\), where \(k\) indicates the corresponding
district and \(t\) the date of measurement.

When we analyzed different adjustment sets given by analysis of the
causal DAG (i.e.~the minimal and optimal adjustment sets), we first
checked if the set included unobserved variables. If this was the case
for the optimal adjustment set, we discarded the unobserved variables
from the set and checked if it was still a valid adjustment set
(function \texttt{gac} in package \texttt{pcalg}
\citep{perkovi2015complete}). If a minimal adjustment set contained
unobserved variables, we discarded the whole set. We conducted a
log-linear regression (function \texttt{glm} with
\texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb}
from the \texttt{MASS} package for the negative binomial regression
\citep{mass}) for every remaining valid adjustment set as regressors and
calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the
sum of squared prediction errors of the current model and \(V_0\) is the
sum of squared prediction errors of the null model (intercept and offset
only). That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of
variance unexplained. Finally, we decided for the model/adjustment set
with the highest pseudo-\(R^2\). We report the exponentiated estimated
coefficients along with 99 percent confidence intervals of the
estimates.

\section{Results}\label{results}

Descriptive statistics for the included variables are presented in Table
\ref{tab:descriptive-vars}.

\begin{table}

\caption{\label{tab:descriptive-vars}Descriptive Statistics for observed variables}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{ll}
\toprule
Variable & mean (SD)\\
\midrule
n & 54937\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & -27.14 (24.59)\\
\hspace{1em}grocery and pharmacy & -4.04 (23.00)\\
\hspace{1em}parks & 46.33 (57.94)\\
\hspace{1em}workplaces & -22.98 (20.56)\\
\hspace{1em}residential & 8.20 (6.54)\\
\hspace{1em}transit stations & -29.83 (21.22)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & 1.85 (3.98)\\
\hspace{1em}Temperature & 10.74 (5.27)\\
\hspace{1em}Humidity & 67.68 (13.10)\\
\hspace{1em}Wind & 3.64 (1.67)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & 0.82 (0.38)\\
\hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
\hspace{1em}Contact restrictions & 0.74 (0.44)\\
\hspace{1em}Mandatory face masks & 0.48 (0.50)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age (pop. 65 and older) & 22.09 (2.74)\\
\hspace{1em}Age (pop. younger 18) & 16.17 (1.25)\\
\hspace{1em}Gender & 0.51 (0.01)\\
\hspace{1em}Population density & 533.75 (701.84)\\
\hspace{1em}Foreign citizens & 10.03 (5.14)\\
\hspace{1em}Foreign citizens (refugees) & 1.88 (1.14)\\
\hspace{1em}Socio-economic status & 30.64 (6.02)\\
\hspace{1em}Turnout & 75.08 (3.79)\\
\hspace{1em}Right-wing populist party votes & 13.39 (5.32)\\
\hspace{1em}Nursing homes & 36.11 (30.69)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 27.24 (18.31)\\
\hspace{1em}COVID-19 burden & 3.55 (10.37)\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
\hspace{1em}Reported new cases COVID-19 & 3.60 (10.39)\\
\hspace{1em}Active cases & 49.83 (121.96)\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{figures/f_timeseries_cases_and_lograte.eps}
\caption{\label{fig:cases-log}Temporal and district level variation of
outcome (log-scale)}
\end{figure}

In the observational period, the number of daily reported COVID-19 cases
increased till the end of March/beginning of April and continually
decreased afterwards till the beginning of June 2020 with a slight
increase and decrease afterwards (Figure \ref{fig:cases-log}A). On the
other hand, the (log-)ratio of reported cases over active cases
decreased steeply till the mid of April and increased steadily
afterwards with a slight decrease close to the end of the observation
period (Figure \ref{fig:cases-log}B). Both figures examplify a
considerable variation among the districts (light blue points are
individual district's data).

In Germany, we observed a rebound in mobility after the initial
political measures, reductions in incident cases were associated with a
diminishing public interest in COVID-19, and temperatures were overall
increasing (cf.~Figure \ref{fig:timeseries}); with correlations between
temporal progression and mobility in retail and recreation \(r_{A,D}=\)
-0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and
temperature \(r_{A,B}=\) 0.79.

\begin{figure}
\centering
\includegraphics{figures/f_timeseries_exposures.eps}
\caption{\label{fig:timeseries}Temporal variation of outcome and main
determinants}
\end{figure}

\subsection{Main results}\label{main-results}

\begin{table}

\caption{\label{tab:causaleffects}Causal effect estimates with 99 percent confidence bands}
\centering
\begin{tabular}[t]{lrrr}
\toprule
cause & effect estimate & CI (lower) & CI (upper)\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & 1.0076 & 1.0056 & 1.0097\\
\hspace{1em}grocery and pharmacy & 0.9951 & 0.9938 & 0.9965\\
\hspace{1em}parks & 0.9998 & 0.9994 & 1.0003\\
\hspace{1em}transit stations & 0.9975 & 0.9946 & 1.0004\\
\hspace{1em}workplaces & 1.0031 & 1.0002 & 1.0059\\
\hspace{1em}residential & 1.0080 & 0.9987 & 1.0174\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
\hspace{1em}COVID-19 burden & 0.9942 & 0.9932 & 0.9951\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Weather}}\\
\hspace{1em}Temperature & 0.9918 & 0.9879 & 0.9958\\
\hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
\hspace{1em}Humidity & 1.0022 & 1.0006 & 1.0039\\
\hspace{1em}Wind & 0.9893 & 0.9802 & 0.9985\\
\bottomrule
\end{tabular}
\end{table}

We list the results of our causal analysis for the effects of our
variables in Table \ref{tab:causaleffects}. The estimates are
multiplicative rates of increase/decrease for a one unit increase of the
respective variable: Values above 1 lead to an increase, below 1 to a
decrease of the infection rate. To put these estimates into perspective,
Figure \ref{fig:effectsrange} shows the relative causal effect of the
different exposure variables on the number of reported COVID-19 cases on
a range of sensible values of the exposure variables (95 percent
quantiles of data points).

\begin{figure}
\centering
\includegraphics{figures/f_exposure_vs_effect.eps}
\caption{\label{fig:effectsrange}Relative causal effects of exposures}
\end{figure}

Within our framework, we saw strong effects for mobility in
retail/recreational areas and essential shopping (grocery and pharmacy).
In the former, an increase of 1 percent point mobility compared to the
reference period (03 January to 06 February 2020) leads to an increase
of the daily reported case number by about 0.8 percent. Contrarily, a
corresponding increase of 1 percent point for the areas of
grocery/pharmacy leads to a decrease in the reported case number by
approximately 0.5 percent. Mobility on workplaces showed a small effect
of 0.3 increase in case numbers for every 1 percent point increase in
mobility. Other causal effects of mobility were insubstantial and not
consistent in their direction (99 percent confidence intervals of
estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
mobility on a range of possible values. Thus, we expect an increase of
daily cases by approximately 23 percent if mobility in retail/recreation
reaches baseline levels of 0 percent difference to the reference period.
On the other hand, an increase of mobility for grocery/pharmacy by 10
percent points compared to the reference period leads to a reduction of
the infection rate by approximately 7 percent.

``Awareness'' had two opposite effects on the outcome in our DAG.
Awareness measured by Google searches for \emph{corona} had a positive
effect on the number of reported cases. An one percent point increase of
the state's Google searches (relative to other states and the
observation period) leads to an increase of approximately 1 percent. For
example, if a district shows 10 percent points more relative searches
for \emph{corona} then another one, we expect approximately 11 percent
more infections for this district after 8 days. \emph{COVID-19 burden}
(reported number of cases on day of exposure) affected the outcome
negatively, where every additional daily case in the district leads to a
0.6 percent decrease in newly reported case numbers. The last plot in
Figure \ref{fig:effectsrange} visualizes this relationship: For a local
outbreak with 20 daily cases as COVID-19 burden, we estimate as total
causal effect a subsequent reduction of infection rate by 9 percent.

Within our model, we observed a causal effect of temperature and all
other weather variables. Every increase of 1 degree Celsius in
temperature leads to a reduction of the daily reported case numbers by
approximately 0.8 percent. Similarly strong was the effect of rainfall:
One millimeter (=1 liter per square meter) more rainfall leads to an
increase of reported case numbers by approximately 1.1 percent. We
observe small effects for humidity and wind as well (higher humidity
leading to fewer cases, stronger wind leading to less cases). In
perspective (Figure \ref{fig:effectsrange}), with temperature we expect
an increase by approximately 9 percent at a daily average temperature of
\(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm
rainfall a corresponding increase of the infection rate by approximately
2 percent.

In all cases we opted to use the reduced optimal adjustment set over the
minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly
above 0.3), except for mobility, where the minimal adjustment set had a
higher pseudo-\(R^2\). Notably, these sets always include most of our
socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets})
as well as the policy variables (except for COVID-19 burden). We also
decided for the lag of 8 days based on the highest pseudo-\(R^2\) values
compared to other lags on the chosen adjustment sets. Similarly,
negative binomial regression was chosen over Poisson regression, because
the latter showed overdispersion and overall lower pseudo-\(R^2\)
values.

\begin{table}

\caption{\label{tab:adjsets}Final adjustment sets for causal analysis}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{llllllll}
\toprule
\rotatebox{90}{ } & \rotatebox{90}{Mobility} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Searches corona} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind}\\
\midrule
Pseudo-R2 & 0.332 & 0.151 & 0.298 & 0.298 & 0.286 & 0.306 & 0.286\\
Weekday & x & x & x & x & x & x & x\\
Holiday (report) & x & x & x & x & x & x & x\\
Holiday (exposure) & x & x & x & x & x & x & x\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation &  &  &  &  &  &  & \\
\hspace{1em}grocery and pharmacy &  &  &  &  &  &  & \\
\hspace{1em}parks &  &  &  &  &  &  & \\
\hspace{1em}workplaces &  &  &  &  &  &  & \\
\hspace{1em}residential &  &  &  &  &  &  & \\
\hspace{1em}transit stations &  &  &  &  &  &  & \\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & x & x & x & x &  & x & x\\
\hspace{1em}Temperature & x & x & x &  &  & x & \\
\hspace{1em}Humidity &  & x & x &  &  &  & \\
\hspace{1em}Wind &  & x & x & x & x & x & \\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & x &  & x & x & x & x & x\\
\hspace{1em}School and kindergarten closures & x &  & x & x & x & x & x\\
\hspace{1em}Contact restrictions & x &  & x & x & x & x & x\\
\hspace{1em}Mandatory face masks &  &  & x & x & x & x & x\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age & x & x & x & x & x & x & x\\
\hspace{1em}Gender & x & x & x & x & x & x & x\\
\hspace{1em}Population density & x & x & x & x & x & x & x\\
\hspace{1em}Foreign citizens &  & x & x &  &  &  & \\
\hspace{1em}Socio-economic status & x & x & x & x & x & x & x\\
\hspace{1em}Turnout &  & x &  &  &  &  & \\
\hspace{1em}Right-wing populist party votes &  & x & x & x & x & x & x\\
\hspace{1em}Nursing homes &  & x & x & x & x & x & x\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & x &  &  & x & x & x & x\\
\hspace{1em}COVID-19 burden & x &  & x & x & x & x & x\\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion}\label{discussion}

\subsection{Main findings}\label{main-findings}

Our objective was to identify causal effects for COVID-19 cases. We
found that weather affects the reported number of infections, especially
temperature (which has a reducing effect on case numbers) and rainfall
(which increases case numbers). We saw that reports of high case numbers
in districts led to a reduction in new infection numbers, which
indicates risk-averse awareness in the population. The overall effect of
mobility showed no consistent effect, however, in specific areas
significant causal effects could be measured: Increasing activity in
retail and recreational areas increased reported case numbers, while
increased movement for essential shopping (grocery and pharmacy) led to
reduced case numbers.

Furthermore, we made a strong case for the use of causal DAGs in
epidemiology and a pandemic like COVID-19: DAGs allow to choose
confounders for the analysis in a principled and statistically correct
way while reducing possible causes for bias. Also, the DAG formalization
allows for discussion about the underlying causal structure.

\subsection{Comparison with previous
research}\label{comparison-with-previous-research}

Most research on determinants affecting case numbers of COVID-19 is
restricted to single aspects
\citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
To reliably identify causal drivers, one needs to adjust for
confounders. To this end, we used an integrated model with variables
from different aspects like mobility, awareness, weather, or
socio-demographics and identified confounders by causal analysis with a
directed acyclic graph. A causal approach is used in another current
COVID-19 analysis \citep{gencoglu2020causal}. There, however, they
identify the causal relationships (reconstruct a DAG), while we
estimated causal effects for a given hypothesized DAG.

Several studies assessing the impact of public health measures on
mobility have each observed a downward trend accompanied by a decrease
in the number of newly reported cases
\citep{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.

Our findings regarding awareness/Google Trends analysis are in good
agreement with the correlations found by
\citet{effenberger_association_2020}, \citet{higgins_correlations_2020},
and \citet{yuan_trends_2020}, who conclude that alertness to COVID-19
rises several days before the highest number of cases are reported. At
this point it should be noted, that awareness is substantially
influenced by public media coverage, which should be considered, if
possible, in future studies \citep{higgins_correlations_2020}. As such,
awareness is difficult to measure and here the number of Google searches
for ``corona'' could only be a proxy for this concept.

In addition, in alignment with other recent published studies, our
results confirm evidence which associated a negative effect of
temperature on new COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
It is however controversial to other scientific literature describing no
effects
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
or even converse correlations
\citep{auler_evidence_2020, xie_association_2020}. The conflicting
results might be explained by different climates and characteristics of
the populations under study. While we are confident that our strict
causal analysis resulted in effect estimates as undistorted as possible,
there might be unconsidered bias in those other studies. Further
research needs to be done to elucidate the biological characteristics of
the novel virus SARS-CoV-2 regarding its ambient temperature survival
and transmission. In regards of humidity, our findings (higher humidity
leading to increased case numbers) are in agreement with previous
research \citep{qi_covid-19_2020, wu_effects_2020, juni_impact_2020}.
Finally, we found a positive causal effect of increment precipitation
and a raise in COVID-19 cases, which supports previous observations
\citep{sobral_association_2020}.

\subsection{Limitations and strengths}\label{limitations-and-strengths}

While use of a causal DAG is itself a strong tool to identify causal
effects (and not just statistical associations), it introduces two
limitations: causal assumptions within the graph (depicted by edges)
need to be well justified, and the statistical regression model that
calculates total causal effects needs to be appropriate for the task at
hand. We endorse our graph as a basis for discussion on residual
confounding.

We observed overdispersion and a substantial increase in model
performance with a negative binomial regression compared to Poisson
regression, which is in line with the results on COVID-19 daily case
counts of \citet{kraemer_effect_2020} and others
\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020}. We did not
model case counts with a differential equation model like the classic
SIR-model \citep{kermack1991contributions} and its successors, since
these are more suited to prediction \citep[e.g.][]{rkimodellierung}
while our choice of a negative binomial regression framework allowed us
to estimate the effects of confounders more reliably. There are more
advanced statistical methods for count data, e.g.~zero-inflated models
and mixed models. We tested both approaches as extensions to the
negative binomial regression and experienced numerical problems and
increased computing time, along with an insubstantial increase in model
performance. Furthermore, our model assumed that all variables have
effects proportional to the size of their measurements. It is possible
that some variables show saturation effects or opposite effects for
lower, medium, or high values. This could be modelled with polynomial or
other transformations of the variables, which we did not employ due to
limited temporal and spatial data availability.

As we were using administrative data for our analysis, the results are
susceptible to the Modifiable Area Unit Problem (MAUP)
\citep{Openshaw1984}. The MAUP postulates that different regional
aggregations of the units of observation may lead to different results
and conclusions. Due to limited available data for the different
variables, there is currently no way to overcome this.

Our observation period was restricted to succession from late winter to
spring and summer (February to July). Nevertheless, this transition with
increasing temperature was a natural experiment that allowed clues on
weather effects.

We could not include data on health care utilization during the pandemic
into our models due to the lack of available resources. This is planned
for a later follow up to this paper since we rank health care
utilization and mobility within health care facilities among the strong
factors for COVID-19 progression: personnel in hospitals and private
practices is particularly exposed to infection, while the lack of
adequate care for other diseases has severe effects on general health of
the population. At the same time, health care facilities are key for
testing and surveillance of COVID-19 patients.

While our analysis focused on Germany and its districts, we assume that
results may be transferred to other countries by adjusting for their
respective weather conditions, mobility habits, socio-demographic
characteristics, and other determining factors.

The code and resources for our analysis are available on Github, we
invite other researchers to replicate our analysis with different
assumptions using the files provided in the repository\footnote{The
  repository is located here: https://github.com/zidatalab/causalcovid19}
of the article.

\subsection{Discussion of causal
effects}\label{discussion-of-causal-effects}

In our analysis, the adverse effect of mobility in retail/recreation and
the favorable effect of mobility in grocery/pharmacy make a strong case
for the successful implementation of policies like contact restrictions
which limit the number of individual interactions that lead to
infections. This is due to retail/recreational areas encompassing mostly
places of social gatherings like restaurants and bars, malls, sports and
music venues, among others, while if people are doing more of their
essential shopping at supermarkets, they will most likely stay at home
with less contact to other people.

The causal effects of awareness measured via searches for ``corona'' and
the COVID-19 burden are harder to interpret. We assume that within our
model, the searches for ``corona'' are an insufficient proxy for
awareness, while the decreasing effect for future case numbers of high
daily COVID-19 burden numbers indicates it affects individual
risk-behavior as a deterrent.

Similarly, the effects of temperature and rainfall can be interpreted as
causal effects for indoor and outdoor activities, such that higher
temperatures and low rainfall indicate more people spending time outdoor
while lower temperatures and high rainfall result in indoor activities,
which lead to more infections. In this light, we advocate for
precautious measures like increased hygiene, face masks, and air
ventilation for unavoidable indoor activities.

\subsection{Conclusions}\label{conclusions}

To the best of our knowledge, this is the most comprehensive analysis of
causes for COVID-19 infections which integrates different data sources
(all publicly available). Causal reasoning with a DAG allows us to
estimate the causal effects reliably.

Our findings suggest that the causal effects of mobility, awareness, and
weather need to be taken strongly into account when deciding for
mitigation and suppression measures, depending on the recent and future
COVID-19 pandemic development.

\renewcommand\refname{References}
\bibliography{mybibfile.bib}

\end{document}

### 2020-07-15 manuscript/v1_medrxiv/results.Rmd
init git repository

# Results

Descriptive statistics for the included variables are presented in Table \@ref(tab:descriptive-vars).

```{r descriptive-vars, echo=FALSE, message=FALSE, warning=FALSE}
library("dplyr")
library("kableExtra")
library("tidyverse")
library("tableone")

Modeldata <- read_csv("../../data/Modeldata_raw.csv") %>%
  rename(`Gender`=`Sex`,
         `Foreign citizens`=`Foreign residents`,
         `Foreign citizens (refugees)`=`Foreign residents (refugees)`) %>%
  filter(date<="2020-07-08")

t1_data <- Modeldata %>% dplyr::select(-contains("Weekday"), -contains("Holiday"), -`Relaxation of measures`,
                     -contains("Raumeinheit"),
                     -contains("Aggregat"),
                     -contains("iso_"),
                     -contains("census_"),
                     -contains("mean"),
                     -id,-bl_id,
                     -cases, -deaths,-recovered,-daycount) 
t1_data_tab <- CreateTableOne(data = t1_data) 
mymeans <- colMeans(t1_data[, -1], na.rm = TRUE)
# t1.df <- t1_data_tab.csv %>% as_tibble() %>% mutate(Variable=as.character(rownames(t1_data_tab.csv))) %>% 
#   dplyr::select(Variable,Descriptives=Overall)
t1.df <- print(t1_data_tab, quote = FALSE, noSpaces = TRUE, printToggle = FALSE, explain=FALSE) 
write.csv(t1.df,"tables/t_descriptives.csv")
t1.df <- read.csv("tables/t_descriptives.csv") %>% rename("Variable"="X") %>%
  mutate(Variable=recode(Variable,
                         "Mobility (retail and recreation)"="retail and recreation",
                                                          "Mobility (grocery and pharmacy)"="grocery and pharmacy",
                                                          "Mobility (parks)"="parks",
                                                          "Mobility (transit stations)"="transit stations",
                                                          "Mobility (workplaces)"="workplaces",
                                                          "Mobility (residential)"="residential",
                                                          "Weather (rainfall)"="Rainfall",
                                                          "Weather (wind)"="Wind",
                                                          "Weather (temperature)"="Temperature",
                                                          "Weather (humidity)"="Humidity",
                                                          "School/Kita closures"="School and kindergarten closures")) %>%
  mutate(Variable =  factor(Variable, ordered=TRUE, levels = c("n",
                                                 "retail and recreation",
                                                 "grocery and pharmacy",
                                                 "parks",
                                                 "workplaces",
                                                 "residential",
                                                 "transit stations",
                                                 "Rainfall",
                                                 "Temperature",
                                                 "Humidity",
                                                 "Wind",
                                                 "Ban of mass gatherings",
                                                 "School and kindergarten closures",
                                                 "Contact restrictions",
                                                 "Mandatory face masks",

                                                 "Age (pop. 65 and older)",
                                                 "Age (pop. younger 18)",
                                                 "Gender",
                                                 "Population density",
                                                 "Foreign citizens",
                                                 "Foreign citizens (refugees)",
                                                 "Socio-economic status",
                                                 "Turnout",
                                                 "Right-wing populist party votes",
                                                 "Nursing homes",
                                                 "Searches corona",
                                                 "COVID-19 burden", 
                                                 "Reported new cases COVID-19",
                                                 "Active cases"))) %>%
  arrange(Variable)    
write.csv(t1.df,"tables/t_descriptives.csv")

# number of districts
ndistricts <- Modeldata %>% count(id) %>% count() %>% pull(n)

# Obs per district
nobs_per_district <- Modeldata %>% group_by(id) %>% count() %>% ungroup() %>% summarise(mean=mean(n),sd=sd(n))

# Correlation of time and main vars
mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"),contains("Weather"),`Searches corona`,contains("Mobility"), `COVID-19 burden`) %>% 
  mutate(outcome=`Reported new cases COVID-19`/(`Active cases`+1),date=as.numeric(date)) %>% 
  dplyr::select(-contains("cases")) %>% filter(!is.na(outcome)) %>% cor(.)

# show table of decriptives
knitr::kable(readr::read_csv("tables/t_descriptives.csv") %>% dplyr::select("Variable","Overall") %>% rename(`mean (SD)`=`Overall`),
             format = "latex", booktabs = TRUE,
             caption = "Descriptive Statistics for observed variables") %>% kable_styling(font_size = 7, full_width = FALSE) %>%
  pack_rows("Mobility", 2, 7) %>%
  pack_rows("Weather", 8, 11) %>%
  pack_rows("Policies", 12, 15) %>%
  pack_rows("Socio-demographic", 16, 25) %>%
  pack_rows("Awareness", 26, 27) %>%
  pack_rows("Case numbers (Outcome and offset)", 28, 29)
```

```{r cases-log, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Temporal and district level variation of outcome (log-scale)"}
knitr::include_graphics(rep('figures/f_timeseries_cases_and_lograte.eps'))
```

In the observational period, the number of daily reported COVID-19 cases increased till the end of March/beginning of April and continually decreased afterwards till the beginning of June 2020 with a slight increase and decrease afterwards (Figure \@ref(fig:cases-log)A). On the other hand, the (log-)ratio of reported cases over active cases decreased steeply till the mid of April and increased steadily afterwards with a slight decrease close to the end of the observation period (Figure \@ref(fig:cases-log)B). Both figures examplify a considerable variation among the districts (light blue points are individual district's data).

In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf. Figure \@ref(fig:timeseries)); with correlations between temporal progression and mobility in retail and recreation $r_{A,D}=$ `r round(mycorrs["date", "Mobility (retail and recreation)"], 2)`, awareness ("Searches corona") $r_{A,C}=$ `r round(mycorrs["date", "Searches corona"], 2)`, and temperature $r_{A,B}=$ `r round(mycorrs["date", "Weather (temperature)"], 2)`.

```{r timeseries, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Temporal variation of outcome and main determinants"}
knitr::include_graphics(rep('figures/f_timeseries_exposures.eps'))
```

## Main results

```{r causaleffects, echo=FALSE, message=FALSE, warning=FALSE}
library("dplyr")
library("kableExtra")
causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_MinAdjSet1.csv"),
                           readr::read_csv("tables/t_effects_Searches corona_RedOptAdjSet.csv"),
                           readr::read_csv("tables/t_effects_COVID-19 burden_RedOptAdjSet.csv"),
                           readr::read_csv("tables/t_effects_Temperature_RedOptAdjSet.csv"),
                           readr::read_csv("tables/t_effects_Rainfall_RedOptAdjSet.csv"),
                           readr::read_csv("tables/t_effects_Humidity_RedOptAdjSet.csv"),
                           readr::read_csv("tables/t_effects_Wind_RedOptAdjSet.csv")) %>%
               rename(cause=variables, `effect estimate`=estimates, `CI (lower)`=conflow, `CI (upper)`=confhigh)
## Note: no special chars in caption...
knitr::kable(causaleffects %>% 
               mutate(cause=stringr::str_remove(stringr::str_remove(cause,"Mobility \\("),"\\)")),
             booktabs = TRUE,
             caption = "Causal effect estimates with 99 percent confidence bands",
             digits=4) %>%
  kable_styling() %>%
  pack_rows("Mobility", 1, 6) %>%
  pack_rows("Awareness", 7, 8) %>%
  pack_rows("Weather", 9, 12)
```

We list the results of our causal analysis for the effects of our variables in Table \@ref(tab:causaleffects). The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate. To put these estimates into perspective, Figure \@ref(fig:effectsrange) shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).

```{r effectsrange, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Relative causal effects of exposures"}
knitr::include_graphics(rep('figures/f_exposure_vs_effect.eps'))
```

Within our framework, we saw strong effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 1)` percent. Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 1)` increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \@ref(fig:effectsrange) shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate"))^(0-mymeans["Mobility (retail and recreation)"])), 0)` percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate"))^(10-mymeans["Mobility (grocery and pharmacy)"])), 0)` percent.

"Awareness" had two opposite effects on the outcome in our DAG. Awareness measured by Google searches for *corona* had a positive effect on the number of reported cases. An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate")), 1)` percent. For example, if a district shows 10 percent points more relative searches for *corona* then another one, we expect approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate"))^(10)), 0)` percent more infections for this district after 8 days.  _COVID-19 burden_ (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a `r round(100*abs(1-causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate")), 1)` percent decrease in newly reported case numbers. The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(20-mymeans["COVID-19 burden"])), 0)` percent.

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 1)` percent. Similarly strong was the effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent. We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases). In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 0)` percent at a daily average temperature of $0^{\circ}C$. For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(0-mymeans["Weather (rainfall)"])), 0)` percent.

In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-$R^2$ values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-$R^2$. Notably, these sets always include most of our socio-demographic variables as confounders (cf. Table \@ref(tab:adjsets)) as well as the policy variables (except for COVID-19 burden). We also decided for the lag of 8 days based on the highest pseudo-$R^2$ values compared to other lags on the chosen adjustment sets. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-$R^2$ values.

```{r adjsets, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:22, 24:28), ] %>% dplyr::select(variable) %>%
  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ")
for (cause in colnames(myvariables)[-c(1,2)]) {
  adjset <- readLines(paste0("tables/t_adjust_", cause, "_RedOptAdjSet.csv"))
  for (as in adjset) {
    myvariables[[cause]][grep(as, myvariables$variable)] <- "x"
  }
}
cause <- colnames(myvariables)[2]
adjset <- readLines(paste0("tables/t_adjust_", cause, "_MinAdjSet1.csv"))
for (as in adjset) {
  myvariables[[cause]][grep(as, myvariables$variable)] <- "x"
}
t_pseudor2s <- read_csv("tables/t_pseudor2s.csv", col_names = c("exposure", "pseudor2"), col_types = "cn") %>%
  mutate(pseudor2=as.character(round(pseudor2, 3))) %>%
  pivot_wider(names_from = "exposure", values_from = "pseudor2")

myvariables <- bind_rows(t_pseudor2s, myvariables)[c(8, 1:7)]
myvariables$variable[1] <- "Pseudo-R2"
kable(myvariables %>% rename(" "=variable),
             format = "latex", booktabs = TRUE, 
             caption = "Final adjustment sets for causal analysis") %>%
  kable_styling(font_size=7) %>% # , latex_options="scale_down"
  row_spec(0, angle = 90) %>%
  pack_rows("Mobility", 5, 10) %>%
  pack_rows("Policies", 15, 18) %>%
  pack_rows("Socio-demographic", 19, 26) %>%
  pack_rows("Awareness", 27, 28) %>%
  pack_rows("Weather", 11, 14) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
```

### 2020-07-15 R/0_1_rkicases.R

@@ -1,0 +1,101 @@
+library(tidyverse)
+library(lubridate)
+
+  # Config Params
+  baseurl = "https://services7.arcgis.com/mOBPykOjAyBO2ZKk/arcgis/rest/services/RKI_COVID19/FeatureServer/0/query?"
+  paramsforids = "where=1%3D1&outFields=*&returnGeometry=false&returnDistinctValues=true&returnIdsOnly=true&outSR=4326&f=json"
+  otherparamsforurl = "&outFields=*&outSR=4326&f=json"
+  
+  # Get objects
+  mytemp = tempfile()
+  myjson = paste0(baseurl,paramsforids)
+  download.file(myjson, mytemp, quiet=TRUE)
+  objectids <- jsonlite::fromJSON(mytemp, flatten = TRUE)
+  myfieldname <- objectids$objectIdFieldName
+  myids <- sort(objectids$objectIds)
+  batchsize <- 4000
+  myidsindexlist <- seq(1,length(myids),batchsize)
+  
+  # Query Objects
+  resultdf <- tibble()
+  
+  # Query all batches except last one
+  for (myid in seq(1,length(myidsindexlist),1)){
+    if (myid < length(myidsindexlist)) {
+      mytemp = tempfile()
+      myjson = paste0(baseurl,"where=","(",myfieldname,">=",myids[myidsindexlist[myid]],")","AND","(",myfieldname,"<",myids[myidsindexlist[myid+1]],")",otherparamsforurl)
+      download.file(myjson, mytemp, quiet=TRUE)
+      myresult <- jsonlite::fromJSON(mytemp, flatten = TRUE)
+    }
+    myresult <- as.data.frame(myresult$features)
+    resultdf <- bind_rows(resultdf,myresult)
+  }
+  
+  # Query last batch
+  mytemp = tempfile()
+  myjson = paste0(baseurl,"where=",myfieldname,">=",myids[myidsindexlist[length(myidsindexlist)]],otherparamsforurl)
+  download.file(myjson, mytemp, quiet=TRUE)
+  myresult <- jsonlite::fromJSON(mytemp, flatten = TRUE)
+  myresult <- as.data.frame(myresult$features)
+  resultdf <- bind_rows(resultdf,myresult)
+  
+  # Remove name prefix
+  newnames <- str_remove(colnames(resultdf),"attributes.")
+  names(resultdf) <- newnames
[…]

### 2020-07-15 R/0_2_mobility.R

@@ -1,0 +1,35 @@
+library(sf)
+library(tidyverse)
+library(geojsonsf)
+library(geojsonio)
+
+mydatapath <- "./data/mobility/"
+
+# google data
+gmobillink <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
+mytemp <- tempfile()
+download.file(gmobillink, mytemp, quiet=TRUE)
+google_german_mobility <- read_csv(mytemp) %>%
+  filter(country_region_code=="DE") %>%
+  dplyr::select(-country_region_code, -country_region, -sub_region_2)
+
+en_rs_map_bl <- c("08", "09", "11", "12",
+                  "04", "02", "06", "03",
+                  "13", "05", "07", "10",
+                  "14", "15","01", "16")
+# names(en_de_map_bl) <- c("Baden-Württemberg", "Bayern", "Berlin", "Brandenburg",
+#                          "Bremen", "Hamburg", "Hessen", "Niedersachsen",
+#                          "Mecklenburg-Vorpommern", "Nordrhein-Westfalen", "Rheinland-Pfalz", "Saarland",
+#                          "Sachsen", "Sachsen-Anhalt","Schleswig-Holstein", "Thüringen")
+names(en_rs_map_bl) <- c("Baden-Württemberg", "Bavaria", "Berlin", "Brandenburg",
+                         "Bremen", "Hamburg", "Hesse", "Lower Saxony",
+                         "Mecklenburg-Vorpommern", "North Rhine-Westphalia", "Rhineland-Palatinate", "Saarland",
+                         "Saxony", "Saxony-Anhalt","Schleswig-Holstein", "Thuringia")
+google_german_mobility <- google_german_mobility %>%
+  mutate(grocery_and_pharmacy_percent_change_from_baseline=replace_na(grocery_and_pharmacy_percent_change_from_baseline, 0)) %>%
+  mutate(parks_percent_change_from_baseline=replace_na(parks_percent_change_from_baseline, 0)) %>%
+  mutate(RS_first2=en_rs_map_bl[sub_region_1]) %>%
+  dplyr::select(-sub_region_1) %>%
+  # drop_na() %>%
+  pivot_longer(cols=ends_with("baseline"), names_to = "areatype", values_to = "perc_change")
+write_csv(google_german_mobility, paste0(mydatapath, "google_mobilitychange_percentages_per_bundesland_and_areatype.csv"))

### 2020-07-15 R/0_3_awareness.R

@@ -1,0 +1,22 @@
+library(tidyverse)
+
+mydatapath <- "./data/awareness/"
+
+trends_alle_bl <- read_delim(paste0(mydatapath, "trends_alle_bl.csv"), 
+                             ";", escape_double = FALSE, col_types = cols(BL_ID = col_integer(), 
+                                                                          Coronatrend = col_double(), Datum = col_date(format = "%Y-%m-%d")), 
+                             locale = locale(encoding = "UTF-8"),
+                             trim_ws = TRUE)
+relativ_gesamt_2020_03_13 <- read_csv(paste0(mydatapath, "relativ_gesamt_2020-03-13.csv"))
+
+relativ_bl_2020_03_13 <- left_join(trends_alle_bl %>% filter(Datum=="2020-03-13"), relativ_gesamt_2020_03_13, by="Bundesland") %>%
+  mutate(baseline20200313=Coronatrend*Relativtrend/100) %>%
+  rename(trend20200313=Coronatrend) %>%
+  dplyr::select(BL_ID, baseline20200313, trend20200313)
+
+relativ_bl <- left_join(trends_alle_bl, relativ_bl_2020_03_13, by="BL_ID") %>%
+  mutate(relativtrend=baseline20200313/trend20200313*Coronatrend) %>%
+  rename(date=Datum) %>%
+  dplyr::select(date, BL_ID, Bundesland, relativtrend)
+
+write_csv(relativ_bl, paste0(mydatapath, "awareness.csv"))

### 2020-07-15 R/0_4_inkar.R

@@ -1,0 +1,26 @@
+library(readr)
+header <- read_delim(file = "data/sociodemographic/inkar_sdg_etc.csv",
+                      delim=";",
+                      col_names = FALSE,
+                      n_max = 1)
+inkar <- read_delim("data/sociodemographic/inkar_sdg_etc.csv", 
+                            ";", escape_double = FALSE, col_names = FALSE, 
+                            locale = locale(decimal_mark = ",", grouping_mark = "."), 
+                            trim_ws = TRUE, skip = 2)
+
+colnames(inkar) <- header[1, ]
+
+inkar <- inkar %>%
+  mutate(`Age (pop. younger 18)`=`Einwohner unter 6 Jahre`+`Einwohner von 6 bis unter 18 Jahren`,
+         Frauenanteil=`Geschlechterproportion insgesamt`/(1+`Geschlechterproportion insgesamt`)) %>%
+  dplyr::select(contains("Haushalte mit niedrigem"),
+                contains("Jahre"),
+                contains("Ausländeranteil"), contains("Schutzsuchende"),
+                contains("Wahlbeteiligung"), contains("AfD"),
+                Einwohnerdichte, Frauenanteil,
+                contains("Age"),
+                Kennziffer, Raumeinheit, Aggregat,
+                -contains("Einwohner unter 6"), -contains("Einwohner von 6")
+  )
+
+write_csv(inkar, "./data/inkar/inkar.csv")

### 2020-07-15 R/0_5_weather.R

@@ -1,0 +1,203 @@
+# the download of weather files takes a while! ~ 1 h
+
+library(readr)
+library(tidyverse)
+
+mydatapath <- "./data/weather/"
+
+opendata_dwd_path <- "https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/recent/"
+
+download.file(paste0(opendata_dwd_path, "KL_Tageswerte_Beschreibung_Stationen.txt"),
+              paste0(mydatapath, "weatherstations.txt"))
+
+header <- read_table2(file = paste0(mydatapath, "weatherstations.txt"),
+                   col_names = FALSE,
+                   n_max = 1)
+weatherdata <- read_table(file = paste0(mydatapath, "weatherstations.txt"),
+                    col_names = FALSE,
+                    locale = locale(encoding = "WINDOWS-1252"),
+                    skip = 2)
+colnames(weatherdata) <- header[1, ]
+
+extractedfiles <- c()
+
+best_bisdatum <- last(names(sort(table(weatherdata$bis_datum))))
+
+weatherdata <- weatherdata %>%
+  filter(bis_datum==as.numeric(best_bisdatum) & von_datum <= 20200101 & Stationshoehe<1000)
+
+for (stationid in weatherdata$Stations_id) {
+  tryCatch({
+    download.file(paste0(opendata_dwd_path, "tageswerte_KL_", stationid, "_akt.zip"),
+                  paste0(mydatapath, "temp.zip"),
+                  quiet=TRUE)
+    cat("downloaded data for", stationid, "successfully\n")
+    listoffilesinzip <- unzip(paste0(mydatapath, "temp.zip"),
+                              list=TRUE)
+    extractedfile <- unzip(paste0(mydatapath, "temp.zip"),
+                           files=last(listoffilesinzip$Name),
+                           exdir=paste0(mydatapath, "temp"))
+    extractedfiles <- c(extractedfiles, extractedfile)
+  }, error = function(e) {
+    cat(stationid, "did not work\n")
+  }, warning = function(w) {
+    cat(stationid, "did not work\n")
+  })
+  
+}
+
+write_lines(extractedfiles,
[…]

### 2020-07-15 R/1_generate_data.R

@@ -1,0 +1,162 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+
+mydatapath <- "./data/" 
+
+google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
+  mutate(id=as.integer(RS_first2)) %>%
+  dplyr::select(-RS_first2) %>%
+  pivot_wider(names_from = areatype, values_from = perc_change) %>%
+  rename(`Mobility (retail and recreation)`=retail_and_recreation_percent_change_from_baseline,
+         `Mobility (grocery and pharmacy)`=grocery_and_pharmacy_percent_change_from_baseline,
+         `Mobility (parks)`=parks_percent_change_from_baseline,
+         `Mobility (transit stations)`=transit_stations_percent_change_from_baseline,
+         `Mobility (workplaces)`=workplaces_percent_change_from_baseline,
+         `Mobility (residential)`=residential_percent_change_from_baseline)
+
+weather <- read_csv(paste0(mydatapath, "weather/mittleres_kreiswetter.csv")) %>%
+  mutate(date=ymd(MESS_DATUM), id=as.integer(RS)*1000) %>%
+  dplyr::select(-MESS_DATUM, -RS) %>%
+  mutate(Wind=replace_na(Wind, mean(Wind, na.rm=TRUE)),
+         Feuchtigkeit=replace_na(Feuchtigkeit, mean(Feuchtigkeit, na.rm=TRUE))) %>%
+  rename(`Weather (rainfall)`=Niederschlag,
+         `Weather (wind)`=Wind,
+         `Weather (temperature)`=Temperatur,
+         `Weather (humidity)`=Feuchtigkeit)
+
+awareness <- read_csv(paste0(mydatapath, "awareness/awareness.csv")) %>%
+  rename(`Searches corona`=relativtrend) %>%
+  dplyr::select(-Bundesland)
+
+pflege_destatis <- read_delim(paste0(mydatapath, "sociodemographic/pflege_destatis.csv"), 
+                              ";", escape_double = FALSE, locale = locale(decimal_mark = ",", 
+                                                                          grouping_mark = ".", encoding = "WINDOWS-1252"), 
+                              trim_ws = TRUE, skip = 6, na = "-") %>%
[…]

### 2020-07-15 R/2_dags_for_covidcases.R

@@ -1,0 +1,270 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+
+# save tables to:
+tablepath <- "./manuscript/v1_medrxiv/tables/"
+
+# load dag
+dagfile <- "./data/bigdag4covid19.txt"
+dagtext <- readChar(dagfile, file.info(dagfile)$size)
+dag <- dagitty::dagitty(dagtext)
+
+unobserved <- c("Access to tests", 
+                "Tracing capacity",
+                "Health care utilization",
+                "Herd immunity",
+                "Onset of COVID-19",
+                "Trust in society",
+                "Susceptibility")
+
+# Load Data
+modeldata <- read_csv("./data/Modeldata.csv") %>%
+  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
+                -`Mobility (mean)`,
+                -`Relaxation of measures`,
+                -contains("iso"), -contains("census")
+                ) %>% 
+  rename(`School and kindergarten closures`=`School/Kita closures`,
+         `Holiday (exposure)`=Holiday,
+         `Foreign citizens`=`Foreign residents`,
+         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
+         `Rainfall`=`Weather (rainfall)`,
+         `Humidity`=`Weather (humidity)`,
+         `Gender`=`Sex`,
+         `Temperature`=`Weather (temperature)`,
+         `Wind`=`Weather (wind)`) %>%
+  filter(date<="2020-07-08") %>%
+  dplyr::select(-date)
+
+# convert dagitty to adjacency matrix
+dagitty_to_adjmatrix <- function(daggity_obj) {
+  edg <- dagitty:::edges(daggity_obj)
+  node_names <- dagitty:::names.dagitty(daggity_obj)
+  ans_mat <- matrix(
+    data = 0, nrow = length(node_names),
+    ncol = length(node_names),
+    dimnames = list(node_names, node_names)
+  )
+  
+  ans_mat[as.matrix(edg[c("v", "w")])] <- 1
+  return(ans_mat)
+}
+
+# function to assemble variable names
+my_vars <- function(modeldata, mynames) {
+  myvars <- character()
[…]

### 2020-07-15 manuscript/v1_medrxiv/manuscript.Rmd
fix abstract

@@
 Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable causal effects of increasing temperature and of increased public mobility for essential shopping (grocery and pharmacy) as well as awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
-Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google and higher rainfall, leading to an increase in new COVID-19 cases.
+Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
 This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

### 2020-07-15 manuscript/v1_medrxiv/manuscript.tex
fix abstract

@@
   effects based on our DAG analysis by negative binomial regression. Our
-  analysis revealed favorable causal effects of increasing temperature and
-  of increased public mobility for essential shopping (grocery and
-  pharmacy) as well as awareness measured by COVID-19 burden, all of them
-  reducing the outcome of newly reported COVID-19 cases. Conversely, we
-  saw adverse effects of public mobility in retail and recreational areas,
-  awareness measured by searches for ``corona'' in Google and higher
-  rainfall, leading to an increase in new COVID-19 cases. This
-  comprehensive causal analysis of a variety of determinants affecting
-  COVID-19 progression gives strong evidence for the driving forces of
-  mobility, public awareness, and temperature, whose implications need to
-  be taken into account for future decisions regarding pandemic
-  management.
+  analysis revealed favorable causal effects of increasing temperature,
+  increased public mobility for essential shopping (grocery and pharmacy),
+  and awareness measured by COVID-19 burden, all of them reducing the
+  outcome of newly reported COVID-19 cases. Conversely, we saw adverse
+  effects of public mobility in retail and recreational areas, awareness
+  measured by searches for ``corona'' in Google, and higher rainfall,
+  leading to an increase in new COVID-19 cases. This comprehensive causal
+  analysis of a variety of determinants affecting COVID-19 progression
+  gives strong evidence for the driving forces of mobility, public
+  awareness, and temperature, whose implications need to be taken into
+  account for future decisions regarding pandemic management.
   \end{abstract}

### 2020-07-21 manuscript/v2_medrxiv/discussion.Rmd
update manuscript for version 2

--- 2020-07-15 manuscript/v1_medrxiv/discussion.Rmd
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-In regards of humidity, our findings (higher humidity leading to increased case numbers) are in agreement with previous research [@qi_covid-19_2020;@wu_effects_2020;@juni_impact_2020].
+In regards of humidity, our findings (higher humidity leading to increased case numbers) are in agreement with some previous research [@qi_covid-19_2020;@juni_impact_2020] but contradict others [@wu_effects_2020].
 Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
@@
 
+We stress the point that our effects were deduced on an aggregate (district) level.
+As such, we cannot make assumptions about effects on an individual level without falling for an ecological fallacy.
+E.g., we found that on district level more humidity leads to more reported cases (similar to the results of )
 As we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984].

### 2020-07-21 manuscript/v2_medrxiv/figures/f_timeseries_exposures.R

@@ -1,0 +1,57 @@
+library("tidyverse")
+library("cowplot")
+
+figurepath <- "./manuscript/v1_medrxiv/figures/"
+
+Modeldata <- read_csv("data/Modeldata_raw.csv")
+Sys.setlocale("LC_TIME", "C")
+mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
+
+mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
+cases <- Modeldata %>% dplyr::select(date,contains("cases"), id) %>%
+    ggplot(aes(x=date,
+                       y=`Reported new cases COVID-19`/(`Active cases`+1)))+
+  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
+  geom_smooth(color=mycolors[1],size=2,se = F) + 
+  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") + labs(x="Date", y="new cases/active cases") + 
+  scale_y_log10() + 
+  theme(axis.title.x = element_blank()) +
+  theme_cowplot(font_size = 8)
+
+temperature <- Modeldata %>% dplyr::select(date,id, contains("temperature")) %>%
+  ggplot(aes(x= date-8,
+             y= `Weather (temperature)`))+
+  geom_point(aes(group=id), color="moccasin", stroke=0, alpha=0.2) +
+  geom_smooth(color=mycolors[2],size=2,se = F) + 
+  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)+2), date_labels = "%d %b") + labs(x="Date", y="Temperature") + 
+  scale_y_continuous() + 
+  theme(axis.title.x = element_blank())+
+  theme_cowplot(font_size = 8)
+
+awareness <- Modeldata %>% dplyr::select(date, id, `Searches corona`) %>%
+  ggplot(aes(x=date-8,
+             y= `Searches corona`)) +
+  geom_point(aes(group=id), color="lightpink", stroke=0, alpha=0.2) +
+  geom_smooth(color=mycolors[3],size=2,se = F) + 
+  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") + labs(x="Date", y="Awareness") + 
+  scale_y_continuous(limits = c(0,100))+
+  labs(y="Awareness") +
+  theme(axis.title.x = element_blank())+
[…]

### 2020-07-21 manuscript/v2_medrxiv/discussion.Rmd
remarks twitter part 1

--- 2020-07-15 manuscript/v1_medrxiv/discussion.Rmd
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-In regards of humidity, our findings (higher humidity leading to increased case numbers) are in agreement with previous research [@qi_covid-19_2020;@wu_effects_2020;@juni_impact_2020].
 Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
@@
 We endorse our graph as a basis for discussion on residual confounding.
+We did not try to construct the DAG from the available data (cf.
+[@gencoglu2020causal]).
+As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.
+between the policies like contact restrictions and mandatory face masks).
 
@@
 This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
+By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers, though.
 
-As we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984].
+We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level.
+As such, we cannot make conclusions about effects on individuals without the possibility for an ecological fallacy.
+Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984].
 The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions.
@@
 
-In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy make a strong case for the successful implementation of policies like contact restrictions which limit the number of individual interactions that lead to infections.
+In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
 This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.
@@
 To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available).
-Causal reasoning with a DAG allows us to estimate the causal effects reliably.
+Causal reasoning with a DAG allows us to estimate the causal effects more reliably.
 

### 2020-07-21 manuscript/v2_medrxiv/results.Rmd
remarks twitter part 1

--- 2020-07-15 manuscript/v1_medrxiv/results.Rmd
@@
 
-Within our framework, we saw strong effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
+Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
 In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent.
@@
 Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 1)` percent.
-Similarly strong was the effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent.
 We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases).

### 2020-07-22 R/2_dags_for_covidcases.R
put dag aux functions into separate file

@@ -8,9 +8,10 @@ library(ggdag)
 library(ggthemes)
 library(pcalg)
 library(pscl)
+source("R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
-tablepath <- "./manuscript/v1_medrxiv/tables/"
+tablepath <- "./manuscript/v2_medrxiv/tables/"
 
 # load dag
 dagfile <- "./data/bigdag4covid19.txt"
@@ -43,169 +44,6 @@ modeldata <- read_csv("./data/Modeldata.csv") %>%
          `Wind`=`Weather (wind)`) %>%
   filter(date<="2020-07-08") %>%
   dplyr::select(-date)
-
-# convert dagitty to adjacency matrix
-dagitty_to_adjmatrix <- function(daggity_obj) {
-  edg <- dagitty:::edges(daggity_obj)
-  node_names <- dagitty:::names.dagitty(daggity_obj)
-  ans_mat <- matrix(
-    data = 0, nrow = length(node_names),
-    ncol = length(node_names),
-    dimnames = list(node_names, node_names)
-  )
-  
-  ans_mat[as.matrix(edg[c("v", "w")])] <- 1
-  return(ans_mat)
-}
-
-# function to assemble variable names
-my_vars <- function(modeldata, mynames) {
-  myvars <- character()
-  allvars <- colnames(modeldata)
-  idx <- 1
-  for (mn in mynames) {
-    thesevars <- allvars[grep(mn, allvars, fixed=TRUE)]
-    for (tvidx in seq(thesevars)) {
-      myvars[idx] <- thesevars[tvidx]
-      idx <- idx+1
-    }
-  }
-  return(myvars)
-}
-
-# function to get optimal adjustment set without unobserved (check for validness)
-get_optadjset <- function(amat, exposure, outcome, unobserved) {
-  x.pos <- which(colnames(amat)==exposure)
-  y.pos <- which(colnames(amat)==outcome)
-  optadjset <- colnames(amat)[optAdjSet(amat, x.pos = x.pos, y.pos = y.pos)]
-  optadjset_red <- setdiff(optadjset, unobserved)
-  optadjset_red_idx <- sapply(optadjset_red, function(z) which(colnames(amat)==z))
-  is_valid <- gac(amat, x.pos, y.pos, optadjset_red_idx, type="dag")
-  if (is_valid$gac) {
[…]

### 2020-07-22 R/z_auxiliary_functions_for_2.R

@@ -1,0 +1,175 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+
+# convert dagitty to adjacency matrix
+dagitty_to_adjmatrix <- function(daggity_obj) {
+  edg <- dagitty:::edges(daggity_obj)
+  node_names <- dagitty:::names.dagitty(daggity_obj)
+  ans_mat <- matrix(
+    data = 0, nrow = length(node_names),
+    ncol = length(node_names),
+    dimnames = list(node_names, node_names)
+  )
+  ans_mat[as.matrix(edg[c("v", "w")])] <- 1
+  ans_mat[as.matrix(edg[edg$e=="--", ][c("w", "v")])] <- 1
+  return(ans_mat)
+}
+
+# function to assemble variable names
+my_vars <- function(modeldata, mynames) {
+  myvars <- character()
+  allvars <- colnames(modeldata)
+  idx <- 1
+  for (mn in mynames) {
+    thesevars <- allvars[grep(mn, allvars, fixed=TRUE)]
+    for (tvidx in seq(thesevars)) {
+      myvars[idx] <- thesevars[tvidx]
+      idx <- idx+1
+    }
+  }
+  return(myvars)
+}
+
+# function to get optimal adjustment set without unobserved (check for validness)
+get_optadjset <- function(amat, exposure, outcome, unobserved) {
+  x.pos <- which(colnames(amat)==exposure)
+  y.pos <- which(colnames(amat)==outcome)
+  optadjset <- colnames(amat)[optAdjSet(amat, x.pos = x.pos, y.pos = y.pos)]
+  optadjset_red <- setdiff(optadjset, unobserved)
+  optadjset_red_idx <- sapply(optadjset_red, function(z) which(colnames(amat)==z))
+  is_valid <- gac(amat, x.pos, y.pos, optadjset_red_idx, type="dag")
+  if (is_valid$gac) {
+    return(list(adjset=optadjset_red, is_reduced=(length(optadjset_red)<length(optadjset))))
+  } else {
+    return(NULL)
+  }
+}
+
+# function to get minimal and optimal adjustment sets without unobserved (check for validness)
+get_adjsets <- function(dag, exposure, outcome, unobserved) {
+  amat <- t(dagitty_to_adjmatrix(dag))
+  x.pos <- which(colnames(amat)==exposure)
[…]

### 2020-07-22 R/2_z_othergraphs.R
mag and markov equivalence class analysis of adjustment sets

@@ -1,0 +1,62 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+source("R/z_auxiliary_functions_for_2.R")
+
+# save tables to:
+tablepath <- "./manuscript/v2_medrxiv/tables/"
+
+# load dag
+dagfile <- "./data/bigdag4covid19.txt"
+dagtext <- readChar(dagfile, file.info(dagfile)$size)
+dag_origin <- dagitty::dagitty(dagtext)
+
+unobserved <- c("Access to tests", 
+                "Tracing capacity",
+                "Health care utilization",
+                "Herd immunity",
+                "Onset of COVID-19",
+                "Trust in society",
+                "Susceptibility")
+
+# Load Data
+modeldata <- read_csv("./data/Modeldata.csv") %>%
+  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
+                -`Mobility (mean)`,
+                -`Relaxation of measures`,
+                -contains("iso"), -contains("census")
+  ) %>% 
+  rename(`School and kindergarten closures`=`School/Kita closures`,
+         `Holiday (exposure)`=Holiday,
+         `Foreign citizens`=`Foreign residents`,
+         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
+         `Rainfall`=`Weather (rainfall)`,
+         `Humidity`=`Weather (humidity)`,
+         `Gender`=`Sex`,
+         `Temperature`=`Weather (temperature)`,
+         `Wind`=`Weather (wind)`) %>%
+  filter(date<="2020-07-08") %>%
+  dplyr::select(-date)
+
+adjsets_list <- vector("list", 3)
+names(adjsets_list) <- c("original", "equivalence", "mag")
+
+exposures <- c("Mobility", "Searches corona", "COVID-19 burden", "Temperature", "Rainfall", "Humidity", "Wind")
+outcome <- "Reported new cases COVID-19"
+
+dag_equiv <- equivalenceClass(dag_origin)
+dag_mag <- toMAG(dag_origin)
+
+# original adjustment sets
+for (expo in exposures) {
[…]

### 2020-07-22 manuscript/v2_medrxiv/discussion.Rmd
update discussion/bibfile with mag/cpdag analysis

@@
 between the policies like contact restrictions and mandatory face masks).
+Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs [@perkovic2017complete].
+For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets.
+A further analysis of these implications is out of the scope of this paper.
 

### 2020-07-22 manuscript/v2_medrxiv/discussion.Rmd
add poisson regr policy paper

@@
 
-We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020].
+We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020;@Islamm2743].
 We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g.
@@
 Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
-By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers, though.
+By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).
 

### 2020-07-22 manuscript/v2_medrxiv/discussion.Rmd
update ecological fallacy/maup paragraph

@@
 We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level.
-As such, we cannot make conclusions about effects on individuals without the possibility for an ecological fallacy.
+As such, conclusions about effects cannot be transferred on individuals without the possibility for an ecological fallacy.
 Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984].
 The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions.
-Due to limited available data for the different variables, there is currently no way to overcome this.
+Due to limited available data for the different variables, there is currently no way to overcome these problems that are inherent to all analyses on aggregated data level.
 

### 2020-07-22 manuscript/v2_medrxiv/manuscript.tex
fix bibfile, compile tex

--- 2020-07-15 manuscript/v1_medrxiv/manuscript.tex
@@
 
-  \title{Causal analysis of COVID-19 observational data in German districts
-reveals effects of mobility, awareness, and temperature}
+  \title{Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
     \author[Zi]{Edgar Steiger}
@@
    \ead{lkroll@zi.de} 
-      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi),
-Salzufer 8, D-10587 Berlin, Germany}
+      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany}
 
   \begin{abstract}
-  Mobility, awareness, and weather are suspected to be causal drivers for
-  new cases of COVID-19 infection. Correcting for possible confounders, we
-  estimated their causal effects on reported case numbers. To this end, we
-  used a directed acyclic graph (DAG) as a graphical representation of the
-  hypothesized causal effects of the aforementioned determinants on new
-  reported cases of COVID-19. Based on this, we computed valid adjustment
-  sets of the possible confounding factors. We collected data for Germany
-  from publicly available sources (e.g.~Robert Koch Institute, Germany's
-  National Meteorological Service, Google) for 401 German districts over
-  the period of 15 February to 8 July 2020, and estimated total causal
-  effects based on our DAG analysis by negative binomial regression. Our
-  analysis revealed favorable causal effects of increasing temperature,
-  increased public mobility for essential shopping (grocery and pharmacy),
-  and awareness measured by COVID-19 burden, all of them reducing the
-  outcome of newly reported COVID-19 cases. Conversely, we saw adverse
-  effects of public mobility in retail and recreational areas, awareness
-  measured by searches for ``corona'' in Google, and higher rainfall,
-  leading to an increase in new COVID-19 cases. This comprehensive causal
-  analysis of a variety of determinants affecting COVID-19 progression
-  gives strong evidence for the driving forces of mobility, public
-  awareness, and temperature, whose implications need to be taken into
-  account for future decisions regarding pandemic management.
+  Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection.
+Correcting for possible confounders, we estimated their causal effects on reported case numbers.
+To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19.
+Based on this, we computed valid adjustment sets of the possible confounding factors.
+We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
+Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
+This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
   \end{abstract}
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-As the COVID-19 pandemic progresses, research on mechanisms behind the
-transmission of SARS-CoV-2 shows conflicting evidence
-\citep{who2020report, Chinazzi395, guan2020clinical}. While effects of
-mobility have been extensively discussed, less is known on other factors
-such as changing awareness in the population
-\citep{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
-or the effects of temperature
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}.
-A limiting factor in many studies is the lack of a causal approach to
-assess the causal contributions of various factors
-\citep{Greenland1999}. This can lead to distorted estimates of the
-causal factors with observational data
-\citep{Greenland1999, schipf_directed_2011, textor_robust_2017}.
+As the COVID-19 pandemic progresses, research on mechanisms behind the transmission of SARS-CoV-2 shows conflicting evidence \citep{who2020report, Chinazzi395, guan2020clinical}.
+While effects of mobility have been extensively discussed, less is known on other factors such as changing awareness in the population \citep{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020} or the effects of temperature \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}.
+A limiting factor in many studies is the lack of a causal approach to assess the causal contributions of various factors \citep{Greenland1999}.
+This can lead to distorted estimates of the causal factors with observational data \citep{Greenland1999, schipf_directed_2011, textor_robust_2017}.
 
-With COVID-19, we find ourselves in a situation in which information on
-the causal contribution of various influencing factors in the population
-is urgently needed to inform politicians and health authories. On the
-other hand, trials cannot be carried out for obvious ethical and legal
-reasons. Therefore, when assessing the effects of determinants of
-SARS-CoV-2 spread, special attention must be paid to strategies for the
-selection of confounding factors.
+With COVID-19, we find ourselves in a situation in which information on the causal contribution of various influencing factors in the population is urgently needed to inform politicians and health authories.
+On the other hand, trials cannot be carried out for obvious ethical and legal reasons.
+Therefore, when assessing the effects of determinants of SARS-CoV-2 spread, special attention must be paid to strategies for the selection of confounding factors.
 
-Another problem with assessing the effects of various determinants of
-SARS-CoV-2 spread is the heterogeneity of the countries and regions
-examined for example in the Johns Hopkins University (JHU) COVID-19
-database \citep{jhucovid19db2020}. The comparison of time series of case
-numbers from different countries and observational periods can be
-strongly distorted by different factors like testing capacities and
-regional variations.
+Another problem with assessing the effects of various determinants of SARS-CoV-2 spread is the heterogeneity of the countries and regions examined for example in the Johns Hopkins University (JHU) COVID-19 database \citep{jhucovid19db2020}.
+The comparison of time series of case numbers from different countries and observational periods can be strongly distorted by different factors like testing capacities and regional variations.
 
-Our objective is to provide estimates of the causal effects of the main
-drivers of the pandemic with reduced bias. We conducted a scoping review
-of the available studies regarding signaling pathways and determinants
-of the spread of SARS-CoV-2 infections and the reported new COVID-19
-cases. Then we integrated the current findings into a directed acyclic
-graph for the progress of the pandemic at the regional level. Using the
-resulting model and the do-calculus we found identifiable effects
-without blocked causal paths whose effects can be analyzed with
-observational data \citep{Pearl_2014}. We used regional time series data
-of all German districts (\(401\)) from various publicly available
-sources to analyze these questions on a regional level. Germany is a
-good choice in this regard, because it has ample data on contributing
-factors on the regional level and has had high testing and treatment
-capacities from early on in the pandemic.
+Our objective is to provide estimates of the causal effects of the main drivers of the pandemic with reduced bias.
+We conducted a scoping review of the available studies regarding signaling pathways and determinants of the spread of SARS-CoV-2 infections and the reported new COVID-19 cases.
+Then we integrated the current findings into a directed acyclic graph for the progress of the pandemic at the regional level.
+Using the resulting model and the do-calculus we found identifiable effects without blocked causal paths whose effects can be analyzed with observational data \citep{Pearl_2014}.
+We used regional time series data of all German districts (\(401\)) from various publicly available sources to analyze these questions on a regional level.
+Germany is a good choice in this regard, because it has ample data on contributing factors on the regional level and has had high testing and treatment capacities from early on in the pandemic.
 
-\section{Causal Model}\label{causal-model}
+\hypertarget{causal-model}{%
+\section{Causal Model}\label{causal-model}}
 
-We used a directed acyclic graph (DAG)
-\citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze
-the causal relationships between several exposures and SARS-CoV-2
-spread. To get an overview on published associations, a scoping review
-was conducted from 20th to 22nd of May 2020 within Pubmed and Google
-scholar. Restrictions were applied to English and German language and
-the publication date in the last one year. The following search terms
-were applied to abstracts and title in Pubmed (``COVID-19'' OR
-``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and
-connected separately in each case with the exposure variables
-(``mobility'', ``public awareness'', ``awareness'', ``google
-trends'',``ambient temperature'', ``temperature''). For ``mobility'', we
-analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together
-with the first ten pages (100 results) in Google scholar
-(``awareness''/``public awareness''/``google trends'' \(n=9\),
-\(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
-We integrated these findings where possible into the construction of our
-DAG, which can be seen in Figure \ref{fig:dag-covid-19}.
+We used a directed acyclic graph (DAG) \citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread.
+To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar.
+Restrictions were applied to English and German language and the publication date in the last one year.
+The following search terms were applied to abstracts and title in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with the exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'',``ambient temperature'', ``temperature'').
+For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
+We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \ref{fig:dag-covid-19}.
 
@@
 \includegraphics{figures/f_full_covid19_dag.eps}
-\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
-cases on the district level. Unobserved variables are light gray,
-variables marked with an asterisk (*) are confounded by
-weekday/holiday.}
+\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19 cases on the district level.
+Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
 \end{figure}
 
-A number of studies report a strong association of \textbf{mobility}
-restrictions on the number of new COVID-19 cases: Restrictive measures
-(e.g. ``stay-at-home'' orders, travel bans, or school closures) are
-shown to possibly reduce the COVID-19 incidence
-\citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
-However, some studies point out the combination of various
-non-pharmaceutical interventions (NPIs) is decisive to prevent new
-infections \citep{juni_impact_2020, lai_effect_2020}.
+A number of studies report a strong association of \textbf{mobility} restrictions on the number of new COVID-19 cases: Restrictive measures (e.g.~``stay-at-home'' orders, travel bans, or school closures) are shown to possibly reduce the COVID-19 incidence \citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
+However, some studies point out the combination of various non-pharmaceutical interventions (NPIs) is decisive to prevent new infections \citep{juni_impact_2020, lai_effect_2020}.
 
-Google Trends \citep{google_trends} data can be used as a tool to get
-insights into public interest (\textbf{awareness}) in the coronavirus
-disease. Several recent studies imply a connection of relative search
-volumes (RSV) indices and reported new COVID-19 cases
-\citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
-Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
-infected cases/total number of cases by roughly 7 to 14 days for
-different countries
-\citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
-Additionally, we acknowledged that individual risk-aware behavior might
-be a reaction to the current COVID-19 burden (measured as reported cases
-at the day of exposure).
+Google Trends \citep{google_trends} data can be used as a tool to get insights into public interest (\textbf{awareness}) in the coronavirus disease.
+Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases \citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
+Some search terms e.g.~``COVID-19'' or ``coronavirus'' predated newly infected cases/total number of cases by roughly 7 to 14 days for different countries \citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
+Additionally, we acknowledged that individual risk-aware behavior might be a reaction to the current COVID-19 burden (measured as reported cases at the day of exposure).
 
-Mixed evidence is available regarding the effect of
-\textbf{temperature}: On the one hand several papers report an
-association between increase in temperature and decrease in newly
-infected COVID-19 cases
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
-On the other hand, also the opposite has been found
-\citep{auler_evidence_2020, xie_association_2020}. Some studies found no
-association at all
-\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
-It should be noted that few studies considered other confounding
-variables than meteorological ones (especially age and population
-density among others
-\citep{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
-In addition, the transferability of results between different climate
-zones is questionable. To avoid possible bias caused by weather
-variables other than temperature, we included rain, wind, and humidity
-in our model.
+Mixed evidence is available regarding the effect of \textbf{temperature}: On the one hand several papers report an association between increase in temperature and decrease in newly infected COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
+On the other hand, also the opposite has been found \citep{auler_evidence_2020, xie_association_2020}.
+Some studies found no association at all \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
+It should be noted that few studies considered other confounding variables than meteorological ones (especially age and population density among others \citep{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
+In addition, the transferability of results between different climate zones is questionable.
+To avoid possible bias caused by weather variables other than temperature, we included rain, wind, and humidity in our model.
 
-When investigating causal determinants of SARS-CoV-2 infections, a
-number of confounders have to be considered. Well-known risk factors for
-SARS-CoV-2 as well as for other infections are demographic factors such
-as age, gender, socio-economic status (SES), population density, and
-foreign citizenship/ethnicity
-\citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
-Germany along with other countries (i.e.~Brazil, USA, or the UK),
-populist parties or politicians and their electorate tend to be more
-sceptical about effects of containment measures than the other part of
-the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
-Therefore we considered both ``right-wing populist party votes'' and
-``voter turnout'' as possible confounders. Public health interventions
-were also taken into account (contact restrictions, school closures
-etc.), as their implementation showed strong correlations with
-controlling the spread of SARS-CoV-2
-\citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
-bias due to reporting delay of case numbers we had to include weekday
-and German holidays. We include some unobserved variables in our DAG
-(e.g. ``Herd immunity''), too. Please note that ``Exposure to
-SARS-CoV-2'' is itself an unobserved variable: German case numbers are
-reported with delay after date of exposure and symptom onset.
-\emph{Exposure to the virus} should not be confused with the formal
-\emph{exposure variables} of the DAG (mobility, awareness, temperature).
+When investigating causal determinants of SARS-CoV-2 infections, a number of confounders have to be considered.
+Well-known risk factors for SARS-CoV-2 as well as for other infections are demographic factors such as age, gender, socio-economic status (SES), population density, and foreign citizenship/ethnicity \citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}.
+In Germany along with other countries (i.e.~Brazil, USA, or the UK), populist parties or politicians and their electorate tend to be more sceptical about effects of containment measures than the other part of the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
+Therefore we considered both ``right-wing populist party votes'' and ``voter turnout'' as possible confounders.
+Public health interventions were also taken into account (contact restrictions, school closures etc.), as their implementation showed strong correlations with controlling the spread of SARS-CoV-2 \citep{cowling2020impact, juni_impact_2020, lai_effect_2020}.
+To avoid bias due to reporting delay of case numbers we had to include weekday and German holidays.
+We include some unobserved variables in our DAG (e.g.~``Herd immunity''), too.
+Please note that ``Exposure to SARS-CoV-2'' is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset.
+\emph{Exposure to the virus} should not be confused with the formal \emph{exposure variables} of the DAG (mobility, awareness, temperature).
 
-\section{Data}\label{data}
+\hypertarget{data}{%
+\section{Data}\label{data}}
 
-We collected and aggregated data on reported COVID-19 cases, regional
-socio-demographic factors, weather, and general mobility on district and
-state level in Germany for the period of 15 February 2020 to 8 July
-2020. Our observation period for the outcome consisted of all dates from
-23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
-days for all confounders. We did not exclude any states or districts
-(\(K=401\)). We analyzed the daily reported number of new cases as
-outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
-predictors was derived from our causal DAG (see Table
-\ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to
-modelling and data limitations, some of the predictors were unobserved
-or were modelled as a construct consisting of several variables. For our
-causal analysis, we computed adjustment sets in three different
-scenarios for separate exposures within the DAG: i) mobility of
-population, ii) awareness of COVID-19 (i.e.~Google searches for
-``corona''), iii) weather (i.e.~temperature).
+We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
+Our observation period for the outcome consisted of all dates from 23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8 days for all confounders.
+We did not exclude any states or districts (\(K=401\)).
+We analyzed the daily reported number of new cases as outcome (\(K\cdot T=54{\ }937\) observations).
+The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}).
+Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables.
+For our causal analysis, we computed adjustment sets in three different scenarios for separate exposures within the DAG: i) mobility of population, ii) awareness of COVID-19 (i.e.~Google searches for ``corona''), iii) weather (i.e.~temperature).
 
-\subsection{Variables}\label{variables}
+\hypertarget{variables}{%
+\subsection{Variables}\label{variables}}
 
-We downloaded German daily case numbers on district level reported by
-Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July
-2020) and aggregated them by date. The number of daily active cases for
-day \(d\) was derived by subtracting the total number of reported cases
-on day \(d\) and day \(d-14\) (14 days as a conservative estimate for
-the infectious period, which corresponds here to the required quarantine
-time in Germany).
+We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July 2020) and aggregated them by date.
+The number of daily active cases for day \(d\) was derived by subtracting the total number of reported cases on day \(d\) and day \(d-14\) (14 days as a conservative estimate for the infectious period, which corresponds here to the required quarantine time in Germany).
 
-To assess the mobility of the German population, we used data publicly
-available on German state level from Google \citep{google_mobility}.
-Measurements are daily relative changes of mobility in percent compared
-to the period of 3 January 2020 to 6 February 2020. Missing values
-(\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the
-state level measurements were passed onto districts within the
-corresponding state. Google mobility data was available for six
-different sectors of daily life (``retail and recreation'', ``grocery
-and pharmacy'', ``parks'', ``transit stations'', ``workplaces'',
-``residential'') which means that ``mobility'' is a construct consisting
-of several variables. All variables but ``residential'' mobility are
-relative changes of daily visitor numbers to the corresponding sectors
-compared to the reference period. ``Residential'' mobility is the
-relative change of daily time spent at residential areas.
+To assess the mobility of the German population, we used data publicly available on German state level from Google \citep{google_mobility}.
+Measurements are daily relative changes of mobility in percent compared to the period of 3 January 2020 to 6 February 2020.
+Missing values (\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the state level measurements were passed onto districts within the corresponding state.
+Google mobility data was available for six different sectors of daily life (``retail and recreation'', ``grocery and pharmacy'', ``parks'', ``transit stations'', ``workplaces'', ``residential'') which means that ``mobility'' is a construct consisting of several variables.
+All variables but ``residential'' mobility are relative changes of daily visitor numbers to the corresponding sectors compared to the reference period.
+``Residential'' mobility is the relative change of daily time spent at residential areas.
 
-The notion of awareness in the population of COVID-19 describes the
-general state of alertness about the new infectious disease. As such, it
-was hard to measure directly. As a proxy, we used the relative interest
-in the topic term ``corona'' as indicated by Google searches. The daily
-data was available on state level \citep{google_trends} and passed onto
-district level. As a second proxy for awareness, we used the daily
-reported number of COVID-19 cases on the day of the exposure: Since
-media reported case numbers prominently, we assumed that this could
-reflect individual awareness, too.
+The notion of awareness in the population of COVID-19 describes the general state of alertness about the new infectious disease.
+As such, it was hard to measure directly.
+As a proxy, we used the relative interest in the topic term ``corona'' as indicated by Google searches.
+The daily data was available on state level \citep{google_trends} and passed onto district level.
+As a second proxy for awareness, we used the daily reported number of COVID-19 cases on the day of the exposure: Since media reported case numbers prominently, we assumed that this could reflect individual awareness, too.
 
-We constructed daily weather from four variables (``temperature'',
-``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
-Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
-stations in Germany below 1000 meters altitude with daily records in
-2020 until 8 July 2020. District level daily weather data was aggregated
-per district by averaging the data from the three nearest weather
-stations (which includes weather stations inside the district). Missing
-values were imputed with mean values (\(n=59\) for wind).
+We constructed daily weather from four variables (``temperature'', ``rainfall'', ``humidity'', ``wind'').
+Weather data was downloaded from Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather stations in Germany below 1000 meters altitude with daily records in 2020 until 8 July 2020.
+District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district).
+Missing values were imputed with mean values (\(n=59\) for wind).
 
-The reported number of COVID-19 cases varied strongly by day of the
-week. Thus, we included ``weekday'' as a categorical variable.
-Similarly, the reported cases and the exposure to the virus were
-affected by official holidays. Within the observation period, this
-included among others Good Friday, Easter Monday, and Labor Day. To
-correct for effects of these days, we included two variables in the
-model, ``Holiday (report)'' (indicates if the day of the report was a
-holiday, because governmental health departments were less likely to be
-on full duty) and ``Holiday (exposure)'' (indicates if the day of
-exposure to the virus was a holiday, because the population behaves
-differently on holidays).
+The reported number of COVID-19 cases varied strongly by day of the week.
+Thus, we included ``weekday'' as a categorical variable.
+Similarly, the reported cases and the exposure to the virus were affected by official holidays.
+Within the observation period, this included among others Good Friday, Easter Monday, and Labor Day.
+To correct for effects of these days, we included two variables in the model, ``Holiday (report)'' (indicates if the day of the report was a holiday, because governmental health departments were less likely to be on full duty) and ``Holiday (exposure)'' (indicates if the day of exposure to the virus was a holiday, because the population behaves differently on holidays).
 
-For different official and political measures we used one-hot encoded
-daily variables, i.e.~ban of mass gatherings, school and kindergarten
-closures and their gradual reopening, contact restrictions, and
-mandatory face masks for shopping and public transport.
+For different official and political measures we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
 
-We included several social, economic, and demographic factors on the
-district level with direct or indirect influence on the risk of exposure
-to SARS-CoV-2 in our analysis. All are readily available from INKAR
-database \citep{inkar}. We used the share of population that is 65 years
-or older and the share of population that is younger than 18 years
-(Age), the share of females in population (Gender), the population
-density, the share of foreign citizenships and the share of the
-population seeking refuge (Foreign citizenship), the share of low-income
-households (Socio-economic status), voter turnout, share of right-wing
-populist party votes, and the number of nursing (retirement) homes.
+We included several social, economic, and demographic factors on the district level with direct or indirect influence on the risk of exposure to SARS-CoV-2 in our analysis.
+All are readily available from INKAR database \citep{inkar}.
+We used the share of population that is 65 years or older and the share of population that is younger than 18 years (Age), the share of females in population (Gender), the population density, the share of foreign citizenships and the share of the population seeking refuge (Foreign citizenship), the share of low-income households (Socio-economic status), voter turnout, share of right-wing populist party votes, and the number of nursing (retirement) homes.
 
@@
 
-All variables but the outcome ``Reported new cases of COVID-19'' and the
-offset ``Active cases'' were centered for numerical stability. We did
-not scale variables to unit variance to maintain interpretability of
-effects on the original scale of variables. Additionally, we lagged the
-effect of all variables (but outcome, offset, and the non-dynamic
-socio-demographic variables) by 8 days (see Section \ref{results}) which
-means that we assumed that their effects on the outcome will be visible
-after 8 days.
+All variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered for numerical stability.
+We did not scale variables to unit variance to maintain interpretability of effects on the original scale of variables.
+Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 8 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 8 days.
 
-\section{Methods}\label{methods}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-\subsection{Causal analysis with DAG and adjustment
-sets}\label{causal-analysis-with-dag-and-adjustment-sets}
+\hypertarget{causal-analysis-with-dag-and-adjustment-sets}{%
+\subsection{Causal analysis with DAG and adjustment sets}\label{causal-analysis-with-dag-and-adjustment-sets}}
 
-We used a directed acyclic graph as a graphical representation of the
-hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
-virus, onset of COVID-19, and finally reports of COVID-19 cases. Every
-node \(v_i\) in the graph is the graphical representation of an observed
-or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow
-from node \(v_i\) to \(v_j\) that implies a direct causal relationship
-from variable \(x_i\) onto variable \(x_j\). The set of all nodes is
-denoted by \(V\), the set of all edges by \(E\), as such, the complete
-DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
-\citep{spirtes2000causation, pearl2009causality} introduce the theory of
-causal analysis, do-calculus, and how to analyze a DAG to estimate the
-total or direct causal effect from a variable \(x_i\) onto a variable
-\(x_j\). The direct effect is the effect associated with the edge
-\(e_{ij}\) only (if it exists), while the total effect takes indirect
-effects via other paths from \(v_i\) to \(v_j\) into account, too. Here
-we estimated total effects only, since most of our variables were not
-hypothesized to have a direct effect on the \emph{reported} number of
-new COVID-19 cases. In contrast to prediction tasks, where one would
-include all variables available, it is actually ill-advised to use all
-available variables to estimate causal effects, due to introducing bias
-by adjusting for unnecessary variables within the causal DAG. This is
-why we need to identify a valid set of necessary variables (an
-adjustment set) to estimate the proper causal effect
-\citep{pearl2009causality}. The ``minimal adjustment set''
-\citep{greenland1999causal} is a valid adjustment set of variables that
-does not contain another valid adjustment set as a subset. However,
-identifying a minimal adjustment set might not be enough to reliably
-estimate the causal effect. Thus, we identified the ``optimal adjustment
-set'' \citep{henckel2019graphical} as the set of variables which is a
-valid adjustment set while having the lowest asymptotic variance in the
-resulting causal effect estimates.
+We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases.
+Every node \(v_i\) in the graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\).
+The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\).
+The seminal works of Spirtes and Pearl \citep{spirtes2000causation, pearl2009causality} introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable \(x_i\) onto a variable \(x_j\).
+The direct effect is the effect associated with the edge \(e_{ij}\) only (if it exists), while the total effect takes indirect effects via other paths from \(v_i\) to \(v_j\) into account, too.
+Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the \emph{reported} number of new COVID-19 cases.
+In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG.
+This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect \citep{pearl2009causality}.
+The ``minimal adjustment set'' \citep{greenland1999causal} is a valid adjustment set of variables that does not contain another valid adjustment set as a subset.
+However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect.
+Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest asymptotic variance in the resulting causal effect estimates.
 
-We analyzed the DAG from Section \ref{causal-model} with the R Software
-\citep{rsoftware} and the R packages \texttt{dagitty} (formal
-representation of the graph and minimal adjustment sets
-\citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
-adjustment set \citep{pcalg}). For the defined exposures and the outcome
-``Reported new cases of COVID-19'', we computed the minimal and optimal
-adjustment sets. Since it was possible that these sets contained
-unobserved variables that needed to be left out of the regression model,
-we chose the valid set with the highest pseudo-\(R^2\) (see next
-section) to estimate the final total causal effect from exposure to
-outcome.
+We analyzed the DAG from Section \ref{causal-model} with the R Software \citep{rsoftware} and the R packages \texttt{dagitty} (formal representation of the graph and minimal adjustment sets \citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal adjustment set \citep{pcalg}).
+For the defined exposures and the outcome ``Reported new cases of COVID-19'', we computed the minimal and optimal adjustment sets.
+Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the highest pseudo-\(R^2\) (see next section) to estimate the final total causal effect from exposure to outcome.
 
-\subsection{Regression with negative binomial
-model}\label{regression-with-negative-binomial-model}
+\hypertarget{regression-with-negative-binomial-model}{%
+\subsection{Regression with negative binomial model}\label{regression-with-negative-binomial-model}}
 
-We can estimate the causal effect from exposure to outcome by regression
-\citep{pearl2009causality}. Since the outcome ``Reported new cases of
-COVID-19'' is a count variable, one should not employ a linear
-regression model with Gaussian errors, but instead we assumed a
-log-linear relationship between the expected value of the outcome \(Y\)
-(new cases) and regressors \(x\), as well as a Poisson or negative
-binomial distribution for \(Y\):
-
+We can estimate the causal effect from exposure to outcome by regression \citep{pearl2009causality}.
+Since the outcome ``Reported new cases of COVID-19'' is a count variable, one should not employ a linear regression model with Gaussian errors, but instead we assumed a log-linear relationship between the expected value of the outcome \(Y\) (new cases) and regressors \(x\), as well as a Poisson or negative binomial distribution for \(Y\):
 \begin{equation}
@@
 \end{equation}
+where \(\alpha\) is the regression intercept, \(S\) is the set of adjustment variables for the exposure \(i^{\ast}\) including the exposure variable itself, \(\beta_i\) are the regression coefficients corresponding to the variables \(x_i\).
+As such \(\beta_{i^{\ast}}\) is the total causal effect from exposure variable \(x_{i^{\ast}}\) on the outcome Y.
 
-where \(\alpha\) is the regression intercept, \(S\) is the set of
-adjustment variables for the exposure \(i^{\ast}\) including the
-exposure variable itself, \(\beta_i\) are the regression coefficients
-corresponding to the variables \(x_i\). As such \(\beta_{i^{\ast}}\) is
-the total causal effect from exposure variable \(x_{i^{\ast}}\) on the
-outcome Y.
+The Poisson regression assumes equality of mean and variance.
+If this is not the case one observes so-called overdispersion (the variance is higher than the mean), this indicates one should use regression with a negative binomial distribution instead to estimate the variance parameter separately from the mean.
 
-The Poisson regression assumes equality of mean and variance. If this is
-not the case one observes so-called overdispersion (the variance is
-higher than the mean), this indicates one should use regression with a
-negative binomial distribution instead to estimate the variance
-parameter separately from the mean.
-
-We needed to account for the fact that our outcome is not counted per
-time unit (one day) only, but depends on the number of active COVID-19
-cases: Holding all other variables fixed, the number of new cases \(Y\)
-is a constant proportion of the number of active cases \(A\). This was
-modeled by including an offset \(\log(A+1)\) in the regression model
-\eqref{eq:regmodel}:
-
+We needed to account for the fact that our outcome is not counted per time unit (one day) only, but depends on the number of active COVID-19 cases: Holding all other variables fixed, the number of new cases \(Y\) is a constant proportion of the number of active cases \(A\).
+This was modeled by including an offset \(\log(A+1)\) in the regression model \eqref{eq:regmodel}:
 \begin{align}
@@
 \end{align}
+Here we added a pseudocount ``\(+1\)'' to ensure a finite logarithm and avoid division by \(0\).
 
-Here we added a pseudocount ``\(+1\)'' to ensure a finite logarithm and
-avoid division by \(0\).
+One can interpret the model as approximating the log-ratio of new cases and active cases by a linear combination of the regressor variables \eqref{eq:logratemodel}.
+If all variables \(x_i\) are centered in \eqref{eq:ratechange}, we have for the baseline \(\forall i\ x_i=0 \Rightarrow \mathbb{E}[Y|x=0]=\exp(\alpha)\cdot (A+1)\).
+In other words, the exponentiated intercept is the baseline daily infection rate (how many people does one infected individual infect in one day).
+If we hold all variables \(x_i\) fixed (e.g.~at baseline 0) in \eqref{eq:ratechange} but now increase the exposure variable \(x_{i^{\ast}}=0\) by one unit to \(x_{i^{\ast}}+1=0+1\), we have \(\mathbb{E}[Y|x']=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}})\), which means the exponentiated coefficient \(\beta_{i^{\ast}}\) describes the rate change of the outcome by one unit increase of the exposure.
 
-One can interpret the model as approximating the log-ratio of new cases
-and active cases by a linear combination of the regressor variables
-\eqref{eq:logratemodel}. If all variables \(x_i\) are centered in
-\eqref{eq:ratechange}, we have for the baseline
-\(\forall i\ x_i=0 \Rightarrow \mathbb{E}[Y|x=0]=\exp(\alpha)\cdot (A+1)\).
-In other words, the exponentiated intercept is the baseline daily
-infection rate (how many people does one infected individual infect in
-one day). If we hold all variables \(x_i\) fixed (e.g.~at baseline 0) in
-\eqref{eq:ratechange} but now increase the exposure variable
-\(x_{i^{\ast}}=0\) by one unit to \(x_{i^{\ast}}+1=0+1\), we have
-\(\mathbb{E}[Y|x']=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}})\),
-which means the exponentiated coefficient \(\beta_{i^{\ast}}\) describes
-the rate change of the outcome by one unit increase of the exposure.
+In practice, given observations of \(Y\) and \(x\) we estimate the regression coefficients \(\alpha\) and \(\beta_i\) by \emph{maximum likelihood} \citep{maxlikelihood}.
+Our observational measurements are \(y_{kt}\) and \(x_{ikt}\), where \(k\) indicates the corresponding district and \(t\) the date of measurement.
 
-In practice, given observations of \(Y\) and \(x\) we estimate the
-regression coefficients \(\alpha\) and \(\beta_i\) by \emph{maximum
-likelihood} \citep{maxlikelihood}. Our observational measurements are
-\(y_{kt}\) and \(x_{ikt}\), where \(k\) indicates the corresponding
-district and \(t\) the date of measurement.
+When we analyzed different adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets), we first checked if the set included unobserved variables.
+If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \citep{perkovi2015complete}).
+If a minimal adjustment set contained unobserved variables, we discarded the whole set.
+We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for every remaining valid adjustment set as regressors and calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the sum of squared prediction errors of the current model and \(V_0\) is the sum of squared prediction errors of the null model (intercept and offset only).
+That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of variance unexplained.
+Finally, we decided for the model/adjustment set with the highest pseudo-\(R^2\).
+We report the exponentiated estimated coefficients along with 99 percent confidence intervals of the estimates.
 
-When we analyzed different adjustment sets given by analysis of the
-causal DAG (i.e.~the minimal and optimal adjustment sets), we first
-checked if the set included unobserved variables. If this was the case
-for the optimal adjustment set, we discarded the unobserved variables
-from the set and checked if it was still a valid adjustment set
-(function \texttt{gac} in package \texttt{pcalg}
-\citep{perkovi2015complete}). If a minimal adjustment set contained
-unobserved variables, we discarded the whole set. We conducted a
-log-linear regression (function \texttt{glm} with
-\texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb}
-from the \texttt{MASS} package for the negative binomial regression
-\citep{mass}) for every remaining valid adjustment set as regressors and
-calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the
-sum of squared prediction errors of the current model and \(V_0\) is the
-sum of squared prediction errors of the null model (intercept and offset
-only). That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of
-variance unexplained. Finally, we decided for the model/adjustment set
-with the highest pseudo-\(R^2\). We report the exponentiated estimated
-coefficients along with 99 percent confidence intervals of the
-estimates.
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-\section{Results}\label{results}
-
-Descriptive statistics for the included variables are presented in Table
-\ref{tab:descriptive-vars}.
+Descriptive statistics for the included variables are presented in Table \ref{tab:descriptive-vars}.
 
@@
 \includegraphics{figures/f_timeseries_cases_and_lograte.eps}
-\caption{\label{fig:cases-log}Temporal and district level variation of
-outcome (log-scale)}
+\caption{\label{fig:cases-log}Temporal and district level variation of outcome (log-scale)}
 \end{figure}
 
-In the observational period, the number of daily reported COVID-19 cases
-increased till the end of March/beginning of April and continually
-decreased afterwards till the beginning of June 2020 with a slight
-increase and decrease afterwards (Figure \ref{fig:cases-log}A). On the
-other hand, the (log-)ratio of reported cases over active cases
-decreased steeply till the mid of April and increased steadily
-afterwards with a slight decrease close to the end of the observation
-period (Figure \ref{fig:cases-log}B). Both figures examplify a
-considerable variation among the districts (light blue points are
-individual district's data).
+In the observational period, the number of daily reported COVID-19 cases increased till the end of March/beginning of April and continually decreased afterwards till the beginning of June 2020 with a slight increase and decrease afterwards (Figure \ref{fig:cases-log}A).
+On the other hand, the (log-)ratio of reported cases over active cases decreased steeply till the mid of April and increased steadily afterwards with a slight decrease close to the end of the observation period (Figure \ref{fig:cases-log}B).
+Both figures examplify a considerable variation among the districts (light blue points are individual district's data).
 
-In Germany, we observed a rebound in mobility after the initial
-political measures, reductions in incident cases were associated with a
-diminishing public interest in COVID-19, and temperatures were overall
-increasing (cf.~Figure \ref{fig:timeseries}); with correlations between
-temporal progression and mobility in retail and recreation \(r_{A,D}=\)
--0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and
-temperature \(r_{A,B}=\) 0.79.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) -0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and temperature \(r_{A,B}=\) 0.79.
 
@@
 \includegraphics{figures/f_timeseries_exposures.eps}
-\caption{\label{fig:timeseries}Temporal variation of outcome and main
-determinants}
+\caption{\label{fig:timeseries}Temporal variation of outcome and main determinants}
 \end{figure}
 
-\subsection{Main results}\label{main-results}
+\hypertarget{main-results}{%
+\subsection{Main results}\label{main-results}}
 
@@
 
-We list the results of our causal analysis for the effects of our
-variables in Table \ref{tab:causaleffects}. The estimates are
-multiplicative rates of increase/decrease for a one unit increase of the
-respective variable: Values above 1 lead to an increase, below 1 to a
-decrease of the infection rate. To put these estimates into perspective,
-Figure \ref{fig:effectsrange} shows the relative causal effect of the
-different exposure variables on the number of reported COVID-19 cases on
-a range of sensible values of the exposure variables (95 percent
-quantiles of data points).
+We list the results of our causal analysis for the effects of our variables in Table \ref{tab:causaleffects}.
+The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate.
+To put these estimates into perspective, Figure \ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).
 
@@
 
-Within our framework, we saw strong effects for mobility in
-retail/recreational areas and essential shopping (grocery and pharmacy).
-In the former, an increase of 1 percent point mobility compared to the
-reference period (03 January to 06 February 2020) leads to an increase
-of the daily reported case number by about 0.8 percent. Contrarily, a
-corresponding increase of 1 percent point for the areas of
-grocery/pharmacy leads to a decrease in the reported case number by
-approximately 0.5 percent. Mobility on workplaces showed a small effect
-of 0.3 increase in case numbers for every 1 percent point increase in
-mobility. Other causal effects of mobility were insubstantial and not
-consistent in their direction (99 percent confidence intervals of
-estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
-mobility on a range of possible values. Thus, we expect an increase of
-daily cases by approximately 23 percent if mobility in retail/recreation
-reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10
-percent points compared to the reference period leads to a reduction of
-the infection rate by approximately 7 percent.
+Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
+In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent.
+Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent.
+Mobility on workplaces showed a small effect of 0.3 increase in case numbers for every 1 percent point increase in mobility.
+Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
+Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
+Thus, we expect an increase of daily cases by approximately 23 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 7 percent.
 
 ``Awareness'' had two opposite effects on the outcome in our DAG.
-Awareness measured by Google searches for \emph{corona} had a positive
-effect on the number of reported cases. An one percent point increase of
-the state's Google searches (relative to other states and the
-observation period) leads to an increase of approximately 1 percent. For
-example, if a district shows 10 percent points more relative searches
-for \emph{corona} then another one, we expect approximately 11 percent
-more infections for this district after 8 days. \emph{COVID-19 burden}
-(reported number of cases on day of exposure) affected the outcome
-negatively, where every additional daily case in the district leads to a
-0.6 percent decrease in newly reported case numbers. The last plot in
-Figure \ref{fig:effectsrange} visualizes this relationship: For a local
-outbreak with 20 daily cases as COVID-19 burden, we estimate as total
-causal effect a subsequent reduction of infection rate by 9 percent.
+Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} then another one, we expect approximately 11 percent more infections for this district after 8 days.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.6 percent decrease in newly reported case numbers.
+The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 9 percent.
 
-Within our model, we observed a causal effect of temperature and all
-other weather variables. Every increase of 1 degree Celsius in
-temperature leads to a reduction of the daily reported case numbers by
-approximately 0.8 percent. Similarly strong was the effect of rainfall:
-One millimeter (=1 liter per square meter) more rainfall leads to an
-increase of reported case numbers by approximately 1.1 percent. We
-observe small effects for humidity and wind as well (higher humidity
-leading to fewer cases, stronger wind leading to less cases). In
-perspective (Figure \ref{fig:effectsrange}), with temperature we expect
-an increase by approximately 9 percent at a daily average temperature of
-\(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm
-rainfall a corresponding increase of the infection rate by approximately
-2 percent.
+Within our model, we observed a causal effect of temperature and all other weather variables.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.8 percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent.
+We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases).
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 9 percent at a daily average temperature of \(0^{\circ}C\).
+For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 2 percent.
 
-In all cases we opted to use the reduced optimal adjustment set over the
-minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly
-above 0.3), except for mobility, where the minimal adjustment set had a
-higher pseudo-\(R^2\). Notably, these sets always include most of our
-socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets})
-as well as the policy variables (except for COVID-19 burden). We also
-decided for the lag of 8 days based on the highest pseudo-\(R^2\) values
-compared to other lags on the chosen adjustment sets. Similarly,
-negative binomial regression was chosen over Poisson regression, because
-the latter showed overdispersion and overall lower pseudo-\(R^2\)
-values.
+In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\).
+Notably, these sets always include most of our socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets}) as well as the policy variables (except for COVID-19 burden).
+We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets.
+Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-\(R^2\) values.
 
@@
 
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
-\subsection{Main findings}\label{main-findings}
+\hypertarget{main-findings}{%
+\subsection{Main findings}\label{main-findings}}
 
-Our objective was to identify causal effects for COVID-19 cases. We
-found that weather affects the reported number of infections, especially
-temperature (which has a reducing effect on case numbers) and rainfall
-(which increases case numbers). We saw that reports of high case numbers
-in districts led to a reduction in new infection numbers, which
-indicates risk-averse awareness in the population. The overall effect of
-mobility showed no consistent effect, however, in specific areas
-significant causal effects could be measured: Increasing activity in
-retail and recreational areas increased reported case numbers, while
-increased movement for essential shopping (grocery and pharmacy) led to
-reduced case numbers.
+Our objective was to identify causal effects for COVID-19 cases.
+We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers).
+We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population.
+The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.
 
-Furthermore, we made a strong case for the use of causal DAGs in
-epidemiology and a pandemic like COVID-19: DAGs allow to choose
-confounders for the analysis in a principled and statistically correct
-way while reducing possible causes for bias. Also, the DAG formalization
-allows for discussion about the underlying causal structure.
+Furthermore, we made a strong case for the use of causal DAGs in epidemiology and a pandemic like COVID-19: DAGs allow to choose confounders for the analysis in a principled and statistically correct way while reducing possible causes for bias.
+Also, the DAG formalization allows for discussion about the underlying causal structure.
 
-\subsection{Comparison with previous
-research}\label{comparison-with-previous-research}
+\hypertarget{comparison-with-previous-research}{%
+\subsection{Comparison with previous research}\label{comparison-with-previous-research}}
 
-Most research on determinants affecting case numbers of COVID-19 is
-restricted to single aspects
-\citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
-To reliably identify causal drivers, one needs to adjust for
-confounders. To this end, we used an integrated model with variables
-from different aspects like mobility, awareness, weather, or
-socio-demographics and identified confounders by causal analysis with a
-directed acyclic graph. A causal approach is used in another current
-COVID-19 analysis \citep{gencoglu2020causal}. There, however, they
-identify the causal relationships (reconstruct a DAG), while we
-estimated causal effects for a given hypothesized DAG.
+Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects \citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
+To reliably identify causal drivers, one needs to adjust for confounders.
+To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
+A causal approach is used in another current COVID-19 analysis \citep{gencoglu2020causal}.
+There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.
 
-Several studies assessing the impact of public health measures on
-mobility have each observed a downward trend accompanied by a decrease
-in the number of newly reported cases
-\citep{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
+Several studies assessing the impact of public health measures on mobility have each observed a downward trend accompanied by a decrease in the number of newly reported cases \citep{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
 
-Our findings regarding awareness/Google Trends analysis are in good
-agreement with the correlations found by
-\citet{effenberger_association_2020}, \citet{higgins_correlations_2020},
-and \citet{yuan_trends_2020}, who conclude that alertness to COVID-19
-rises several days before the highest number of cases are reported. At
-this point it should be noted, that awareness is substantially
-influenced by public media coverage, which should be considered, if
-possible, in future studies \citep{higgins_correlations_2020}. As such,
-awareness is difficult to measure and here the number of Google searches
-for ``corona'' could only be a proxy for this concept.
+Our findings regarding awareness/Google Trends analysis are in good agreement with the correlations found by \citet{effenberger_association_2020}, \citet{higgins_correlations_2020}, and \citet{yuan_trends_2020}, who conclude that alertness to COVID-19 rises several days before the highest number of cases are reported.
+At this point it should be noted, that awareness is substantially influenced by public media coverage, which should be considered, if possible, in future studies \citep{higgins_correlations_2020}.
+As such, awareness is difficult to measure and here the number of Google searches for ``corona'' could only be a proxy for this concept.
 
-In addition, in alignment with other recent published studies, our
-results confirm evidence which associated a negative effect of
-temperature on new COVID-19 cases
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
-It is however controversial to other scientific literature describing no
-effects
-\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
-or even converse correlations
-\citep{auler_evidence_2020, xie_association_2020}. The conflicting
-results might be explained by different climates and characteristics of
-the populations under study. While we are confident that our strict
-causal analysis resulted in effect estimates as undistorted as possible,
-there might be unconsidered bias in those other studies. Further
-research needs to be done to elucidate the biological characteristics of
-the novel virus SARS-CoV-2 regarding its ambient temperature survival
-and transmission. In regards of humidity, our findings (higher humidity
-leading to increased case numbers) are in agreement with previous
-research \citep{qi_covid-19_2020, wu_effects_2020, juni_impact_2020}.
-Finally, we found a positive causal effect of increment precipitation
-and a raise in COVID-19 cases, which supports previous observations
-\citep{sobral_association_2020}.
+In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
+It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}.
+The conflicting results might be explained by different climates and characteristics of the populations under study.
+While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies.
+Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
+Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
 
-\subsection{Limitations and strengths}\label{limitations-and-strengths}
+\hypertarget{limitations-and-strengths}{%
+\subsection{Limitations and strengths}\label{limitations-and-strengths}}
 
-While use of a causal DAG is itself a strong tool to identify causal
-effects (and not just statistical associations), it introduces two
-limitations: causal assumptions within the graph (depicted by edges)
-need to be well justified, and the statistical regression model that
-calculates total causal effects needs to be appropriate for the task at
-hand. We endorse our graph as a basis for discussion on residual
-confounding.
+While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
+We endorse our graph as a basis for discussion on residual confounding.
+We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}).
+As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks).
+Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}.
+For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets.
+A further analysis of these implications is out of the scope of this paper.
 
-We observed overdispersion and a substantial increase in model
-performance with a negative binomial regression compared to Poisson
-regression, which is in line with the results on COVID-19 daily case
-counts of \citet{kraemer_effect_2020} and others
-\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020}. We did not
-model case counts with a differential equation model like the classic
-SIR-model \citep{kermack1991contributions} and its successors, since
-these are more suited to prediction \citep[e.g.][]{rkimodellierung}
-while our choice of a negative binomial regression framework allowed us
-to estimate the effects of confounders more reliably. There are more
-advanced statistical methods for count data, e.g.~zero-inflated models
-and mixed models. We tested both approaches as extensions to the
-negative binomial regression and experienced numerical problems and
-increased computing time, along with an insubstantial increase in model
-performance. Furthermore, our model assumed that all variables have
-effects proportional to the size of their measurements. It is possible
-that some variables show saturation effects or opposite effects for
-lower, medium, or high values. This could be modelled with polynomial or
-other transformations of the variables, which we did not employ due to
-limited temporal and spatial data availability.
+We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
+We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
+There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models.
+We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance.
+Furthermore, our model assumed that all variables have effects proportional to the size of their measurements.
+It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values.
+This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
+By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).
 
-As we were using administrative data for our analysis, the results are
-susceptible to the Modifiable Area Unit Problem (MAUP)
-\citep{Openshaw1984}. The MAUP postulates that different regional
-aggregations of the units of observation may lead to different results
-and conclusions. Due to limited available data for the different
-variables, there is currently no way to overcome this.
+We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level.
+As such, conclusions about effects cannot be transferred on individuals without the possibility for an ecological fallacy.
+Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) \citep{Openshaw1984}.
+The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions.
+Due to limited available data for the different variables, there is currently no way to overcome these problems that are inherent to all analyses on aggregated data level.
 
-Our observation period was restricted to succession from late winter to
-spring and summer (February to July). Nevertheless, this transition with
-increasing temperature was a natural experiment that allowed clues on
-weather effects.
+Our observation period was restricted to succession from late winter to spring and summer (February to July).
+Nevertheless, this transition with increasing temperature was a natural experiment that allowed clues on weather effects.
 
-We could not include data on health care utilization during the pandemic
-into our models due to the lack of available resources. This is planned
-for a later follow up to this paper since we rank health care
-utilization and mobility within health care facilities among the strong
-factors for COVID-19 progression: personnel in hospitals and private
-practices is particularly exposed to infection, while the lack of
-adequate care for other diseases has severe effects on general health of
-the population. At the same time, health care facilities are key for
-testing and surveillance of COVID-19 patients.
+We could not include data on health care utilization during the pandemic into our models due to the lack of available resources.
+This is planned for a later follow up to this paper since we rank health care utilization and mobility within health care facilities among the strong factors for COVID-19 progression: personnel in hospitals and private practices is particularly exposed to infection, while the lack of adequate care for other diseases has severe effects on general health of the population.
+At the same time, health care facilities are key for testing and surveillance of COVID-19 patients.
 
-While our analysis focused on Germany and its districts, we assume that
-results may be transferred to other countries by adjusting for their
-respective weather conditions, mobility habits, socio-demographic
-characteristics, and other determining factors.
+While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.
 
-The code and resources for our analysis are available on Github, we
-invite other researchers to replicate our analysis with different
-assumptions using the files provided in the repository\footnote{The
-  repository is located here: https://github.com/zidatalab/causalcovid19}
-of the article.
+The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository\footnote{The repository is located here: https://github.com/zidatalab/causalcovid19} of the article.
 
-\subsection{Discussion of causal
-effects}\label{discussion-of-causal-effects}
+\hypertarget{discussion-of-causal-effects}{%
+\subsection{Discussion of causal effects}\label{discussion-of-causal-effects}}
 
-In our analysis, the adverse effect of mobility in retail/recreation and
-the favorable effect of mobility in grocery/pharmacy make a strong case
-for the successful implementation of policies like contact restrictions
-which limit the number of individual interactions that lead to
-infections. This is due to retail/recreational areas encompassing mostly
-places of social gatherings like restaurants and bars, malls, sports and
-music venues, among others, while if people are doing more of their
-essential shopping at supermarkets, they will most likely stay at home
-with less contact to other people.
+In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
+This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.
 
-The causal effects of awareness measured via searches for ``corona'' and
-the COVID-19 burden are harder to interpret. We assume that within our
-model, the searches for ``corona'' are an insufficient proxy for
-awareness, while the decreasing effect for future case numbers of high
-daily COVID-19 burden numbers indicates it affects individual
-risk-behavior as a deterrent.
+The causal effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret.
+We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.
 
-Similarly, the effects of temperature and rainfall can be interpreted as
-causal effects for indoor and outdoor activities, such that higher
-temperatures and low rainfall indicate more people spending time outdoor
-while lower temperatures and high rainfall result in indoor activities,
-which lead to more infections. In this light, we advocate for
-precautious measures like increased hygiene, face masks, and air
-ventilation for unavoidable indoor activities.
+Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections.
+In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.
 
-\subsection{Conclusions}\label{conclusions}
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
-To the best of our knowledge, this is the most comprehensive analysis of
-causes for COVID-19 infections which integrates different data sources
-(all publicly available). Causal reasoning with a DAG allows us to
-estimate the causal effects reliably.
+To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available).
+Causal reasoning with a DAG allows us to estimate the causal effects more reliably.
 
-Our findings suggest that the causal effects of mobility, awareness, and
-weather need to be taken strongly into account when deciding for
-mitigation and suppression measures, depending on the recent and future
-COVID-19 pandemic development.
+Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
 

### 2020-07-22 manuscript/v2_medrxiv/manuscript.tex
insert acknowledgments

@@
 Furthermore, we made a strong case for the use of causal DAGs in epidemiology and a pandemic like COVID-19: DAGs allow to choose confounders for the analysis in a principled and statistically correct way while reducing possible causes for bias.
-Also, the DAG formalization allows for discussion about the underlying causal structure.
+Also, the DAG formalization allows for discussion about the underlying causal assumptions.
 
@@
 
+\hypertarget{acknowledgments}{%
+\section*{Acknowledgments}\label{acknowledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+We are thankful for feedback from Thomas Czihal, Johannes Textor, and Ralph Brinks, who gave helpful suggestions on an earlier draft of the manuscript.
+
 \renewcommand\refname{References}

### 2020-07-23 manuscript/v2_medrxiv/manuscript.Rmd
insert acknowledgments
fixes tobias and aerosol reference WHO

--- 2020-07-15 manuscript/v1_medrxiv/manuscript.Rmd
@@
 
-With COVID-19, we find ourselves in a situation in which information on the causal contribution of various influencing factors in the population is urgently needed to inform politicians and health authories.
+With COVID-19, we find ourselves in a situation in which information on the causal contribution of various influencing factors in the population is urgently needed to inform politicians and health authorities.
 On the other hand, trials cannot be carried out for obvious ethical and legal reasons.
@@
 
+# Acknowledgments {-}
+
+```{r child = 'acknowledgments.Rmd'}
+```
+
 # References

### 2020-07-23 manuscript/v2_medrxiv/results.Rmd
fixes tobias and aerosol reference WHO

@@
 Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent.
+For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent.
 Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 1)` percent.
@@
 An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate")), 1)` percent.
-For example, if a district shows 10 percent points more relative searches for *corona* then another one, we expect approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate"))^(10)), 0)` percent more infections for this district after 8 days.
+For example, if a district shows 10 percent points more relative searches for *corona* than another one, we expect approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate"))^(10)), 0)` percent more infections for this district after 8 days.
 _COVID-19 burden_ (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a `r round(100*abs(1-causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate")), 1)` percent decrease in newly reported case numbers.
-The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(20-mymeans["COVID-19 burden"])), 0)` percent.
+The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 25 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(25-mymeans["COVID-19 burden"])), 1)` percent.
 
@@
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent.
-We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases).
-In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 0)` percent at a daily average temperature of $0^{\circ}C$.
-For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(0-mymeans["Weather (rainfall)"])), 0)` percent.
+We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
+In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 1)` percent at a daily average temperature of $0^{\circ}C$.
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(10-mymeans["Weather (rainfall)"])), 1)` percent.
 
 In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-$R^2$ values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-$R^2$.
-Notably, these sets always include most of our socio-demographic variables as confounders (cf.
-Table \@ref(tab:adjsets)) as well as the policy variables (except for COVID-19 burden).
+Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.
+Table \@ref(tab:adjsets), with the exception being COVID-19 burden).
 We also decided for the lag of 8 days based on the highest pseudo-$R^2$ values compared to other lags on the chosen adjustment sets.

### 2020-07-24 manuscript/v3_submission/data_methods.Rmd
init version for peer review submission

--- 2020-07-15 manuscript/v1_medrxiv/data_methods.Rmd
@@
 We constructed daily weather from four variables ("temperature", "rainfall", "humidity", "wind").
-Weather data was downloaded from Deutscher Wetterdienst (DWD, [@dwd_weather]) for all weather stations in Germany below 1000 meters altitude with daily records in 2020 until 8 July 2020.
+Weather data was downloaded from Deutscher Wetterdienst (DWD, [@dwd_weather]) for all weather stations in Germany below 1000 meters altitude with daily records for our observation period.
 District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district).

### 2020-07-24 manuscript/v3_submission/discussion.Rmd
init version for peer review submission

--- 2020-07-22 manuscript/v2_medrxiv/discussion.Rmd
@@
 We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers).
-We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population.
+We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population and/or effective public health measures to suppress a local outbreak.
 The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.
@@
 Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects [@fowler_effect_2020;@li_retrospective_2020;@shi_impact_2020;@Wang2020temperature].
-To reliably identify causal drivers, one needs to adjust for confounders.
+To reliably identify causal drivers, one must adjust for confounders.
 To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
@@
 Furthermore, our model assumed that all variables have effects proportional to the size of their measurements.
-It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values.
+It is possible that some variables show saturation effects or opposite effects for low, medium, or high values.
 This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
@@
 The causal effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret.
-We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.
+We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.
 
 Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections.
+Current research suggests this to be due to the prevalent airborne and respiratory droplets and aerosol transmission of the SARS-CoV-2 virus [@world2020transmission].
 In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

### 2020-07-24 manuscript/v3_submission/manuscript.tex
init version for peer review submission

--- 2020-07-15 manuscript/v1_medrxiv/manuscript.tex
@@
 the causal contribution of various influencing factors in the population
-is urgently needed to inform politicians and health authories. On the
+is urgently needed to inform politicians and health authorities. On the
 other hand, trials cannot be carried out for obvious ethical and legal
@@
 Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
-stations in Germany below 1000 meters altitude with daily records in
-2020 until 8 July 2020. District level daily weather data was aggregated
+stations in Germany below 1000 meters altitude with daily records for
+our observation period. District level daily weather data was aggregated
 per district by averaging the data from the three nearest weather
@@
 
-Within our framework, we saw strong effects for mobility in
+Within our framework, we saw significant effects for mobility in
 retail/recreational areas and essential shopping (grocery and pharmacy).
-In the former, an increase of 1 percent point mobility compared to the
-reference period (03 January to 06 February 2020) leads to an increase
-of the daily reported case number by about 0.8 percent. Contrarily, a
-corresponding increase of 1 percent point for the areas of
+For retail/recreation, an increase of 1 percent point mobility compared
+to the reference period (03 January to 06 February 2020) leads to an
+increase of the daily reported case number by about 0.8 percent.
+Contrarily, a corresponding increase of 1 percent point for the areas of
 grocery/pharmacy leads to a decrease in the reported case number by
@@
 example, if a district shows 10 percent points more relative searches
-for \emph{corona} then another one, we expect approximately 11 percent
+for \emph{corona} than another one, we expect approximately 11 percent
 more infections for this district after 8 days. \emph{COVID-19 burden}
@@
 Figure \ref{fig:effectsrange} visualizes this relationship: For a local
-outbreak with 20 daily cases as COVID-19 burden, we estimate as total
-causal effect a subsequent reduction of infection rate by 9 percent.
+outbreak with 25 daily cases as COVID-19 burden, we estimate as total
+causal effect a subsequent reduction of infection rate by 11.8 percent.
 
@@
 temperature leads to a reduction of the daily reported case numbers by
-approximately 0.8 percent. Similarly strong was the effect of rainfall:
-One millimeter (=1 liter per square meter) more rainfall leads to an
-increase of reported case numbers by approximately 1.1 percent. We
-observe small effects for humidity and wind as well (higher humidity
-leading to fewer cases, stronger wind leading to less cases). In
+approximately 0.8 percent. On the other hand, we found an increasing
+effect of rainfall: One millimeter (=1 liter per square meter) more
+rainfall leads to an increase of reported case numbers by approximately
+1.1 percent. We observe effects for humidity and wind as well (higher
+humidity leading to more cases, stronger wind leading to less cases). In
 perspective (Figure \ref{fig:effectsrange}), with temperature we expect
-an increase by approximately 9 percent at a daily average temperature of
-\(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm
+an increase by approximately 9.2 percent at a daily average temperature
+of \(0^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm
 rainfall a corresponding increase of the infection rate by approximately
-2 percent.
+9.2 percent.
 
@@
 higher pseudo-\(R^2\). Notably, these sets always include most of our
-socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets})
-as well as the policy variables (except for COVID-19 burden). We also
-decided for the lag of 8 days based on the highest pseudo-\(R^2\) values
-compared to other lags on the chosen adjustment sets. Similarly,
-negative binomial regression was chosen over Poisson regression, because
-the latter showed overdispersion and overall lower pseudo-\(R^2\)
-values.
+socio-demographic variables as confounders as well as the policy
+variables (cf.~Table \ref{tab:adjsets}, with the exception being
+COVID-19 burden). We also decided for the lag of 8 days based on the
+highest pseudo-\(R^2\) values compared to other lags on the chosen
+adjustment sets. Similarly, negative binomial regression was chosen over
+Poisson regression, because the latter showed overdispersion and overall
+lower pseudo-\(R^2\) values.
 
@@
 in districts led to a reduction in new infection numbers, which
-indicates risk-averse awareness in the population. The overall effect of
-mobility showed no consistent effect, however, in specific areas
+indicates risk-averse awareness in the population and/or effective
+public health measures to suppress a local outbreak. The overall effect
+of mobility showed no consistent effect, however, in specific areas
 significant causal effects could be measured: Increasing activity in
@@
 way while reducing possible causes for bias. Also, the DAG formalization
-allows for discussion about the underlying causal structure.
+allows for discussion about the underlying causal assumptions.
 
@@
 \citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
-To reliably identify causal drivers, one needs to adjust for
-confounders. To this end, we used an integrated model with variables
-from different aspects like mobility, awareness, weather, or
-socio-demographics and identified confounders by causal analysis with a
-directed acyclic graph. A causal approach is used in another current
-COVID-19 analysis \citep{gencoglu2020causal}. There, however, they
-identify the causal relationships (reconstruct a DAG), while we
-estimated causal effects for a given hypothesized DAG.
+To reliably identify causal drivers, one must adjust for confounders. To
+this end, we used an integrated model with variables from different
+aspects like mobility, awareness, weather, or socio-demographics and
+identified confounders by causal analysis with a directed acyclic graph.
+A causal approach is used in another current COVID-19 analysis
+\citep{gencoglu2020causal}. There, however, they identify the causal
+relationships (reconstruct a DAG), while we estimated causal effects for
+a given hypothesized DAG.
 
@@
 the novel virus SARS-CoV-2 regarding its ambient temperature survival
-and transmission. In regards of humidity, our findings (higher humidity
-leading to increased case numbers) are in agreement with previous
-research \citep{qi_covid-19_2020, wu_effects_2020, juni_impact_2020}.
-Finally, we found a positive causal effect of increment precipitation
-and a raise in COVID-19 cases, which supports previous observations
-\citep{sobral_association_2020}.
+and transmission. Finally, we found a positive causal effect of
+increment precipitation and a raise in COVID-19 cases, which supports
+previous observations \citep{sobral_association_2020}.
 
@@
 hand. We endorse our graph as a basis for discussion on residual
-confounding.
+confounding. We did not try to construct the DAG from the available data
+(cf. \citep{gencoglu2020causal}). As such, our proposed DAG is not
+entirely consistent with the data and there are conditional dependencies
+between variables that cannot be dissolved by adding edges to the DAG
+(e.g.~between the policies like contact restrictions and mandatory face
+masks). Another way to identify potential problems in the proposed DAG
+is to perform a sensitivity analysis of its structure by inspecting its
+maximal ancestral graph (MAG) or its Markov equivalence class
+represented by a complete partially DAG (CPDAG) and the existence of
+valid adjustment sets for these generalized graphs
+\citep{perkovic2017complete}. For the MAG derived from our DAG, only the
+effects for exposures mobility and searches for corona can be estimated
+with valid adjustment sets, while for the Markov equivalence class all
+exposures but COVID-19 burden lead to valid adjustments sets. A further
+analysis of these implications is out of the scope of this paper.
 
@@
 counts of \citet{kraemer_effect_2020} and others
-\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020}. We did not
-model case counts with a differential equation model like the classic
-SIR-model \citep{kermack1991contributions} and its successors, since
-these are more suited to prediction \citep[e.g.][]{rkimodellierung}
-while our choice of a negative binomial regression framework allowed us
-to estimate the effects of confounders more reliably. There are more
-advanced statistical methods for count data, e.g.~zero-inflated models
-and mixed models. We tested both approaches as extensions to the
-negative binomial regression and experienced numerical problems and
-increased computing time, along with an insubstantial increase in model
-performance. Furthermore, our model assumed that all variables have
-effects proportional to the size of their measurements. It is possible
-that some variables show saturation effects or opposite effects for
-lower, medium, or high values. This could be modelled with polynomial or
-other transformations of the variables, which we did not employ due to
-limited temporal and spatial data availability.
-
-As we were using administrative data for our analysis, the results are
-susceptible to the Modifiable Area Unit Problem (MAUP)
+\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
+We did not model case counts with a differential equation model like the
+classic SIR-model \citep{kermack1991contributions} and its successors,
+since these are more suited to prediction
+\citep[e.g.][]{rkimodellierung} while our choice of a negative binomial
+regression framework allowed us to estimate the effects of confounders
+more reliably. There are more advanced statistical methods for count
+data, e.g.~zero-inflated models and mixed models. We tested both
+approaches as extensions to the negative binomial regression and
+experienced numerical problems and increased computing time, along with
+an insubstantial increase in model performance. Furthermore, our model
+assumed that all variables have effects proportional to the size of
+their measurements. It is possible that some variables show saturation
+effects or opposite effects for low, medium, or high values. This could
+be modelled with polynomial or other transformations of the variables,
+which we did not employ due to limited temporal and spatial data
+availability. Use of a fixed DAG with effect estimation via regression
+assumes that data was generated by the same underlying process for the
+observation period. By inclusion of the successive mitigation policies
+as binary variables we were able to explain some of the variance caused
+by the changing dynamics of case numbers (similar to
+\citep{Islamm2743}).
+
+We stress the point that our effects were deduced on an aggregate
+(district) level in the absence of available data on an individual
+level. As such, conclusions about effects cannot be transferred on
+individuals without the possibility for an ecological fallacy.
+Furthermore, as we were using administrative data for our analysis, the
+results are susceptible to the Modifiable Area Unit Problem (MAUP)
 \citep{Openshaw1984}. The MAUP postulates that different regional
@@
 and conclusions. Due to limited available data for the different
-variables, there is currently no way to overcome this.
+variables, there is currently no way to overcome these problems that are
+inherent to all analyses on aggregated data level.
 
@@
 In our analysis, the adverse effect of mobility in retail/recreation and
-the favorable effect of mobility in grocery/pharmacy make a strong case
-for the successful implementation of policies like contact restrictions
-which limit the number of individual interactions that lead to
-infections. This is due to retail/recreational areas encompassing mostly
-places of social gatherings like restaurants and bars, malls, sports and
-music venues, among others, while if people are doing more of their
-essential shopping at supermarkets, they will most likely stay at home
-with less contact to other people.
+the favorable effect of mobility in grocery/pharmacy indicate that
+policies like contact restrictions which limit the number of individual
+interactions can lead to reduced infection numbers. This is due to
+retail/recreational areas encompassing mostly places of social
+gatherings like restaurants and bars, malls, sports and music venues,
+among others, while if people are doing more of their essential shopping
+at supermarkets, they will most likely stay at home with less contact to
+other people.
 
@@
 awareness, while the decreasing effect for future case numbers of high
-daily COVID-19 burden numbers indicates it affects individual
-risk-behavior as a deterrent.
+daily COVID-19 burden indicates it affects individual risk-behavior and
+entails effective non-pharmaceutical interventions.
 
@@
 while lower temperatures and high rainfall result in indoor activities,
-which lead to more infections. In this light, we advocate for
-precautious measures like increased hygiene, face masks, and air
-ventilation for unavoidable indoor activities.
+which lead to more infections. Current research suggests this to be due
+to the prevalent airborne and respiratory droplets and aerosol
+transmission of the SARS-CoV-2 virus \citep{world2020transmission}. In
+this light, we advocate for precautious measures like increased hygiene,
+face masks, and air ventilation for unavoidable indoor activities.
 
@@
 (all publicly available). Causal reasoning with a DAG allows us to
-estimate the causal effects reliably.
+estimate the causal effects more reliably.
 
@@
 
+\section*{Acknowledgments}\label{acknowledgments}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+We are thankful for feedback from Thomas Czihal, Johannes Textor, and
+Ralph Brinks, who gave helpful suggestions on an earlier draft of the
+manuscript.
+
 \renewcommand\refname{References}

### 2020-07-24 manuscript/v3_submission/figures/f_full_covid19_dag.R

@@ -1,0 +1,63 @@
+library(tidyverse)
+library(ggdag)
+library(cowplot)
+library(stringr)
+
+figurepath <- "./manuscript/v1_medrxiv/figures/"
+
+# Set theme
+theme_set(theme_cowplot())
+mycolors <- c("#268bd2","#dc322f","#d33682","#859900","grey60","lightgrey")
+
+dagfile <- "./data/bigdag4covid19_plot.txt"
+dagtext <- readChar(dagfile, file.info(dagfile)$size)
+dag <- dagitty::dagitty(dagtext,layout = F)
+tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
+
+tdc <- tidy_dag %>% mutate(curves = 0,
+                           curves = ifelse(name=="Age"&to=="Onset of COVID-19",
+                                           -0.2,
+                                           curves),
+                           curves = ifelse(name=="Age"&to=="Exposure to SARS-COV-2",
+                                           +0.1,
+                                           curves),
+                           curves = ifelse(name=="Gender"&to=="Turnout",
+                                           +0.2,
+                                           curves),
+                           curves = ifelse(name=="Foreign\ncitizens"&to=="Right-wing populist party votes",
+                                           +0.3,
+                                           curves),
+                           curves = ifelse(name=="Health\ncare\nutilization*"&to=="Exposure to SARS-COV-2",
+                                           -0.2,
+                                           curves))
+
+tidy_dag %>% 
+  group_by(name) %>% mutate(connections=n(),
+                                         farbe = case_when(str_detect(name,"Mobility")~"4",
+                                                         # str_detect(name,"burden")~"3",
+                                                         str_detect(name,"Searches")~"3",
+                                                         str_detect(name,"Temperature")~"2",
+                                                         str_detect(name,"new cases")~"1",                                  
+                                                         str_detect(name,"Tracing")~"6",
+                                                         str_detect(name,"Herd")~"6",
+                                                         str_detect(name,"Access")~"6",
+                                                         str_detect(name,"Susceptibility")~"6",
+                                                         str_detect(name,"Exposure")~"6",
+                                                         str_detect(name,"Onset")~"6",
+                                                         str_detect(name,"utiliz")~"6",
[…]

### 2020-07-24 manuscript/v3_submission/figures/f_exposure_vs_effect.R
cleanup to copy from tex to tex

@@ -1,96 +1,0 @@
-library("tidyverse")
-library("cowplot")
-
-figurepath <- "./manuscript/v1_medrxiv/figures/"
-
-modeldata_raw <- read_csv("data/Modeldata_raw.csv")
-
-varnames <- c("Mobility (retail and recreation)",
-              "Mobility (grocery and pharmacy)",
-              "Mobility (parks)",
-              "Mobility (transit stations)",
-              "Mobility (workplaces)",
-              "Mobility (residential)",
-              "Weather (temperature)", "Weather (rainfall)", "Weather (humidity)", "Weather (wind)",
-              "Searches corona", "COVID-19 burden") 
-varnames <- factor(varnames, levels=varnames, ordered=TRUE)
-vardata <- tibble(varnames=varnames, means=0, effects=0, conflow=0, confhigh=0, rangelow=0, rangehigh=0)
-
-ranges <- apply(modeldata_raw[, varnames], 2, quantile, probs=c(0.025, 0.975))
-
-vardata <- vardata %>%
-  mutate(means=sapply(varnames, function(v) mean(modeldata_raw %>% pull(v) )),
-         rangelow=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[1]),
-         rangehigh=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[2]))
-vardata$myrangelow <- ranges[1, varnames] # c(-100, -25, -0, -100, -50, -10, -10, 0, 0, 0, 0, 0)
-vardata$myrangehigh <- ranges[2, varnames] # c(+50, +25, +100, +50, +0, +20, 30, 20, 100, 15, 100, 100)
-vardata[1:6, 3:5] <- cbind(read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$estimates,
-                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$conflow,
-                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$confhigh)
-vardata[11, 3:5] <- cbind(read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$estimates,
-                      read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$conflow,
[…]

### 2020-07-24 manuscript/v3_submission/figures/f_full_covid19_dag.R

@@ -1,63 +1,0 @@
-library(tidyverse)
-library(ggdag)
-library(cowplot)
-library(stringr)
-
-figurepath <- "./manuscript/v1_medrxiv/figures/"
-
-# Set theme
-theme_set(theme_cowplot())
-mycolors <- c("#268bd2","#dc322f","#d33682","#859900","grey60","lightgrey")
-
-dagfile <- "./data/bigdag4covid19_plot.txt"
-dagtext <- readChar(dagfile, file.info(dagfile)$size)
-dag <- dagitty::dagitty(dagtext,layout = F)
-tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
-
-tdc <- tidy_dag %>% mutate(curves = 0,
-                           curves = ifelse(name=="Age"&to=="Onset of COVID-19",
-                                           -0.2,
-                                           curves),
-                           curves = ifelse(name=="Age"&to=="Exposure to SARS-COV-2",
-                                           +0.1,
-                                           curves),
-                           curves = ifelse(name=="Gender"&to=="Turnout",
-                                           +0.2,
-                                           curves),
-                           curves = ifelse(name=="Foreign\ncitizens"&to=="Right-wing populist party votes",
-                                           +0.3,
-                                           curves),
-                           curves = ifelse(name=="Health\ncare\nutilization*"&to=="Exposure to SARS-COV-2",
-                                           -0.2,
-                                           curves))
-
-tidy_dag %>% 
-  group_by(name) %>% mutate(connections=n(),
-                                         farbe = case_when(str_detect(name,"Mobility")~"4",
-                                                         # str_detect(name,"burden")~"3",
-                                                         str_detect(name,"Searches")~"3",
-                                                         str_detect(name,"Temperature")~"2",
-                                                         str_detect(name,"new cases")~"1",                                  
-                                                         str_detect(name,"Tracing")~"6",
-                                                         str_detect(name,"Herd")~"6",
-                                                         str_detect(name,"Access")~"6",
-                                                         str_detect(name,"Susceptibility")~"6",
-                                                         str_detect(name,"Exposure")~"6",
-                                                         str_detect(name,"Onset")~"6",
-                                                         str_detect(name,"utiliz")~"6",
[…]

### 2020-07-24 manuscript/v3_submission/figures/f_timeseries_cases_and_lograte.R

@@ -1,40 +1,0 @@
-library("tidyverse")
-library("cowplot")
-
-figurepath <- "./manuscript/v1_medrxiv/figures/"
-
-Modeldata <- read_csv("data/Modeldata_raw.csv")
-Sys.setlocale("LC_TIME", "C")
-mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
-
-mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
-
-cases <- Modeldata %>% dplyr::select(date, `Reported new cases COVID-19`, id) %>%
-  ggplot(aes(x=date,
-             y=`Reported new cases COVID-19`)) + 
-  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
-  geom_smooth(color=mycolors[1],size=2,se = F) + 
-  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") +
-  labs(x="Date", y="new cases") + 
-  scale_y_log10() +
-  theme(axis.title.x = element_blank()) +
-  theme_cowplot(font_size = 8)
-
-logrates <- Modeldata %>% dplyr::select(date, `Reported new cases COVID-19`, `Active cases`, id) %>%
-  ggplot(aes(x=date,
-             y=`Reported new cases COVID-19`/(`Active cases`+1))) + 
-  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
-  geom_smooth(color=mycolors[1], size=2, se = FALSE) + 
-  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") +
-  labs(x="Date", y="new cases/active cases") + 
-  scale_y_log10() +
-  theme(axis.title.x = element_blank()) +
-  theme_cowplot(font_size = 8)
-
-plot_grid(cases, logrates, labels = "AUTO",label_size = 12)
-
-ggsave(filename = paste0(figurepath, "f_timeseries_cases_and_lograte.eps"),
-       dpi=600, width=15,height = 7, unit="cm", device=cairo_ps, fallback_resolution = 600)
-
-# ggsave(filename = paste0(figurepath, "f_timeseries_cases_and_lograte.png"),
-#        dpi=600, width=15,height = 7, unit="cm", device=cairo_ps, fallback_resolution = 600)

### 2020-07-24 manuscript/v3_submission/figures/f_timeseries_exposures.R

@@ -1,57 +1,0 @@
-library("tidyverse")
-library("cowplot")
-
-figurepath <- "./manuscript/v1_medrxiv/figures/"
-
-Modeldata <- read_csv("data/Modeldata_raw.csv")
-Sys.setlocale("LC_TIME", "C")
-mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
-
-mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
-cases <- Modeldata %>% dplyr::select(date,contains("cases"), id) %>%
-    ggplot(aes(x=date,
-                       y=`Reported new cases COVID-19`/(`Active cases`+1)))+
-  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
-  geom_smooth(color=mycolors[1],size=2,se = F) + 
-  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") + labs(x="Date", y="new cases/active cases") + 
-  scale_y_log10() + 
-  theme(axis.title.x = element_blank()) +
-  theme_cowplot(font_size = 8)
-
-temperature <- Modeldata %>% dplyr::select(date,id, contains("temperature")) %>%
-  ggplot(aes(x= date-8,
-             y= `Weather (temperature)`))+
-  geom_point(aes(group=id), color="moccasin", stroke=0, alpha=0.2) +
-  geom_smooth(color=mycolors[2],size=2,se = F) + 
-  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)+2), date_labels = "%d %b") + labs(x="Date", y="Temperature") + 
-  scale_y_continuous() + 
-  theme(axis.title.x = element_blank())+
-  theme_cowplot(font_size = 8)
-
-awareness <- Modeldata %>% dplyr::select(date, id, `Searches corona`) %>%
-  ggplot(aes(x=date-8,
-             y= `Searches corona`)) +
-  geom_point(aes(group=id), color="lightpink", stroke=0, alpha=0.2) +
-  geom_smooth(color=mycolors[3],size=2,se = F) + 
-  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") + labs(x="Date", y="Awareness") + 
-  scale_y_continuous(limits = c(0,100))+
-  labs(y="Awareness") +
-  theme(axis.title.x = element_blank())+
[…]

### 2020-07-28 manuscript/v2_medrxiv/results.Rmd
small fixes (range of exposure variables)

@@
 _COVID-19 burden_ (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a `r round(100*abs(1-causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate")), 1)` percent decrease in newly reported case numbers.
-The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 25 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(25-mymeans["COVID-19 burden"])), 1)` percent.
+The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(20-0)), 1)` percent.
 
@@
 We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
-In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 1)` percent at a daily average temperature of $0^{\circ}C$.
-For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(10-mymeans["Weather (rainfall)"])), 1)` percent.
+In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-20)), 1)` percent at a daily average temperature of $0^{\circ}C$ compared to a day with $20^{\circ}C$.
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(10-0)), 1)` percent compared to a day with no precipitation.
 

### 2021-03-01 manuscript/v2_medrxiv/manuscript.Rmd
small change in title to test difflatex

@@
 ---
-title: Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
+title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
 author:

### 2021-03-01 manuscript/v2_medrxiv/manuscript.tex
small change in title to test difflatex

@@
 \usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
-\makeatother
-\let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
+            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
             colorlinks=false,
@@
 % Pandoc toggle for numbering sections (defaults to be off)
+
+% Pandoc citation processing
 
@@
 
-  \title{Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
+  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
     \author[Zi]{Edgar Steiger}
@@
 We constructed daily weather from four variables (``temperature'', ``rainfall'', ``humidity'', ``wind'').
-Weather data was downloaded from Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather stations in Germany below 1000 meters altitude with daily records in 2020 until 8 July 2020.
+Weather data was downloaded from Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather stations in Germany below 1000 meters altitude with daily records for our observation period.
 District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district).
@@
 Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent.
+For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent.
 Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent.
@@
 An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent.
-For example, if a district shows 10 percent points more relative searches for \emph{corona} then another one, we expect approximately 11 percent more infections for this district after 8 days.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 11 percent more infections for this district after 8 days.
 \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.6 percent decrease in newly reported case numbers.
-The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 9 percent.
+The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 11 percent.
 
@@
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent.
-We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases).
-In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 9 percent at a daily average temperature of \(0^{\circ}C\).
-For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 2 percent.
+We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 17.8 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 11.4 percent compared to a day with no precipitation.
 
 In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\).
-Notably, these sets always include most of our socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets}) as well as the policy variables (except for COVID-19 burden).
+Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.~Table \ref{tab:adjsets}, with the exception being COVID-19 burden).
 We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets.
@@
 Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects \citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
-To reliably identify causal drivers, one needs to adjust for confounders.
+To reliably identify causal drivers, one must adjust for confounders.
 To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
+Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
 
@@
 We endorse our graph as a basis for discussion on residual confounding.
-We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}).
+We did not try to construct the DAG from the available data (cf.
+\citep{gencoglu2020causal}).
 As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks).
@@
 We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
-We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
+We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
 There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models.
@@
 The causal effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret.
-We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.
+We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.
 
 Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections.
+Current research suggests this to be due to the prevalent airborne and respiratory droplets and aerosol transmission of the SARS-CoV-2 virus \citep{world2020transmission}.
 In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.
@@
 
-Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
+Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
 

### 2021-03-01 manuscript/v2_medrxiv/manuscript.tex
revert title

@@
             pdfauthor={},
-            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
+            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
             colorlinks=false,
@@
 
-  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
+  \title{Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
     \author[Zi]{Edgar Steiger}

### 2021-03-01 manuscript/v4_medrxiv_revision/causal_model.Rmd
add new folders for revision, test latexdiff

# Causal Model

We used a directed acyclic graph (DAG) [@schipf_directed_2011;@textor_robust_2017] as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread. To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar. Restrictions were applied to English and German language and the publication date in the last one year. The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with the exposure variables ("mobility", “public awareness”, "awareness", “google trends”,“ambient temperature”, "temperature"). For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$). We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \@ref(fig:dag-covid-19).

```{r dag-covid-19, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="DAG of determinants of reported COVID-19 cases on the district level. Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.", fig.height=5}
knitr::include_graphics(rep('figures/f_full_covid19_dag.eps'))
```

A number of studies report a strong association of **mobility** restrictions on the number of new COVID-19 cases: Restrictive measures (e.g. “stay-at-home” orders, travel bans, or school closures) are shown to possibly reduce the COVID-19 incidence [@chang_modeling_2020;@Chinazzi395;@fowler_effect_2020;@kraemer_effect_2020;@lasry_timing_2020;@linka_outbreak_2020;@mazzoli_effects_2020;@xiong_data-driven_2020]. However, some studies point out the combination of various non-pharmaceutical interventions (NPIs) is decisive to prevent new infections [@juni_impact_2020;@lai_effect_2020].

Google Trends [@google_trends] data can be used as a tool to get insights into public interest (**awareness**) in the coronavirus disease. Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases [@ayyoubzadeh_predicting_2020;@effenberger_association_2020;@higgins_correlations_2020;@li_retrospective_2020;@lin_google_2020;@mavragani_tracking_2020;@walker_use_2020;@yuan_trends_2020;@zhou_effects_2020]. Some search terms e.g. "COVID-19" or "coronavirus" predated newly infected cases/total number of cases by roughly 7 to 14 days for different countries [@effenberger_association_2020; @higgins_correlations_2020;@li_retrospective_2020;@yuan_trends_2020]. Additionally, we acknowledged that individual risk-aware behavior might be a reaction to the current COVID-19 burden (measured as reported cases at the day of exposure).

Mixed evidence is available regarding the effect of **temperature**: On the one hand several papers report an association between increase in temperature and decrease in newly infected COVID-19 cases [@bannister-tyrrell_preliminary_2020;@demongeot_temperature_2020;@liu_impact_2020;@qi_covid-19_2020;@shi_impact_2020;@sobral_association_2020;@tosepu_correlation_2020;@Wang2020temperature;@wu_effects_2020]. On the other hand, also the opposite has been found [@auler_evidence_2020;@xie_association_2020]. Some studies found no association at all [@briz-redon_spatio-temporal_2020;@iqbal_nexus_2020;@jahangiri_sensitivity_2020;@juni_impact_2020;@yao_no_2020]. It should be noted that few studies considered other confounding variables than meteorological ones (especially age and population density among others [@briz-redon_spatio-temporal_2020;@juni_impact_2020;@wu_effects_2020]). In addition, the transferability of results between different climate zones is questionable. To avoid possible bias caused by weather variables other than temperature, we included rain, wind, and humidity in our model.

When investigating causal determinants of SARS-CoV-2 infections, a number of confounders have to be considered. Well-known risk factors for SARS-CoV-2 as well as for other infections are demographic factors such as age, gender, socio-economic status (SES), population density, and foreign citizenship/ethnicity [@de2020risk;@Dragano2020.06.17.20133918;@jhucovid19db2020]. In Germany along with other countries (i.e. Brazil, USA, or the UK), populist parties or politicians and their electorate tend to be more sceptical about effects of containment measures than the other part of the electorate [@dohle_wingen_schreiber_2020;@engle_staying_2020]. Therefore we considered both "right-wing populist party votes" and "voter turnout" as possible confounders. Public health interventions were also taken into account (contact restrictions, school closures etc.), as their implementation showed strong correlations with controlling the spread of SARS-CoV-2 [@cowling2020impact;@juni_impact_2020;@lai_effect_2020]. To avoid bias due to reporting delay of case numbers we had to include weekday and German holidays. We include some unobserved variables in our DAG (e.g. "Herd immunity"), too. Please note that "Exposure to SARS-CoV-2" is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset. *Exposure to the virus* should not be confused with the formal *exposure variables* of the DAG (mobility, awareness, temperature).

### 2021-03-01 manuscript/v4_medrxiv_revision/discussion.Rmd
add new folders for revision, test latexdiff

--- 2020-07-24 manuscript/v3_submission/discussion.Rmd
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
+Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
 
@@
 
-Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
+Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.

### 2021-03-01 manuscript/v5_submission_revision/manuscript_tex4submission.tex
add new folders for revision, test latexdiff

--- 2020-07-24 manuscript/v3_submission/manuscript.tex
@@
-\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
-%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
-\usepackage[hyphens]{url}
-
-  \journal{medRxiv} % Sets Journal name
-
-\usepackage{lineno} % add
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
-%%%%%%%%%%%%%%%% end my additions to header
-
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{fontspec}
-  \ifxetex
-    \usepackage{xltxtra,xunicode}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use microtype if available
-\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\usepackage{natbib}
-\bibliographystyle{plainnat}
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% my packages and configs
 \usepackage{longtable}
-\usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
-\makeatother
-\let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
-            colorlinks=false,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-
-\setcounter{secnumdepth}{5}
-% Pandoc toggle for numbering sections (defaults to be off)
-
-% Pandoc header
-\usepackage{geometry}
-% \usepackage[numbers,sort&compress]{natbib}
 \usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
+\usepackage{url}
+\def\UrlBreaks{\do\/\do-}
+
+%% END MACROS SECTION
 
 \begin{document}
-\begin{frontmatter}
-
-  \title{Causal analysis of COVID-19 observational data in German districts
-reveals effects of mobility, awareness, and temperature}
-    \author[Zi]{Edgar Steiger}
-   \ead{esteiger@zi.de} 
-    \author[Zi]{Tobias Mußgnug}
-   \ead{tmussgnug@zi.de} 
-    \author[Zi]{Lars Eric Kroll}
-   \ead{lkroll@zi.de} 
-      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi),
-Salzufer 8, D-10587 Berlin, Germany}
-
-  \begin{abstract}
-  Mobility, awareness, and weather are suspected to be causal drivers for
-  new cases of COVID-19 infection. Correcting for possible confounders, we
-  estimated their causal effects on reported case numbers. To this end, we
-  used a directed acyclic graph (DAG) as a graphical representation of the
-  hypothesized causal effects of the aforementioned determinants on new
-  reported cases of COVID-19. Based on this, we computed valid adjustment
-  sets of the possible confounding factors. We collected data for Germany
-  from publicly available sources (e.g.~Robert Koch Institute, Germany's
-  National Meteorological Service, Google) for 401 German districts over
-  the period of 15 February to 8 July 2020, and estimated total causal
-  effects based on our DAG analysis by negative binomial regression. Our
-  analysis revealed favorable causal effects of increasing temperature,
-  increased public mobility for essential shopping (grocery and pharmacy),
-  and awareness measured by COVID-19 burden, all of them reducing the
-  outcome of newly reported COVID-19 cases. Conversely, we saw adverse
-  effects of public mobility in retail and recreational areas, awareness
-  measured by searches for ``corona'' in Google, and higher rainfall,
-  leading to an increase in new COVID-19 cases. This comprehensive causal
-  analysis of a variety of determinants affecting COVID-19 progression
-  gives strong evidence for the driving forces of mobility, public
-  awareness, and temperature, whose implications need to be taken into
-  account for future decisions regarding pandemic management.
-  \end{abstract}
-
- \end{frontmatter}
-
-\section{Introduction}\label{introduction}
-
-As the COVID-19 pandemic progresses, research on mechanisms behind the
-transmission of SARS-CoV-2 shows conflicting evidence
-\citep{who2020report, Chinazzi395, guan2020clinical}. While effects of
-mobility have been extensively discussed, less is known on other factors
-such as changing awareness in the population
-\citep{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Causal analysis of COVID-19 observational data in German districts
+reveals effects of mobility, awareness, and temperature} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Edgar Steiger\textsuperscript{1*},
+Tobias Mußgnug\textsuperscript{1},
+Lars Eric Kroll\textsuperscript{1}%,
+%Name4 Surname\textsuperscript{2},
+%Name5 Surname\textsuperscript{2\ddag},
+%Name6 Surname\textsuperscript{2\ddag},
+%Name7 Surname\textsuperscript{1,2,3*},
+%with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
+\\
+\bigskip
+\textbf{1} Central Research Institute of Ambulatory Health Care in Germany (Zi),
+Salzufer 8, D-10587 Berlin, Germany
+\\
+%\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+%\\
+%\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+%\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+%\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+%\ddag These authors also contributed equally to this work.
+
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* esteiger@zi.de
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection.
+Correcting for possible confounders, we estimated their causal effects on reported case numbers.
+To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19.
+Based on this, we computed valid adjustment sets of the possible confounding factors.
+We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
+Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
+This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+%\section*{Author summary}
+%Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}\label{introduction}
+As the COVID-19 pandemic progresses, research on mechanisms behind the transmission of SARS-CoV-2 shows conflicting evidence \cite{who2020report, Chinazzi395, guan2020clinical}.
+While effects of mobility have been extensively discussed, less is known on other factors such as changing awareness in the population
+\cite{higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}
 or the effects of temperature
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}.
-A limiting factor in many studies is the lack of a causal approach to
+\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020}. A limiting factor in many studies is the lack of a causal approach to
 assess the causal contributions of various factors
-\citep{Greenland1999}. This can lead to distorted estimates of the
+\cite{Greenland1999}. This can lead to distorted estimates of the
 causal factors with observational data
-\citep{Greenland1999, schipf_directed_2011, textor_robust_2017}.
+\cite{Greenland1999, schipf_directed_2011, textor_robust_2017}.
 
@@
 examined for example in the Johns Hopkins University (JHU) COVID-19
-database \citep{jhucovid19db2020}. The comparison of time series of case
+database \cite{jhucovid19db2020}. The comparison of time series of case
 numbers from different countries and observational periods can be
@@
 without blocked causal paths whose effects can be analyzed with
-observational data \citep{Pearl_2014}. We used regional time series data
-of all German districts (\(401\)) from various publicly available
+observational data \cite{Pearl_2014}. We used regional time series data
+of all 401 German districts from various publicly available
 sources to analyze these questions on a regional level. Germany is a
@@
 
-\section{Causal Model}\label{causal-model}
+\section*{Causal Model}\label{causal-model}
 
 We used a directed acyclic graph (DAG)
-\citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze
+\cite{schipf_directed_2011, textor_robust_2017} as a tool to analyze
 the causal relationships between several exposures and SARS-CoV-2
@@
 We integrated these findings where possible into the construction of our
-DAG, which can be seen in Figure \ref{fig:dag-covid-19}.
-
-\begin{figure}
-\centering
-\includegraphics{figures/f_full_covid19_dag.eps}
-\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
-cases on the district level. Unobserved variables are light gray,
-variables marked with an asterisk (*) are confounded by
-weekday/holiday.}
+DAG, which can be seen in Fig~\ref{fig:dag-covid-19}.
+
+%\begin{figure}
+%\centering
+%\includegraphics{figures/f_full_covid19_dag.eps}
+%\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19
+%cases on the district level. Unobserved variables are light gray,
+%variables marked with an asterisk (*) are confounded by
+%weekday/holiday.}
+%\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf DAG of determinants of reported COVID-19 cases on the district level.}
+Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
+\label{fig:dag-covid-19}
 \end{figure}
@@
 shown to possibly reduce the COVID-19 incidence
-\citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
+\cite{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
 However, some studies point out the combination of various
 non-pharmaceutical interventions (NPIs) is decisive to prevent new
-infections \citep{juni_impact_2020, lai_effect_2020}.
-
-Google Trends \citep{google_trends} data can be used as a tool to get
+infections \cite{juni_impact_2020, lai_effect_2020}.
+
+Google Trends \cite{google_trends} data can be used as a tool to get
 insights into public interest (\textbf{awareness}) in the coronavirus
@@
 volumes (RSV) indices and reported new COVID-19 cases
-\citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
+\cite{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
 Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
@@
 different countries
-\citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
+\cite{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
 Additionally, we acknowledged that individual risk-aware behavior might
@@
 infected COVID-19 cases
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
+\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
 On the other hand, also the opposite has been found
-\citep{auler_evidence_2020, xie_association_2020}. Some studies found no
+\cite{auler_evidence_2020, xie_association_2020}. Some studies found no
 association at all
-\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
+\cite{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}.
 It should be noted that few studies considered other confounding
@@
 density among others
-\citep{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
+\cite{briz-redon_spatio-temporal_2020, juni_impact_2020, wu_effects_2020}).
 In addition, the transferability of results between different climate
@@
 foreign citizenship/ethnicity
-\citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
+\cite{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
 Germany along with other countries (i.e.~Brazil, USA, or the UK),
@@
 sceptical about effects of containment measures than the other part of
-the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
+the electorate \cite{dohle_wingen_schreiber_2020, engle_staying_2020}.
 Therefore we considered both ``right-wing populist party votes'' and
@@
 controlling the spread of SARS-CoV-2
-\citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
+\cite{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
 bias due to reporting delay of case numbers we had to include weekday
-and German holidays. We include some unobserved variables in our DAG
+and German holidays. We included some unobserved variables in our DAG
 (e.g. ``Herd immunity''), too. Please note that ``Exposure to
@@
 
-\section{Data}\label{data}
+%Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam Eq~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+%\begin{eqnarray}
+%\label{eq:schemeP}
+%	\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}},
+%\end{eqnarray}
+
+\section*{Materials and methods}
+\subsection*{Data}\label{data}
 
@@
 outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
-predictors was derived from our causal DAG (see Table
-\ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to
+predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
 modelling and data limitations, some of the predictors were unobserved
@@
 
-\subsection{Variables}\label{variables}
+\subsubsection*{Variables}\label{variables}
 
 We downloaded German daily case numbers on district level reported by
-Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July
-2020) and aggregated them by date. The number of daily active cases for
+Robert Koch Institute (RKI, \cite{casenumbers_rki}) and aggregated them by date. The number of daily active cases for
 day \(d\) was derived by subtracting the total number of reported cases
@@
 To assess the mobility of the German population, we used data publicly
-available on German state level from Google \citep{google_mobility}.
+available on German state level from Google \cite{google_mobility}.
 Measurements are daily relative changes of mobility in percent compared
@@
 in the topic term ``corona'' as indicated by Google searches. The daily
-data was available on state level \citep{google_trends} and passed onto
+data was available on state level \cite{google_trends} and passed onto
 district level. As a second proxy for awareness, we used the daily
@@
 ``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
-Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
+Deutscher Wetterdienst (DWD, \cite{dwd_weather}) for all weather
 stations in Germany below 1000 meters altitude with daily records for
@@
 to SARS-CoV-2 in our analysis. All are readily available from INKAR
-database \citep{inkar}. We used the share of population that is 65 years
+database \cite{inkar}. We used the share of population that is 65 years
 or older and the share of population that is younger than 18 years
@@
 
-\begin{table}
-
-\caption{\label{tab:model-vars}Observed model variables}
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\fontsize{6}{8}\selectfont
-\begin{tabular}[t]{llll>{\raggedright\arraybackslash}p{10em}l}
+\caption{
+{\bf Observed model variables.}}
+\fontsize{8}{10}\selectfont
+\begin{tabular}[t]{llll>{\raggedright\arraybackslash}p{20em}l}
 \toprule
@@
 \multicolumn{6}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
-\hspace{1em}grocery and pharmacy & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
-\hspace{1em}parks & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
-\hspace{1em}workplaces & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
-\hspace{1em}residential & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
-\hspace{1em}transit stations & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
+\hspace{1em}retail and recreation & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
+\hspace{1em}grocery and pharmacy & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
+\hspace{1em}parks & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
+\hspace{1em}workplaces & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
+\hspace{1em}residential & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
+\hspace{1em}transit stations & daily & state & numeric & percent change compared to reference period & Google \cite{google_mobility}\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & daily & district & numeric & mm (l/sqm) & DWD \citep{dwd_weather}\\
-\hspace{1em}Temperature & daily & district & numeric & °C & DWD \citep{dwd_weather}\\
-\hspace{1em}Humidity & daily & district & numeric & relative humidity (\%) & DWD \citep{dwd_weather}\\
-\hspace{1em}Wind & daily & district & numeric & m/s & DWD \citep{dwd_weather}\\
+\hspace{1em}Rainfall & daily & district & numeric & mm (l/sqm) & DWD \cite{dwd_weather}\\
+\hspace{1em}Temperature & daily & district & numeric & °C & DWD \cite{dwd_weather}\\
+\hspace{1em}Humidity & daily & district & numeric & relative humidity (\%) & DWD \cite{dwd_weather}\\
+\hspace{1em}Wind & daily & district & numeric & m/s & DWD \cite{dwd_weather}\\
 \addlinespace[0.3em]
@@
 \hspace{1em}Contact restrictions & daily & national & binary & - & -\\
-\hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
+\hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \cite{mitze2020face}\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Socio-demographic}}\\
-\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \citep{inkar}\\
-\hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \citep{inkar}\\
-\hspace{1em}Population density & constant & district & numeric & population per sqkm & INKAR \citep{inkar}\\
-\hspace{1em}Foreign citizens & constant & district & numeric & 2 variables: share of foreign citizens \& of population seeking refuge & INKAR \citep{inkar}\\
-\hspace{1em}Socio-economic status & constant & district & numeric & share of households with low income & INKAR \citep{inkar}\\
-\hspace{1em}Turnout & constant & district & numeric & voter turnout in last election & INKAR \citep{inkar}\\
-\hspace{1em}Right-wing populist party votes & constant & district & numeric & share of votes for AfD in last election & INKAR \citep{inkar}\\
-\hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \citep{inkar}\\
+\hspace{1em}Age & constant & district & numeric & 2 variables: share of population $>=65$ years \& $<18$ years & INKAR \cite{inkar}\\
+\hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \cite{inkar}\\
+\hspace{1em}Population density & constant & district & numeric & population per sqkm & INKAR \cite{inkar}\\
+\hspace{1em}Foreign citizens & constant & district & numeric & 2 variables: share of foreign citizens \& of population seeking refuge & INKAR \cite{inkar}\\
+\hspace{1em}Socio-economic status & constant & district & numeric & share of households with low income & INKAR \cite{inkar}\\
+\hspace{1em}Turnout & constant & district & numeric & voter turnout in last election & INKAR \cite{inkar}\\
+\hspace{1em}Right-wing populist party votes & constant & district & numeric & share of votes for AfD in last election & INKAR \cite{inkar}\\
+\hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \cite{inkar}\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
-\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
+\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \cite{google_trends}\\
+\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \cite{casenumbers_rki}\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Case numbers}}\\
-\hspace{1em}Reported new cases of COVID-19 & daily & district & numeric & - & RKI \citep{casenumbers_rki}\\
-\hspace{1em}Active cases & daily & district & numeric & active cases on day of report & RKI \citep{casenumbers_rki}\\
+\hspace{1em}Reported new cases of COVID-19 & daily & district & numeric & - & RKI \cite{casenumbers_rki}\\
+\hspace{1em}Active cases & daily & district & numeric & active cases on day of report & RKI \cite{casenumbers_rki}\\
 \bottomrule
 \end{tabular}
+\label{tab:model-vars}
+\end{adjustwidth}
 \end{table}
@@
 effect of all variables (but outcome, offset, and the non-dynamic
-socio-demographic variables) by 8 days (see Section \ref{results}) which
+socio-demographic variables) by 8 days (see Section Results) which
 means that we assumed that their effects on the outcome will be visible
@@
 
-\section{Methods}\label{methods}
-
-\subsection{Causal analysis with DAG and adjustment
+\subsection*{Methods}\label{methods}
+
+\subsubsection*{Causal analysis with DAG and adjustment
 sets}\label{causal-analysis-with-dag-and-adjustment-sets}
@@
 DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl
-\citep{spirtes2000causation, pearl2009causality} introduce the theory of
+\cite{spirtes2000causation, pearl2009causality} introduce the theory of
 causal analysis, do-calculus, and how to analyze a DAG to estimate the
@@
 adjustment set) to estimate the proper causal effect
-\citep{pearl2009causality}. The ``minimal adjustment set''
-\citep{greenland1999causal} is a valid adjustment set of variables that
+\cite{pearl2009causality}. The ``minimal adjustment set''
+\cite{greenland1999causal} is a valid adjustment set of variables that
 does not contain another valid adjustment set as a subset. However,
@@
 estimate the causal effect. Thus, we identified the ``optimal adjustment
-set'' \citep{henckel2019graphical} as the set of variables which is a
+set'' \cite{henckel2019graphical} as the set of variables which is a
 valid adjustment set while having the lowest asymptotic variance in the
@@
 
-We analyzed the DAG from Section \ref{causal-model} with the R Software
-\citep{rsoftware} and the R packages \texttt{dagitty} (formal
+We analyzed the DAG from Fig~\ref{fig:dag-covid-19} with the R Software
+\cite{rsoftware} and the R packages \texttt{dagitty} (formal
 representation of the graph and minimal adjustment sets
-\citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
-adjustment set \citep{pcalg}). For the defined exposures and the outcome
+\cite{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal
+adjustment set \cite{pcalg}). For the defined exposures and the outcome
 ``Reported new cases of COVID-19'', we computed the minimal and optimal
@@
 
-\subsection{Regression with negative binomial
+\subsubsection*{Regression with negative binomial
 model}\label{regression-with-negative-binomial-model}
@@
 We can estimate the causal effect from exposure to outcome by regression
-\citep{pearl2009causality}. Since the outcome ``Reported new cases of
+\cite{pearl2009causality}. Since the outcome ``Reported new cases of
 COVID-19'' is a count variable, one should not employ a linear
@@
 is a constant proportion of the number of active cases \(A\). This was
-modeled by including an offset \(\log(A+1)\) in the regression model
-\eqref{eq:regmodel}:
+modeled by including an offset \(\log(A+1)\) in the regression model Eq~\eqref{eq:regmodel}:
 
@@
 One can interpret the model as approximating the log-ratio of new cases
-and active cases by a linear combination of the regressor variables
-\eqref{eq:logratemodel}. If all variables \(x_i\) are centered in
-\eqref{eq:ratechange}, we have for the baseline
+and active cases by a linear combination of the regressor variables in Eq~\eqref{eq:logratemodel}. If all variables \(x_i\) are centered in Eq~\eqref{eq:ratechange}, we have for the baseline
 \(\forall i\ x_i=0 \Rightarrow \mathbb{E}[Y|x=0]=\exp(\alpha)\cdot (A+1)\).
@@
 infection rate (how many people does one infected individual infect in
-one day). If we hold all variables \(x_i\) fixed (e.g.~at baseline 0) in
-\eqref{eq:ratechange} but now increase the exposure variable
+one day). If we hold all variables \(x_i\) fixed (e.g.~at baseline 0) in Eq~\eqref{eq:ratechange} but now increase the exposure variable
 \(x_{i^{\ast}}=0\) by one unit to \(x_{i^{\ast}}+1=0+1\), we have
-\(\mathbb{E}[Y|x']=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0=\exp(\alpha)\cdot(A+1)\exp(\beta_{i^{\ast}})\),
+\begin{align*}
+\mathbb{E}[Y|x']&=\exp(\alpha)\cdot(A+1)\cdot\exp(\beta_{i^{\ast}}^{x_{i^{\ast}}+1})\prod_{i\neq i^{\ast}}\exp(\beta_i)^0 \\
+&=\exp(\alpha)\cdot(A+1)\cdot\exp(\beta_{i^{\ast}}),
+\end{align*}
 which means the exponentiated coefficient \(\beta_{i^{\ast}}\) describes
@@
 In practice, given observations of \(Y\) and \(x\) we estimate the
-regression coefficients \(\alpha\) and \(\beta_i\) by \emph{maximum
-likelihood} \citep{maxlikelihood}. Our observational measurements are
+regression coefficients \(\alpha\) and \(\beta_i\) by maximum
+likelihood \cite{maxlikelihood}. Our observational measurements are
 \(y_{kt}\) and \(x_{ikt}\), where \(k\) indicates the corresponding
@@
 (function \texttt{gac} in package \texttt{pcalg}
-\citep{perkovi2015complete}). If a minimal adjustment set contained
+\cite{perkovi2015complete}). If a minimal adjustment set contained
 unobserved variables, we discarded the whole set. We conducted a
@@
 from the \texttt{MASS} package for the negative binomial regression
-\citep{mass}) for every remaining valid adjustment set as regressors and
+\cite{mass}) for every remaining valid adjustment set as regressors and
 calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the
@@
 
-\section{Results}\label{results}
-
-Descriptive statistics for the included variables are presented in Table
-\ref{tab:descriptive-vars}.
-
-\begin{table}
-
-\caption{\label{tab:descriptive-vars}Descriptive Statistics for observed variables}
+%% For figure citations, please use "Fig" instead of "Figure".
+%Nulla mi mi, Fig~\ref{fig1} venenatis sed ipsum varius, volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+%
+%% Place figure captions after the first paragraph in which they are cited.
+%\begin{figure}[!h]
+%\caption{{\bf Bold the figure title.}
+%Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+%\label{fig1}
+%\end{figure}
+
+% Results and Discussion can be combined.
+\section*{Results}\label{results}
+
+Descriptive statistics for the included variables are presented in Table~\ref{tab:descriptive-vars}.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\fontsize{7}{9}\selectfont
+\caption{
+{\bf Descriptive statistics for observed variables.}}
+%\fontsize{8}{10}\selectfont
 \begin{tabular}[t]{ll}
@@
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
+\multicolumn{2}{l}{\textbf{Case numbers (outcome and offset)}}\\
 \hspace{1em}Reported new cases COVID-19 & 3.60 (10.39)\\
@@
 \end{tabular}
+\label{tab:descriptive-vars}
+%\end{adjustwidth}
 \end{table}
-
-\begin{figure}
-\centering
-\includegraphics{figures/f_timeseries_cases_and_lograte.eps}
-\caption{\label{fig:cases-log}Temporal and district level variation of
-outcome (log-scale)}
-\end{figure}
 
@@
 decreased afterwards till the beginning of June 2020 with a slight
-increase and decrease afterwards (Figure \ref{fig:cases-log}A). On the
+increase and decrease afterwards (Fig~\ref{fig:cases-log}A). On the
 other hand, the (log-)ratio of reported cases over active cases
@@
 afterwards with a slight decrease close to the end of the observation
-period (Figure \ref{fig:cases-log}B). Both figures examplify a
+period (Fig~\ref{fig:cases-log}B). Both figures examplify a
 considerable variation among the districts (light blue points are
 individual district's data).
+
+\begin{figure}[!h]
+\caption{{\bf Temporal and district level variation of outcome (log-scale).}}
+\label{fig:cases-log}
+\end{figure}
 
@@
 diminishing public interest in COVID-19, and temperatures were overall
-increasing (cf.~Figure \ref{fig:timeseries}); with correlations between
+increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between
 temporal progression and mobility in retail and recreation \(r_{A,D}=\)
@@
 
-\begin{figure}
+\begin{figure}[!h]
+\caption{{\bf Temporal variation of outcome and main
+determinants.}}
+\label{fig:timeseries}
+\end{figure}
+
+\subsection{Main results}\label{main-results}
+
+We list the results of our causal analysis for the effects of our
+variables in Table~\ref{tab:causaleffects}. The estimates are
+multiplicative rates of increase/decrease for a one unit increase of the
+respective variable: Values above 1 lead to an increase, below 1 to a
+decrease of the infection rate. To put these estimates into perspective,
+Fig~\ref{fig:effectsrange} shows the relative causal effect of the
+different exposure variables on the number of reported COVID-19 cases on
+a range of sensible values of the exposure variables (95 percent
+quantiles of data points).
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\includegraphics{figures/f_timeseries_exposures.eps}
-\caption{\label{fig:timeseries}Temporal variation of outcome and main
-determinants}
-\end{figure}
-
-\subsection{Main results}\label{main-results}
-
-\begin{table}
-
-\caption{\label{tab:causaleffects}Causal effect estimates with 99 percent confidence bands}
-\centering
+\caption{
+{\bf Causal effect estimates with 99 percent confidence bands.}}
+%\fontsize{8}{10}\selectfont
 \begin{tabular}[t]{lrrr}
@@
 \end{tabular}
+\label{tab:causaleffects}
+%\end{adjustwidth}
 \end{table}
 
-We list the results of our causal analysis for the effects of our
-variables in Table \ref{tab:causaleffects}. The estimates are
-multiplicative rates of increase/decrease for a one unit increase of the
-respective variable: Values above 1 lead to an increase, below 1 to a
-decrease of the infection rate. To put these estimates into perspective,
-Figure \ref{fig:effectsrange} shows the relative causal effect of the
-different exposure variables on the number of reported COVID-19 cases on
-a range of sensible values of the exposure variables (95 percent
-quantiles of data points).
-
-\begin{figure}
-\centering
-\includegraphics{figures/f_exposure_vs_effect.eps}
-\caption{\label{fig:effectsrange}Relative causal effects of exposures}
+\begin{figure}[!h]
+\caption{{\bf Relative causal effects of exposures.}}
+\label{fig:effectsrange}
 \end{figure}
@@
 consistent in their direction (99 percent confidence intervals of
-estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
+estimates include 1). Fig~\ref{fig:effectsrange} shows the effects of
 mobility on a range of possible values. Thus, we expect an increase of
@@
 0.6 percent decrease in newly reported case numbers. The last plot in
-Figure \ref{fig:effectsrange} visualizes this relationship: For a local
-outbreak with 25 daily cases as COVID-19 burden, we estimate as total
-causal effect a subsequent reduction of infection rate by 11.8 percent.
+Fig~\ref{fig:effectsrange} visualizes this relationship: For a local
+outbreak with 20 daily cases as COVID-19 burden, we estimate as total
+causal effect a subsequent reduction of infection rate by 11 percent.
 
@@
 humidity leading to more cases, stronger wind leading to less cases). In
-perspective (Figure \ref{fig:effectsrange}), with temperature we expect
-an increase by approximately 9.2 percent at a daily average temperature
-of \(0^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm
-rainfall a corresponding increase of the infection rate by approximately
-9.2 percent.
+perspective (Fig~\ref{fig:effectsrange}), with temperature we expect
+an increase by approximately 17.8 percent at a daily average temperature
+of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately
+11.4 percent compared to a day with no precipitation.
 
@@
 socio-demographic variables as confounders as well as the policy
-variables (cf.~Table \ref{tab:adjsets}, with the exception being
+variables (cf.~Table~\ref{tab:adjsets}, with the exception being
 COVID-19 burden). We also decided for the lag of 8 days based on the
@@
 
-\begin{table}
-
-\caption{\label{tab:adjsets}Final adjustment sets for causal analysis}
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\fontsize{7}{9}\selectfont
+\caption{
+{\bf Final adjustment sets for causal analysis.}}
+%\fontsize{8}{10}\selectfont
 \begin{tabular}[t]{llllllll}
@@
 \midrule
-Pseudo-R2 & 0.332 & 0.151 & 0.298 & 0.298 & 0.286 & 0.306 & 0.286\\
+Pseudo-$R^2$ & 0.332 & 0.151 & 0.298 & 0.298 & 0.286 & 0.306 & 0.286\\
 Weekday & x & x & x & x & x & x & x\\
@@
 \end{tabular}
+\label{tab:adjsets}
+\end{adjustwidth}
 \end{table}
 
-\section{Discussion}\label{discussion}
-
-\subsection{Main findings}\label{main-findings}
+%Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+
+%% Place tables after the first paragraph in which they are cited.
+%\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+%\centering
+%\caption{
+%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+%\begin{tabular}{|l+l|l|l|l|l|l|l|}
+%\hline
+%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+%\end{tabular}
+%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+%\end{flushleft}
+%\label{table1}
+%\end{adjustwidth}
+%\end{table}
+
+%%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
+%\subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
+%\subsubsection*{3rd level heading} 
+%Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
+Sed non aliquet felis.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris commodo justo ac dui pretium imperdiet.
+Sed suscipit iaculis mi at feugiat.
+%
+%\begin{enumerate}
+%	\item{react}
+%	\item{diffuse free particles}
+%	\item{increment time by dt and go to 1}
+%\end{enumerate}
+%
+%
+%\subsection*{Sed ac quam id nisi malesuada congue}
+%
+%Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero.
+%
+%\begin{itemize}
+%	\item First bulleted item.
+%	\item Second bulleted item.
+%	\item Third bulleted item.
+%\end{itemize}
+
+\section*{Discussion}\label{discussion}
+
+\subsection*{Main findings}\label{main-findings}
 
@@
 
-\subsection{Comparison with previous
+\subsection*{Comparison with previous
 research}\label{comparison-with-previous-research}
@@
 restricted to single aspects
-\citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
+\cite{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
 To reliably identify causal drivers, one must adjust for confounders. To
@@
 A causal approach is used in another current COVID-19 analysis
-\citep{gencoglu2020causal}. There, however, they identify the causal
+\cite{gencoglu2020causal}. There, however, they identify the causal
 relationships (reconstruct a DAG), while we estimated causal effects for
@@
 in the number of newly reported cases
-\citep{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
+\cite{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
 
 Our findings regarding awareness/Google Trends analysis are in good
-agreement with the correlations found by
-\citet{effenberger_association_2020}, \citet{higgins_correlations_2020},
-and \citet{yuan_trends_2020}, who conclude that alertness to COVID-19
+agreement with the correlations found by others \cite{effenberger_association_2020,higgins_correlations_2020,yuan_trends_2020}, who conclude that alertness to COVID-19
 rises several days before the highest number of cases are reported. At
@@
 influenced by public media coverage, which should be considered, if
-possible, in future studies \citep{higgins_correlations_2020}. As such,
+possible, in future studies \cite{higgins_correlations_2020}. As such,
 awareness is difficult to measure and here the number of Google searches
@@
 temperature on new COVID-19 cases
-\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
+\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
 It is however controversial to other scientific literature describing no
 effects
-\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
+\cite{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
 or even converse correlations
-\citep{auler_evidence_2020, xie_association_2020}. The conflicting
+\cite{auler_evidence_2020, xie_association_2020}. The conflicting
 results might be explained by different climates and characteristics of
@@
 and transmission. Finally, we found a positive causal effect of
-increment precipitation and a raise in COVID-19 cases, which supports
-previous observations \citep{sobral_association_2020}.
-
-\subsection{Limitations and strengths}\label{limitations-and-strengths}
+increased precipitation (rainfall) and a raise in COVID-19 cases, which supports
+previous observations \cite{sobral_association_2020}.
+
+\subsection*{Limitations and strengths}\label{limitations-and-strengths}
 
@@
 confounding. We did not try to construct the DAG from the available data
-(cf. \citep{gencoglu2020causal}). As such, our proposed DAG is not
+(cf.~\cite{gencoglu2020causal}). As such, our proposed DAG is not
 entirely consistent with the data and there are conditional dependencies
@@
 valid adjustment sets for these generalized graphs
-\citep{perkovic2017complete}. For the MAG derived from our DAG, only the
+\cite{perkovic2017complete}. For the MAG derived from our DAG, only the
 effects for exposures mobility and searches for corona can be estimated
@@
 regression, which is in line with the results on COVID-19 daily case
-counts of \citet{kraemer_effect_2020} and others
-\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
+counts of others \cite{kraemer_effect_2020,liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
 We did not model case counts with a differential equation model like the
-classic SIR-model \citep{kermack1991contributions} and its successors,
-since these are more suited to prediction
-\citep[e.g.][]{rkimodellierung} while our choice of a negative binomial
-regression framework allowed us to estimate the effects of confounders
+classic SIR-model \cite{kermack1991contributions} and its successors,
+since these are more suited to prediction, e.g.~\cite{rkimodellierung}, while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders
 more reliably. There are more advanced statistical methods for count
@@
 by the changing dynamics of case numbers (similar to
-\citep{Islamm2743}).
+\cite{Islamm2743}).
 
@@
 results are susceptible to the Modifiable Area Unit Problem (MAUP)
-\citep{Openshaw1984}. The MAUP postulates that different regional
+\cite{Openshaw1984}. The MAUP postulates that different regional
 aggregations of the units of observation may lead to different results
@@
 
-\subsection{Discussion of causal
+\subsection*{Discussion of causal
 effects}\label{discussion-of-causal-effects}
@@
 to the prevalent airborne and respiratory droplets and aerosol
-transmission of the SARS-CoV-2 virus \citep{world2020transmission}. In
+transmission of the SARS-CoV-2 virus \cite{world2020transmission}. In
 this light, we advocate for precautious measures like increased hygiene,
@@
 
-\subsection{Conclusions}\label{conclusions}
+\section*{Conclusion}\label{conclusions}
 
@@
 Our findings suggest that the causal effects of mobility, awareness, and
-weather need to be taken strongly into account when deciding for
+weather need to be taken into account when deciding for
 mitigation and suppression measures, depending on the recent and future
@@
 
+%Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
+Proin rutrum vel massa non gravida.
+Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Morbi at justo vitae nulla elementum commodo eu id massa.
+In vitae diam ac augue semper tincidunt eu ut eros.
+Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque.
+Praesent faucibus semper libero%~\cite{bib3}.
+
+%\section*{Conclusion}
+%
+%CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
+%
+%Sed non aliquet felis.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris commodo justo ac dui pretium imperdiet.
+Sed suscipit iaculis mi at feugiat.
+Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
+Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+Nam condimentum sem eget mollis euismod.
+Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
+Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
+Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
+For more information, see \nameref{S1_Appendix}.
+
+%\section*{Supporting information}
+%
+%% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+%\paragraph*{S1 Fig.}
+%\label{S1_Fig}
+%{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+%
+%\paragraph*{S2 Fig.}
+%\label{S2_Fig}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 File.}
+%\label{S1_File}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 Video.}
+%\label{S1_Video}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 Appendix.}
+%\label{S1_Appendix}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+%
+%\paragraph*{S1 Table.}
+%\label{S1_Table}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
 \section*{Acknowledgments}\label{acknowledgments}
-\addcontentsline{toc}{section}{Acknowledgments}
-
 We are thankful for feedback from Thomas Czihal, Johannes Textor, and
@@
 
-\renewcommand\refname{References}
-\bibliography{mybibfile.bib}
+\nolinenumbers
+
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+% 
+%\begin{thebibliography}{10}
+%
+%\bibitem{bib1}
+%Conant GC, Wolfe KH.
+%\newblock {{T}urning a hobby into a job: how duplicated genes find new
+%  functions}.
+%\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
+%
+%\bibitem{bib2}
+%Ohno S.
+%\newblock Evolution by gene duplication.
+%\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+%  Springer-Verlag.; 1970.
+%
+%\bibitem{bib3}
+%Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+%\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+%  infection through a transposon insertion followed by a {D}uplication}.
+%\newblock PLoS Genet. 2011 Oct;7(10):e1002337.
+%
+%\end{thebibliography}
+
+\bibliography{mybibfile}
 

### 2021-03-01 manuscript/v4_medrxiv_revision/figures/f_exposure_vs_effect.R

@@ -1,0 +1,96 @@
+library("tidyverse")
+library("cowplot")
+
+figurepath <- "./manuscript/v1_medrxiv/figures/"
+
+modeldata_raw <- read_csv("data/Modeldata_raw.csv")
+
+varnames <- c("Mobility (retail and recreation)",
+              "Mobility (grocery and pharmacy)",
+              "Mobility (parks)",
+              "Mobility (transit stations)",
+              "Mobility (workplaces)",
+              "Mobility (residential)",
+              "Weather (temperature)", "Weather (rainfall)", "Weather (humidity)", "Weather (wind)",
+              "Searches corona", "COVID-19 burden") 
+varnames <- factor(varnames, levels=varnames, ordered=TRUE)
+vardata <- tibble(varnames=varnames, means=0, effects=0, conflow=0, confhigh=0, rangelow=0, rangehigh=0)
+
+ranges <- apply(modeldata_raw[, varnames], 2, quantile, probs=c(0.025, 0.975))
+
+vardata <- vardata %>%
+  mutate(means=sapply(varnames, function(v) mean(modeldata_raw %>% pull(v) )),
+         rangelow=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[1]),
+         rangehigh=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[2]))
+vardata$myrangelow <- ranges[1, varnames] # c(-100, -25, -0, -100, -50, -10, -10, 0, 0, 0, 0, 0)
+vardata$myrangehigh <- ranges[2, varnames] # c(+50, +25, +100, +50, +0, +20, 30, 20, 100, 15, 100, 100)
+vardata[1:6, 3:5] <- cbind(read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$estimates,
+                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$conflow,
+                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$confhigh)
+vardata[11, 3:5] <- cbind(read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$estimates,
+                      read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$conflow,
[…]

### 2021-03-01 manuscript/v4_medrxiv_revision/figures/f_timeseries_cases_and_lograte.R

@@ -1,0 +1,40 @@
+library("tidyverse")
+library("cowplot")
+
+figurepath <- "./manuscript/v1_medrxiv/figures/"
+
+Modeldata <- read_csv("data/Modeldata_raw.csv")
+Sys.setlocale("LC_TIME", "C")
+mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
+
+mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
+
+cases <- Modeldata %>% dplyr::select(date, `Reported new cases COVID-19`, id) %>%
+  ggplot(aes(x=date,
+             y=`Reported new cases COVID-19`)) + 
+  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
+  geom_smooth(color=mycolors[1],size=2,se = F) + 
+  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") +
+  labs(x="Date", y="new cases") + 
+  scale_y_log10() +
+  theme(axis.title.x = element_blank()) +
+  theme_cowplot(font_size = 8)
+
+logrates <- Modeldata %>% dplyr::select(date, `Reported new cases COVID-19`, `Active cases`, id) %>%
+  ggplot(aes(x=date,
+             y=`Reported new cases COVID-19`/(`Active cases`+1))) + 
+  geom_point(aes(group=id), color="lightblue", stroke=0, alpha=0.2) +
+  geom_smooth(color=mycolors[1], size=2, se = FALSE) + 
+  scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)), date_labels = "%d %b") +
+  labs(x="Date", y="new cases/active cases") + 
+  scale_y_log10() +
+  theme(axis.title.x = element_blank()) +
+  theme_cowplot(font_size = 8)
+
+plot_grid(cases, logrates, labels = "AUTO",label_size = 12)
+
+ggsave(filename = paste0(figurepath, "f_timeseries_cases_and_lograte.eps"),
+       dpi=600, width=15,height = 7, unit="cm", device=cairo_ps, fallback_resolution = 600)
+
+# ggsave(filename = paste0(figurepath, "f_timeseries_cases_and_lograte.png"),
+#        dpi=600, width=15,height = 7, unit="cm", device=cairo_ps, fallback_resolution = 600)

### 2021-03-01 manuscript/v4_medrxiv_revision/manuscript.Rmd
add new folders for revision, test latexdiff
use latexdiffr

--- 2021-03-01 manuscript/v2_medrxiv/manuscript.Rmd
@@
 
+```{r rundifflatex, message=FALSE, warning=FALSE, include=FALSE}
+# careful, it's always one version behind
+library(latexdiffr)
+latexdiff("../v2_medrxiv/manuscript.tex", "manuscript.tex", "latexdiffrRmd")
+# system("latexdiff ../v2_medrxiv/manuscript.tex manuscript.tex > diffRmd.tex")
+# tinytex::pdflatex("diffRmd.tex")
+```
+
 # References

### 2021-03-01 R/2_dags_for_covidcases.R
check for linearity diagnostics

@@ -11,10 +11,10 @@ library(pscl)
 source("R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
-tablepath <- "./manuscript/v2_medrxiv/tables/"
+tablepath <- "manuscript/v2_medrxiv/tables/"
 
 # load dag
-dagfile <- "./data/bigdag4covid19.txt"
+dagfile <- "data/bigdag4covid19.txt"
 dagtext <- readChar(dagfile, file.info(dagfile)$size)
 dag <- dagitty::dagitty(dagtext)
 
@@ -27,7 +27,7 @@ unobserved <- c("Access to tests",
                 "Susceptibility")
 
 # Load Data
-modeldata <- read_csv("./data/Modeldata.csv") %>%
+modeldata <- read_csv("data/Modeldata.csv") %>%
   dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
                 -`Mobility (mean)`,
                 -`Relaxation of measures`,

### 2021-03-01 R/3_checklinearity.R

@@ -1,0 +1,58 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+source("R/z_auxiliary_functions_for_2.R")
+
+unobserved <- c("Access to tests", 
+                "Tracing capacity",
+                "Health care utilization",
+                "Herd immunity",
+                "Onset of COVID-19",
+                "Trust in society",
+                "Susceptibility")
+
+# Load Data
+modeldata <- read_csv("data/Modeldata.csv") %>%
+  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
+                -`Mobility (mean)`,
+                -`Relaxation of measures`,
+                -contains("iso"), -contains("census")
+  ) %>% 
+  rename(`School and kindergarten closures`=`School/Kita closures`,
+         `Holiday (exposure)`=Holiday,
+         `Foreign citizens`=`Foreign residents`,
+         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
+         `Rainfall`=`Weather (rainfall)`,
+         `Humidity`=`Weather (humidity)`,
+         `Gender`=`Sex`,
+         `Temperature`=`Weather (temperature)`,
+         `Wind`=`Weather (wind)`) %>%
+  filter(date<="2020-07-08") %>%
+  dplyr::select(-date)
+
+modeldata_cont <- modeldata %>%
+  dplyr::select(`Mobility (retail and recreation)`, `Mobility (grocery and pharmacy)`,
+         `Mobility (parks)`, `Mobility (transit stations)`,
+         `Mobility (workplaces)`, `Mobility (residential)`,
+         Rainfall, Temperature, Humidity, Wind,
+         `Searches corona`,
+         `Socio-economic status`, `Age (pop. 65 and older)`,
+         `Foreign citizens`, `Foreign citizens (refugees)`,
+         Turnout, `Right-wing populist party votes`,
+         `Population density`, Gender,
+         `Age (pop. younger 18)`, `Nursing homes`, 
+         `COVID-19 burden`, `Active cases`)
+myY <- log1p(modeldata$`Reported new cases COVID-19`/(1+modeldata$`Active cases`))
+myYagainstcont <- bind_cols(modeldata_cont, Outcome=myY) %>%
[…]

### 2021-03-09 R/1_generate_data.R
pca for mobility, sum of measures as variable

@@ -4,6 +4,8 @@ library(DBI)
 library(broom)
 library(MASS)
 
+maxanalysisdate <- as_date("2020-07-08")
+
 mydatapath <- "./data/" 
 
 google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
@@ -95,7 +97,9 @@ measures_zeroone_all <- measures_zeroone_all %>%
          `Mandatory face masks`=ifelse(id==8325049 & date>=as.Date("2020-04-17"), 1, `Mandatory face masks`), # rottweil
          `Mandatory face masks`=ifelse(id==16062041  & date>=as.Date("2020-04-14"), 1, `Mandatory face masks`), # nordhausen
          `Mandatory face masks`=ifelse(id==16053000  & date>=as.Date("2020-04-06"), 1, `Mandatory face masks`) # jena
-         )
+         ) %>% 
+  mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
+  dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
 
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
@@ -143,20 +147,62 @@ modeldata_raw <- left_join(brd_timeseries, google_mobility %>%
   left_join(., inkar, by=c("id"="id")) %>%
   left_join(., pflege_destatis, by=c("id"="id")) %>%
   left_join(., brd_burden, by=c('date'='date', 'id'='id')) %>%
-  filter(id>16 & date>=startdate & date<=enddate)# %>% drop_na()
+  filter(id>16 & date>=startdate & date<=enddate) %>%
+  rename(# `School and kindergarten closures`=`School/Kita closures`,
+         `Holiday (exposure)`=Holiday,
+         `Foreign citizens`=`Foreign residents`,
+         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
+         `Rainfall`=`Weather (rainfall)`,
+         `Humidity`=`Weather (humidity)`,
+         `Gender`=`Sex`,
+         `Temperature`=`Weather (temperature)`,
+         `Wind`=`Weather (wind)`) %>%
[…]

### 2021-03-09 R/2_dags_for_covidcases.R

@@ -27,23 +27,7 @@ unobserved <- c("Access to tests",
                 "Susceptibility")
 
 # Load Data
-modeldata <- read_csv("data/Modeldata.csv") %>%
-  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
-                -`Mobility (mean)`,
-                -`Relaxation of measures`,
-                -contains("iso"), -contains("census")
-                ) %>% 
-  rename(`School and kindergarten closures`=`School/Kita closures`,
-         `Holiday (exposure)`=Holiday,
-         `Foreign citizens`=`Foreign residents`,
-         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
-         `Rainfall`=`Weather (rainfall)`,
-         `Humidity`=`Weather (humidity)`,
-         `Gender`=`Sex`,
-         `Temperature`=`Weather (temperature)`,
-         `Wind`=`Weather (wind)`) %>%
-  filter(date<="2020-07-08") %>%
-  dplyr::select(-date)
+modeldata <- read_csv("data/Modeldata_scaled_pcamobility.csv")
 
 # all variables
 res <- my_causal(dag, modeldata, exposure=NULL, unobserved)

### 2021-03-09 R/3_modelassumptions.R

@@ -11,49 +11,19 @@ library(pscl)
 source("R/z_auxiliary_functions_for_2.R")
 
 # Load Data
-modeldata <- read_csv("data/Modeldata.csv") %>%
-  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
-                -`Mobility (mean)`,
-                -`Relaxation of measures`,
-                -contains("iso"), -contains("census")
-  ) %>% 
-  rename(`School and kindergarten closures`=`School/Kita closures`,
-         `Holiday (exposure)`=Holiday,
-         `Foreign citizens`=`Foreign residents`,
-         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
-         `Rainfall`=`Weather (rainfall)`,
-         `Humidity`=`Weather (humidity)`,
-         `Gender`=`Sex`,
-         `Temperature`=`Weather (temperature)`,
-         `Wind`=`Weather (wind)`) %>%
-  filter(date<="2020-07-08") %>%
-  dplyr::select(-date) %>%
-  mutate(dummy = 1) %>% # column with single value
-  spread(
-    key = "Weekday (exposure)", # column to spread
-    value = dummy,
-    fill = 0
-  ) %>%
-  dplyr::select(-"Weekday (report)", -"7Sun")
+modeldata <- bind_cols(read_csv("data/Modeldata_scaled_pcamobility.csv"),
+                       read_csv("data/Modeldata_raw.csv") %>%
+                         dplyr::select(`Reported new cases COVID-19`, `Active cases`))
 
-id_daycount <- read_csv("data/Modeldata.csv") %>%
+id_daycount <- read_csv("data/Modeldata_raw.csv") %>%
   dplyr::select(id, daycount, date, bl_id) %>%
-  filter(date<="2020-07-08") %>%
   dplyr::select(-date)
   
 
 modeldata_cont <- modeldata %>%
-  dplyr::select(`Mobility (retail and recreation)`, `Mobility (grocery and pharmacy)`,
-         `Mobility (parks)`, `Mobility (transit stations)`,
-         `Mobility (workplaces)`, `Mobility (residential)`,
-         Rainfall, Temperature, Humidity, Wind,
-         `Searches corona`,
[…]

### 2021-03-11 R/1_generate_data.R
save params for scaling

@@ -185,12 +185,12 @@ modeldata_X_cont <- modeldata_X %>% dplyr::select(`Mobility (retail and recreati
 scale_params <- tibble(variable=colnames(modeldata_X_cont),
                        mymean=colMeans(modeldata_X_cont),
                        mysd=apply(modeldata_X_cont, 2, sd))
+write_csv(scale_params, paste0(mydatapath, "scale_params.csv"))
 modeldata_X_cont_scaled <- sapply(seq(dim(modeldata_X_cont)[2]),
                                   function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], 2*scale_params[i, 3])) # gelman
 colnames(modeldata_X_cont_scaled) <- colnames(modeldata_X_cont)
 modeldata_X_cont_scaled <- as_tibble(modeldata_X_cont_scaled)
-modeldata_X_bin <- modeldata_X %>% dplyr::select(`Holiday (report)`, `Holiday (exposure)`, Interventions,
-                                          "1Mo", "2Di", "3Mi", "5Fr", "6Sa", "7So")
+modeldata_X_bin <- modeldata_X %>% dplyr::select(-colnames(modeldata_X_cont))
 
 modeldata_scaled <- bind_cols(modeldata_X_bin,
                               modeldata_X_cont_scaled)
@@ -202,7 +202,12 @@ modeldata_scaled_pca_mobility <- modeldata_scaled %>%
   bind_cols("Mobility (PC1)"=pca_mobility$x[,1],
             "Mobility (PC2)"=pca_mobility$x[,2],
             "Mobility (PC3)"=pca_mobility$x[,3],
-            "Mobility (PC4)"=pca_mobility$x[,4])
+            "Mobility (PC4)"=pca_mobility$x[,4],
+            "Mobility (PC5)"=pca_mobility$x[,5],
+            "Mobility (PC6)"=pca_mobility$x[,6])
+write_csv(as_tibble(pca_mobility$rotation, rownames="original_variable"), paste0(mydatapath, "pca_mobility_loadings.csv"))
 
-write_csv(modeldata_scaled, paste0(mydatapath,"Modeldata_scaled.csv"))
-write_csv(modeldata_scaled_pca_mobility,paste0(mydatapath,"Modeldata_scaled_pcamobility.csv"))
+write_csv(modeldata_scaled %>%
[…]

### 2021-03-11 R/z_auxiliary_functions_for_2.R
use aic for model selection

@@ -116,6 +116,7 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
   coefficients <- vector("list", nf)
   causaleffects <- vector("list", nf)
   pseudor2s <- rep(0, nf)
+  aics <- rep(0, nf)
   mynullglm <- glm.nb(as.formula(mynullformula),
                       data=modeldata)
   for (idxf in seq(nf)) {
@@ -127,7 +128,9 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
     #                                     `Weekday (report)`+`Holiday (report)`")), data = modeldata)
     # pseudo RSquared
     pseudor2s[idxf] <- 1-sum((myglm$fitted.values-myglm$y)^2)/sum((mynullglm$fitted.values-mynullglm$y)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
+    aics[idxf] <- AIC(myglm)
     cat("   pseudo R2:", pseudor2s[idxf], "\n")
+    cat("   AIC:", aics[idxf], "\n")
     coefficients[[idxf]] <- tidy(myglm, exponentiate=TRUE, conf.int=TRUE, conf.level = 0.99)
     if (!is.null(exposure)) {
       effects <- tail(coefficients[[idxf]]$estimate, nexp)
@@ -143,7 +146,7 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
       causaleffects[[idxf]] <- coefficients[[idxf]]
     }
   }
-  return(list(pseudor2s=pseudor2s, causaleffects=causaleffects))
+  return(list(pseudor2s=pseudor2s, aics=aics, causaleffects=causaleffects))
 }
 
 # all-in-one function
@@ -161,15 +164,18 @@ my_causal <- function(dag, modeldata, exposure, unobserved) {
 # write final causal effects to csv
 write_cause <- function(res, exposure) {
   whichmaxr2 <- which.max(res$pseudor2s)
+  whichminaic <- which.max(res$aics)
   line <- paste0(exposure, ",", max(res$pseudor2s))
   write(line, file=paste0(tablepath, "t_pseudor2s.csv"), append=TRUE)
-  write_csv(res$causaleffects[[whichmaxr2]],
+  line <- paste0(exposure, ",", min(res$aics))
+  write(line, file=paste0(tablepath, "t_aics.csv"), append=TRUE)
+  write_csv(res$causaleffects[[whichminaic]],
[…]

### 2021-03-11 R/3_modelassumptions.R
use all 6 principal components of mobility for regression

@@ -29,7 +29,7 @@ exposure=NULL
 adjsets=NULL
 myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
 myglm <- glm.nb(as.formula(myformula),
-                  data=modeldata)
+                  data=modeldata) # %>%dplyr::select(-`Mobility (PC4)`)
 mynullformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))" 
 mynullglm <- glm.nb(as.formula(mynullformula),
                     data=modeldata)
@@ -84,3 +84,17 @@ corrplot(corrplotscores$r,
 library(car)
 myvifs <- vif(myglm)
 myvifs[myvifs>=5]
+
+# rescale principal components of mobility to original mobility variables
+pcaloadings <- read_csv("data/pca_mobility_loadings.csv")
+orig_mobi <- as.vector(exp(as.matrix(pcaloadings[, 2:7]) %*% myglm$coefficients[c(
+  "`Mobility (PC1)`",
+  "`Mobility (PC2)`",
+  "`Mobility (PC3)`",
+  "`Mobility (PC4)`",
+  "`Mobility (PC5)`",
+  "`Mobility (PC6)`"
+)]))
+names(orig_mobi) <- pcaloadings$original_variable
+orig_mobi
+coeffs <- tidy(myglm, exponentiate=TRUE) # , conf.int=TRUE, conf.level = 0.99

### 2021-03-11 R/2_dags_for_covidcases.R
interventions as sum

@@ -11,7 +11,7 @@ library(pscl)
 source("R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
-tablepath <- "manuscript/v2_medrxiv/tables/"
+tablepath <- "manuscript/v4_medrxiv_revision/tables/"
 
 # load dag
 dagfile <- "data/bigdag4covid19.txt"
@@ -33,6 +33,7 @@ modeldata <- read_csv("data/Modeldata_scaled_pcamobility.csv")
 res <- my_causal(dag, modeldata, exposure=NULL, unobserved)
 
 file.create(paste0(tablepath, "t_pseudor2s.csv"))
+file.create(paste0(tablepath, "t_aics.csv"))
 # mobility
 res <- my_causal(dag, modeldata, exposure="Mobility", unobserved=unobserved)
 write_cause(res, exposure="Mobility")
@@ -49,22 +50,8 @@ for (myweather in c("Temperature", "Rainfall", "Humidity", "Wind")) {
   write_cause(res, exposure=myweather)
 }
 
-# maskenpflicht
-res <- my_causal(dag, modeldata, exposure="Mandatory face masks", unobserved=unobserved)
-
-# kontaktsperre 
-res <- my_causal(dag, modeldata, exposure="Contact restrictions", unobserved=unobserved)
-
-# school/kita closure adjustment 
-res <- my_causal(dag, modeldata, exposure="School and kindergarten closures", unobserved=unobserved)
-
-
-# "cluster" Pflegeheime adjustment 
-res <- my_causal(dag, modeldata, exposure="Nursing homes", unobserved=unobserved)
-
-# "cluster" Age? --> schools
-res <- my_causal(dag, modeldata, exposure="Age", unobserved=unobserved)
-
+# interventions
+res <- my_causal(dag, modeldata, exposure="Interventions", unobserved=unobserved)
 
 # tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
 # ggdag(tidy_dag,text_size = 1.5,color="grey",

### 2021-03-11 manuscript/v5_submission_revision/replyreviewer.Rmd
reply to reviewer

---
title: "reply to reviewer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*This is an interesting manuscript that aims to identify "causal" relationships" for COVID-19 transmission based on a DAG analysis and empirical data from other studies for variable selection.*

Yes, we aim to estimate causal effects based on a DAG that was constructed based on other publications. Empirical data was extracted and aggregated from different publicly available databases: RKI, Google mobility/Trends, INKAR, DWD. Variable selection for the different models was based on the proposed DAG using the well-studied methods from Pearl etc.

*The DAG analysis is a multivariate analysis to select a subset of variables that are correlated from a large number of candidates. The first select the sets of variables that are associated with the outcome using DAG analysis and then put them into the regression. The DAG analysis returns different sets of variables, using different selection criterion ( for example, the most parsimonious set or the most variation explained ). So the authors assumedly use R-squared to select the final sets of variables in the regression model.*

Yes, we identify subsets of variables using DAG analysis. We stress the point that we do this identification process separately for every variable of interest where we are interested in the causal effect. We always regress on the same outcome Y (Reported COVID-19 cases) using negative-binomial regression. That is, we select one variable Xi where we are interested in the effect of Xi on Y. Then we identify all possible subsets of variables in the DAG that allow to estimate the causal effect of Xi on Y (based on Pearl etc). Out of these subsets we choose the "best" one in regards to pseudo-R2 (or AIC). We repeat this procedure for several variables Xi.

*The challenge I have here (and having reviewed this as well with our statistical team) is that the claims of causal relationships are not convincing, even if they identify associative relationships that other studies have also found. Specifically: 1) The model does not address the nonlinearity of variables (such as temperature); other studies have found that temperature has a u-shaped effect at low and high temperatures. Humidity effects may also be nonlinear. The assumption of non-linearity on continuous variables needs to be considered.*

Yes, non-linearity might be a problem for our generalized linear model (negative-binomial regression). We added a complete check of model diagnostics that addresses linearity assumptions, model specification, independence of observations, and multi-collinearity checks (see below). We did not find significant problems with non-linear relationships (see plot XX supplementary?).

*2) There are no tests for multi-collinearity that are presented, which presents significant concerns for over-fitting the model. Temperature and humidity are an example here. So is the mobility data.*

This an important remark. We added a correlation plot for the continuous variables and assessed variance inflation factors (VIFs) for the full model. We observed high VIFs for the mobility variables and the different intervention variables. We decided to transform the original mobility variables by PCA and used the set of these principal components instead of the original mobility variables. Furthermore, we summarised the intervention variables into one "Intervention" variable (compare new DAG) by summing these binary variables. Analysing the VIFs on the full model with these transformed variables, we are left with only one VIF>5 (PC Mobi 1=5.1xxx), which is a huge improvement on the original model. We want to add that causal graph analysis for regression is more robust to multi-collinearity than traditional regression approaches (cite paper).

*2) Feels like they cherry-pick the variables of interest in the end? Not sure how they arrived at the final variable list for causal effects. Whey were the restrictions not included in that analysis?*

*3) The term “causal analysis” is a bit strong for what they have done here. The basis of the work is the proposed DAG (Figure 1), but the diagram was constructed from other association studies. Other than the DAG analysis, they did not do anything to ensure the results are “causal relationship”. So I am not convinced that the analysis or results are causal.How do you avoid over-fitting the model or including mediating variables in the analysis?*

*5) What about interaction terms? For example, residential mobility and colder weather? Or rain? These relationships are not simply a straightforward multivariable model.*

*6) The most interesting terms in their model were the interventions (restrictions) themselves, yet they were dropped from the model. No discussion is really pursued on that. Why were public health interventions removed?*

### 2021-03-12 manuscript/v5_submission_revision/replyreviewer.Rmd
update reviewer reply

@@
 
+At the time of writing, we were more interested in variables that could be prognostive of future developments (especially weather, since weather is not modifiable).
+In light of this remark, we added an overview of the effect estimates derived from the causal graph for all observed variables to the supplementary and discussed them briefly in the main paper.
+
 *3) The term “causal analysis” is a bit strong for what they have done here.
@@
 
+We use the term "causal" the same way that Pearl (cite XXX) and the growing body of work based on causal analysis with DAGs use it.
+We modified the title of our paper to "Causal graph analysis of [...]" to reflect the fundamental assumption of a causal graph to be able to do causal analysis.
+We are happy to discuss the structure of our proposed DAG.
+
+Proper identification of valid variable subsets for causal effect estimation takes care of the mediating variable problem, since this is at the heart of causal DAG analysis (see Pearl page xxx?).
+
+Overfitting might still be a problem.
+Since the time of submission the first version of the paper, we were able to "collect" more data of the COVID-19 pandemic.
+For a sensitivity analysis in regards to overfitting, we re-ran our analysis pipeline on the completely separate dataset from XX-07-2020 to XX-0X-2021 and discuss the corresponding effect estimates in comparison to the results of the original dataset (main paper and supplementary XXX).
+
 *5) What about interaction terms? For example, residential mobility and colder weather? Or rain? These relationships are not simply a straightforward multivariable model.*
+
+DAG analysis explicitly takes care of interaction between variables by identifying valid subsets of variables that need to be included or excluded in the multivariable model, see (cite Pearl some page/chapter?).
+We added a remark on interaction to the paper (see ...) to reflect on this remark.
 
@@
 Why were public health interventions removed?*
+
+See our answer x) above. We added effect estimates for all variables.

### 2021-03-12 R/1_generate_data.R
fix weekday variable and functions such that adjustment sets and non-identifiable problems are aved properly

@@ -159,12 +159,13 @@ modeldata_raw <- left_join(brd_timeseries, google_mobility %>%
          `Wind`=`Weather (wind)`) %>%
   filter(date<=maxanalysisdate) %>%
   mutate(dummy = 1) %>% # column with single value
-  spread(
-    key = "Weekday (exposure)", # column to spread
-    value = dummy,
-    fill = 0
+  pivot_wider(
+    names_from = "Weekday (exposure)", # column to spread
+    names_prefix = "Weekday ",
+    values_from = dummy,
+    values_fill = 0
   ) %>%
-  dplyr::select(-"Weekday (report)", -"4Do", -contains("iso"), -contains("census"))
+  dplyr::select(-"Weekday (report)", -"Weekday 4Do", -contains("iso"), -contains("census"))
 
 write_csv(modeldata_raw,paste0(mydatapath,"Modeldata_raw.csv"))
 

### 2021-03-12 R/2_dags_for_covidcases.R

@@ -21,6 +21,7 @@ dag <- dagitty::dagitty(dagtext)
 unobserved <- c("Access to tests", 
                 "Tracing capacity",
                 "Health care utilization",
+                "Exposure to SARS-COV-2",
                 "Herd immunity",
                 "Onset of COVID-19",
                 "Trust in society",
@@ -53,27 +54,11 @@ for (myweather in c("Temperature", "Rainfall", "Humidity", "Wind")) {
 # interventions
 res <- my_causal(dag, modeldata, exposure="Interventions", unobserved=unobserved)
 
-# tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
-# ggdag(tidy_dag,text_size = 1.5,color="grey",
-#       edge_type = "link",node_size = 13,label_size = 1.0) + 
-#   ggdag::theme_dag_blank(base_size = 4)
-# 
-# tidy_dag %>% 
-#   ggplot(aes(x = x, y = y, xend = xend, yend = yend))+
-#   geom_dag_edges_diagonal(check_overlap = TRUE) +
-#   geom_dag_point(size=10,color="lightgrey") +
-#   geom_dag_text(col = "black",size=1.2) +
-#   theme_solid()
-
-# library(igraph)
-# data <- dagitty_to_adjmatrix(dag)
-# network <- graph_from_adjacency_matrix(data , mode='directed', diag=FALSE)
-# 
-# plot(network, layout=layout.sphere, main="sphere")
-# plot(network, layout=layout.circle, main="circle")
-# plot(network, layout=layout.random, main="random")
-# plot(network, layout=layout.fruchterman.reingold,
-#      vertex.size=10,
-#      edge.arrow.size=0.5,
-#      margin=-0.5,
-#      vertex.label.cex=0.5)
+# other variables
+othervariables <- setdiff(names(dag), c("Reported new cases COVID-19", unobserved,
+                                        "Mobility", "COVID-19 burden", "Searches corona",
+                                        "Temperature", "Rainfall", "Humidity", "Wind"))
+for (ov in othervariables) {
+  res <- my_causal(dag, modeldata, exposure=ov, unobserved=unobserved)
+  write_cause(res, exposure=ov)
+}

### 2021-03-12 R/z_auxiliary_functions_for_2.R

@@ -73,10 +73,12 @@ get_adjsets <- function(dag, exposure, outcome, unobserved) {
   if (length(adjsets)>0) {
     names(adjsets) <- paste0("MinAdjSet", seq(valid_adjsets))
     optadjset <- get_optadjset(amat, exposure, outcome, unobserved)
-    if (optadjset[[2]]) { # is a reduced set
-      adjsets[["RedOptAdjSet"]] <- optadjset[[1]]
-    } else { # is actually real optimal adjustment set
-      adjsets[["OptAdjSet"]] <- optadjset[[1]]
+    if (!is.null(optadjset)) {
+      if (optadjset[[2]]) { # is a reduced set
+        adjsets[["RedOptAdjSet"]] <- optadjset[[1]]
+      } else { # is actually real optimal adjustment set
+        adjsets[["OptAdjSet"]] <- optadjset[[1]]
+      }
     }
   }
   return(adjsets)
@@ -156,26 +158,34 @@ my_causal <- function(dag, modeldata, exposure, unobserved) {
   } else {
     adjsets <- get_adjsets(dag, exposure, outcome="Reported new cases COVID-19", unobserved)
   }
-  res <- my_negbin(modeldata, exposure=exposure, adjsets=adjsets)
-  res[["adjustmentsets"]] <- adjsets
+  if (length(adjsets)>0 | is.null(exposure)) {
+    res <- my_negbin(modeldata, exposure=exposure, adjsets=adjsets)
+    res[["adjustmentsets"]] <- adjsets
+  } else {
+    res <- NULL
+  }
   return(res)
 }
 
 # write final causal effects to csv
 write_cause <- function(res, exposure) {
-  whichmaxr2 <- which.max(res$pseudor2s)
-  whichminaic <- which.max(res$aics)
-  line <- paste0(exposure, ",", max(res$pseudor2s))
-  write(line, file=paste0(tablepath, "t_pseudor2s.csv"), append=TRUE)
-  line <- paste0(exposure, ",", min(res$aics))
-  write(line, file=paste0(tablepath, "t_aics.csv"), append=TRUE)
-  write_csv(res$causaleffects[[whichminaic]],
-            paste0(tablepath, "t_effects_",
-                   exposure, "_", 
-                   names(res$adjustmentsets[whichminaic]), ".csv"))
-  write(res$adjustmentsets[[whichminaic]],
-        paste0(tablepath, "t_adjust_",
-               exposure, "_", 
[…]

### 2021-03-12 secondwave_analysis/R/0_1_rkicases.R
add second wave analysis

@@ -1,0 +1,64 @@
+library(tidyverse)
+library(lubridate)
+
+csvurl <- "https://opendata.arcgis.com/datasets/dd4580c810204019a7b8eb3e0b329dd6_0.csv"
+mytemp <- tempfile()
+options(timeout=300)
+download.file(csvurl, mytemp, quiet=TRUE)
+myresult <- read_csv(mytemp)
+localrki <- myresult %>%
+  mutate(Meldedatum= as.character(as_date(Meldedatum)),
+         Refdatum= as.character(as_date(Refdatum)),
+         Datenstand=as.character(as.Date(Datenstand, format="%d.%m.%Y")))
+  
+  rkitimeframe <- localrki %>% summarise(mindate=min(date(Meldedatum)),maxdate=max(date(Meldedatum)))
+  rkidays <- date(rkitimeframe$maxdate)-date(rkitimeframe$mindate)
+  rkidates = date(now())-seq(1,rkidays+1)
+  
+  Kreise <- localrki %>%  mutate(date=date(Meldedatum)) %>%
+    rename(id=IdLandkreis,cases=AnzahlFall,deaths=AnzahlTodesfall,recovered=AnzahlGenesen) %>%
+    dplyr::select(id,date,cases,deaths,recovered) %>% group_by(id,date) %>%
+    summarise(cases=sum(cases,na.rm=T),deaths=sum(deaths,na.rm=T),recovered=sum(recovered,na.rm=T)) %>%
+    ungroup() %>% mutate(id=as.integer(id)*1000) %>%
+    filter(!is.na(id)) %>% arrange(id,date)
+  
+  Kreise <- expand.grid("id"=unique(Kreise$id),"date"=rkidates) %>% as_tibble(.) %>%
+    arrange(id,date) %>%
+    left_join(Kreise,by=c("id","date")) %>%
+    mutate(cases=ifelse(is.na(cases),0,cases),
+           deaths=ifelse(is.na(deaths),0,deaths),
+           recovered=ifelse(is.na(recovered),0,recovered)) %>%
+    mutate(date=as.character(date))
+  
+  Laender <- localrki %>% mutate(date=date(Meldedatum)) %>%
+    rename(id=IdBundesland,cases=AnzahlFall,deaths=AnzahlTodesfall,recovered=AnzahlGenesen) %>%
+    dplyr::select(id,date,cases,deaths,recovered) %>% group_by(id,date) %>%
[…]

### 2021-03-12 secondwave_analysis/R/0_2_mobility.R

@@ -1,0 +1,41 @@
+library(sf)
+library(tidyverse)
+library(geojsonsf)
+library(geojsonio)
+
+mydatapath <- "./data/mobility/"
+
+# google data
+gmobillink <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
+download.file(gmobillink, 
+              paste0(mydatapath, "googlemobility_raw.csv"),
+              quiet=TRUE)
+google_german_mobility <- read_csv(paste0(mydatapath, "googlemobility_raw.csv")) %>%
+  filter(country_region_code=="DE") %>%
+  dplyr::select(-country_region_code, -country_region, -sub_region_2, -metro_area,
+                -census_fips_code)
+
+en_rs_map_bl <- c("08", "09", "11", "12",
+                  "04", "02", "06", "03",
+                  "13", "05", "07", "10",
+                  "14", "15","01", "16",
+                  "00")
+# names(en_de_map_bl) <- c("Baden-Württemberg", "Bayern", "Berlin", "Brandenburg",
+#                          "Bremen", "Hamburg", "Hessen", "Niedersachsen",
+#                          "Mecklenburg-Vorpommern", "Nordrhein-Westfalen", "Rheinland-Pfalz", "Saarland",
+#                          "Sachsen", "Sachsen-Anhalt","Schleswig-Holstein", "Thüringen")
+names(en_rs_map_bl) <- c("Baden-Württemberg", "Bavaria", "Berlin", "Brandenburg",
+                         "Bremen", "Hamburg", "Hessen", "Lower Saxony",
+                         "Mecklenburg-Vorpommern", "North Rhine-Westphalia", "Rhineland-Palatinate", "Saarland",
+                         "Saxony", "Saxony-Anhalt","Schleswig-Holstein", "Thuringia",
+                         "Gesamt")
+google_german_mobility <- google_german_mobility %>%
+  mutate(sub_region_1=ifelse(is.na(sub_region_1), "Gesamt", sub_region_1)) %>%
+  mutate(grocery_and_pharmacy_percent_change_from_baseline=replace_na(grocery_and_pharmacy_percent_change_from_baseline, 0)) %>%
+  mutate(parks_percent_change_from_baseline=replace_na(parks_percent_change_from_baseline, 0)) %>%
[…]

### 2021-03-12 secondwave_analysis/R/0_3_awareness.R

@@ -1,0 +1,92 @@
+library(tidyverse)
+library(lubridate)
+
+mydatapath <- "./data/awareness/"
+
+bundeslaender_maps <- tibble(
+  BL_ID=1:16,
+  Bundesland=c(
+    "Schleswig-Holstein",
+    "Hamburg",
+    "Niedersachsen",
+    "Bremen",
+    "Nordrhein-Westfalen",
+    "Hessen",
+    "Rheinland-Pfalz",
+    "Baden-Württemberg",
+    "Bayern",
+    "Saarland",
+    "Berlin",
+    "Brandenburg",
+    "Mecklenburg-Vorpommern",
+    "Sachsen",
+    "Sachsen-Anhalt",
+    "Thüringen"
+  ),
+  Kuerzel=c(
+    "SH",
+    "HH",
+    "NI",
+    "HB",
+    "NW",
+    "HE",
+    "RP",
+    "BW",
+    "BY",
+    "SL",
+    "BE",
+    "BB",
+    "MV",
+    "SN",
+    "ST",
+    "TH"
+  )
+)
+
+trends_alle_bl <- tibble()
+
+for (kuerzel in bundeslaender_maps$Kuerzel) {
+  this_bl <- read_csv(paste0("data/awareness/trends_", kuerzel, ".csv"), 
+                      col_names = c("woche", "searches"),
+                      col_types = cols(woche=col_date(format = "%Y-%m-%d"), 
+                                       searches=col_double()), 
+                      skip = 2) %>%
+    mutate(searches=ifelse(is.na(searches), 0, searches),
+           BL_ID=bundeslaender_maps%>%filter(Kuerzel==kuerzel)%>%pull(BL_ID),
+           Kuerzel=kuerzel,
+           Bundesland=bundeslaender_maps%>%filter(Kuerzel==kuerzel)%>%pull(Bundesland)) %>%
+    drop_na()
+  trends_alle_bl <- bind_rows(trends_alle_bl,
+                              this_bl)
+}
+
+relativ_gesamt_2020_03_15 <- read_csv(paste0(mydatapath, 
+                                             "relativ_gesamt_2020-03-15.csv"))
+
+relativ_bl_2020_03_15 <- left_join(trends_alle_bl %>% 
+                                     filter(woche=="2020-03-15"), 
+                                   relativ_gesamt_2020_03_15,
+                                   by="Bundesland") %>%
+  mutate(baseline20200315=searches*Relativtrend/100) %>%
[…]

### 2021-03-12 secondwave_analysis/R/0_5_weather.R

@@ -1,0 +1,203 @@
+# the download of weather files takes a while! ~ 1 h
+
+library(readr)
+library(tidyverse)
+
+mydatapath <- "./data/weather/"
+
+opendata_dwd_path <- "https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/recent/"
+
+download.file(paste0(opendata_dwd_path, "KL_Tageswerte_Beschreibung_Stationen.txt"),
+              paste0(mydatapath, "weatherstations.txt"))
+
+header <- read_table2(file = paste0(mydatapath, "weatherstations.txt"),
+                   col_names = FALSE,
+                   n_max = 1)
+weatherdata <- read_table(file = paste0(mydatapath, "weatherstations.txt"),
+                    col_names = FALSE,
+                    locale = locale(encoding = "WINDOWS-1252"),
+                    skip = 2)
+colnames(weatherdata) <- header[1, ]
+
+extractedfiles <- c()
+
+best_bisdatum <- last(names(sort(table(weatherdata$bis_datum))))
+
+weatherdata <- weatherdata %>%
+  filter(bis_datum==as.numeric(best_bisdatum) & von_datum <= 20200101 & Stationshoehe<1000)
+
+for (stationid in weatherdata$Stations_id) {
+  tryCatch({
+    download.file(paste0(opendata_dwd_path, "tageswerte_KL_", stationid, "_akt.zip"),
+                  paste0(mydatapath, "temp.zip"),
+                  quiet=TRUE)
+    cat("downloaded data for", stationid, "successfully\n")
+    listoffilesinzip <- unzip(paste0(mydatapath, "temp.zip"),
+                              list=TRUE)
+    extractedfile <- unzip(paste0(mydatapath, "temp.zip"),
+                           files=last(listoffilesinzip$Name),
+                           exdir=paste0(mydatapath, "temp"))
+    extractedfiles <- c(extractedfiles, extractedfile)
+  }, error = function(e) {
+    cat(stationid, "did not work\n")
+  }, warning = function(w) {
+    cat(stationid, "did not work\n")
+  })
+  
+}
+
+write_lines(extractedfiles,
[…]

### 2021-03-12 secondwave_analysis/R/0_z_testlags.R

@@ -1,0 +1,357 @@
+mylags <- 1:20
+
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+
+mydatapath <- "./data/" 
+
+google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
+  mutate(id=as.integer(RS_first2)) %>%
+  dplyr::select(-RS_first2) %>%
+  pivot_wider(names_from = areatype, values_from = perc_change) %>%
+  rename(`Mobility (retail and recreation)`=retail_and_recreation_percent_change_from_baseline,
+         `Mobility (grocery and pharmacy)`=grocery_and_pharmacy_percent_change_from_baseline,
+         `Mobility (parks)`=parks_percent_change_from_baseline,
+         `Mobility (transit stations)`=transit_stations_percent_change_from_baseline,
+         `Mobility (workplaces)`=workplaces_percent_change_from_baseline,
+         `Mobility (residential)`=residential_percent_change_from_baseline)
+
+weather <- read_csv(paste0(mydatapath, "weather/mittleres_kreiswetter.csv")) %>%
+  mutate(date=ymd(MESS_DATUM), id=as.integer(RS)*1000) %>%
+  dplyr::select(-MESS_DATUM, -RS) %>%
+  mutate(Wind=replace_na(Wind, mean(Wind, na.rm=TRUE)),
+         Feuchtigkeit=replace_na(Feuchtigkeit, mean(Feuchtigkeit, na.rm=TRUE))) %>%
+  rename(`Weather (rainfall)`=Niederschlag,
+         `Weather (wind)`=Wind,
+         `Weather (temperature)`=Temperatur,
+         `Weather (humidity)`=Feuchtigkeit)
+
+awareness <- read_csv(paste0(mydatapath, "awareness/awareness.csv")) %>%
+  rename(`Searches corona`=relativtrend) %>%
+  dplyr::select(-Bundesland)
+
+pflege_destatis <- read_delim(paste0(mydatapath, "sociodemographic/pflege_destatis.csv"), 
+                              ";", escape_double = FALSE, locale = locale(decimal_mark = ",", 
+                                                                          grouping_mark = ".", encoding = "WINDOWS-1252"), 
+                              trim_ws = TRUE, skip = 6, na = "-") %>%
[…]

### 2021-03-12 secondwave_analysis/R/1_generate_data.R

@@ -1,0 +1,216 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+source("R/z_auxiliary_functions_for_1.R")
+
+mydatapath <- "./data/" 
+
+google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
+  mutate(id=as.integer(RS_first2)) %>%
+  dplyr::select(-RS_first2) %>%
+  pivot_wider(names_from = areatype, values_from = perc_change) %>%
+  rename(`Mobility (retail and recreation)`=retail_and_recreation_percent_change_from_baseline,
+         `Mobility (grocery and pharmacy)`=grocery_and_pharmacy_percent_change_from_baseline,
+         `Mobility (parks)`=parks_percent_change_from_baseline,
+         `Mobility (transit stations)`=transit_stations_percent_change_from_baseline,
+         `Mobility (workplaces)`=workplaces_percent_change_from_baseline,
+         `Mobility (residential)`=residential_percent_change_from_baseline)
+
+weather <- read_csv(paste0(mydatapath, "weather/mittleres_kreiswetter.csv")) %>%
+  mutate(date=ymd(MESS_DATUM), id=as.integer(RS)*1000) %>%
+  dplyr::select(-MESS_DATUM, -RS) %>%
+  mutate(Wind=replace_na(Wind, mean(Wind, na.rm=TRUE)),
+         Feuchtigkeit=replace_na(Feuchtigkeit, mean(Feuchtigkeit, na.rm=TRUE)),
+         Luftdruck=replace_na(Luftdruck, mean(Luftdruck, na.rm=TRUE))) %>%
+  rowwise() %>%
+  mutate(Indoorfeuchtigkeit=qfun(Feuchtigkeit, Temperatur, Luftdruck)) %>%
+  rename(`Weather (rainfall)`=Niederschlag,
+         `Weather (wind)`=Wind,
+         `Weather (temperature)`=Temperatur,
+         `Weather (humidity)`=Feuchtigkeit,
+         `Weather (indoor humidity)`=Indoorfeuchtigkeit) %>%
+  dplyr::select(-Luftdruck)
+
+awareness <- read_csv(paste0(mydatapath, "awareness/awareness.csv")) %>%
+  rename(`Searches corona`=relativtrend) %>%
[…]

### 2021-03-12 secondwave_analysis/R/2_dags_for_covidcases.R

@@ -1,0 +1,120 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+source("R/z_auxiliary_functions_for_2.R")
+
+# save tables to:
+tablepath <- "./manuscript/v4_indoorhumidity/tables/"
+
+# load dag
+dagfile <- "./data/bigdag4covid19.txt" # "./data/bigdag4covid19_indoortemp.txt" # 
+dagtext <- readChar(dagfile, file.info(dagfile)$size)
+dag <- dagitty::dagitty(dagtext)
+
+unobserved <- c("Access to tests", 
+                "Tracing capacity",
+                "Health care utilization",
+                "Herd immunity",
+                "Onset of COVID-19",
+                "Trust in society",
+                "Susceptibility")
+
+# Load Data
+modeldata <- read_csv("./data/Modeldata_scaled.csv") %>%
+  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
+                -`Mobility (mean)`,
+                # -`Relaxation of measures`,
+                -contains("iso"), -contains("census")
+                ) %>% 
+  rename(`School and kindergarten closures`=`School/Kita closures`,
+         `Holiday (exposure)`=Holiday,
+         `Foreign citizens`=`Foreign residents`,
+         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
+         `Rainfall`=`Weather (rainfall)`,
+         `Humidity`=`Weather (humidity)`,
+         `Indoor humidity`=`Weather (indoor humidity)`,
+         `Gender`=`Sex`,
+         `Temperature`=`Weather (temperature)`,
+         `Wind`=`Weather (wind)`) %>%
+  # filter(date<="2020-12-31") %>%
+  dplyr::select(-date)
+
+# all variables
+res <- my_causal(dag, modeldata, exposure=NULL, unobserved)
+
+file.create(paste0(tablepath, "t_pseudor2s.csv"))
+# mobility
+res <- my_causal(dag, modeldata, exposure="Mobility", unobserved=unobserved)
+write_cause(res, exposure="Mobility")
+
+# awareness
[…]

### 2021-03-12 secondwave_analysis/R/z_auxiliary_functions_for_1.R

@@ -1,0 +1,17 @@
+# transform relative outdoor humidity and temperature to indoor humidity
+qfun <- function(humidity, temperature, pressure, maxq=8){
+  rh <- humidity/100
+  t <- temperature+273.15
+  P <- pressure*100
+  resq <- min(maxq, 1000*5/8*1/P*rh*611.2*exp(17.62*(t-273.15)/(t-30.03)))
+  return(resq)
+}
+
+qlevel <- function(humidity, temperature, pressure, cutoffq=6){
+  rh <- humidity/100
+  t <- temperature+273.15
+  P <- pressure*100
+  resq <- 1000*5/8*1/P*rh*611.2*exp(17.62*(t-273.15)/(t-30.03))
+  rlevel <- ifelse(resq<cutoffq, 0, 1)
+  return(rlevel)
+}

### 2021-03-12 secondwave_analysis/R/1_generate_data.R
prepare second wave analysis

@@ -3,9 +3,9 @@ library(tidyverse)
 library(DBI)
 library(broom)
 library(MASS)
-source("R/z_auxiliary_functions_for_1.R")
+source("secondwave_analysis/R/z_auxiliary_functions_for_1.R")
 
-mydatapath <- "./data/" 
+mydatapath <- "secondwave_analysis/data/" 
 
 google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
   mutate(id=as.integer(RS_first2)) %>%

### 2021-03-12 secondwave_analysis/R/2_dags_for_covidcases.R

@@ -8,13 +8,13 @@ library(ggdag)
 library(ggthemes)
 library(pcalg)
 library(pscl)
-source("R/z_auxiliary_functions_for_2.R")
+source("secondwave_analysis/R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
-tablepath <- "./manuscript/v4_indoorhumidity/tables/"
+tablepath <- "manuscript/v4_revision/tables/secondwave_analysis/"
 
 # load dag
-dagfile <- "./data/bigdag4covid19.txt" # "./data/bigdag4covid19_indoortemp.txt" # 
+dagfile <- "data/bigdag4covid19.txt"
 dagtext <- readChar(dagfile, file.info(dagfile)$size)
 dag <- dagitty::dagitty(dagtext)
 
@@ -27,7 +27,7 @@ unobserved <- c("Access to tests",
                 "Susceptibility")
 
 # Load Data
-modeldata <- read_csv("./data/Modeldata_scaled.csv") %>%
+modeldata <- read_csv("secondwave_analysis/data/Modeldata_scaled.csv") %>%
   dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
                 -`Mobility (mean)`,
                 # -`Relaxation of measures`,
@@ -82,39 +82,3 @@ res <- my_causal(dag, modeldata, exposure="Nursing homes", unobserved=unobserved
 # "cluster" Age? --> schools
 res <- my_causal(dag, modeldata, exposure="Age", unobserved=unobserved)
 
-
-# tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
-# ggdag(tidy_dag,text_size = 1.5,color="grey",
-#       edge_type = "link",node_size = 13,label_size = 1.0) + 
-#   ggdag::theme_dag_blank(base_size = 4)
-# 
-# tidy_dag %>% 
-#   ggplot(aes(x = x, y = y, xend = xend, yend = yend))+
-#   geom_dag_edges_diagonal(check_overlap = TRUE) +
-#   geom_dag_point(size=10,color="lightgrey") +
-#   geom_dag_text(col = "black",size=1.2) +
-#   theme_solid()
-
-# library(igraph)
-# data <- dagitty_to_adjmatrix(dag)
-# network <- graph_from_adjacency_matrix(data , mode='directed', diag=FALSE)
-# 
-# plot(network, layout=layout.sphere, main="sphere")
[…]

### 2021-03-18 R/0_z_testlags.R
secondwave analysis etc

@@ -1,5 +1,8 @@
 mylags <- 1:20
 
+myaic <- rep(0, 20)
+myglms <- vector("list", 20)
+
 library(lubridate)
 library(tidyverse)
 library(DBI)
@@ -97,46 +100,46 @@ measures_zeroone_all <- measures_zeroone_all %>%
          `Mandatory face masks`=ifelse(id==8325049 & date>=as.Date("2020-04-17"), 1, `Mandatory face masks`), # rottweil
          `Mandatory face masks`=ifelse(id==16062041  & date>=as.Date("2020-04-14"), 1, `Mandatory face masks`), # nordhausen
          `Mandatory face masks`=ifelse(id==16053000  & date>=as.Date("2020-04-06"), 1, `Mandatory face masks`) # jena
-  )
+  ) %>% 
+  mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
+  dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
 
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
 
+brd_timeseries <- brd_timeseries %>%
+  mutate(id=replace(id, id==11, 11000000)) %>%
+  filter(id<=11000000 | id>=12000000) %>%
+  group_by(id) %>%
+  mutate(`Reported new cases COVID-19`=cases-lag(cases),
+         `Reported new cases COVID-19`=ifelse(`Reported new cases COVID-19`<0, 0, `Reported new cases COVID-19`),
+         `Active cases`=cases-lag(cases, 14),
+         `Active cases`=ifelse(`Active cases`<0, 0, `Active cases`)) %>%
+  ungroup() %>%
+  filter(date >= min(dateset)) %>%
+  mutate(daycount=as.integer(date-min(dateset)),
+         `Weekday (report)`=as.character(paste0(wday(date, week_start=5), wday(date, week_start=1, label=TRUE))),
+         `Holiday (report)`=ifelse(date %in% germanholidays, 1, 0),
+         bl_id=floor(id/1000/1000)) %>%
+  filter(id>16)
+
+google_mobility <- google_mobility %>% filter(date %in% dateset)
+weather <- weather %>% filter(date %in% dateset)
[…]

### 2021-03-18 R/1_generate_data.R

@@ -4,7 +4,7 @@ library(DBI)
 library(broom)
 library(MASS)
 
-maxanalysisdate <- as_date("2020-07-08")
+mylag <- 12
 
 mydatapath <- "./data/" 
 
@@ -104,9 +104,8 @@ measures_zeroone_all <- measures_zeroone_all %>%
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
 
-mylag <- 8
-startdate <- min(dateset)+mylag
-enddate <- max(brd_timeseries$date)
+startdate <- min(dateset) + 14 # +mylag # +max(mylags)
+enddate <- as.Date("2020-07-08") # startdate + days(100) # 
 
 lagweather <- mylag
 lagmobility <- mylag
@@ -157,7 +156,7 @@ modeldata_raw <- left_join(brd_timeseries, google_mobility %>%
          `Gender`=`Sex`,
          `Temperature`=`Weather (temperature)`,
          `Wind`=`Weather (wind)`) %>%
-  filter(date<=maxanalysisdate) %>%
+  filter(date<=enddate) %>%
   mutate(dummy = 1) %>% # column with single value
   pivot_wider(
     names_from = "Weekday (exposure)", # column to spread

### 2021-03-18 R/3_modelassumptions.R

@@ -11,9 +11,7 @@ library(pscl)
 source("R/z_auxiliary_functions_for_2.R")
 
 # Load Data
-modeldata <- bind_cols(read_csv("data/Modeldata_scaled_pcamobility.csv"),
-                       read_csv("data/Modeldata_raw.csv") %>%
-                         dplyr::select(`Reported new cases COVID-19`, `Active cases`))
+modeldata <- bind_cols(read_csv("data/Modeldata_scaled_pcamobility.csv"))
 
 id_daycount <- read_csv("data/Modeldata_raw.csv") %>%
   dplyr::select(id, daycount, date, bl_id) %>%
@@ -21,7 +19,7 @@ id_daycount <- read_csv("data/Modeldata_raw.csv") %>%
   
 
 modeldata_cont <- modeldata %>%
-  dplyr::select(-c("1Mo", "2Di", "3Mi", "5Fr", "6Sa", "7So",
+  dplyr::select(-c("Weekday 1Mo", "Weekday 2Di", "Weekday 3Mi", "Weekday 5Fr", "Weekday 6Sa", "Weekday 7So",
                    "Holiday (exposure)", "Holiday (report)",
                    "Reported new cases COVID-19"))
 
@@ -40,7 +38,7 @@ devresids <- residuals(myglm, type="deviance")
 library(statmod)
 qresids <- qresid(myglm)
 
-# linearity # [-1989]
+# linearity
 scatter.smooth(predict(myglm, type="link"), devresids, col='gray')
 
 myresidagainstcont <- bind_cols(modeldata_cont, devresids=devresids) %>%
@@ -51,13 +49,13 @@ ggplot(myresidagainstcont, aes(x=value, y=devresids)) +
   geom_smooth(se=FALSE)
 
 # proper distribution
-qqnorm(qresids)
-qqline(qresids)
-hist(qresids, freq=FALSE)
-lines(density(qresids),
+qqnorm(qresids[-1971])
+qqline(qresids[-1971])
+hist(qresids[-1971], freq=FALSE)
+lines(density(qresids[-1971]),
       lwd = 2,
       col = "red")
-lines(density(rnorm(length(qresids))),
+lines(density(rnorm(length(qresids[-1971]))),
       lwd = 2,
       col = "blue")
 
@@ -97,4 +95,5 @@ orig_mobi <- as.vector(exp(as.matrix(pcaloadings[, 2:7]) %*% myglm$coefficients[
[…]

### 2021-03-18 secondwave_analysis/R/0_z_testlags.R

@@ -1,12 +1,15 @@
 mylags <- 1:20
 
+myaic <- rep(0, 20)
+myglms <- vector("list", 20)
+
 library(lubridate)
 library(tidyverse)
 library(DBI)
 library(broom)
 library(MASS)
 
-mydatapath <- "./data/" 
+mydatapath <- "secondwave_analysis/data/" 
 
 google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
   mutate(id=as.integer(RS_first2)) %>%
@@ -63,14 +66,32 @@ germanholidays <- as.Date(c("2020-04-10", #karfreitag
                             "2020-04-13", #ostermontag
                             "2020-05-01", # tag der arbeit
                             "2020-05-21", # himmelfahrt
-                            "2020-06-01")) # pfingstmontag
-nobigevents <- as.Date("2020-03-10")
+                            "2020-06-01", # pfingstmontag
+                            "2020-10-03", # day of unity
+                            "2020-12-24", "2020-12-25", "2020-12-26", # christmas
+                            "2020-12-31", # new years eve
+                            "2021-01-01" # new year
+)) 
+nobigevents <- as.Date("2020-03-08")
 noschool <- as.Date("2020-03-16")
 contactrestrict <- as.Date("2020-03-22")
-measures_zeroone <- tibble(date=as.Date("2020-01-01")+0:360) %>%
+somerelaxations <- as.Date("2020-05-06")
+lockdownlight <- as.Date("2020-11-02")
+lockdownharder <- as.Date("2020-12-13")
+measures_zeroone <- tibble(date=as_date(as_date("2020-01-01"):as_date("2021-01-29"))) %>%
   mutate(`Ban of mass gatherings`=ifelse(date<nobigevents, 0, 1),
-         `School/Kita closures`=ifelse(date<noschool, 0, 1),
-         `Contact restrictions`=ifelse(date<contactrestrict, 0, 1),
+         `School/Kita closures`=case_when(
+           date<noschool ~ 0,
+           date>=noschool & date<somerelaxations ~ 1,
+           date>=somerelaxations & date<lockdownharder ~ 0.5,
+           date>=lockdownharder ~ 1
+         ),
[…]

### 2021-03-18 secondwave_analysis/R/1_generate_data.R

@@ -5,6 +5,8 @@ library(broom)
 library(MASS)
 source("secondwave_analysis/R/z_auxiliary_functions_for_1.R")
 
+mylag <- 7
+
 mydatapath <- "secondwave_analysis/data/" 
 
 google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
@@ -159,9 +161,9 @@ measures_zeroone_all <- measures_zeroone_all %>%
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
 
-mylag <- 8
-startdate <- as.Date("2020-07-09")
-enddate <- max(brd_timeseries$date)
+enddate <- as.Date("2021-01-25") # as.Date("2020-12-16") # max(brd_timeseries$date) # 
+
+startdate <- enddate - days(100)# as.Date("2020-07-16") # +max(mylags)
 
 lagweather <- mylag
 lagmobility <- mylag

### 2021-03-18 secondwave_analysis/R/2_dags_for_covidcases.R

@@ -11,45 +11,30 @@ library(pscl)
 source("secondwave_analysis/R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
-tablepath <- "manuscript/v4_revision/tables/secondwave_analysis/"
+tablepath <- "manuscript/v4_medrxiv_revision/tables/secondwave/"
 
 # load dag
-dagfile <- "data/bigdag4covid19.txt"
+dagfile <- "secondwave_analysis/data/bigdag4covid19.txt"
 dagtext <- readChar(dagfile, file.info(dagfile)$size)
 dag <- dagitty::dagitty(dagtext)
 
 unobserved <- c("Access to tests", 
                 "Tracing capacity",
                 "Health care utilization",
+                "Exposure to SARS-COV-2",
                 "Herd immunity",
                 "Onset of COVID-19",
                 "Trust in society",
                 "Susceptibility")
 
 # Load Data
-modeldata <- read_csv("secondwave_analysis/data/Modeldata_scaled.csv") %>%
-  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
-                -`Mobility (mean)`,
-                # -`Relaxation of measures`,
-                -contains("iso"), -contains("census")
-                ) %>% 
-  rename(`School and kindergarten closures`=`School/Kita closures`,
-         `Holiday (exposure)`=Holiday,
-         `Foreign citizens`=`Foreign residents`,
-         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
-         `Rainfall`=`Weather (rainfall)`,
-         `Humidity`=`Weather (humidity)`,
-         `Indoor humidity`=`Weather (indoor humidity)`,
-         `Gender`=`Sex`,
-         `Temperature`=`Weather (temperature)`,
-         `Wind`=`Weather (wind)`) %>%
-  # filter(date<="2020-12-31") %>%
-  dplyr::select(-date)
+modeldata <- read_csv("secondwave_analysis/data/Modeldata_scaled_pcamobility.csv")
 
 # all variables
 res <- my_causal(dag, modeldata, exposure=NULL, unobserved)
 
 file.create(paste0(tablepath, "t_pseudor2s.csv"))
+file.create(paste0(tablepath, "t_aics.csv"))
 # mobility
 res <- my_causal(dag, modeldata, exposure="Mobility", unobserved=unobserved)
 write_cause(res, exposure="Mobility")
[…]

### 2021-03-18 secondwave_analysis/R/3_modelassumptions.R

@@ -1,0 +1,99 @@
+library(lubridate)
+library(tidyverse)
+library(DBI)
+library(broom)
+library(MASS)
+library(dagitty)
+library(ggdag)
+library(ggthemes)
+library(pcalg)
+library(pscl)
+source("secondwave_analysis/R/z_auxiliary_functions_for_2.R")
+
+# Load Data
+modeldata <- bind_cols(read_csv("secondwave_analysis/data/Modeldata_scaled_pcamobility.csv"))
+
+id_daycount <- read_csv("secondwave_analysis/data/Modeldata_raw.csv") %>%
+  dplyr::select(id, daycount, date, bl_id) %>%
+  dplyr::select(-date)
+  
+
+modeldata_cont <- modeldata %>%
+  dplyr::select(-c("Weekday 1Mo", "Weekday 2Di", "Weekday 3Mi", "Weekday 5Fr", "Weekday 6Sa", "Weekday 7So",
+                   "Holiday (exposure)", "Holiday (report)",
+                   "Reported new cases COVID-19"))
+
+exposure=NULL
+adjsets=NULL
+myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
+myglm <- glm.nb(as.formula(myformula),
+                  data=modeldata) # %>%dplyr::select(-`Mobility (PC4)`)
+mynullformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))" 
+mynullglm <- glm.nb(as.formula(mynullformula),
+                    data=modeldata)
+pseudor2 <- 1-sum((myglm$fitted.values-myglm$y)^2)/sum((mynullglm$fitted.values-mynullglm$y)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
+myaic <- AIC(myglm)
+mynullaic <- AIC(mynullglm)
+devresids <- residuals(myglm, type="deviance")
+library(statmod)
+qresids <- qresid(myglm)
+
+# linearity # [-1989]
+scatter.smooth(predict(myglm, type="link"), devresids, col='gray')
+
+myresidagainstcont <- bind_cols(modeldata_cont, devresids=devresids) %>%
+  pivot_longer(cols=`Rainfall`:`Active cases`)
+ggplot(myresidagainstcont, aes(x=value, y=devresids)) +
+  geom_point() +
[…]

### 2021-03-18 secondwave_analysis/R/z_auxiliary_functions_for_2.R

@@ -73,10 +73,12 @@ get_adjsets <- function(dag, exposure, outcome, unobserved) {
   if (length(adjsets)>0) {
     names(adjsets) <- paste0("MinAdjSet", seq(valid_adjsets))
     optadjset <- get_optadjset(amat, exposure, outcome, unobserved)
-    if (optadjset[[2]]) { # is a reduced set
-      adjsets[["RedOptAdjSet"]] <- optadjset[[1]]
-    } else { # is actually real optimal adjustment set
-      adjsets[["OptAdjSet"]] <- optadjset[[1]]
+    if (!is.null(optadjset)) {
+      if (optadjset[[2]]) { # is a reduced set
+        adjsets[["RedOptAdjSet"]] <- optadjset[[1]]
+      } else { # is actually real optimal adjustment set
+        adjsets[["OptAdjSet"]] <- optadjset[[1]]
+      }
     }
   }
   return(adjsets)
@@ -116,6 +118,7 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
   coefficients <- vector("list", nf)
   causaleffects <- vector("list", nf)
   pseudor2s <- rep(0, nf)
+  aics <- rep(0, nf)
   mynullglm <- glm.nb(as.formula(mynullformula),
                       data=modeldata)
   for (idxf in seq(nf)) {
@@ -127,7 +130,9 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
     #                                     `Weekday (report)`+`Holiday (report)`")), data = modeldata)
     # pseudo RSquared
     pseudor2s[idxf] <- 1-sum((myglm$fitted.values-myglm$y)^2)/sum((mynullglm$fitted.values-mynullglm$y)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
+    aics[idxf] <- AIC(myglm)
     cat("   pseudo R2:", pseudor2s[idxf], "\n")
+    cat("   AIC:", aics[idxf], "\n")
     coefficients[[idxf]] <- tidy(myglm, exponentiate=TRUE, conf.int=TRUE, conf.level = 0.99)
     if (!is.null(exposure)) {
       effects <- tail(coefficients[[idxf]]$estimate, nexp)
@@ -143,7 +148,7 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
       causaleffects[[idxf]] <- coefficients[[idxf]]
     }
   }
[…]

### 2021-03-18 secondwave_analysis/R/0_z_testlags.R
original scale for secondwave analysis

@@ -241,9 +241,11 @@ for (mylag in mylags) {
                                                     `Right-wing populist party votes`, `Population density`,
                                                     Gender, `Nursing homes`,
                                                     `COVID-19 burden`)
-  scale_params <- tibble(variable=colnames(modeldata_X_cont),
-                         mymean=colMeans(modeldata_X_cont),
-                         mysd=apply(modeldata_X_cont, 2, sd))
+  # scale_params <- tibble(variable=colnames(modeldata_X_cont),
+  #                        mymean=colMeans(modeldata_X_cont),
+  #                        mysd=apply(modeldata_X_cont, 2, sd))
+  # write_csv(scale_params, paste0(mydatapath, "scale_params.csv"))
+  scale_params <- read_csv("data/scale_params.csv") # original scaling first wave!
   modeldata_X_cont_scaled <- sapply(seq(dim(modeldata_X_cont)[2]),
                                     function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], 2*scale_params[i, 3])) # gelman
   colnames(modeldata_X_cont_scaled) <- colnames(modeldata_X_cont)
@@ -253,16 +255,27 @@ for (mylag in mylags) {
   modeldata_scaled <- bind_cols(modeldata_X_bin,
                                 modeldata_X_cont_scaled)
   
-  pca_mobility <- prcomp(modeldata_scaled %>% dplyr::select(contains("Mobility")))
-  cumsum(pca_mobility$sdev)/sum(pca_mobility$sdev)
-  modeldata_scaled_pca_mobility <- modeldata_scaled %>% 
+  # pca_mobility <- prcomp(modeldata_scaled %>% dplyr::select(contains("Mobility")))
+  # cumsum(pca_mobility$sdev)/sum(pca_mobility$sdev)
+  # modeldata_scaled_pca_mobility <- modeldata_scaled %>% 
+  #   dplyr::select(-contains("Mobility")) %>%
+  #   bind_cols("Mobility (PC1)"=pca_mobility$x[,1],
+  #             "Mobility (PC2)"=pca_mobility$x[,2],
+  #             "Mobility (PC3)"=pca_mobility$x[,3],
+  #             "Mobility (PC4)"=pca_mobility$x[,4],
+  #             "Mobility (PC5)"=pca_mobility$x[,5],
[…]

### 2021-03-18 secondwave_analysis/R/1_generate_data.R

@@ -240,10 +240,11 @@ modeldata_X_cont <- modeldata_X %>% dplyr::select(`Mobility (retail and recreati
                                                   `Right-wing populist party votes`, `Population density`,
                                                   Gender, `Nursing homes`,
                                                   `COVID-19 burden`)
-scale_params <- tibble(variable=colnames(modeldata_X_cont),
-                       mymean=colMeans(modeldata_X_cont),
-                       mysd=apply(modeldata_X_cont, 2, sd))
-write_csv(scale_params, paste0(mydatapath, "scale_params.csv"))
+# scale_params <- tibble(variable=colnames(modeldata_X_cont),
+#                        mymean=colMeans(modeldata_X_cont),
+#                        mysd=apply(modeldata_X_cont, 2, sd))
+# write_csv(scale_params, paste0(mydatapath, "scale_params.csv"))
+scale_params <- read_csv("data/scale_params.csv") # original scaling first wave!
 modeldata_X_cont_scaled <- sapply(seq(dim(modeldata_X_cont)[2]),
                                   function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], 2*scale_params[i, 3])) # gelman
 colnames(modeldata_X_cont_scaled) <- colnames(modeldata_X_cont)
@@ -253,17 +254,27 @@ modeldata_X_bin <- modeldata_X %>% dplyr::select(-colnames(modeldata_X_cont))
 modeldata_scaled <- bind_cols(modeldata_X_bin,
                               modeldata_X_cont_scaled)
 
-pca_mobility <- prcomp(modeldata_scaled %>% dplyr::select(contains("Mobility")))
-cumsum(pca_mobility$sdev)/sum(pca_mobility$sdev)
-modeldata_scaled_pca_mobility <- modeldata_scaled %>% 
+# pca_mobility <- prcomp(modeldata_scaled %>% dplyr::select(contains("Mobility")))
+# cumsum(pca_mobility$sdev)/sum(pca_mobility$sdev)
+# modeldata_scaled_pca_mobility <- modeldata_scaled %>% 
+#   dplyr::select(-contains("Mobility")) %>%
+#   bind_cols("Mobility (PC1)"=pca_mobility$x[,1],
+#             "Mobility (PC2)"=pca_mobility$x[,2],
[…]

### 2021-03-18 R/1_generate_data.R
update all data with original scale params

@@ -104,7 +104,7 @@ measures_zeroone_all <- measures_zeroone_all %>%
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
 
-startdate <- min(dateset) + 14 # +mylag # +max(mylags)
+startdate <- min(dateset) + mylag # +mylag # +max(mylags)
 enddate <- as.Date("2020-07-08") # startdate + days(100) # 
 
 lagweather <- mylag

### 2021-03-18 R/2_dags_for_covidcases.R

@@ -51,9 +51,6 @@ for (myweather in c("Temperature", "Rainfall", "Humidity", "Wind")) {
   write_cause(res, exposure=myweather)
 }
 
-# interventions
-res <- my_causal(dag, modeldata, exposure="Interventions", unobserved=unobserved)
-
 # other variables
 othervariables <- setdiff(names(dag), c("Reported new cases COVID-19", unobserved,
                                         "Mobility", "COVID-19 burden", "Searches corona",

### 2021-03-18 secondwave_analysis/R/1_generate_data.R

@@ -92,6 +92,7 @@ measures_zeroone <- tibble(date=as_date(as_date("2020-01-01"):as_date("2021-01-2
            date>=lockdownlight ~ 1
          ),
          Holiday=ifelse(date %in% germanholidays, 1, 0),
+         Holiday=ifelse(date>="2020-12-24" & date<="2021-01-01", 1, Holiday),
          `Mandatory face masks`=0,
          id=kreisids[1])
 measures_zeroone_all <- measures_zeroone
@@ -163,7 +164,7 @@ dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, 
 
 enddate <- as.Date("2021-01-25") # as.Date("2020-12-16") # max(brd_timeseries$date) # 
 
-startdate <- enddate - days(100)# as.Date("2020-07-16") # +max(mylags)
+startdate <- enddate - days(132)# as.Date("2020-07-16") # +max(mylags)
 
 lagweather <- mylag
 lagmobility <- mylag

### 2021-03-18 secondwave_analysis/R/z_auxiliary_functions_for_2.R

@@ -121,6 +121,8 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
   aics <- rep(0, nf)
   mynullglm <- glm.nb(as.formula(mynullformula),
                       data=modeldata)
+  scale_params <- read_csv("data/scale_params.csv") # original scaling first wave!
+  to_rescale <- scale_params$variable
   for (idxf in seq(nf)) {
     cat("   Exposure:", exposure, "\n")
     cat("   Adjustment set:", adjsets[[idxf]], "\n")
@@ -134,6 +136,15 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
     cat("   pseudo R2:", pseudor2s[idxf], "\n")
     cat("   AIC:", aics[idxf], "\n")
     coefficients[[idxf]] <- tidy(myglm, exponentiate=TRUE, conf.int=TRUE, conf.level = 0.99)
+    for (expovar in coefficients[[idxf]]$term) {
+      expovar_clean <- str_remove_all(expovar, "`")
+      if (expovar_clean %in% to_rescale) {
+        # coefficients[[idxf]]$estimate[coefficients[[idxf]]$term=="(Intercept)"] <- coefficients[[idxf]]$estimate[coefficients[[idxf]]$term=="(Intercept)"]*exp(-coefficients[[idxf]]$estimate[coefficients[[idxf]]$term==expovar]*scale_params$mymean[scale_params$variable==expovar_clean]/scale_params$mysd[scale_params$variable==expovar_clean])
+        coefficients[[idxf]]$estimate[coefficients[[idxf]]$term==expovar] <- coefficients[[idxf]]$estimate[coefficients[[idxf]]$term==expovar] ^(1/scale_params$mysd[scale_params$variable==expovar_clean])
+        coefficients[[idxf]]$conf.low[coefficients[[idxf]]$term==expovar] <- coefficients[[idxf]]$conf.low[coefficients[[idxf]]$term==expovar] ^(1/scale_params$mysd[scale_params$variable==expovar_clean])
+        coefficients[[idxf]]$conf.high[coefficients[[idxf]]$term==expovar] <- coefficients[[idxf]]$conf.high[coefficients[[idxf]]$term==expovar] ^(1/scale_params$mysd[scale_params$variable==expovar_clean])
+      }
+    }
     if (!is.null(exposure)) {
[…]

### 2021-03-18 R/3_modelassumptions.R
cv for regularization (ridge)

@@ -8,6 +8,7 @@ library(ggdag)
 library(ggthemes)
 library(pcalg)
 library(pscl)
+library(glmnet)
 source("R/z_auxiliary_functions_for_2.R")
 
 # Load Data
@@ -16,30 +17,53 @@ modeldata <- bind_cols(read_csv("data/Modeldata_scaled_pcamobility.csv"))
 id_daycount <- read_csv("data/Modeldata_raw.csv") %>%
   dplyr::select(id, daycount, date, bl_id) %>%
   dplyr::select(-date)
-  
 
-modeldata_cont <- modeldata %>%
-  dplyr::select(-c("Weekday 1Mo", "Weekday 2Di", "Weekday 3Mi", "Weekday 5Fr", "Weekday 6Sa", "Weekday 7So",
-                   "Holiday (exposure)", "Holiday (report)",
-                   "Reported new cases COVID-19"))
-
-exposure=NULL
-adjsets=NULL
-myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
+# run full model (and null model)
+myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))+."
 myglm <- glm.nb(as.formula(myformula),
-                  data=modeldata) # %>%dplyr::select(-`Mobility (PC4)`)
+                data=modeldata) # %>%dplyr::select(-`Mobility (PC4)`)
 mynullformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))" 
 mynullglm <- glm.nb(as.formula(mynullformula),
                     data=modeldata)
 pseudor2 <- 1-sum((myglm$fitted.values-myglm$y)^2)/sum((mynullglm$fitted.values-mynullglm$y)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
 myaic <- AIC(myglm)
 mynullaic <- AIC(mynullglm)
+
+# regularized, cv
+mytheta <- myglm$theta
+myglm_fixtheta <- glm(as.formula(myformula),
+              data=modeldata, 
+              family=negative.binomial(mytheta))
+myy <- modeldata$`Reported new cases COVID-19`
+myx <- as.matrix(modeldata %>% dplyr::select(-`Reported new cases COVID-19`))
+
+myfoldids <- (id_daycount %>% group_indices(id)) %% 10 + 1
+
+myglm_ridge_cv <- cv.glmnet(x=myx, y=myy, family = negative.binomial(mytheta),
[…]

### 2021-03-19 R/3_modelassumptions.R
cv and lambda for all effects

@@ -39,12 +39,15 @@ myx <- as.matrix(modeldata %>% dplyr::select(-`Reported new cases COVID-19`))
 
 myfoldids <- (id_daycount %>% group_indices(id)) %% 10 + 1
 
+set.seed(1502)
 myglm_ridge_cv <- cv.glmnet(x=myx, y=myy, family = negative.binomial(mytheta),
                             offset=log(modeldata%>%pull(`Active cases`)+1),
                             standardize=FALSE,
                             alpha=0, nfolds=10, foldid = myfoldids) # 
 
 mylambda <- myglm_ridge_cv$lambda.1se
+
+write_csv(tibble(mytheta=mytheta, mylambda=mylambda), "data/myglmnetparams.csv")
 
 myglm_ridge <- glmnet(myx, myy, family = negative.binomial(mytheta),
                       offset=log(modeldata%>%pull(`Active cases`)+1),
@@ -54,8 +57,8 @@ myglm_ridge <- glmnet(myx, myy, family = negative.binomial(mytheta),
 # residuals
 
 devresids <- residuals(myglm, type="deviance")
-library(statmod)
-qresids <- qresid(myglm)
+# library(statmod)
+# qresids <- qresid(myglm_ridge)
 
 # linearity
 scatter.smooth(predict(myglm, type="link"), devresids, col='gray')

### 2021-03-19 R/z_auxiliary_functions_for_2.R

@@ -38,6 +38,15 @@ my_vars <- function(modeldata, mynames) {
   return(myvars)
 }
 
+# get AIC from glmnet
+aic_glmnet <- function(myglmnet) {
+  tLL <- -deviance(myglmnet) # myglmnet$nulldev*myglmnet$dev.ratio # myglmnet$nulldev - deviance(myglmnet)
+  k <- myglmnet$df
+  n <- myglmnet$nobs
+  AICc <- -tLL+2*k+2*k*(k+1)/(n-k-1)
+  return(AICc)
+}
+
 # function to get optimal adjustment set without unobserved (check for validness)
 get_optadjset <- function(amat, exposure, outcome, unobserved) {
   x.pos <- which(colnames(amat)==exposure)
@@ -162,6 +171,98 @@ my_negbin <- function(modeldata, exposure=NULL, adjsets=NULL) {
   return(list(pseudor2s=pseudor2s, aics=aics, causaleffects=causaleffects))
 }
 
+# function for neg-binomial model based on modeldata and adjustment set from dag, but with ridge regularization
+my_negbin_ridge <- function(modeldata, exposure=NULL, adjsets=NULL) {
+  myglmnetparams <- read_csv("data/myglmnetparams.csv")
+  mytheta <- myglmnetparams %>% pull(mytheta)
+  mylambda <- myglmnetparams %>% pull(mylambda)
+  myy <- modeldata$`Reported new cases COVID-19`
+  myx_data <- modeldata %>% dplyr::select(-`Reported new cases COVID-19`)
+  mynullx <- cbind(0, as.matrix(myx_data %>% dplyr::select(`Active cases`))) # "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))" 
+  if (is.null(exposure) & is.null(adjsets)) { # adjustment set not given
+    myxcols <- vector("list", 1)
+    myxcols[[1]] <- colnames(myx_data) # "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
+    nf <- 1
+  } else if (is.null(exposure)+is.null(adjsets)==1) { # exactly one is different from NULL: not allowed
+    myxcols <- vector("list", 1)
+    myxcols[[1]] <- colnames(myx_data) # "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
[…]

### 2021-03-19 secondwave_analysis/R/2_dags_for_covidcases.R

@@ -8,13 +8,13 @@ library(ggdag)
 library(ggthemes)
 library(pcalg)
 library(pscl)
-source("secondwave_analysis/R/z_auxiliary_functions_for_2.R")
+source("R/z_auxiliary_functions_for_2.R")
 
 # save tables to:
 tablepath <- "manuscript/v4_medrxiv_revision/tables/secondwave/"
 
 # load dag
-dagfile <- "secondwave_analysis/data/bigdag4covid19.txt"
+dagfile <- "data/bigdag4covid19.txt"
 dagtext <- readChar(dagfile, file.info(dagfile)$size)
 dag <- dagitty::dagitty(dagtext)
 

### 2021-03-19 secondwave_analysis/R/z_auxiliary_functions_for_2.R

@@ -1,202 +1,0 @@
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-library(dagitty)
-library(ggdag)
-library(ggthemes)
-library(pcalg)
-library(pscl)
-
-# convert dagitty to adjacency matrix
-dagitty_to_adjmatrix <- function(daggity_obj) {
-  edg <- dagitty:::edges(daggity_obj)
-  node_names <- dagitty:::names.dagitty(daggity_obj)
-  ans_mat <- matrix(
-    data = 0, nrow = length(node_names),
-    ncol = length(node_names),
-    dimnames = list(node_names, node_names)
-  )
-  ans_mat[as.matrix(edg[c("v", "w")])] <- 1
-  ans_mat[as.matrix(edg[edg$e=="--", ][c("w", "v")])] <- 1
-  return(ans_mat)
-}
-
-# function to assemble variable names
-my_vars <- function(modeldata, mynames) {
-  myvars <- character()
-  allvars <- colnames(modeldata)
-  idx <- 1
-  for (mn in mynames) {
-    thesevars <- allvars[grep(mn, allvars, fixed=TRUE)]
-    for (tvidx in seq(thesevars)) {
-      myvars[idx] <- thesevars[tvidx]
-      idx <- idx+1
-    }
-  }
-  return(myvars)
-}
-
-# function to get optimal adjustment set without unobserved (check for validness)
-get_optadjset <- function(amat, exposure, outcome, unobserved) {
-  x.pos <- which(colnames(amat)==exposure)
-  y.pos <- which(colnames(amat)==outcome)
-  optadjset <- colnames(amat)[optAdjSet(amat, x.pos = x.pos, y.pos = y.pos)]
-  optadjset_red <- setdiff(optadjset, unobserved)
-  optadjset_red_idx <- sapply(optadjset_red, function(z) which(colnames(amat)==z))
-  is_valid <- gac(amat, x.pos, y.pos, optadjset_red_idx, type="dag")
-  if (is_valid$gac) {
-    return(list(adjset=optadjset_red, is_reduced=(length(optadjset_red)<length(optadjset))))
-  } else {
-    return(NULL)
-  }
-}
-
-# function to get minimal and optimal adjustment sets without unobserved (check for validness)
-get_adjsets <- function(dag, exposure, outcome, unobserved) {
-  amat <- t(dagitty_to_adjmatrix(dag))
-  x.pos <- which(colnames(amat)==exposure)
[…]

### 2021-03-19 manuscript/v4_medrxiv_revision/causal_model.Rmd
careful wording "causal"

@@
 Restrictions were applied to English and German language and the publication date in the last one year.
-The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with the exposure variables ("mobility", “public awareness”, "awareness", “google trends”,“ambient temperature”, "temperature").
-For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$).
+The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with a selection of exposure variables ("mobility", “public awareness”, "awareness", “google trends”, “ambient temperature”, "temperature").
+For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$).
 We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \@ref(fig:dag-covid-19).
@@
 
-Google Trends [@google_trends] data can be used as a tool to get insights into public interest (**awareness**) in the coronavirus disease.
+Google Trends [@google_trends] data can be used as a tool to get insights into public interest (**awareness**) in COVID-19.
 Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases [@ayyoubzadeh_predicting_2020;@effenberger_association_2020;@higgins_correlations_2020;@li_retrospective_2020;@lin_google_2020;@mavragani_tracking_2020;@walker_use_2020;@yuan_trends_2020;@zhou_effects_2020].
@@
 Please note that "Exposure to SARS-CoV-2" is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset.
-*Exposure to the virus* should not be confused with the formal *exposure variables* of the DAG (mobility, awareness, temperature).
+*Exposure to the virus* should not be confused with the formal *exposure variables* of the DAG.

### 2021-03-19 manuscript/v4_medrxiv_revision/manuscript.Rmd
more careful abstract, changed title, "causal"

@@
 ---
-title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
+title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases
 author:
@@
     address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
-abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection.
-Correcting for possible confounders, we estimated their causal effects on reported case numbers.
-To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19.
+abstract: Several determinants are suspected to be causal drivers for new cases of COVID-19 infection.
+Correcting for possible confounders, we estimated the effects of the most prominent determining factors on reported case numbers.
+To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the determinants on new reported cases of COVID-19.
 Based on this, we computed valid adjustment sets of the possible confounding factors.
@@
 Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
 Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
-This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
+This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
 
@@
 
-Our objective is to provide estimates of the causal effects of the main drivers of the pandemic with reduced bias.
+Our objective is to provide valid estimates of the effects of the main drivers of the pandemic with a causal graph approach.
 We conducted a scoping review of the available studies regarding signaling pathways and determinants of the spread of SARS-CoV-2 infections and the reported new COVID-19 cases.

### 2021-03-19 manuscript/v4_medrxiv_revision/data_methods.Rmd
update methods, intervention variable, glmnet, ridge, cv, ...

--- 2020-07-24 manuscript/v3_submission/data_methods.Rmd
@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 23 February 2020 to 8 July 2020 ($T=137$), since we used a lag of 8 days for all confounders.
+Our observation period for the outcome consisted of all dates from 28 February 2020 to 8 July 2020 ($T=132$), since we used a lag of 12 days for all confounders.
 We did not exclude any states or districts ($K=401$).
@@
 Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables.
-For our causal analysis, we computed adjustment sets in three different scenarios for separate exposures within the DAG: i) mobility of population, ii) awareness of COVID-19 (i.e.
-Google searches for "corona"), iii) weather (i.e.
-temperature).
+For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
 
@@
 "Residential" mobility is the relative change of daily time spent at residential areas.
+The six mobility variables showed high correlations among each other and with other variables.
+To reduce multicollinearity, we transformed them by principal component analysis (PCA) onto six uncorrelated principal components which were used in place of the original variables.
 
@@
 
-For different official and political measures we used one-hot encoded daily variables, i.e.
+For different official and political measures on a day and district level we used one-hot encoded daily variables, i.e.
 ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
+These variables were highly correlated with each other and other variables (especially mobility).
+To reduce multicollinearity, we aggregated the different official/political measures by summing them into one variable "interventions" which describes the current severity of interventions on a given day in a given district.
 
@@
 All variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered for numerical stability.
-We did not scale variables to unit variance to maintain interpretability of effects on the original scale of variables.
-Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 8 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 8 days.
+We scaled continuous variables by dividing by 2 times their standard deviation, while we left binary variables as-is [@gelman2008scaling].
+After estimating their effects, we re-scaled variables to their original scale and for mobility, we re-transformed the effects of the principal components to the original mobility variables.
+Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 12 days.
 
@@
 We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases.
-Every node $v_i$ in the graph is the graphical representation of an observed or unobserved variable $x_i$, a directed edge $e_{ij}$ is an arrow from node $v_i$ to $v_j$ that implies a direct causal relationship from variable $x_i$ onto variable $x_j$.
+We use the terms "causal effect" or "causal relationship" for effect estimates that are based on this causal graph framework.
+Every node $v_i$ in the causal graph is the graphical representation of an observed or unobserved variable $x_i$, a directed edge $e_{ij}$ is an arrow from node $v_i$ to $v_j$ that implies a direct causal relationship from variable $x_i$ onto variable $x_j$.
 The set of all nodes is denoted by $V$, the set of all edges by $E$, as such, the complete DAG is the tuple $G=(V,E)$.
@@
 
-When we analyzed different adjustment sets given by analysis of the causal DAG (i.e.
-the minimal and optimal adjustment sets), we first checked if the set included unobserved variables.
+To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach [@ridge1970], which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlated variables on their respective regression coefficients.
+The hyper-parameter $\lambda$ of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the $401$ districts.
+
+We conducted a log-linear regression (function `glm` with `family=poisson()` for Poisson regression, and `glm.nb` from the `MASS` package for the negative binomial regression [@mass]) for the full data set to assess general model adequacy and to estimate the $\theta$ parameter of the negative binomial.
+We used this parameter with the `cv.glmnet` function from the R package `glmnet` [@glmnet2010] with `family=negative.binomial(theta)` and chose the $\lambda$ value within one standard deviation from the minimal $\lambda$ as regularization hyper-parameter.
+Afterwards, we calculated the effects of separate exposures on the outcome.
+For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.
+the minimal and optimal adjustment sets).
+Then, we first checked if the respective set included unobserved variables.
 If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function `gac` in package `pcalg` [@perkovi2015complete]).
 If a minimal adjustment set contained unobserved variables, we discarded the whole set.
-We conducted a log-linear regression (function `glm` with `family=poisson()` for Poisson regression, and `glm.nb` from the `MASS` package for the negative binomial regression [@mass]) for every remaining valid adjustment set as regressors and calculated a Pseudo-$R^2$ given by $1-V_m/V_0$, where $V_m$ is the sum of squared prediction errors of the current model and $V_0$ is the sum of squared prediction errors of the null model (intercept and offset only).
-That is, our Pseudo-$R^2$ is $1$ minus the fraction of variance unexplained.
-Finally, we decided for the model/adjustment set with the highest pseudo-$R^2$.
-We report the exponentiated estimated coefficients along with 99 percent confidence intervals of the estimates.
+If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph.
+We used the function `glmnet` with the parameters $\theta$ and $\lambda$ as above on every remaining valid adjustment set as regressors and calculated the Akaike information criterion (AIC) for this set of regressors.
+Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC.
+We report the exponentiated estimated coefficients.

### 2021-03-19 manuscript/v4_medrxiv_revision/results.Rmd
fix results

--- 2020-07-28 manuscript/v2_medrxiv/results.Rmd
@@
 Modeldata <- read_csv("../../data/Modeldata_raw.csv") %>%
-  rename(`Gender`=`Sex`,
-         `Foreign citizens`=`Foreign residents`,
-         `Foreign citizens (refugees)`=`Foreign residents (refugees)`) %>%
   filter(date<="2020-07-08")
 
-t1_data <- Modeldata %>% dplyr::select(-contains("Weekday"), -contains("Holiday"), -`Relaxation of measures`,
+t1_data <- Modeldata %>% dplyr::select(-contains("Weekday"), -contains("Holiday"),
                      -contains("Raumeinheit"),
@@
                      -id,-bl_id,
-                     -cases, -deaths,-recovered,-daycount) 
+                     -cases, -deaths,-recovered,-daycount,-date) 
 t1_data_tab <- CreateTableOne(data = t1_data) 

### 2021-03-19 manuscript/v4_medrxiv_revision/causal_model.Rmd
update manuscript

@@
 We used a directed acyclic graph (DAG) [@schipf_directed_2011;@textor_robust_2017] as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread.
-To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar.
+To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google Scholar.
 Restrictions were applied to English and German language and the publication date in the last one year.
-The following search terms were applied to abstracts and title in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with a selection of exposure variables ("mobility", “public awareness”, "awareness", “google trends”, “ambient temperature”, "temperature").
+The following search terms were applied to abstracts and titles in Pubmed ("COVID-19" OR "COVID19" OR "Corona" OR "Coronavirus" OR "SARS-CoV-2") and connected separately in each case with a selection of exposure variables ("mobility", “public awareness”, "awareness", “google trends”, “ambient temperature”, "temperature").
 For "mobility", we analyzed $n=8$ studies, $N=103$ were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar ("awareness"/"public awareness"/"google trends" $n=9$, $N=215$; "temperature"/"ambient temperature" $n=16$, $N=235$).

### 2021-03-19 manuscript/v4_medrxiv_revision/data_methods.Rmd
update manuscript

@@
 The six mobility variables showed high correlations among each other and with other variables.
-To reduce multicollinearity, we transformed them by principal component analysis (PCA) onto six uncorrelated principal components which were used in place of the original variables.
+To reduce multicollinearity, we transformed them by principal component analysis (PCA) into six uncorrelated principal components which were used in place of the original variables.
 
@@
 If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph.
-We used the function `glmnet` with the parameters $\theta$ and $\lambda$ as above on every remaining valid adjustment set as regressors and calculated the Akaike information criterion (AIC) for this set of regressors.
+We used the function `glmnet` with the parameters $\theta$ and $\lambda$ as above on every remaining valid adjustment set as regressors (that is, we applied ridge regression) and calculated the Akaike information criterion (AIC) for this model/set of regressors.
 Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC.
-We report the exponentiated estimated coefficients.
+We report the exponentiated estimated coefficients for the separate exposures on their original scale.

### 2021-03-19 manuscript/v4_medrxiv_revision/manuscript.Rmd
update manuscript

@@
 # careful, it's always one version behind
-library(latexdiffr)
-latexdiff("../v2_medrxiv/manuscript.tex", "manuscript.tex", "latexdiffrRmd")
+# library(latexdiffr)
+# latexdiff("../v2_medrxiv/manuscript.tex", "manuscript.tex", "latexdiffrRmd")
 # system("latexdiff ../v2_medrxiv/manuscript.tex manuscript.tex > diffRmd.tex")

### 2021-03-19 manuscript/v4_medrxiv_revision/manuscript.tex
update manuscript

--- 2021-03-01 manuscript/v2_medrxiv/manuscript.tex
@@
             pdfauthor={},
-            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
+            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases},
             colorlinks=false,
@@
 
-  \title{Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
+  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases}
     \author[Zi]{Edgar Steiger}
@@
   \begin{abstract}
-  Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection.
-Correcting for possible confounders, we estimated their causal effects on reported case numbers.
-To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19.
+  Several determinants are suspected to be causal drivers for new cases of COVID-19 infection.
+Correcting for possible confounders, we estimated the effects of the most prominent determining factors on reported case numbers.
+To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the determinants on new reported cases of COVID-19.
 Based on this, we computed valid adjustment sets of the possible confounding factors.
 We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
 Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
-This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
+This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
   \end{abstract}
@@
 Restrictions were applied to English and German language and the publication date in the last one year.
-The following search terms were applied to abstracts and title in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with the exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'',``ambient temperature'', ``temperature'').
-For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
+The following search terms were applied to abstracts and title in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with a selection of exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'', ``ambient temperature'', ``temperature'').
+For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
 We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \ref{fig:dag-covid-19}.
@@
 
-Google Trends \citep{google_trends} data can be used as a tool to get insights into public interest (\textbf{awareness}) in the coronavirus disease.
+Google Trends \citep{google_trends} data can be used as a tool to get insights into public interest (\textbf{awareness}) in COVID-19.
 Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases \citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
@@
 Please note that ``Exposure to SARS-CoV-2'' is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset.
-\emph{Exposure to the virus} should not be confused with the formal \emph{exposure variables} of the DAG (mobility, awareness, temperature).
+\emph{Exposure to the virus} should not be confused with the formal \emph{exposure variables} of the DAG.
 
@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8 days for all confounders.
+Our observation period for the outcome consisted of all dates from 28 February 2020 to 8 July 2020 (\(T=132\)), since we used a lag of 12 days for all confounders.
 We did not exclude any states or districts (\(K=401\)).
@@
 Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables.
-For our causal analysis, we computed adjustment sets in three different scenarios for separate exposures within the DAG: i) mobility of population, ii) awareness of COVID-19 (i.e.~Google searches for ``corona''), iii) weather (i.e.~temperature).
+For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
 
@@
 ``Residential'' mobility is the relative change of daily time spent at residential areas.
+The six mobility variables showed high correlations among each other and with other variables.
+To reduce multicollinearity, we transformed them by principal component analysis (PCA) onto six uncorrelated principal components which were used in place of the original variables.
 
@@
 
-For different official and political measures we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
+For different official and political measures on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
+These variables were highly correlated with each other and other variables (especially mobility).
+To reduce multicollinearity, we aggregated the different official/political measures by summing them into one variable ``interventions'' which describes the current severity of interventions on a given day in a given district.
 
@@
 All variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered for numerical stability.
-We did not scale variables to unit variance to maintain interpretability of effects on the original scale of variables.
-Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 8 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 8 days.
+We scaled continuous variables by dividing by 2 times their standard deviation, while we left binary variables as-is \citep{gelman2008scaling}.
+After estimating their effects, we re-scaled variables to their original scale and for mobility, we re-transformed the effects of the principal components to the original mobility variables.
+Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.
 
@@
 We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases.
-Every node \(v_i\) in the graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\).
+We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework.
+Every node \(v_i\) in the causal graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\).
 The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\).
@@
 
-When we analyzed different adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets), we first checked if the set included unobserved variables.
+To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlated variables on their respective regression coefficients.
+The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the \(401\) districts.
+
+We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial.
+We used this parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \citep{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter.
+Afterwards, we calculated the effects of separate exposures on the outcome.
+For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets).
+Then, we first checked if the respective set included unobserved variables.
 If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \citep{perkovi2015complete}).
 If a minimal adjustment set contained unobserved variables, we discarded the whole set.
-We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for every remaining valid adjustment set as regressors and calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the sum of squared prediction errors of the current model and \(V_0\) is the sum of squared prediction errors of the null model (intercept and offset only).
-That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of variance unexplained.
-Finally, we decided for the model/adjustment set with the highest pseudo-\(R^2\).
-We report the exponentiated estimated coefficients along with 99 percent confidence intervals of the estimates.
+If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph.
+We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors and calculated the Akaike information criterion (AIC) for this set of regressors.
+Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC.
+We report the exponentiated estimated coefficients.
 
@@
 \midrule
-n & 54937\\
+n & 53333\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & -27.14 (24.59)\\
-\hspace{1em}grocery and pharmacy & -4.04 (23.00)\\
-\hspace{1em}parks & 46.33 (57.94)\\
-\hspace{1em}workplaces & -22.98 (20.56)\\
-\hspace{1em}residential & 8.20 (6.54)\\
-\hspace{1em}transit stations & -29.83 (21.22)\\
+\hspace{1em}retail and recreation & -27.70 (24.65)\\
+\hspace{1em}grocery and pharmacy & -4.36 (22.96)\\
+\hspace{1em}parks & 44.77 (57.40)\\
+\hspace{1em}workplaces & -23.45 (20.52)\\
+\hspace{1em}residential & 8.36 (6.56)\\
+\hspace{1em}transit stations & -30.25 (21.31)\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & 1.85 (3.98)\\
-\hspace{1em}Temperature & 10.74 (5.27)\\
-\hspace{1em}Humidity & 67.68 (13.10)\\
-\hspace{1em}Wind & 3.64 (1.67)\\
+\hspace{1em}Rainfall & 1.83 (3.95)\\
+\hspace{1em}Temperature & 10.50 (5.15)\\
+\hspace{1em}Humidity & 67.62 (13.23)\\
+\hspace{1em}Wind & 3.64 (1.68)\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Policies}}\\
-\hspace{1em}Ban of mass gatherings & 0.82 (0.38)\\
-\hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
-\hspace{1em}Contact restrictions & 0.74 (0.44)\\
-\hspace{1em}Mandatory face masks & 0.48 (0.50)\\
+\hspace{1em}Interventions & 2.56 (1.33)\\
 \addlinespace[0.3em]
@@
 \multicolumn{2}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 27.24 (18.31)\\
-\hspace{1em}COVID-19 burden & 3.55 (10.37)\\
+\hspace{1em}Searches corona & 27.64 (18.43)\\
+\hspace{1em}COVID-19 burden & 3.63 (10.49)\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
-\hspace{1em}Reported new cases COVID-19 & 3.60 (10.39)\\
-\hspace{1em}Active cases & 49.83 (121.96)\\
+\hspace{1em}Reported new cases COVID-19 & 3.71 (10.52)\\
+\hspace{1em}Active cases & 51.33 (123.47)\\
 \bottomrule
@@
 
-In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) -0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and temperature \(r_{A,B}=\) 0.79.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) -0.06, awareness (``Searches corona'') \(r_{A,C}=\) -0.26, and temperature \(r_{A,B}=\) 0.77.
 
@@
 \multicolumn{4}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & 1.0076 & 1.0056 & 1.0097\\
-\hspace{1em}grocery and pharmacy & 0.9951 & 0.9938 & 0.9965\\
-\hspace{1em}parks & 0.9998 & 0.9994 & 1.0003\\
-\hspace{1em}transit stations & 0.9975 & 0.9946 & 1.0004\\
-\hspace{1em}workplaces & 1.0031 & 1.0002 & 1.0059\\
-\hspace{1em}residential & 1.0080 & 0.9987 & 1.0174\\
+\hspace{1em}PC1 & 0.9044 & - & -\\
+\hspace{1em}PC2 & 0.9823 & - & -\\
+\hspace{1em}PC3 & 1.0465 & - & -\\
+\hspace{1em}PC4 & 1.0247 & - & -\\
+\hspace{1em}PC5 & 1.0037 & - & -\\
+\hspace{1em}PC6 & 0.9999 & - & -\\
 \addlinespace[0.3em]
@@
 \hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
-\hspace{1em}COVID-19 burden & 0.9942 & 0.9932 & 0.9951\\
+\hspace{1em}COVID-19 burden & 0.9005 & 0.8819 & 0.9202\\
 \addlinespace[0.3em]
 \multicolumn{4}{l}{\textbf{Weather}}\\
-\hspace{1em}Temperature & 0.9918 & 0.9879 & 0.9958\\
+\hspace{1em}Temperature & 0.9460 & 0.9129 & 0.9803\\
 \hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
@@
 Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent.
-Mobility on workplaces showed a small effect of 0.3 increase in case numbers for every 1 percent point increase in mobility.
+For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about percent.
+Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately percent.
+Mobility on workplaces showed a small effect of increase in case numbers for every 1 percent point increase in mobility.
 Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
 Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
-Thus, we expect an increase of daily cases by approximately 23 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 7 percent.
+Thus, we expect an increase of daily cases by approximately percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately percent.
 
@@
 For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 11 percent more infections for this district after 8 days.
-\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.6 percent decrease in newly reported case numbers.
-The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 11 percent.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 9.9 percent decrease in newly reported case numbers.
+The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 87.7 percent.
 
 Within our model, we observed a causal effect of temperature and all other weather variables.
-Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.8 percent.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 5.4 percent.
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent.
 We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
-In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 17.8 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 203.4 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
 For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 11.4 percent compared to a day with no precipitation.
@@
 \fontsize{7}{9}\selectfont
-\begin{tabular}[t]{llllllll}
+\begin{tabular}[t]{lllllllll}
 \toprule
-\rotatebox{90}{ } & \rotatebox{90}{Mobility} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Searches corona} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind}\\
+\rotatebox{90}{Foreign citizens} & \rotatebox{90}{Mobility} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Searches corona} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind} & \rotatebox{90}{ }\\
 \midrule
-Pseudo-R2 & 0.332 & 0.151 & 0.298 & 0.298 & 0.286 & 0.306 & 0.286\\
-Weekday & x & x & x & x & x & x & x\\
-Holiday (report) & x & x & x & x & x & x & x\\
-Holiday (exposure) & x & x & x & x & x & x & x\\
+0.005 & 0.006 & 0.002 & 0.005 & 0.004 & 0.005 & 0.006 & 0.005 & Pseudo-R2\\
+- & x & x & x & x & x & x & x & \vphantom{6} Pseudo-R2\\
+- & x & x & x & x & x & x & x & \vphantom{5} Pseudo-R2\\
+- & x & x & x & x & x & x & x & \vphantom{4} Pseudo-R2\\
 \addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation &  &  &  &  &  &  & \\
-\hspace{1em}grocery and pharmacy &  &  &  &  &  &  & \\
-\hspace{1em}parks &  &  &  &  &  &  & \\
-\hspace{1em}workplaces &  &  &  &  &  &  & \\
-\hspace{1em}residential &  &  &  &  &  &  & \\
-\hspace{1em}transit stations &  &  &  &  &  &  & \\
+\multicolumn{9}{l}{\textbf{Mobility}}\\
+\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{5} Pseudo-R2\\
+\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{4} Pseudo-R2\\
+\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{3} Pseudo-R2\\
+\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{2} Pseudo-R2\\
+\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{1} Pseudo-R2\\
+\hspace{1em}- &  &  &  &  &  &  &  & Pseudo-R2\\
 \addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & x & x & x & x &  & x & x\\
-\hspace{1em}Temperature & x & x & x &  &  & x & \\
-\hspace{1em}Humidity &  & x & x &  &  &  & \\
-\hspace{1em}Wind &  & x & x & x & x & x & \\
+\multicolumn{9}{l}{\textbf{Weather}}\\
+\hspace{1em}- & x & x & x & x &  & x & x & Pseudo-R2\\
+\hspace{1em}- & x & x & x &  &  & x &  & Pseudo-R2\\
+\hspace{1em}- &  & x & x &  &  &  &  & \vphantom{1} Pseudo-R2\\
+\hspace{1em}- &  & x & x & x & x & x &  & Pseudo-R2\\
 \addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Policies}}\\
-\hspace{1em}Ban of mass gatherings & x &  & x & x & x & x & x\\
-\hspace{1em}School and kindergarten closures & x &  & x & x & x & x & x\\
-\hspace{1em}Contact restrictions & x &  & x & x & x & x & x\\
-\hspace{1em}Mandatory face masks &  &  & x & x & x & x & x\\
+\multicolumn{9}{l}{\textbf{Policies}}\\
+\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{3} Pseudo-R2\\
+\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{2} Pseudo-R2\\
+\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{1} Pseudo-R2\\
+\hspace{1em}- &  &  & x &  & x & x & x & Pseudo-R2\\
 \addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Socio-demographic}}\\
-\hspace{1em}Age & x & x & x & x & x & x & x\\
-\hspace{1em}Gender & x & x & x & x & x & x & x\\
-\hspace{1em}Population density & x & x & x & x & x & x & x\\
-\hspace{1em}Foreign citizens &  & x & x &  &  &  & \\
-\hspace{1em}Socio-economic status & x & x & x & x & x & x & x\\
-\hspace{1em}Turnout &  & x &  &  &  &  & \\
-\hspace{1em}Right-wing populist party votes &  & x & x & x & x & x & x\\
-\hspace{1em}Nursing homes &  & x & x & x & x & x & x\\
+\multicolumn{9}{l}{\textbf{Socio-demographic}}\\
+\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{3} Pseudo-R2\\
+\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{2} Pseudo-R2\\
+\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{1} Pseudo-R2\\
+\hspace{1em}- &  & x & x &  &  &  &  & Pseudo-R2\\
+\hspace{1em}- & x & x & x & x & x & x & x & Pseudo-R2\\
+\hspace{1em}- &  & x &  &  &  &  &  & Pseudo-R2\\
+\hspace{1em}- &  & x & x & x & x & x & x & \vphantom{1} Pseudo-R2\\
+\hspace{1em}- &  & x & x & x & x & x & x & Pseudo-R2\\
 \addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & x &  &  & x & x & x & x\\
-\hspace{1em}COVID-19 burden & x &  & x & x & x & x & x\\
+\multicolumn{9}{l}{\textbf{Awareness}}\\
+\hspace{1em}- & x &  &  & x & x & x & x & Pseudo-R2\\
+\hspace{1em}- & x &  & x & x & x & x & x & Pseudo-R2\\
 \bottomrule
@@
 
-We are thankful for feedback from Thomas Czihal, Johannes Textor, and Ralph Brinks, who gave helpful suggestions on an earlier draft of the manuscript.
+We are thankful for feedback from Thomas Czihal, Johannes Textor, Ralph Brinks, and an anonymous reviewer who gave helpful suggestions on earlier versions of the manuscript.
 

### 2021-03-19 manuscript/v4_medrxiv_revision/results.Rmd
update manuscript

@@
                                                           "Weather (temperature)"="Temperature",
-                                                          "Weather (humidity)"="Humidity",
-                                                          "School/Kita closures"="School and kindergarten closures")) %>%
+                                                          "Weather (humidity)"="Humidity")) %>%
   mutate(Variable =  factor(Variable, ordered=TRUE, levels = c("n",
@@
                                                  "Wind",
-                                                 "Ban of mass gatherings",
-                                                 "School and kindergarten closures",
-                                                 "Contact restrictions",
-                                                 "Mandatory face masks",
-
+                                                 "Interventions",
                                                  "Age (pop. 65 and older)",
@@
 # Correlation of time and main vars
-mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"),contains("Weather"),`Searches corona`,contains("Mobility"), `COVID-19 burden`) %>% 
+mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"), "Rainfall", "Temperature", "Humidity", "Wind", `Searches corona`,contains("Mobility"), `COVID-19 burden`) %>% 
   mutate(outcome=`Reported new cases COVID-19`/(`Active cases`+1),date=as.numeric(date)) %>% 
@@
   pack_rows("Weather", 8, 11) %>%
-  pack_rows("Policies", 12, 15) %>%
-  pack_rows("Socio-demographic", 16, 25) %>%
-  pack_rows("Awareness", 26, 27) %>%
-  pack_rows("Case numbers (Outcome and offset)", 28, 29)
+  pack_rows("Policies", 12, 12) %>%
+  pack_rows("Socio-demographic", 13, 22) %>%
+  pack_rows("Awareness", 23, 24) %>%
+  pack_rows("Case numbers (Outcome and offset)", 25, 26)
 ```
@@
 In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.
-Figure \@ref(fig:timeseries)); with correlations between temporal progression and mobility in retail and recreation $r_{A,D}=$ `r round(mycorrs["date", "Mobility (retail and recreation)"], 2)`, awareness ("Searches corona") $r_{A,C}=$ `r round(mycorrs["date", "Searches corona"], 2)`, and temperature $r_{A,B}=$ `r round(mycorrs["date", "Weather (temperature)"], 2)`.
+Figure \@ref(fig:timeseries)); with correlations between temporal progression and mobility in retail and recreation $r_{A,D}=$ `r round(mycorrs["date", "Mobility (retail and recreation)"], 2)`, awareness ("Searches corona") $r_{A,C}=$ `r round(mycorrs["date", "Searches corona"], 2)`, and temperature $r_{A,B}=$ `r round(mycorrs["date", "Temperature"], 2)`.
 

### 2021-03-19 R/1_generate_data.R
fix one sd scale

@@ -187,7 +187,7 @@ scale_params <- tibble(variable=colnames(modeldata_X_cont),
                        mysd=apply(modeldata_X_cont, 2, sd))
 write_csv(scale_params, paste0(mydatapath, "scale_params.csv"))
 modeldata_X_cont_scaled <- sapply(seq(dim(modeldata_X_cont)[2]),
-                                  function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], 2*scale_params[i, 3])) # gelman
+                                  function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], scale_params[i, 3]))
 colnames(modeldata_X_cont_scaled) <- colnames(modeldata_X_cont)
 modeldata_X_cont_scaled <- as_tibble(modeldata_X_cont_scaled)
 modeldata_X_bin <- modeldata_X %>% dplyr::select(-colnames(modeldata_X_cont))

### 2021-03-19 R/z_auxiliary_functions_for_2.R

@@ -8,6 +8,7 @@ library(ggdag)
 library(ggthemes)
 library(pcalg)
 library(pscl)
+library(glmnet)
 
 # convert dagitty to adjacency matrix
 dagitty_to_adjmatrix <- function(daggity_obj) {
@@ -231,8 +232,8 @@ my_negbin_ridge <- function(modeldata, exposure=NULL, adjsets=NULL) {
                           offset=log(modeldata%>%pull(`Active cases`)+1),
                           standardize=FALSE,
                           alpha=0, lambda=mylambda)
-    myglm_fitted.values <- predict(myglm, myx, newoffset=log(modeldata%>%pull(`Active cases`)+1))[,1]
-    mynullglm_fitted.values <- predict(mynullglm, mynullx, newoffset=log(modeldata%>%pull(`Active cases`)+1))[,1]
+    myglm_fitted.values <- exp(predict(myglm, myx, newoffset=log(modeldata%>%pull(`Active cases`)+1))[,1])
+    mynullglm_fitted.values <- exp(predict(mynullglm, mynullx, newoffset=log(modeldata%>%pull(`Active cases`)+1))[,1])
     pseudor2s[idxf] <- 1-sum((myglm_fitted.values-myy)^2)/sum((mynullglm_fitted.values-myy)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
     aics[idxf] <- aic_glmnet(myglm)
     cat("   pseudo R2:", pseudor2s[idxf], "\n")
@@ -248,6 +249,25 @@ my_negbin_ridge <- function(modeldata, exposure=NULL, adjsets=NULL) {
     }
     if (!is.null(exposure)) {
       effects <- tail(coefficients[[idxf]]$estimate, nexp)
+      if (exposure=="Mobility") {
+        pcaloadings <- read_csv("data/pca_mobility_loadings.csv")
+        mobi_coeffs_pca <- coefficients[[idxf]] %>% filter(term %in% c(
+          "Mobility (PC1)",
+          "Mobility (PC2)",
+          "Mobility (PC3)",
+          "Mobility (PC4)",
+          "Mobility (PC5)",
+          "Mobility (PC6)"
+        )) %>% pull(estimate)
+        orig_mobi <- sapply(1:6, function(s) prod(mobi_coeffs_pca^as.numeric(pcaloadings[s, -1])))# as.vector(exp(as.matrix(pcaloadings[, 2:7]) %*% mobi_coeffs_pca))
[…]

### 2021-03-19 manuscript/v4_medrxiv_revision/data_methods.Rmd
updte dag plot and adjustment sets and effects

@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 28 February 2020 to 8 July 2020 ($T=132$), since we used a lag of 12 days for all confounders.
+Our observation period for the outcome consisted of all dates from 27 February 2020 to 8 July 2020 ($T=133$), since we used a lag of 12 days for all confounders.
 We did not exclude any states or districts ($K=401$).
-We analyzed the daily reported number of new cases as outcome ($K\cdot T=54{\ }937$ observations).
+We analyzed the daily reported number of new cases as outcome ($K\cdot T=53{\ }333$ observations).
 The set of possible predictors was derived from our causal DAG (see Table \@ref(tab:model-vars) and Figure \@ref(fig:dag-covid-19)).
@@
 
-All variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered for numerical stability.
-We scaled continuous variables by dividing by 2 times their standard deviation, while we left binary variables as-is [@gelman2008scaling].
-After estimating their effects, we re-scaled variables to their original scale and for mobility, we re-transformed the effects of the principal components to the original mobility variables.
-Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 12 days.
+All continuous variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is.
+After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale.
+Additionally for mobility, we re-transformed the effects of the principal components to the original mobility variables.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 12 days.
 
@@
 However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect.
-Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest asymptotic variance in the resulting causal effect estimates.
+Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest AIC.
 
@@
 For the defined exposures and the outcome "Reported new cases of COVID-19", we computed the minimal and optimal adjustment sets.
-Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the highest pseudo-$R^2$ (see next section) to estimate the final total causal effect from exposure to outcome.
+Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the lowest AIC (see next section) to estimate the final total causal effect from exposure to outcome.
 
@@
 
-To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach [@ridge1970], which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlated variables on their respective regression coefficients.
+To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach [@ridge1970], which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlation between variables on their respective regression coefficients.
 The hyper-parameter $\lambda$ of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the $401$ districts.

### 2021-03-19 manuscript/v4_medrxiv_revision/results.Rmd
updte dag plot and adjustment sets and effects

@@
 # Correlation of time and main vars
-mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"), "Rainfall", "Temperature", "Humidity", "Wind", `Searches corona`,contains("Mobility"), `COVID-19 burden`) %>% 
+mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"), "Rainfall", "Temperature", "Humidity", "Wind", `Searches corona`,contains("Mobility"), `COVID-19 burden`, "Interventions") %>% 
   mutate(outcome=`Reported new cases COVID-19`/(`Active cases`+1),date=as.numeric(date)) %>% 
@@
 library("kableExtra")
-causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_MinAdjSet1.csv"),
+causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_RedAdjSet1.csv"),
                            readr::read_csv("tables/t_effects_Searches corona_RedOptAdjSet.csv"),
@@
                            readr::read_csv("tables/t_effects_Humidity_RedOptAdjSet.csv"),
-                           readr::read_csv("tables/t_effects_Wind_RedOptAdjSet.csv")) %>%
-               rename(cause=variables, `effect estimate`=estimates, `CI (lower)`=conflow, `CI (upper)`=confhigh)
+                           readr::read_csv("tables/t_effects_Wind_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Interventions_RedOptAdjSet.csv")) %>%
+               rename(cause=variables, `effect estimate`=estimate)
 ## Note: no special chars in caption...
@@
 
-We list the results of our causal analysis for the effects of our variables in Table \@ref(tab:causaleffects).
+We list the results of our causal analysis for the effects of different exposure variables in Table \@ref(tab:causaleffects).
 The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate.
@@
 Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 1)` percent.
-Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 1)` increase in case numbers for every 1 percent point increase in mobility.
+For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 2)` percent.
+Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 2)` percent.
+Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase in mobility.
 Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
@@
 
-In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-$R^2$ values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-$R^2$.
-Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.
-Table \@ref(tab:adjsets), with the exception being COVID-19 burden).
-We also decided for the lag of 8 days based on the highest pseudo-$R^2$ values compared to other lags on the chosen adjustment sets.
-Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-$R^2$ values.
+The variable "interventions" showed ...
+
+The effects of the different socio-demographic factors ...
+
+In all cases we opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", where the minimal adjustment set had a lower AIC.
+We found that there were no valid adjustment sets for the unidentifiable variables turnout and right-wing populist party votes.
+
+We decided for the lag of 12 days based on the lowest AIC value compared to other lags between 1 and 20 days on the full model.
+Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
 

### 2021-03-19 R/0_z_testlags.R

@@ -189,7 +189,7 @@ for (mylag in mylags) {
                          mymean=colMeans(modeldata_X_cont),
                          mysd=apply(modeldata_X_cont, 2, sd))
   modeldata_X_cont_scaled <- sapply(seq(dim(modeldata_X_cont)[2]),
-                                    function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], 2*scale_params[i, 3])) # gelman
+                                    function(i) scale(modeldata_X_cont[, i], scale_params[i, 2], scale_params[i, 3]))
   colnames(modeldata_X_cont_scaled) <- colnames(modeldata_X_cont)
   modeldata_X_cont_scaled <- as_tibble(modeldata_X_cont_scaled)
   modeldata_X_bin <- modeldata_X %>% dplyr::select(-colnames(modeldata_X_cont))

### 2021-03-19 R/z_auxiliary_functions_for_2.R

@@ -303,7 +303,7 @@ my_causal <- function(dag, modeldata, exposure, unobserved) {
 write_cause <- function(res, exposure) {
   if (!is.null(res)) {
     whichmaxr2 <- which.max(res$pseudor2s)
-    whichminaic <- which.max(res$aics)
+    whichminaic <- which.min(res$aics)
     line <- paste0(exposure, ",", max(res$pseudor2s))
     write(line, file=paste0(tablepath, "t_pseudor2s.csv"), append=TRUE)
     line <- paste0(exposure, ",", min(res$aics))

### 2021-03-19 manuscript/v4_medrxiv_revision/figures/f_full_covid19_dag.R

@@ -3,13 +3,13 @@ library(ggdag)
 library(cowplot)
 library(stringr)
 
-figurepath <- "./manuscript/v1_medrxiv/figures/"
+figurepath <- "manuscript/v4_medrxiv_revision/figures/"
 
 # Set theme
 theme_set(theme_cowplot())
-mycolors <- c("#268bd2","#dc322f","#d33682","#859900","grey60","lightgrey")
+mycolors <- c("#268bd2","#dc322f","#d33682","#859900","grey60","lightgrey", "#ffa500")
 
-dagfile <- "./data/bigdag4covid19_plot.txt"
+dagfile <- "data/bigdag4covid19_plot.txt"
 dagtext <- readChar(dagfile, file.info(dagfile)$size)
 dag <- dagitty::dagitty(dagtext,layout = F)
 tidy_dag <- tidy_dagitty(dag) %>% mutate(name=str_wrap(name,width = 10)) 
@@ -34,10 +34,14 @@ tdc <- tidy_dag %>% mutate(curves = 0,
 tidy_dag %>% 
   group_by(name) %>% mutate(connections=n(),
                                          farbe = case_when(str_detect(name,"Mobility")~"4",
-                                                         # str_detect(name,"burden")~"3",
+                                                         str_detect(name,"burden")~"3",
                                                          str_detect(name,"Searches")~"3",
                                                          str_detect(name,"Temperature")~"2",
-                                                         str_detect(name,"new cases")~"1",                                  
+                                                         str_detect(name,"Rainfall")~"2",
+                                                         str_detect(name,"Humidity")~"2",
+                                                         str_detect(name,"Wind")~"2",
+                                                         str_detect(name,"new cases")~"1",  
+                                                         str_detect(name,"Interventions")~"7", 
                                                          str_detect(name,"Tracing")~"6",
                                                          str_detect(name,"Herd")~"6",
                                                          str_detect(name,"Access")~"6",

### 2021-03-19 manuscript/v4_medrxiv_revision/data_methods.Rmd
update table adj sets

@@
   pack_rows("Weather", 10, 13) %>%
-  pack_rows("Policies", 14, 17) %>%
-  pack_rows("Socio-demographic", 18, 25) %>%
-  pack_rows("Awareness", 26, 27) %>%
-  pack_rows("Case numbers", 28, 29) %>%
+  pack_rows("Policies", 14, 18) %>%
+  pack_rows("Socio-demographic", 19, 26) %>%
+  pack_rows("Awareness", 27, 28) %>%
+  pack_rows("Case numbers", 29, 30) %>%
   # pack_rows("Unobserved", 31, 36) %>%

### 2021-03-19 manuscript/v4_medrxiv_revision/manuscript.tex
update table adj sets

@@
 We used a directed acyclic graph (DAG) \citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread.
-To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar.
+To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google Scholar.
 Restrictions were applied to English and German language and the publication date in the last one year.
-The following search terms were applied to abstracts and title in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with a selection of exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'', ``ambient temperature'', ``temperature'').
+The following search terms were applied to abstracts and titles in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with a selection of exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'', ``ambient temperature'', ``temperature'').
 For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)).
@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 28 February 2020 to 8 July 2020 (\(T=132\)), since we used a lag of 12 days for all confounders.
+Our observation period for the outcome consisted of all dates from 27 February 2020 to 8 July 2020 (\(T=133\)), since we used a lag of 12 days for all confounders.
 We did not exclude any states or districts (\(K=401\)).
-We analyzed the daily reported number of new cases as outcome (\(K\cdot T=54{\ }937\) observations).
+We analyzed the daily reported number of new cases as outcome (\(K\cdot T=53{\ }333\) observations).
 The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}).
@@
 \hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
+\hspace{1em}Interventions & daily & district & numeric & sum of the above & -\\
 \addlinespace[0.3em]
@@
 
-All variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered for numerical stability.
-We scaled continuous variables by dividing by 2 times their standard deviation, while we left binary variables as-is \citep{gelman2008scaling}.
-After estimating their effects, we re-scaled variables to their original scale and for mobility, we re-transformed the effects of the principal components to the original mobility variables.
-Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.
+All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is.
+After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale.
+Additionally for mobility, we re-transformed the effects of the principal components to the original mobility variables.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.
 
@@
 However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect.
-Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest asymptotic variance in the resulting causal effect estimates.
+Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest AIC.
 
@@
 For the defined exposures and the outcome ``Reported new cases of COVID-19'', we computed the minimal and optimal adjustment sets.
-Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the highest pseudo-\(R^2\) (see next section) to estimate the final total causal effect from exposure to outcome.
+Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the lowest AIC (see next section) to estimate the final total causal effect from exposure to outcome.
 
@@
 
-To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlated variables on their respective regression coefficients.
+To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlation between variables on their respective regression coefficients.
 The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the \(401\) districts.
@@
 If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph.
-We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors and calculated the Akaike information criterion (AIC) for this set of regressors.
+We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors (that is, we applied ridge regression) and calculated the Akaike information criterion (AIC) for this model/set of regressors.
 Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC.
-We report the exponentiated estimated coefficients.
+We report the exponentiated estimated coefficients for the separate exposures on their original scale.
 
@@
 
-\caption{\label{tab:causaleffects}Causal effect estimates with 99 percent confidence bands}
+\caption{\label{tab:causaleffects}Effect estimates from causal graph analysis}
 \centering
-\begin{tabular}[t]{lrrr}
+\begin{tabular}[t]{lr}
 \toprule
-cause & effect estimate & CI (lower) & CI (upper)\\
+cause & effect estimate\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Mobility}}\\
-\hspace{1em}PC1 & 0.9044 & - & -\\
-\hspace{1em}PC2 & 0.9823 & - & -\\
-\hspace{1em}PC3 & 1.0465 & - & -\\
-\hspace{1em}PC4 & 1.0247 & - & -\\
-\hspace{1em}PC5 & 1.0037 & - & -\\
-\hspace{1em}PC6 & 0.9999 & - & -\\
+\multicolumn{2}{l}{\textbf{Mobility}}\\
+\hspace{1em}retail and recreation & 1.0010\\
+\hspace{1em}grocery and pharmacy & 0.9988\\
+\hspace{1em}parks & 0.9997\\
+\hspace{1em}transit stations & 1.0024\\
+\hspace{1em}workplaces & 1.0025\\
+\hspace{1em}residential & 0.9914\\
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
-\hspace{1em}COVID-19 burden & 0.9005 & 0.8819 & 0.9202\\
+\multicolumn{2}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & 1.0007\\
+\hspace{1em}COVID-19 burden & 0.9922\\
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Weather}}\\
-\hspace{1em}Temperature & 0.9460 & 0.9129 & 0.9803\\
-\hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
-\hspace{1em}Humidity & 1.0022 & 1.0006 & 1.0039\\
-\hspace{1em}Wind & 0.9893 & 0.9802 & 0.9985\\
+\multicolumn{2}{l}{\textbf{Weather}}\\
+\hspace{1em}Temperature & 0.9959\\
+\hspace{1em}Rainfall & 1.0100\\
+\hspace{1em}Humidity & 1.0042\\
+\hspace{1em}Wind & 1.0164\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Policies}}\\
+\hspace{1em}Interventions & 0.8037\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Demographic}}\\
+\hspace{1em}Age (pop. 65 and older & 0.9976\\
+\hspace{1em}Age (pop. younger 18 & 1.0097\\
+\hspace{1em}Foreign citizens & 1.0034\\
+\hspace{1em}Foreign citizens (refugees & 1.0010\\
+\hspace{1em}Gender & 0.5232\\
+\hspace{1em}Nursing homes & 1.0007\\
+\hspace{1em}Population density & 1.0000\\
+\hspace{1em}Socio-economic status & 0.9993\\
 \bottomrule
@@
 
-We list the results of our causal analysis for the effects of our variables in Table \ref{tab:causaleffects}.
+We list the results of our causal analysis for the effects of different exposure variables in Table \ref{tab:causaleffects}.
 The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate.
@@
 Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately percent.
-Mobility on workplaces showed a small effect of increase in case numbers for every 1 percent point increase in mobility.
+For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.1 percent.
+Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.12 percent.
+Mobility on workplaces showed a small effect of 0.25 increase in case numbers for every 1 percent point increase in mobility.
 Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
 Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
-Thus, we expect an increase of daily cases by approximately percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately percent.
+Thus, we expect an increase of daily cases by approximately 3 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 2 percent.
 
@@
 Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases.
-An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent.
-For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 11 percent more infections for this district after 8 days.
-\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 9.9 percent decrease in newly reported case numbers.
-The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 87.7 percent.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.1 percent.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 1 percent more infections for this district after 8 days.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.8 percent decrease in newly reported case numbers.
+The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 14.6 percent.
 
 Within our model, we observed a causal effect of temperature and all other weather variables.
-Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 5.4 percent.
-On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.4 percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1 percent.
 We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
-In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 203.4 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
-For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 11.4 percent compared to a day with no precipitation.
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 8.5 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 10.5 percent compared to a day with no precipitation.
 
-In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\).
-Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.~Table \ref{tab:adjsets}, with the exception being COVID-19 burden).
-We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets.
-Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-\(R^2\) values.
+The variable ``interventions'' showed \ldots{}
+
+The effects of the different socio-demographic factors \ldots{}
+
+In all cases we opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC.
+We found that there were no valid adjustment sets for the unidentifiable variables turnout and right-wing populist party votes.
+
+We decided for the lag of 12 days based on the lowest AIC value compared to other lags between 1 and 20 days on the full model.
+Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
 
@@
 \fontsize{7}{9}\selectfont
-\begin{tabular}[t]{lllllllll}
+\begin{tabular}[t]{lllllllllllllll}
 \toprule
-\rotatebox{90}{Foreign citizens} & \rotatebox{90}{Mobility} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Searches corona} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind} & \rotatebox{90}{ }\\
+\rotatebox{90}{ } & \rotatebox{90}{Mobility} & \rotatebox{90}{Searches corona} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind} & \rotatebox{90}{Interventions} & \rotatebox{90}{Age} & \rotatebox{90}{Foreign citizens} & \rotatebox{90}{Gender} & \rotatebox{90}{Nursing homes} & \rotatebox{90}{Population density} & \rotatebox{90}{Socio-economic status}\\
 \midrule
-0.005 & 0.006 & 0.002 & 0.005 & 0.004 & 0.005 & 0.006 & 0.005 & Pseudo-R2\\
-- & x & x & x & x & x & x & x & \vphantom{6} Pseudo-R2\\
-- & x & x & x & x & x & x & x & \vphantom{5} Pseudo-R2\\
-- & x & x & x & x & x & x & x & \vphantom{4} Pseudo-R2\\
+Weekday & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
+Holiday (report) & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
+Holiday (exposure) & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
 \addlinespace[0.3em]
-\multicolumn{9}{l}{\textbf{Mobility}}\\
-\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{5} Pseudo-R2\\
-\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{4} Pseudo-R2\\
-\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{3} Pseudo-R2\\
-\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{2} Pseudo-R2\\
-\hspace{1em}- &  &  &  &  &  &  &  & \vphantom{1} Pseudo-R2\\
-\hspace{1em}- &  &  &  &  &  &  &  & Pseudo-R2\\
+\multicolumn{15}{l}{\textbf{Mobility}}\\
+\hspace{1em}retail and recreation &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}grocery and pharmacy &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}parks &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}workplaces &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}residential &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}transit stations &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
 \addlinespace[0.3em]
-\multicolumn{9}{l}{\textbf{Weather}}\\
-\hspace{1em}- & x & x & x & x &  & x & x & Pseudo-R2\\
-\hspace{1em}- & x & x & x &  &  & x &  & Pseudo-R2\\
-\hspace{1em}- &  & x & x &  &  &  &  & \vphantom{1} Pseudo-R2\\
-\hspace{1em}- &  & x & x & x & x & x &  & Pseudo-R2\\
+\multicolumn{15}{l}{\textbf{Weather}}\\
+\hspace{1em}Rainfall & x & x & x & x &  & x & x & x & x & x & x &  & x & x\\
+\hspace{1em}Temperature & x & x & x &  &  & x &  & x & x & x & x &  & x & x\\
+\hspace{1em}Humidity & x & x & x &  &  &  &  & x & x & x & x &  & x & x\\
+\hspace{1em}Wind & x & x & x & x & x & x &  & x & x & x & x &  & x & x\\
 \addlinespace[0.3em]
-\multicolumn{9}{l}{\textbf{Policies}}\\
-\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{3} Pseudo-R2\\
-\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{2} Pseudo-R2\\
-\hspace{1em}- &  &  & x &  & x & x & x & \vphantom{1} Pseudo-R2\\
-\hspace{1em}- &  &  & x &  & x & x & x & Pseudo-R2\\
+\multicolumn{15}{l}{\textbf{Policies}}\\
+\hspace{1em}Interventions & x & x &  & x & x & x & x &  & x & x & x & x & x & x\\
 \addlinespace[0.3em]
-\multicolumn{9}{l}{\textbf{Socio-demographic}}\\
-\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{3} Pseudo-R2\\
-\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{2} Pseudo-R2\\
-\hspace{1em}- & x & x & x & x & x & x & x & \vphantom{1} Pseudo-R2\\
-\hspace{1em}- &  & x & x &  &  &  &  & Pseudo-R2\\
-\hspace{1em}- & x & x & x & x & x & x & x & Pseudo-R2\\
-\hspace{1em}- &  & x &  &  &  &  &  & Pseudo-R2\\
-\hspace{1em}- &  & x & x & x & x & x & x & \vphantom{1} Pseudo-R2\\
-\hspace{1em}- &  & x & x & x & x & x & x & Pseudo-R2\\
+\multicolumn{15}{l}{\textbf{Socio-demographic}}\\
+\hspace{1em}Age & x & x & x & x & x & x & x & x &  &  &  & x & x & x\\
+\hspace{1em}Gender & x & x & x & x & x & x & x & x & x & x &  & x & x & x\\
+\hspace{1em}Population density & x & x & x & x & x & x & x & x & x & x & x & x &  & x\\
+\hspace{1em}Foreign citizens &  & x & x &  &  &  &  & x & x &  & x &  & x & x\\
+\hspace{1em}Socio-economic status & x & x & x & x & x & x & x & x &  &  &  & x & x & \\
+\hspace{1em}Trust in society &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}Turnout &  &  & x &  &  &  &  & x &  &  &  &  &  & \\
+\hspace{1em}Right-wing populist party votes & x & x & x & x & x & x & x & x &  &  &  & x &  & \\
+\hspace{1em}Nursing homes & x & x & x & x & x & x & x & x &  &  &  &  & x & \\
 \addlinespace[0.3em]
-\multicolumn{9}{l}{\textbf{Awareness}}\\
-\hspace{1em}- & x &  &  & x & x & x & x & Pseudo-R2\\
-\hspace{1em}- & x &  & x & x & x & x & x & Pseudo-R2\\
+\multicolumn{15}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & x &  &  & x & x & x & x &  &  &  &  & x &  & \\
+\hspace{1em}COVID-19 burden & x & x &  & x & x & x & x & x & x & x & x & x & x & x\\
 \bottomrule

### 2021-03-19 manuscript/v4_medrxiv_revision/results.Rmd
update table adj sets

@@
 library("kableExtra")
-causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_RedAdjSet1.csv"),
+causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_RedOptAdjSet.csv"),
                            readr::read_csv("tables/t_effects_Searches corona_RedOptAdjSet.csv"),
@@
                            readr::read_csv("tables/t_effects_Wind_RedOptAdjSet.csv"),
-                           readr::read_csv("tables/t_effects_Interventions_RedOptAdjSet.csv")) %>%
-               rename(cause=variables, `effect estimate`=estimate)
+                           readr::read_csv("tables/t_effects_Interventions_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Age_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Foreign citizens_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Gender_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Nursing homes_MinAdjSet1.csv"),
+                           readr::read_csv("tables/t_effects_Population density_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Socio-economic status_RedOptAdjSet.csv")) %>%
+               rename(cause=variables, `effect estimate`=estimates)
 ## Note: no special chars in caption...
@@
              booktabs = TRUE,
-             caption = "Causal effect estimates with 99 percent confidence bands",
+             caption = "Effect estimates from causal graph analysis",
              digits=4) %>%
@@
   pack_rows("Awareness", 7, 8) %>%
-  pack_rows("Weather", 9, 12)
+  pack_rows("Weather", 9, 12) %>%
+  pack_rows("Policies", 13, 13) %>%
+  pack_rows("Demographic", 14, 21)
 ```
@@
 library(kableExtra)
-myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:22, 24:28), ] %>% dplyr::select(variable) %>%
-  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ")
-for (cause in colnames(myvariables)[-c(1,2)]) {
+myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:13, 18:29), ] %>% dplyr::select(variable) %>%
+  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ",
+         Interventions= " ", Age=" ", `Foreign citizens`= " ", Gender=" ", `Nursing homes`=" ", `Population density`=" ", `Socio-economic status`=" ")
+for (cause in colnames(myvariables)[-c(1,13)]) {
   adjset <- readLines(paste0("tables/t_adjust_", cause, "_RedOptAdjSet.csv"))
@@
 }
-cause <- colnames(myvariables)[2]
+cause <- colnames(myvariables)[13]
 adjset <- readLines(paste0("tables/t_adjust_", cause, "_MinAdjSet1.csv"))
@@
 }
-t_pseudor2s <- read_csv("tables/t_pseudor2s.csv", col_names = c("exposure", "pseudor2"), col_types = "cn") %>%
-  mutate(pseudor2=as.character(round(pseudor2, 3))) %>%
-  pivot_wider(names_from = "exposure", values_from = "pseudor2")
+# t_pseudor2s <- read_csv("tables/t_pseudor2s.csv", col_names = c("exposure", "pseudor2"), col_types = "cn") %>%
+#   mutate(pseudor2=as.character(round(pseudor2, 3))) %>%
+#   pivot_wider(names_from = "exposure", values_from = "pseudor2")
 
-myvariables <- bind_rows(t_pseudor2s, myvariables)[c(8, 1:7)]
-myvariables$variable[1] <- "Pseudo-R2"
+# myvariables <- bind_rows(t_pseudor2s, myvariables)[c(8, 1:7)]
+# myvariables$variable[1] <- "Pseudo-R2"
 kable(myvariables %>% rename(" "=variable),
@@
   row_spec(0, angle = 90) %>%
-  pack_rows("Mobility", 5, 10) %>%
-  pack_rows("Policies", 15, 18) %>%
-  pack_rows("Socio-demographic", 19, 26) %>%
-  pack_rows("Awareness", 27, 28) %>%
-  pack_rows("Weather", 11, 14) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
+  pack_rows("Mobility", 4, 9) %>%
+  pack_rows("Policies", 14, 14) %>%
+  pack_rows("Socio-demographic", 15, 23) %>%
+  pack_rows("Awareness", 24, 25) %>%
+  pack_rows("Weather", 10, 13) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
 ```

### 2021-03-20 R/1_generate_data.R
update gender scaling (percentage like foreign citizens and age)

@@ -46,6 +46,7 @@ inkar <- read_csv(paste0(mydatapath, "sociodemographic/inkar.csv")) %>%
   mutate(id=as.integer(Kennziffer)*1000) %>%
   dplyr::select(-Kennziffer, -Raumeinheit, -Aggregat) %>%
   mutate(`Anteil Schutzsuchender an Bevölkerung`=replace_na(`Anteil Schutzsuchender an Bevölkerung`, mean(`Anteil Schutzsuchender an Bevölkerung`, na.rm = TRUE))) %>%
+  mutate(Frauenanteil=Frauenanteil*100) %>% 
   rename(`Age (pop. 65 and older)`=`Einwohner 65 Jahre und älter`,
          `Population density`=Einwohnerdichte,
          `Sex`=Frauenanteil,

### 2021-03-20 manuscript/v4_medrxiv_revision/results.Rmd
cv for lag (5), update data accordingly, interventions splitted again (ridge should take care of multicollinearity)

--- 2021-03-19 manuscript/v4_medrxiv_revision/results.Rmd
@@
 library("kableExtra")
-causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_RedAdjSet1.csv"),
+library(stringr)
+causaleffects <- bind_rows(readr::read_csv("tables/t_effects_Mobility_RedOptAdjSet.csv"),
                            readr::read_csv("tables/t_effects_Searches corona_RedOptAdjSet.csv"),
@@
                            readr::read_csv("tables/t_effects_Wind_RedOptAdjSet.csv"),
-                           readr::read_csv("tables/t_effects_Interventions_RedOptAdjSet.csv")) %>%
-               rename(cause=variables, `effect estimate`=estimate)
+                           readr::read_csv("tables/t_effects_Interventions_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Age_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Foreign citizens_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Gender_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Nursing homes_MinAdjSet1.csv"),
+                           readr::read_csv("tables/t_effects_Population density_RedOptAdjSet.csv"),
+                           readr::read_csv("tables/t_effects_Socio-economic status_RedOptAdjSet.csv")) %>%
+               rename(cause=variables, `effect estimate`=estimates)
 ## Note: no special chars in caption...
 knitr::kable(causaleffects %>% 
-               mutate(cause=stringr::str_remove(stringr::str_remove(cause,"Mobility \\("),"\\)")),
+               mutate(cause=ifelse(str_starts(cause, "Mobility"), stringr::str_remove(stringr::str_remove(cause,"Mobility \\("),"\\)"), cause)),
              booktabs = TRUE,
-             caption = "Causal effect estimates with 99 percent confidence bands",
+             caption = "Effect estimates from causal graph analysis",
              digits=4) %>%
@@
 library(kableExtra)
-myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:22, 24:28), ] %>% dplyr::select(variable) %>%
-  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ")
-for (cause in colnames(myvariables)[-c(1,2)]) {
+myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:13, 18:23, 25:29), ] %>% dplyr::select(variable) %>%
+  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ",
+         Interventions= " ", Age=" ", `Foreign citizens`= " ", Gender=" ", `Nursing homes`=" ", `Population density`=" ", `Socio-economic status`=" ")
+for (cause in colnames(myvariables)[-c(1,13)]) {
   adjset <- readLines(paste0("tables/t_adjust_", cause, "_RedOptAdjSet.csv"))
@@
   row_spec(0, angle = 90) %>%
-  pack_rows("Mobility", 5, 10) %>%
-  pack_rows("Policies", 15, 18) %>%
-  pack_rows("Socio-demographic", 19, 26) %>%
-  pack_rows("Awareness", 27, 28) %>%
-  pack_rows("Weather", 11, 14) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
+  pack_rows("Mobility", 4, 9) %>%
+  pack_rows("Policies", 14, 14) %>%
+  pack_rows("Socio-demographic", 15, 22) %>%
+  pack_rows("Awareness", 23, 24) %>%
+  pack_rows("Weather", 10, 13) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
 ```

### 2021-03-20 R/0_z_testlags.R

@@ -1,6 +1,8 @@
 mylags <- 1:20
 
 myaic <- rep(0, 20)
+myaic_ridge <- myaic
+myaic_ridge0 <- myaic
 myglms <- vector("list", 20)
 
 library(lubridate)
@@ -8,6 +10,8 @@ library(tidyverse)
 library(DBI)
 library(broom)
 library(MASS)
+
+source("R/z_auxiliary_functions_for_2.R")
 
 mydatapath <- "./data/" 
 
@@ -100,9 +104,9 @@ measures_zeroone_all <- measures_zeroone_all %>%
          `Mandatory face masks`=ifelse(id==8325049 & date>=as.Date("2020-04-17"), 1, `Mandatory face masks`), # rottweil
          `Mandatory face masks`=ifelse(id==16062041  & date>=as.Date("2020-04-14"), 1, `Mandatory face masks`), # nordhausen
          `Mandatory face masks`=ifelse(id==16053000  & date>=as.Date("2020-04-06"), 1, `Mandatory face masks`) # jena
-  ) %>% 
-  mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
-  dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
+  ) #%>% 
+  # mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
+  # dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
 
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
@@ -133,6 +137,15 @@ google_mobility <- google_mobility %>%
 startdate <- min(dateset) + max(mylags)
 enddate <- as.Date("2020-07-08") # startdate + days(100) # 
 
+set.seed(3141)
+myids <- sample(unique(brd_timeseries$id))
+myrandomfoldids <- c(rep(1:9, each=40), rep(10, 41))
+myidmatch <- tibble(id=myids, randomfoldid=myrandomfoldids)
+
+myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
+
+cverror <- matrix(0, 20, 10)
+
[…]

### 2021-03-20 R/1_generate_data.R

@@ -4,7 +4,7 @@ library(DBI)
 library(broom)
 library(MASS)
 
-mylag <- 12
+mylag <- 5
 
 mydatapath <- "./data/" 
 
@@ -98,9 +98,9 @@ measures_zeroone_all <- measures_zeroone_all %>%
          `Mandatory face masks`=ifelse(id==8325049 & date>=as.Date("2020-04-17"), 1, `Mandatory face masks`), # rottweil
          `Mandatory face masks`=ifelse(id==16062041  & date>=as.Date("2020-04-14"), 1, `Mandatory face masks`), # nordhausen
          `Mandatory face masks`=ifelse(id==16053000  & date>=as.Date("2020-04-06"), 1, `Mandatory face masks`) # jena
-         ) %>% 
-  mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
-  dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
+         ) # %>% 
+  # mutate(Interventions=`Ban of mass gatherings`+`School/Kita closures`+`Contact restrictions`+`Mandatory face masks`) %>%
+  # dplyr::select(-c(`Ban of mass gatherings`, `School/Kita closures`, `Contact restrictions`, `Mandatory face masks`))
 
 dateset <- as.Date(Reduce(intersect, list(awareness$date, google_mobility$date, weather$date, brd_timeseries$date)),
                    origin="1970-01-01")
@@ -148,7 +148,10 @@ modeldata_raw <- left_join(brd_timeseries, google_mobility %>%
   left_join(., pflege_destatis, by=c("id"="id")) %>%
   left_join(., brd_burden, by=c('date'='date', 'id'='id')) %>%
   filter(id>16 & date>=startdate & date<=enddate) %>%
-  rename(# `School and kindergarten closures`=`School/Kita closures`,
+  rename(`Interventions (school and kindergarten closures)`=`School/Kita closures`,
+         `Interventions (mandatory face masks)`=`Mandatory face masks`,
+         `Interventions (ban of mass gatherings)`=`Ban of mass gatherings`,
+         `Interventions (contact restrictions)`=`Contact restrictions`,
          `Holiday (exposure)`=Holiday,
          `Foreign citizens`=`Foreign residents`,
          `Foreign citizens (refugees)`=`Foreign residents (refugees)`,

### 2021-03-20 R/3_modelassumptions.R

@@ -39,7 +39,7 @@ myx <- as.matrix(modeldata %>% dplyr::select(-`Reported new cases COVID-19`))
 
 myfoldids <- (id_daycount %>% group_indices(id)) %% 10 + 1
 
-set.seed(1502)
+set.seed(2407)
 myglm_ridge_cv <- cv.glmnet(x=myx, y=myy, family = negative.binomial(mytheta),
                             offset=log(modeldata%>%pull(`Active cases`)+1),
                             standardize=FALSE,
@@ -57,8 +57,8 @@ myglm_ridge <- glmnet(myx, myy, family = negative.binomial(mytheta),
 # residuals
 
 devresids <- residuals(myglm, type="deviance")
-# library(statmod)
-# qresids <- qresid(myglm_ridge)
+library(statmod)
+qresids <- qresid(myglm)
 
 # linearity
 scatter.smooth(predict(myglm, type="link"), devresids, col='gray')
@@ -76,13 +76,13 @@ ggplot(myresidagainstcont, aes(x=value, y=devresids)) +
   geom_smooth(se=FALSE)
 
 # proper distribution
-qqnorm(qresids[-1971])
-qqline(qresids[-1971])
-hist(qresids[-1971], freq=FALSE)
-lines(density(qresids[-1971]),
+qqnorm(qresids)
+qqline(qresids)
+hist(qresids, freq=FALSE)
+lines(density(qresids),
       lwd = 2,
       col = "red")
-lines(density(rnorm(length(qresids[-1971]))),
+lines(density(rnorm(length(qresids))),
       lwd = 2,
       col = "blue")
 
@@ -110,18 +110,3 @@ corrplot(corrplotscores$r,
 library(car)
 myvifs <- vif(myglm)
 myvifs[myvifs>=5]
-
-# rescale principal components of mobility to original mobility variables
-pcaloadings <- read_csv("data/pca_mobility_loadings.csv")
-orig_mobi <- as.vector(exp(as.matrix(pcaloadings[, 2:7]) %*% myglm$coefficients[c(
-  "`Mobility (PC1)`",
-  "`Mobility (PC2)`",
-  "`Mobility (PC3)`",
-  "`Mobility (PC4)`",
-  "`Mobility (PC5)`",
-  "`Mobility (PC6)`"
-)]))
-names(orig_mobi) <- pcaloadings$original_variable
-orig_mobi
-
-coeffs <- tidy(myglm, exponentiate=TRUE) # , conf.int=TRUE, conf.level = 0.99

### 2021-03-20 R/z_auxiliary_functions_for_2.R

@@ -317,7 +317,7 @@ write_cause <- function(res, exposure) {
                  exposure, "_", 
                  names(res$adjustmentsets[whichminaic]), ".csv"))
   } else {
-    cat("NO ADJUSTMENT SET FOR ", exposure, " (non-identifiable)\n")
+    cat("\n NO ADJUSTMENT SET FOR ", exposure, " (non-identifiable) \n \n ")
   }
   return(NULL)
 }

### 2021-03-20 manuscript/v4_medrxiv_revision/figures/f_exposure_vs_effect.R

@@ -1,7 +1,7 @@
 library("tidyverse")
 library("cowplot")
 
-figurepath <- "./manuscript/v1_medrxiv/figures/"
+figurepath <- "manuscript/v4_medrxiv_revision/figures/"
 
 modeldata_raw <- read_csv("data/Modeldata_raw.csv")
 
@@ -11,10 +11,15 @@ varnames <- c("Mobility (retail and recreation)",
               "Mobility (transit stations)",
               "Mobility (workplaces)",
               "Mobility (residential)",
-              "Weather (temperature)", "Weather (rainfall)", "Weather (humidity)", "Weather (wind)",
-              "Searches corona", "COVID-19 burden") 
+              "Temperature", "Rainfall", "Humidity", "Wind",
+              "Searches corona", "COVID-19 burden",
+              "Interventions",
+              "Age (pop. 65 and older)", "Age (pop. younger 18)", 
+              "Foreign citizens", "Foreign citizens (refugees)",
+              "Gender", "Nursing homes",
+              "Population density", "Socio-economic status") 
 varnames <- factor(varnames, levels=varnames, ordered=TRUE)
-vardata <- tibble(varnames=varnames, means=0, effects=0, conflow=0, confhigh=0, rangelow=0, rangehigh=0)
+vardata <- tibble(varnames=varnames, means=0, effects=0, rangelow=0, rangehigh=0)
 
 ranges <- apply(modeldata_raw[, varnames], 2, quantile, probs=c(0.025, 0.975))
 
@@ -24,50 +29,52 @@ vardata <- vardata %>%
          rangehigh=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[2]))
 vardata$myrangelow <- ranges[1, varnames] # c(-100, -25, -0, -100, -50, -10, -10, 0, 0, 0, 0, 0)
 vardata$myrangehigh <- ranges[2, varnames] # c(+50, +25, +100, +50, +0, +20, 30, 20, 100, 15, 100, 100)
-vardata[1:6, 3:5] <- cbind(read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$estimates,
-                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$conflow,
-                        read_csv(paste0(figurepath, "../tables/t_effects_Mobility_MinAdjSet1.csv"))$confhigh)
[…]

### 2021-03-20 manuscript/v4_medrxiv_revision/data_methods.Rmd
finish manuscript for revision

@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 27 February 2020 to 8 July 2020 ($T=133$), since we used a lag of 12 days for all confounders.
+Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 ($T=140$), since we used a lag of 5 days for all confounders.
 We did not exclude any states or districts ($K=401$).
-We analyzed the daily reported number of new cases as outcome ($K\cdot T=53{\ }333$ observations).
+We analyzed the daily reported number of new cases as outcome ($K\cdot T=56{\ }140$ observations).
 The set of possible predictors was derived from our causal DAG (see Table \@ref(tab:model-vars) and Figure \@ref(fig:dag-covid-19)).
@@
 
-For different official and political measures on a day and district level we used one-hot encoded daily variables, i.e.
+For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.
 ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
-These variables were highly correlated with each other and other variables (especially mobility).
-To reduce multicollinearity, we aggregated the different official/political measures by summing them into one variable "interventions" which describes the current severity of interventions on a given day in a given district.
 
@@
   pack_rows("Mobility", 4, 9) %>%
-  pack_rows("Weather", 10, 13) %>%
-  pack_rows("Policies", 14, 18) %>%
-  pack_rows("Socio-demographic", 19, 26) %>%
-  pack_rows("Awareness", 27, 28) %>%
-  pack_rows("Case numbers", 29, 30) %>%
+  pack_rows("Awareness", 10, 11) %>%
+  pack_rows("Weather", 12, 15) %>%
+  pack_rows("Interventions", 16, 19) %>%
+  pack_rows("Socio-demographic", 20, 27) %>%
+  pack_rows("Case numbers", 28, 29) %>%
   # pack_rows("Unobserved", 31, 36) %>%
@@
 After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale.
-Additionally for mobility, we re-transformed the effects of the principal components to the original mobility variables.
-Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 12 days.
+Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 
-To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach [@ridge1970], which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlation between variables on their respective regression coefficients.
-The hyper-parameter $\lambda$ of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the $401$ districts.
-
 We conducted a log-linear regression (function `glm` with `family=poisson()` for Poisson regression, and `glm.nb` from the `MASS` package for the negative binomial regression [@mass]) for the full data set to assess general model adequacy and to estimate the $\theta$ parameter of the negative binomial.
-We used this parameter with the `cv.glmnet` function from the R package `glmnet` [@glmnet2010] with `family=negative.binomial(theta)` and chose the $\lambda$ value within one standard deviation from the minimal $\lambda$ as regularization hyper-parameter.
+The proper lag between exposures and outcome was found by 10-fold cross-validation on different lags between 1 and 20 days.
+Model diagnostics on the final full model did not show severe problems with model assumptions (linearity, distribution of residuals, independence of observations).
+Analysis of variance inflation factors revealed some problems with multicollinearity.
+To reduce the effects of multicollinearity, first we transformed the highly correlated mobility variables by PCA as described above.
+Second, we used a ridge regression approach [@ridge1970], which is a regularization method that shrinks regression coefficients and alleviates the effect of correlation between variables on their respective regression coefficients.
+Furthermore, regularized regression allows for better fits on unseen data, thus preventing overfitting the data, too.
+The hyper-parameter $\lambda$ of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from random subsets of the $401$ districts.
+We used this hyper-parameter with the `cv.glmnet` function from the R package `glmnet` [@glmnet2010] with `family=negative.binomial(theta)` and chose the $\lambda$ value within one standard deviation from the minimal $\lambda$ as regularization hyper-parameter.
 Afterwards, we calculated the effects of separate exposures on the outcome.

### 2021-03-20 manuscript/v4_medrxiv_revision/discussion.Rmd
finish manuscript for revision

@@
 
-Our objective was to identify causal effects for COVID-19 cases.
+Our objective was to identify effects of determining factors for COVID-19 cases within a causal framework.
 We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers).
 We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population and/or effective public health measures to suppress a local outbreak.
-The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.
+Mobility showed distinct effects: Increasing activity in retail and recreational areas, as well as transit stations and workplaces increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) and in parks or residential areas led to reduced case numbers.
+All interventions considered (ban of mass gatherings, school/kindergarten closures, contact restrictions, mandatory face masks) reduced case numbers considerably.
+Socio-demographic variables had small effects individually, but in conjunction they explained larger case numbers in (urban) areas with younger population, lower socio-economic status, and higher population density.
 
@@
 Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects [@fowler_effect_2020;@li_retrospective_2020;@shi_impact_2020;@Wang2020temperature].
-To reliably identify causal drivers, one must adjust for confounders.
+To reliably identify (causal) drivers, one must adjust for confounders.
 To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
 A causal approach is used in another current COVID-19 analysis [@gencoglu2020causal].
-There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.
+There, however, they identify the causal relationships (reconstruct a DAG), while we estimated effects for a given hypothesized causal DAG.
 
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
+Finally, we found a positive effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
 
@@
 
-While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
+While use of a causal DAG is itself a strong tool to identify *causal* effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
 We endorse our graph as a basis for discussion on residual confounding.
@@
 As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.
-between the policies like contact restrictions and mandatory face masks).
+between the interventions like contact restrictions and mandatory face masks).
 Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs [@perkovic2017complete].
@@
 It is possible that some variables show saturation effects or opposite effects for low, medium, or high values.
-This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
 Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
-By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).
+By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).
 
@@
 
-In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
-This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.
+In our analysis, the adverse effects of mobility in retail/recreation and workplaces and the favorable effect of mobility in grocery/pharmacy and residential areas indicate that interventions like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
+This is due to retail/recreational and workplace areas encompassing mostly places of (social) gatherings, while if people are doing more of their essential shopping at supermarkets and stay at home with less contact to other people, they are less likely to come in contact with infected individuals.
 
-The causal effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret.
+The effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret.
 We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.
@@
 
+Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions.
+To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
+
 ## Conclusions
@@
 To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available).
-Causal reasoning with a DAG allows us to estimate the causal effects more reliably.
+Causal reasoning with a DAG allows us to estimate the possible causal effects more reliably.
 
-Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
+Our findings suggest that the infection-driving effects of mobility, awareness, and weather (and to some extent socio-demographic factors) need to be taken into account when deciding for mitigation and suppression interventions, depending on the recent and future COVID-19 pandemic development.

### 2021-03-20 manuscript/v4_medrxiv_revision/manuscript.Rmd
finish manuscript for revision

@@
 ---
-title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases
+title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers
 author:
@@
 Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
-Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
+Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy) or within residential areas, and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects leading to an increase in new COVID-19 cases for public mobility in retail and recreational areas or workplaces, awareness measured by searches for “corona” in Google, higher rainfall, and some socio-demographic factors.
+Non-pharmaceutical interventions were found to be effective in reducing case numbers.
 This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

### 2021-03-20 manuscript/v4_medrxiv_revision/manuscript.tex
cv for lag (5), update data accordingly, interventions splitted again (ridge should take care of multicollinearity)
finish manuscript for revision

@@
             pdfauthor={},
-            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases},
+            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers},
             colorlinks=false,
@@
 
-  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases}
+  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers}
     \author[Zi]{Edgar Steiger}
@@
 We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
-Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
+Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy) or within residential areas, and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects leading to an increase in new COVID-19 cases for public mobility in retail and recreational areas or workplaces, awareness measured by searches for ``corona'' in Google, higher rainfall, and some socio-demographic factors.
+Non-pharmaceutical interventions were found to be effective in reducing case numbers.
 This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
@@
 We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020.
-Our observation period for the outcome consisted of all dates from 27 February 2020 to 8 July 2020 (\(T=133\)), since we used a lag of 12 days for all confounders.
+Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5 days for all confounders.
 We did not exclude any states or districts (\(K=401\)).
-We analyzed the daily reported number of new cases as outcome (\(K\cdot T=53{\ }333\) observations).
+We analyzed the daily reported number of new cases as outcome (\(K\cdot T=56{\ }140\) observations).
 The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}).
@@
 
-For different official and political measures on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
-These variables were highly correlated with each other and other variables (especially mobility).
-To reduce multicollinearity, we aggregated the different official/political measures by summing them into one variable ``interventions'' which describes the current severity of interventions on a given day in a given district.
+For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
 
@@
 \addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
+\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
+\addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Weather}}\\
@@
 \addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Policies}}\\
+\multicolumn{6}{l}{\textbf{Interventions}}\\
 \hspace{1em}Ban of mass gatherings & daily & national & binary & - & -\\
@@
 \hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
-\hspace{1em}Interventions & daily & district & numeric & sum of the above & -\\
 \addlinespace[0.3em]
@@
 \addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
-\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
-\addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Case numbers}}\\
@@
 After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale.
-Additionally for mobility, we re-transformed the effects of the principal components to the original mobility variables.
-Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.
+Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 
-To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlation between variables on their respective regression coefficients.
-The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the \(401\) districts.
-
 We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial.
-We used this parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \citep{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter.
+The proper lag between exposures and outcome was found by 10-fold cross-validation on different lags between 1 and 20 days.
+Model diagnostics on the final full model did not show severe problems with model assumptions (linearity, distribution of residuals, independence of observations).
+Analysis of variance inflation factors revealed some problems with multicollinearity.
+To reduce the effects of multicollinearity, first we transformed the highly correlated mobility variables by PCA as described above.
+Second, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients and alleviates the effect of correlation between variables on their respective regression coefficients.
+Furthermore, regularized regression allows for better fits on unseen data, thus preventing overfitting the data, too.
+The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from random subsets of the \(401\) districts.
+We used this hyper-parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \citep{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter.
 Afterwards, we calculated the effects of separate exposures on the outcome.
@@
 \midrule
-n & 53333\\
+n & 56140\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & -27.70 (24.65)\\
-\hspace{1em}grocery and pharmacy & -4.36 (22.96)\\
-\hspace{1em}parks & 44.77 (57.40)\\
-\hspace{1em}workplaces & -23.45 (20.52)\\
-\hspace{1em}residential & 8.36 (6.56)\\
-\hspace{1em}transit stations & -30.25 (21.31)\\
+\hspace{1em}retail and recreation & -26.62 (24.60)\\
+\hspace{1em}grocery and pharmacy & -3.94 (22.77)\\
+\hspace{1em}parks & 47.26 (58.20)\\
+\hspace{1em}workplaces & -22.96 (20.35)\\
+\hspace{1em}residential & 8.13 (6.49)\\
+\hspace{1em}transit stations & -29.58 (21.11)\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & 26.94 (18.23)\\
+\hspace{1em}COVID-19 burden & 3.50 (10.28)\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & 1.83 (3.95)\\
-\hspace{1em}Temperature & 10.50 (5.15)\\
-\hspace{1em}Humidity & 67.62 (13.23)\\
-\hspace{1em}Wind & 3.64 (1.68)\\
+\hspace{1em}Rainfall & 1.89 (4.01)\\
+\hspace{1em}Temperature & 10.90 (5.33)\\
+\hspace{1em}Humidity & 67.81 (13.03)\\
+\hspace{1em}Wind & 3.63 (1.66)\\
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Policies}}\\
-\hspace{1em}Interventions & 2.56 (1.33)\\
+\multicolumn{2}{l}{\textbf{Interventions}}\\
+\hspace{1em}Ban of mass gatherings & 0.83 (0.38)\\
+\hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
+\hspace{1em}Contact restrictions & 0.74 (0.44)\\
+\hspace{1em}Mandatory face masks & 0.49 (0.50)\\
 \addlinespace[0.3em]
@@
 \hspace{1em}Age (pop. younger 18) & 16.17 (1.25)\\
-\hspace{1em}Gender & 0.51 (0.01)\\
+\hspace{1em}Gender & 50.59 (0.64)\\
 \hspace{1em}Population density & 533.75 (701.84)\\
@@
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 27.64 (18.43)\\
-\hspace{1em}COVID-19 burden & 3.63 (10.49)\\
-\addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
-\hspace{1em}Reported new cases COVID-19 & 3.71 (10.52)\\
-\hspace{1em}Active cases & 51.33 (123.47)\\
+\hspace{1em}Reported new cases COVID-19 & 3.53 (10.29)\\
+\hspace{1em}Active cases & 48.76 (120.86)\\
 \bottomrule
@@
 
-In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) -0.06, awareness (``Searches corona'') \(r_{A,C}=\) -0.26, and temperature \(r_{A,B}=\) 0.77.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,B}=\) 0.8.
 
@@
 \multicolumn{2}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & 1.0010\\
-\hspace{1em}grocery and pharmacy & 0.9988\\
-\hspace{1em}parks & 0.9997\\
-\hspace{1em}transit stations & 1.0024\\
-\hspace{1em}workplaces & 1.0025\\
-\hspace{1em}residential & 0.9914\\
+\hspace{1em}retail and recreation & 1.0011\\
+\hspace{1em}grocery and pharmacy & 0.9979\\
+\hspace{1em}parks & 0.9996\\
+\hspace{1em}transit stations & 1.0025\\
+\hspace{1em}workplaces & 1.0032\\
+\hspace{1em}residential & 0.9906\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 1.0007\\
-\hspace{1em}COVID-19 burden & 0.9922\\
+\hspace{1em}Searches corona & 1.0087\\
+\hspace{1em}COVID-19 burden & 0.9981\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Weather}}\\
-\hspace{1em}Temperature & 0.9959\\
-\hspace{1em}Rainfall & 1.0100\\
-\hspace{1em}Humidity & 1.0042\\
-\hspace{1em}Wind & 1.0164\\
+\hspace{1em}Temperature & 0.9905\\
+\hspace{1em}Rainfall & 1.0117\\
+\hspace{1em}Humidity & 1.0056\\
+\hspace{1em}Wind & 1.0328\\
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Policies}}\\
-\hspace{1em}Interventions & 0.8037\\
+\multicolumn{2}{l}{\textbf{Interventions}}\\
+\hspace{1em}Interventions (ban of mass gatherings) & 0.9733\\
+\hspace{1em}Interventions (school and kindergarten closures) & 0.9315\\
+\hspace{1em}Interventions (contact restrictions) & 0.8400\\
+\hspace{1em}Interventions (mandatory face masks) & 0.9109\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Demographic}}\\
-\hspace{1em}Age (pop. 65 and older & 0.9976\\
-\hspace{1em}Age (pop. younger 18 & 1.0097\\
-\hspace{1em}Foreign citizens & 1.0034\\
-\hspace{1em}Foreign citizens (refugees & 1.0010\\
-\hspace{1em}Gender & 0.5232\\
-\hspace{1em}Nursing homes & 1.0007\\
+\hspace{1em}Age (pop. 65 and older) & 0.9952\\
+\hspace{1em}Age (pop. younger 18) & 1.0116\\
+\hspace{1em}Foreign citizens & 1.0047\\
+\hspace{1em}Foreign citizens (refugees) & 0.9987\\
+\hspace{1em}Gender & 0.9929\\
+\hspace{1em}Nursing homes & 1.0011\\
 \hspace{1em}Population density & 1.0000\\
-\hspace{1em}Socio-economic status & 0.9993\\
+\hspace{1em}Socio-economic status & 0.9983\\
 \bottomrule
@@
 
-Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.1 percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.12 percent.
-Mobility on workplaces showed a small effect of 0.25 increase in case numbers for every 1 percent point increase in mobility.
-Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
+Within our framework, we saw very different effects for mobility variables.
+For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent.
+Similarly, Mmbility on workplaces showed an effect of 0.32 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.25 increase in case numbers for every 1 percent point increase.
+Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.
+An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.21 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.04 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.94 percent.
 Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
-Thus, we expect an increase of daily cases by approximately 3 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 2 percent.
+Thus, we expect an increase of daily cases by approximately 7.5 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 1.8 percent.
 
@@
 Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases.
-An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.1 percent.
-For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 1 percent more infections for this district after 8 days.
-\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.8 percent decrease in newly reported case numbers.
-The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 14.6 percent.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.87 percent.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 9 percent more infections for this district after 5 days.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.19 percent decrease in newly reported case numbers.
+The corresponding plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.8 percent.
 
-Within our model, we observed a causal effect of temperature and all other weather variables.
-Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.4 percent.
-On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1 percent.
-We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
-In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 8.5 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
-For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 10.5 percent compared to a day with no precipitation.
+Within our model, we observed effects of temperature and all other weather variables.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.17 percent.
+We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21.1 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.3 percent compared to a day with no precipitation.
 
-The variable ``interventions'' showed \ldots{}
+The different intervention variables showed the strongest effects in our analysis, see Table \ref{tab:causaleffects}.
+While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 6.8 percent and mandatory face masks reduced this by another 8.9 percent.
+The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.
 
-The effects of the different socio-demographic factors \ldots{}
+The effects of the different socio-demographic factors are quite small in comparison to the effects described above.
+We see an increasing effect on case numbers by additional nursing homes between districts.
+Districts with a younger population, more foreign citizens, higher population density and a lower average social-economic status showed higher case numbers, too.
 
-In all cases we opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC.
-We found that there were no valid adjustment sets for the unidentifiable variables turnout and right-wing populist party votes.
+For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC.
+We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.
 
-We decided for the lag of 12 days based on the lowest AIC value compared to other lags between 1 and 20 days on the full model.
-Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
+We decided for a lag of 5 days based on cross-validation. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
 
@@
 \multicolumn{15}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
-\hspace{1em}grocery and pharmacy &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
-\hspace{1em}parks &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
-\hspace{1em}workplaces &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
-\hspace{1em}residential &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
-\hspace{1em}transit stations &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
+\hspace{1em}Mobility &  &  &  &  &  & x &  &  &  &  &  &  &  & \\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & x &  &  & x & x & x & x &  &  &  &  & x &  & \\
+\hspace{1em}COVID-19 burden & x & x &  & x & x & x & x & x & x & x & x & x & x & x\\
 \addlinespace[0.3em]
 \multicolumn{15}{l}{\textbf{Weather}}\\
+\hspace{1em}Temperature & x & x & x &  &  & x &  & x & x & x & x &  & x & x\\
 \hspace{1em}Rainfall & x & x & x & x &  & x & x & x & x & x & x &  & x & x\\
-\hspace{1em}Temperature & x & x & x &  &  & x &  & x & x & x & x &  & x & x\\
 \hspace{1em}Humidity & x & x & x &  &  &  &  & x & x & x & x &  & x & x\\
@@
 \addlinespace[0.3em]
-\multicolumn{15}{l}{\textbf{Policies}}\\
+\multicolumn{15}{l}{\textbf{Interventions}}\\
 \hspace{1em}Interventions & x & x &  & x & x & x & x &  & x & x & x & x & x & x\\
@@
 \hspace{1em}Socio-economic status & x & x & x & x & x & x & x & x &  &  &  & x & x & \\
-\hspace{1em}Trust in society &  &  &  &  &  &  &  &  &  &  &  &  &  & \\
 \hspace{1em}Turnout &  &  & x &  &  &  &  & x &  &  &  &  &  & \\
@@
 \hspace{1em}Nursing homes & x & x & x & x & x & x & x & x &  &  &  &  & x & \\
-\addlinespace[0.3em]
-\multicolumn{15}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & x &  &  & x & x & x & x &  &  &  &  & x &  & \\
-\hspace{1em}COVID-19 burden & x & x &  & x & x & x & x & x & x & x & x & x & x & x\\
 \bottomrule
@@
 Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects \citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
-To reliably identify causal drivers, one must adjust for confounders.
+To reliably identify (causal) drivers, one must adjust for confounders.
 To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
 A causal approach is used in another current COVID-19 analysis \citep{gencoglu2020causal}.
-There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.
+There, however, they identify the causal relationships (reconstruct a DAG), while we estimated effects for a given hypothesized causal DAG.
 
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
+Finally, we found a positive effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
 
@@
 
-While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
+While use of a causal DAG is itself a strong tool to identify \emph{causal} effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
 We endorse our graph as a basis for discussion on residual confounding.
@@
 \citep{gencoglu2020causal}).
-As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks).
+As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the interventions like contact restrictions and mandatory face masks).
 Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}.
@@
 It is possible that some variables show saturation effects or opposite effects for low, medium, or high values.
-This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
 Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
-By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).
+By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).
 
@@
 
-In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
-This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.
+In our analysis, the adverse effects of mobility in retail/recreation and workplaces and the favorable effect of mobility in grocery/pharmacy and residential areas indicate that interventions like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
+This is due to retail/recreational and workplace areas encompassing mostly places of (social) gatherings, while if people are doing more of their essential shopping at supermarkets and stay at home with less contact to other people, they are less likely to come in contact with infected individuals.
 
-The causal effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret.
+The effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret.
 We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.
@@
 
+Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions.
+To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
+
 \hypertarget{conclusions}{%
@@
 To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available).
-Causal reasoning with a DAG allows us to estimate the causal effects more reliably.
+Causal reasoning with a DAG allows us to estimate the possible causal effects more reliably.
 
-Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
+Our findings suggest that the infection-driving effects of mobility, awareness, and weather (and to some extent socio-demographic factors) need to be taken into account when deciding for mitigation and suppression interventions, depending on the recent and future COVID-19 pandemic development.
 

### 2021-03-20 manuscript/v4_medrxiv_revision/results.Rmd
finish manuscript for revision

@@
 t1.df <- read.csv("tables/t_descriptives.csv") %>% rename("Variable"="X") %>%
-  mutate(Variable=recode(Variable,
+  mutate(Variable=dplyr::recode(Variable,
                          "Mobility (retail and recreation)"="retail and recreation",
@@
                                                           "Weather (temperature)"="Temperature",
-                                                          "Weather (humidity)"="Humidity")) %>%
+                                                          "Weather (humidity)"="Humidity",
+                                                          "Interventions (ban of mass gatherings)"="Ban of mass gatherings",
+                                                          "Interventions (school and kindergarten closures)"="School and kindergarten closures",
+                                                          "Interventions (contact restrictions)"="Contact restrictions",
+                                                          "Interventions (mandatory face masks)"="Mandatory face masks")) %>%
   mutate(Variable =  factor(Variable, ordered=TRUE, levels = c("n",
@@
                                                  "transit stations",
+                                                 "Searches corona",
+                                                 "COVID-19 burden", 
                                                  "Rainfall",
@@
                                                  "Wind",
-                                                 "Interventions",
+                                                 "Ban of mass gatherings",
+                                                 "School and kindergarten closures",
+                                                 "Contact restrictions",
+                                                 "Mandatory face masks",
                                                  "Age (pop. 65 and older)",
@@
                                                  "Nursing homes",
-                                                 "Searches corona",
-                                                 "COVID-19 burden", 
                                                  "Reported new cases COVID-19",
@@
 # Correlation of time and main vars
-mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"), "Rainfall", "Temperature", "Humidity", "Wind", `Searches corona`,contains("Mobility"), `COVID-19 burden`, "Interventions") %>% 
+mycorrs <- Modeldata %>% dplyr::select(date,contains("cases"), "Rainfall", "Temperature", "Humidity", "Wind", `Searches corona`,contains("Mobility"), `COVID-19 burden`) %>% 
   mutate(outcome=`Reported new cases COVID-19`/(`Active cases`+1),date=as.numeric(date)) %>% 
@@
   pack_rows("Mobility", 2, 7) %>%
-  pack_rows("Weather", 8, 11) %>%
-  pack_rows("Policies", 12, 12) %>%
-  pack_rows("Socio-demographic", 13, 22) %>%
-  pack_rows("Awareness", 23, 24) %>%
-  pack_rows("Case numbers (Outcome and offset)", 25, 26)
+  pack_rows("Awareness", 8, 9) %>%
+  pack_rows("Weather", 10, 13) %>%
+  pack_rows("Interventions", 14, 17) %>%
+  pack_rows("Socio-demographic", 18, 27) %>%
+  pack_rows("Case numbers (Outcome and offset)", 28, 29)
 ```
@@
   pack_rows("Weather", 9, 12) %>%
-  pack_rows("Policies", 13, 13) %>%
-  pack_rows("Demographic", 14, 21)
+  pack_rows("Interventions", 13, 16) %>%
+  pack_rows("Demographic", 17, 24)
 ```
@@
 
-Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 2)` percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 2)` percent.
-Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase in mobility.
-Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1).
+Within our framework, we saw very different effects for mobility variables.
+For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 2)` percent.
+Similarly, Mmbility on workplaces showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (transit stations)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase.
+Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.
+An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 2)` percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (parks)")%>%pull("effect estimate")), 2)` percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (residential)")%>%pull("effect estimate")), 2)` percent.
 Figure \@ref(fig:effectsrange) shows the effects of mobility on a range of possible values.
-Thus, we expect an increase of daily cases by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate"))^(0-mymeans["Mobility (retail and recreation)"])), 0)` percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate"))^(10-mymeans["Mobility (grocery and pharmacy)"])), 0)` percent.
+Thus, we expect an increase of daily cases by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate"))^(0-mymeans["Mobility (workplaces)"])), 1)` percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (residential)")%>%pull("effect estimate"))^(10-mymeans["Mobility (residential)"])), 1)` percent.
 
@@
 Awareness measured by Google searches for *corona* had a positive effect on the number of reported cases.
-An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate")), 1)` percent.
-For example, if a district shows 10 percent points more relative searches for *corona* than another one, we expect approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate"))^(10)), 0)` percent more infections for this district after 8 days.
-_COVID-19 burden_ (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a `r round(100*abs(1-causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate")), 1)` percent decrease in newly reported case numbers.
-The last plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(20-0)), 1)` percent.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate")), 2)` percent.
+For example, if a district shows 10 percent points more relative searches for *corona* than another one, we expect approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Searches corona")%>%pull("effect estimate"))^(10)), 1)` percent more infections for this district after 5 days.
+_COVID-19 burden_ (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a `r round(100*abs(1-causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate")), 2)` percent decrease in newly reported case numbers.
+The corresponding plot in Figure \@ref(fig:effectsrange) visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by `r round(100*abs(1-(causaleffects%>%filter(cause=="COVID-19 burden")%>%pull("effect estimate"))^(20-0)), 1)` percent.
 
-Within our model, we observed a causal effect of temperature and all other weather variables.
-Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 1)` percent.
-On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent.
-We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases).
+Within our model, we observed effects of temperature and all other weather variables.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 2)` percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 2)` percent.
+We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
 In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-20)), 1)` percent at a daily average temperature of $0^{\circ}C$ compared to a day with $20^{\circ}C$.
@@
 
-The variable "interventions" showed ...
+The different intervention variables showed the strongest effects in our analysis, see Table \@ref(tab:causaleffects).
+While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (ban of mass gatherings)")%>%pull("effect estimate")), 1)` percent, the closure of schools/kindergartens reduced infections by an additional `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (school and kindergarten closures)")%>%pull("effect estimate")), 1)` percent and mandatory face masks reduced this by another `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (mandatory face masks)")%>%pull("effect estimate")), 1)` percent.
+The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (contact restrictions)")%>%pull("effect estimate")), 1)`.
 
-The effects of the different socio-demographic factors ...
+The effects of the different socio-demographic factors are quite small in comparison to the effects described above.
+We see an increasing effect on case numbers by additional nursing homes between districts.
+Districts with a younger population, more foreign citizens, higher population density and a lower average social-economic status showed higher case numbers, too.
 
-In all cases we opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", where the minimal adjustment set had a lower AIC.
-We found that there were no valid adjustment sets for the unidentifiable variables turnout and right-wing populist party votes.
+For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", where the minimal adjustment set had a lower AIC.
+We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.
 
-We decided for the lag of 12 days based on the lowest AIC value compared to other lags between 1 and 20 days on the full model.
-Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
+We decided for a lag of 5 days based on cross-validation. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
 
@@
 library(kableExtra)
-myvariables <- readr::read_delim("tables/t_variables.csv", delim=";")[c(1:13, 18:23, 25:29), ] %>% dplyr::select(variable) %>%
-  mutate(Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ",
+myvariables <- tibble(variable=c("Weekday", "Holiday (report)", "Holiday (exposure)",
+                                 "Mobility", "Searches corona", "COVID-19 burden", 
+                                 "Temperature", "Rainfall", "Humidity", "Wind",
+                                 "Interventions",
+                                 "Age", "Gender", "Population density", "Foreign citizens",
+                                 "Socio-economic status", "Turnout", "Right-wing populist party votes",
+                                 "Nursing homes"),
+                      Mobility=" ", `Searches corona`=" ", `COVID-19 burden`=" ", Temperature=" ", Rainfall=" ", Humidity=" ", Wind=" ",
          Interventions= " ", Age=" ", `Foreign citizens`= " ", Gender=" ", `Nursing homes`=" ", `Population density`=" ", `Socio-economic status`=" ")
@@
   row_spec(0, angle = 90) %>%
-  pack_rows("Mobility", 4, 9) %>%
-  pack_rows("Policies", 14, 14) %>%
-  pack_rows("Socio-demographic", 15, 22) %>%
-  pack_rows("Awareness", 23, 24) %>%
-  pack_rows("Weather", 10, 13) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
+  pack_rows("Mobility", 4, 4) %>%
+  pack_rows("Awareness", 5, 6) %>%
+  pack_rows("Weather", 7, 10) %>%
+  pack_rows("Interventions", 11, 11) %>%
+  pack_rows("Socio-demographic", 12, 19) # %>%  add_header_above(c(" " = 2, "Awareness" = 2, "Weather" = 4))
 ```

### 2021-03-20 manuscript/v4_medrxiv_revision/figures/f_exposure_vs_effect.R

@@ -11,9 +11,12 @@ varnames <- c("Mobility (retail and recreation)",
               "Mobility (transit stations)",
               "Mobility (workplaces)",
               "Mobility (residential)",
-              "Temperature", "Rainfall", "Humidity", "Wind",
               "Searches corona", "COVID-19 burden",
-              "Interventions",
+              "Temperature", "Rainfall", "Humidity", "Wind",
+              # "Interventions (ban of mass gatherings)",
+              # "Interventions (school and kindergarten closures)",
+              # "Interventions (contact restrictions)",
+              # "Interventions (mandatory face masks)",
               "Age (pop. 65 and older)", "Age (pop. younger 18)", 
               "Foreign citizens", "Foreign citizens (refugees)",
               "Gender", "Nursing homes",
@@ -30,18 +33,18 @@ vardata <- vardata %>%
 vardata$myrangelow <- ranges[1, varnames] # c(-100, -25, -0, -100, -50, -10, -10, 0, 0, 0, 0, 0)
 vardata$myrangehigh <- ranges[2, varnames] # c(+50, +25, +100, +50, +0, +20, 30, 20, 100, 15, 100, 100)
 vardata[1:6, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Mobility_RedOptAdjSet.csv"))$estimates
-vardata[11, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$estimates
-vardata[12, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_COVID-19 burden_RedOptAdjSet.csv"))$estimates
-vardata[13, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Interventions_RedOptAdjSet.csv"))$estimates
-vardata[14:15, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Age_RedOptAdjSet.csv"))$estimates
-vardata[16:17, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Foreign citizens_RedOptAdjSet.csv"))$estimates
-vardata[18, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Gender_RedOptAdjSet.csv"))$estimates
[…]

### 2021-03-20 manuscript/v4_medrxiv_revision/figures/f_timeseries_cases_and_lograte.R

@@ -1,11 +1,11 @@
 library("tidyverse")
 library("cowplot")
 
-figurepath <- "./manuscript/v1_medrxiv/figures/"
+figurepath <- "manuscript/v4_medrxiv_revision/figures/"
 
 Modeldata <- read_csv("data/Modeldata_raw.csv")
 Sys.setlocale("LC_TIME", "C")
-mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
+mybreaks <- as.Date("2020-02-20")+seq(0, 139, length.out = 5)
 
 mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
 
@@ -31,7 +31,7 @@ logrates <- Modeldata %>% dplyr::select(date, `Reported new cases COVID-19`, `Ac
   theme(axis.title.x = element_blank()) +
   theme_cowplot(font_size = 8)
 
-plot_grid(cases, logrates, labels = "AUTO",label_size = 12)
+plot_grid(cases, logrates, labels = "AUTO", label_size = 12)
 
 ggsave(filename = paste0(figurepath, "f_timeseries_cases_and_lograte.eps"),
        dpi=600, width=15,height = 7, unit="cm", device=cairo_ps, fallback_resolution = 600)

### 2021-03-20 manuscript/v4_medrxiv_revision/figures/f_timeseries_exposures.R

@@ -1,11 +1,11 @@
 library("tidyverse")
 library("cowplot")
 
-figurepath <- "./manuscript/v1_medrxiv/figures/"
+figurepath <- "manuscript/v4_medrxiv_revision/figures/"
 
 Modeldata <- read_csv("data/Modeldata_raw.csv")
 Sys.setlocale("LC_TIME", "C")
-mybreaks <- as.Date("2020-02-15")+seq(0, 144, length.out = 5)
+mybreaks <- as.Date("2020-02-20")+seq(0, 139, length.out = 5)
 
 mycolors <- c("#268bd2","#dc322f","#d33682","#859900")
 cases <- Modeldata %>% dplyr::select(date,contains("cases"), id) %>%
@@ -18,9 +18,9 @@ cases <- Modeldata %>% dplyr::select(date,contains("cases"), id) %>%
   theme(axis.title.x = element_blank()) +
   theme_cowplot(font_size = 8)
 
-temperature <- Modeldata %>% dplyr::select(date,id, contains("temperature")) %>%
-  ggplot(aes(x= date-8,
-             y= `Weather (temperature)`))+
+temperature <- Modeldata %>% dplyr::select(date,id, contains("Temperature")) %>%
+  ggplot(aes(x= date-5,
+             y= `Temperature`))+
   geom_point(aes(group=id), color="moccasin", stroke=0, alpha=0.2) +
   geom_smooth(color=mycolors[2],size=2,se = F) + 
   scale_x_date(breaks=mybreaks, limits=c(min(mybreaks), max(mybreaks)+2), date_labels = "%d %b") + labs(x="Date", y="Temperature") + 
@@ -29,7 +29,7 @@ temperature <- Modeldata %>% dplyr::select(date,id, contains("temperature")) %>%
   theme_cowplot(font_size = 8)
 
 awareness <- Modeldata %>% dplyr::select(date, id, `Searches corona`) %>%
-  ggplot(aes(x=date-8,
+  ggplot(aes(x=date-5,
              y= `Searches corona`)) +
   geom_point(aes(group=id), color="lightpink", stroke=0, alpha=0.2) +
   geom_smooth(color=mycolors[3],size=2,se = F) + 
@@ -40,7 +40,7 @@ awareness <- Modeldata %>% dplyr::select(date, id, `Searches corona`) %>%
   theme_cowplot(font_size = 8)
 
[…]

### 2021-03-20 manuscript/v2_medrxiv/manuscript.tex
fix manuscript v2?

@@
 \usepackage{graphicx}
+% We will generate all images so they have a width \maxwidth. This means
+% that they will get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+\else\Gin@nat@width\fi}
+\makeatother
+\let\Oldincludegraphics\includegraphics
+\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 \ifxetex
@@
 % Pandoc toggle for numbering sections (defaults to be off)
-
-% Pandoc citation processing
 
@@
 Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
-Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
+Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.
 
@@
 We endorse our graph as a basis for discussion on residual confounding.
-We did not try to construct the DAG from the available data (cf.
-\citep{gencoglu2020causal}).
+We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}).
 As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks).
@@
 We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
-We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
+We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
 There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models.
@@
 
-Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
+Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
 

### 2021-03-21 R/3_modelassumptions.R
update lambda with fixed folds, parallel cv (?)

@@ -9,6 +9,7 @@ library(ggthemes)
 library(pcalg)
 library(pscl)
 library(glmnet)
+library(doMC)
 source("R/z_auxiliary_functions_for_2.R")
 
 # Load Data
@@ -37,13 +38,21 @@ myglm_fixtheta <- glm(as.formula(myformula),
 myy <- modeldata$`Reported new cases COVID-19`
 myx <- as.matrix(modeldata %>% dplyr::select(-`Reported new cases COVID-19`))
 
-myfoldids <- (id_daycount %>% group_indices(id)) %% 10 + 1
+set.seed(3141)
+myids <- sample(unique(id_daycount$id))
+myrandomfoldids <- c(rep(1:9, each=40), rep(10, 41))
+myidmatch <- tibble(id=myids, randomfoldid=myrandomfoldids)
+myfoldids <- read_csv("data/Modeldata_raw.csv") %>%
+  dplyr::select(id, daycount, date, bl_id) %>%
+  dplyr::select(-date) %>%
+  left_join(myidmatch, by="id") %>% pull(randomfoldid)
 
-set.seed(2407)
+registerDoMC(cores=4)
 myglm_ridge_cv <- cv.glmnet(x=myx, y=myy, family = negative.binomial(mytheta),
                             offset=log(modeldata%>%pull(`Active cases`)+1),
                             standardize=FALSE,
-                            alpha=0, nfolds=10, foldid = myfoldids) # 
+                            alpha=0, nfolds=10, foldid = myfoldids,
+                            parallel = TRUE) # 
 
 mylambda <- myglm_ridge_cv$lambda.1se
 

### 2021-03-21 manuscript/v4_medrxiv_revision/manuscript.tex
update with lambda fixed folds

@@
 \hspace{1em}retail and recreation & 1.0011\\
-\hspace{1em}grocery and pharmacy & 0.9979\\
-\hspace{1em}parks & 0.9996\\
-\hspace{1em}transit stations & 1.0025\\
-\hspace{1em}workplaces & 1.0032\\
-\hspace{1em}residential & 0.9906\\
+\hspace{1em}grocery and pharmacy & 0.9977\\
+\hspace{1em}parks & 0.9997\\
+\hspace{1em}transit stations & 1.0026\\
+\hspace{1em}workplaces & 1.0033\\
+\hspace{1em}residential & 0.9903\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 1.0087\\
-\hspace{1em}COVID-19 burden & 0.9981\\
+\hspace{1em}Searches corona & 1.0089\\
+\hspace{1em}COVID-19 burden & 0.9980\\
 \addlinespace[0.3em]
@@
 \hspace{1em}Temperature & 0.9905\\
-\hspace{1em}Rainfall & 1.0117\\
-\hspace{1em}Humidity & 1.0056\\
-\hspace{1em}Wind & 1.0328\\
+\hspace{1em}Rainfall & 1.0121\\
+\hspace{1em}Humidity & 1.0057\\
+\hspace{1em}Wind & 1.0329\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Interventions}}\\
-\hspace{1em}Interventions (ban of mass gatherings) & 0.9733\\
-\hspace{1em}Interventions (school and kindergarten closures) & 0.9315\\
-\hspace{1em}Interventions (contact restrictions) & 0.8400\\
-\hspace{1em}Interventions (mandatory face masks) & 0.9109\\
+\hspace{1em}Interventions (ban of mass gatherings) & 0.9729\\
+\hspace{1em}Interventions (school and kindergarten closures) & 0.9277\\
+\hspace{1em}Interventions (contact restrictions) & 0.8314\\
+\hspace{1em}Interventions (mandatory face masks) & 0.9064\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Demographic}}\\
-\hspace{1em}Age (pop. 65 and older) & 0.9952\\
-\hspace{1em}Age (pop. younger 18) & 1.0116\\
-\hspace{1em}Foreign citizens & 1.0047\\
-\hspace{1em}Foreign citizens (refugees) & 0.9987\\
-\hspace{1em}Gender & 0.9929\\
+\hspace{1em}Age (pop. 65 and older) & 0.9953\\
+\hspace{1em}Age (pop. younger 18) & 1.0120\\
+\hspace{1em}Foreign citizens & 1.0048\\
+\hspace{1em}Foreign citizens (refugees) & 0.9985\\
+\hspace{1em}Gender & 0.9925\\
 \hspace{1em}Nursing homes & 1.0011\\
 \hspace{1em}Population density & 1.0000\\
-\hspace{1em}Socio-economic status & 0.9983\\
+\hspace{1em}Socio-economic status & 0.9982\\
 \bottomrule
@@
 For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent.
-Similarly, Mmbility on workplaces showed an effect of 0.32 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.25 increase in case numbers for every 1 percent point increase.
+Similarly, Mmbility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase.
 Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.
-An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.21 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.04 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.94 percent.
+An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.23 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.03 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.97 percent.
 Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
-Thus, we expect an increase of daily cases by approximately 7.5 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
+Thus, we expect an increase of daily cases by approximately 7.8 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
 On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 1.8 percent.
@@
 Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases.
-An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.87 percent.
-For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 9 percent more infections for this district after 5 days.
-\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.19 percent decrease in newly reported case numbers.
-The corresponding plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.8 percent.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.89 percent.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 9.3 percent more infections for this district after 5 days.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.2 percent decrease in newly reported case numbers.
+The corresponding plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.9 percent.
 
@@
 Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent.
-On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.17 percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent.
 We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
-In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21.1 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
-For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.3 percent compared to a day with no precipitation.
+In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.
 
 The different intervention variables showed the strongest effects in our analysis, see Table \ref{tab:causaleffects}.
-While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 6.8 percent and mandatory face masks reduced this by another 8.9 percent.
-The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.
+While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 7.2 percent and mandatory face masks reduced this by another 9.4 percent.
+The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.
 

### 2021-03-21 manuscript/v4_medrxiv_revision/discussion.Rmd
add reference causal collinearity, and discussion points

@@
 This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+Interaction effects of variables and confounding effects or mediating variables are explicitly taken care of by deriving the valid adjustment sets for a given exposure based on the causal DAG.
 Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
 By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).
+While multicollinearity of variables poses less of a problem for a proper causal graph analysis [@schisterman2017collinearitycausal], we addressed the problem of multicollinearity in our predictors by two approaches: principal component analysis for the highly collinear mobility variables as well as a regularized regression approach (ridge regression).
+The latter (in conjunction with cross-validation) also reduced the problem of overfitting.
 

### 2021-03-21 manuscript/v4_medrxiv_revision/manuscript.tex
update rmd tex

@@
 This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+Interaction effects of variables and confounding effects or mediating variables are explicitly taken care of by deriving the valid adjustment sets for a given exposure based on the causal DAG.
 Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
 By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).
+While multicollinearity of variables poses less of a problem for a proper causal graph analysis \citep{schisterman2017collinearitycausal}, we addressed the problem of multicollinearity in our predictors by two approaches: principal component analysis for the highly collinear mobility variables as well as a regularized regression approach (ridge regression).
+The latter (in conjunction with cross-validation) also reduced the problem of overfitting.
 

### 2021-03-21 manuscript/v4_medrxiv_revision/manuscript.tex
fix order figure timeseries in correlation text

@@
 
-In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,D}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,B}=\) 0.8.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Figure \ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,B}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,D}=\) 0.8.
 

### 2021-03-21 manuscript/v4_medrxiv_revision/results.Rmd
fix order figure timeseries in correlation text

@@
 In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.
-Figure \@ref(fig:timeseries)); with correlations between temporal progression and mobility in retail and recreation $r_{A,D}=$ `r round(mycorrs["date", "Mobility (retail and recreation)"], 2)`, awareness ("Searches corona") $r_{A,C}=$ `r round(mycorrs["date", "Searches corona"], 2)`, and temperature $r_{A,B}=$ `r round(mycorrs["date", "Temperature"], 2)`.
+Figure \@ref(fig:timeseries)); with correlations between temporal progression and mobility in retail and recreation $r_{A,B}=$ `r round(mycorrs["date", "Mobility (retail and recreation)"], 2)`, awareness ("Searches corona") $r_{A,C}=$ `r round(mycorrs["date", "Searches corona"], 2)`, and temperature $r_{A,D}=$ `r round(mycorrs["date", "Temperature"], 2)`.
 

### 2021-03-21 manuscript/v5_submission_revision/manuscript_tex4submission.tex
update manuscript for submission

@@
 {\Large
-\textbf\newline{Causal analysis of COVID-19 observational data in German districts
-reveals effects of mobility, awareness, and temperature} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \section*{Abstract}
-Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection.
-Correcting for possible confounders, we estimated their causal effects on reported case numbers.
-To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19.
+Several determinants are suspected to be causal drivers for new cases of COVID-19 infection.
+Correcting for possible confounders, we estimated the effects of the most prominent determining factors on reported case numbers.
+To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the determinants on new reported cases of COVID-19.
 Based on this, we computed valid adjustment sets of the possible confounding factors.
 We collected data for Germany from publicly available sources (e.g.~Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression.
-Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
-Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for ``corona'' in Google, and higher rainfall, leading to an increase in new COVID-19 cases.
-This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
+Our analysis revealed favorable effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy) or within residential areas, and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases.
+Conversely, we saw adverse effects leading to an increase in new COVID-19 cases for public mobility in retail and recreational areas or workplaces, awareness measured by searches for ``corona'' in Google, higher rainfall, and some socio-demographic factors.
+Non-pharmaceutical interventions were found to be effective in reducing case numbers.
+This comprehensive causal graph analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.
 
@@
 
-Our objective is to provide estimates of the causal effects of the main
-drivers of the pandemic with reduced bias. We conducted a scoping review
-of the available studies regarding signaling pathways and determinants
-of the spread of SARS-CoV-2 infections and the reported new COVID-19
-cases. Then we integrated the current findings into a directed acyclic
-graph for the progress of the pandemic at the regional level. Using the
-resulting model and the do-calculus we found identifiable effects
-without blocked causal paths whose effects can be analyzed with
-observational data \cite{Pearl_2014}. We used regional time series data
-of all 401 German districts from various publicly available
-sources to analyze these questions on a regional level. Germany is a
-good choice in this regard, because it has ample data on contributing
-factors on the regional level and has had high testing and treatment
-capacities from early on in the pandemic.
+Our objective is to provide valid estimates of the effects of the main drivers of the pandemic with a causal graph approach.
+We conducted a scoping review of the available studies regarding signaling pathways and determinants of the spread of SARS-CoV-2 infections and the reported new COVID-19 cases.
+Then we integrated the current findings into a directed acyclic graph for the progress of the pandemic at the regional level.
+Using the resulting model and the do-calculus we found identifiable effects without blocked causal paths whose effects can be analyzed with observational data \cite{Pearl_2014}.
+We used regional time series data of all German districts (\(401\)) from various publicly available sources to analyze these questions on a regional level.
+Germany is a good choice in this regard, because it has ample data on contributing factors on the regional level and has had high testing and treatment capacities from early on in the pandemic.
 
@@
 and German holidays. We included some unobserved variables in our DAG
-(e.g. ``Herd immunity''), too. Please note that ``Exposure to
+(e.g.~``Herd immunity''), too. Please note that ``Exposure to
 SARS-CoV-2'' is itself an unobserved variable: German case numbers are
@@
 \emph{Exposure to the virus} should not be confused with the formal
-\emph{exposure variables} of the DAG (mobility, awareness, temperature).
+\emph{exposure variables} of the DAG.
 
@@
 2020. Our observation period for the outcome consisted of all dates from
-23 February 2020 to 8 July 2020 (\(T=137\)), since we used a lag of 8
+20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5
 days for all confounders. We did not exclude any states or districts
 (\(K=401\)). We analyzed the daily reported number of new cases as
-outcome (\(K\cdot T=54{\ }937\) observations). The set of possible
+outcome (\(K\cdot T=56{\ }140\) observations). The set of possible
 predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
 modelling and data limitations, some of the predictors were unobserved
-or were modelled as a construct consisting of several variables. For our
-causal analysis, we computed adjustment sets in three different
-scenarios for separate exposures within the DAG: i) mobility of
-population, ii) awareness of COVID-19 (i.e.~Google searches for
-``corona''), iii) weather (i.e.~temperature).
+or were modelled as a construct consisting of several variables.
+For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
 
@@
 relative change of daily time spent at residential areas.
+The six mobility variables showed high correlations among each other and with other variables.
+To reduce multicollinearity, we transformed them by principal component analysis (PCA) into six uncorrelated principal components which were used in place of the original variables.
 
@@
 
-For different official and political measures we used one-hot encoded
-daily variables, i.e.~ban of mass gatherings, school and kindergarten
-closures and their gradual reopening, contact restrictions, and
-mandatory face masks for shopping and public transport.
+For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
 
@@
 \addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \cite{google_trends}\\
+\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \cite{casenumbers_rki}\\
+\addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Weather}}\\
@@
 \multicolumn{6}{l}{\textbf{Socio-demographic}}\\
-\hspace{1em}Age & constant & district & numeric & 2 variables: share of population $>=65$ years \& $<18$ years & INKAR \cite{inkar}\\
+\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \cite{inkar}\\
 \hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \cite{inkar}\\
@@
 \hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \cite{inkar}\\
-\addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \cite{google_trends}\\
-\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \cite{casenumbers_rki}\\
 \addlinespace[0.3em]
@@
 
-All variables but the outcome ``Reported new cases of COVID-19'' and the
-offset ``Active cases'' were centered for numerical stability. We did
-not scale variables to unit variance to maintain interpretability of
-effects on the original scale of variables. Additionally, we lagged the
-effect of all variables (but outcome, offset, and the non-dynamic
-socio-demographic variables) by 8 days (see Section Results) which
-means that we assumed that their effects on the outcome will be visible
-after 8 days.
+All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is.
+After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale.
+Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section Results) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 hypothesized causal reasoning that leads to exposure to the SARS-CoV-2
-virus, onset of COVID-19, and finally reports of COVID-19 cases. Every
+virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every
 node \(v_i\) in the graph is the graphical representation of an observed
@@
 set'' \cite{henckel2019graphical} as the set of variables which is a
-valid adjustment set while having the lowest asymptotic variance in the
-resulting causal effect estimates.
+valid adjustment set while having the lowest AIC.
 
@@
 ``Reported new cases of COVID-19'', we computed the minimal and optimal
-adjustment sets. Since it was possible that these sets contained
-unobserved variables that needed to be left out of the regression model,
-we chose the valid set with the highest pseudo-\(R^2\) (see next
-section) to estimate the final total causal effect from exposure to
-outcome.
+adjustment sets.
+Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the lowest AIC (see next section) to estimate the final total causal effect from exposure to outcome.
 
@@
 
-When we analyzed different adjustment sets given by analysis of the
-causal DAG (i.e.~the minimal and optimal adjustment sets), we first
-checked if the set included unobserved variables. If this was the case
-for the optimal adjustment set, we discarded the unobserved variables
-from the set and checked if it was still a valid adjustment set
-(function \texttt{gac} in package \texttt{pcalg}
-\cite{perkovi2015complete}). If a minimal adjustment set contained
-unobserved variables, we discarded the whole set. We conducted a
-log-linear regression (function \texttt{glm} with
-\texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb}
-from the \texttt{MASS} package for the negative binomial regression
-\cite{mass}) for every remaining valid adjustment set as regressors and
-calculated a Pseudo-\(R^2\) given by \(1-V_m/V_0\), where \(V_m\) is the
-sum of squared prediction errors of the current model and \(V_0\) is the
-sum of squared prediction errors of the null model (intercept and offset
-only). That is, our Pseudo-\(R^2\) is \(1\) minus the fraction of
-variance unexplained. Finally, we decided for the model/adjustment set
-with the highest pseudo-\(R^2\). We report the exponentiated estimated
-coefficients along with 99 percent confidence intervals of the
-estimates.
+We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \cite{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial.
+The proper lag between exposures and outcome was found by 10-fold cross-validation on different lags between 1 and 20 days.
+Model diagnostics on the final full model did not show severe problems with model assumptions (linearity, distribution of residuals, independence of observations).
+Analysis of variance inflation factors revealed some problems with multicollinearity.
+To reduce the effects of multicollinearity, first we transformed the highly correlated mobility variables by PCA as described above.
+Second, we used a ridge regression approach \cite{ridge1970}, which is a regularization method that shrinks regression coefficients and alleviates the effect of correlation between variables on their respective regression coefficients.
+Furthermore, regularized regression allows for better fits on unseen data, thus preventing overfitting the data, too.
+The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from random subsets of the \(401\) districts.
+We used this hyper-parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \cite{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter.
+Afterwards, we calculated the effects of separate exposures on the outcome.
+For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets).
+Then, we first checked if the respective set included unobserved variables.
+If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \cite{perkovi2015complete}).
+If a minimal adjustment set contained unobserved variables, we discarded the whole set.
+If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph.
+We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors (that is, we applied ridge regression) and calculated the Akaike information criterion (AIC) for this model/set of regressors.
+Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC.
+We report the exponentiated estimated coefficients for the separate exposures on their original scale.
 
@@
 \midrule
-n & 54937\\
+n & 56140\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & -27.14 (24.59)\\
-\hspace{1em}grocery and pharmacy & -4.04 (23.00)\\
-\hspace{1em}parks & 46.33 (57.94)\\
-\hspace{1em}workplaces & -22.98 (20.56)\\
-\hspace{1em}residential & 8.20 (6.54)\\
-\hspace{1em}transit stations & -29.83 (21.22)\\
+\hspace{1em}retail and recreation & -26.62 (24.60)\\
+\hspace{1em}grocery and pharmacy & -3.94 (22.77)\\
+\hspace{1em}parks & 47.26 (58.20)\\
+\hspace{1em}workplaces & -22.96 (20.35)\\
+\hspace{1em}residential & 8.13 (6.49)\\
+\hspace{1em}transit stations & -29.58 (21.11)\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & 26.94 (18.23)\\
+\hspace{1em}COVID-19 burden & 3.50 (10.28)\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & 1.85 (3.98)\\
-\hspace{1em}Temperature & 10.74 (5.27)\\
-\hspace{1em}Humidity & 67.68 (13.10)\\
-\hspace{1em}Wind & 3.64 (1.67)\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Policies}}\\
-\hspace{1em}Ban of mass gatherings & 0.82 (0.38)\\
+\hspace{1em}Rainfall & 1.89 (4.01)\\
+\hspace{1em}Temperature & 10.90 (5.33)\\
+\hspace{1em}Humidity & 67.81 (13.03)\\
+\hspace{1em}Wind & 3.63 (1.66)\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Interventions}}\\
+\hspace{1em}Ban of mass gatherings & 0.83 (0.38)\\
 \hspace{1em}School and kindergarten closures & 0.54 (0.36)\\
 \hspace{1em}Contact restrictions & 0.74 (0.44)\\
-\hspace{1em}Mandatory face masks & 0.48 (0.50)\\
+\hspace{1em}Mandatory face masks & 0.49 (0.50)\\
 \addlinespace[0.3em]
@@
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 27.24 (18.31)\\
-\hspace{1em}COVID-19 burden & 3.55 (10.37)\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{Case numbers (outcome and offset)}}\\
-\hspace{1em}Reported new cases COVID-19 & 3.60 (10.39)\\
-\hspace{1em}Active cases & 49.83 (121.96)\\
+\multicolumn{2}{l}{\textbf{Case numbers (Outcome and offset)}}\\
+\hspace{1em}Reported new cases COVID-19 & 3.53 (10.29)\\
+\hspace{1em}Active cases & 48.76 (120.86)\\
 \bottomrule
@@
 
-In the observational period, the number of daily reported COVID-19 cases
-increased till the end of March/beginning of April and continually
-decreased afterwards till the beginning of June 2020 with a slight
-increase and decrease afterwards (Fig~\ref{fig:cases-log}A). On the
-other hand, the (log-)ratio of reported cases over active cases
-decreased steeply till the mid of April and increased steadily
-afterwards with a slight decrease close to the end of the observation
-period (Fig~\ref{fig:cases-log}B). Both figures examplify a
-considerable variation among the districts (light blue points are
-individual district's data).
+In the observational period, the number of daily reported COVID-19 cases increased till the end of March/beginning of April and continually decreased afterwards till the beginning of June 2020 with a slight increase and decrease afterwards (Fig~\ref{fig:cases-log}A).
+On the other hand, the (log-)ratio of reported cases over active cases decreased steeply till the mid of April and increased steadily afterwards with a slight decrease close to the end of the observation period (Fig~\ref{fig:cases-log}B).
+Both figures examplify a considerable variation among the districts (light blue points are individual district's data).
 
@@
 
-In Germany, we observed a rebound in mobility after the initial
-political measures, reductions in incident cases were associated with a
-diminishing public interest in COVID-19, and temperatures were overall
-increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between
-temporal progression and mobility in retail and recreation \(r_{A,D}=\)
--0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.29, and
-temperature \(r_{A,B}=\) 0.79.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,B}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,D}=\) 0.8.
 
@@
 
-We list the results of our causal analysis for the effects of our
-variables in Table~\ref{tab:causaleffects}. The estimates are
-multiplicative rates of increase/decrease for a one unit increase of the
-respective variable: Values above 1 lead to an increase, below 1 to a
-decrease of the infection rate. To put these estimates into perspective,
-Fig~\ref{fig:effectsrange} shows the relative causal effect of the
-different exposure variables on the number of reported COVID-19 cases on
-a range of sensible values of the exposure variables (95 percent
-quantiles of data points).
+We list the results of our causal analysis for the effects of different exposure variables in Table~\ref{tab:causaleffects}.
+The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate.
+To put these estimates into perspective, Fig~\ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).
 
@@
 \caption{
-{\bf Causal effect estimates with 99 percent confidence bands.}}
+{\bf Effect estimates from causal graph analysis.}}
 %\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{lrrr}
+\begin{tabular}[t]{lr}
 \toprule
-cause & effect estimate & CI (lower) & CI (upper)\\
+cause & effect estimate\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation & 1.0076 & 1.0056 & 1.0097\\
-\hspace{1em}grocery and pharmacy & 0.9951 & 0.9938 & 0.9965\\
-\hspace{1em}parks & 0.9998 & 0.9994 & 1.0003\\
-\hspace{1em}transit stations & 0.9975 & 0.9946 & 1.0004\\
-\hspace{1em}workplaces & 1.0031 & 1.0002 & 1.0059\\
-\hspace{1em}residential & 1.0080 & 0.9987 & 1.0174\\
-\addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
-\hspace{1em}COVID-19 burden & 0.9942 & 0.9932 & 0.9951\\
-\addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Weather}}\\
-\hspace{1em}Temperature & 0.9918 & 0.9879 & 0.9958\\
-\hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
-\hspace{1em}Humidity & 1.0022 & 1.0006 & 1.0039\\
-\hspace{1em}Wind & 0.9893 & 0.9802 & 0.9985\\
+\multicolumn{2}{l}{\textbf{Mobility}}\\
+\hspace{1em}retail and recreation & 1.0011\\
+\hspace{1em}grocery and pharmacy & 0.9977\\
+\hspace{1em}parks & 0.9997\\
+\hspace{1em}transit stations & 1.0026\\
+\hspace{1em}workplaces & 1.0033\\
+\hspace{1em}residential & 0.9903\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & 1.0089\\
+\hspace{1em}COVID-19 burden & 0.9980\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Weather}}\\
+\hspace{1em}Temperature & 0.9905\\
+\hspace{1em}Rainfall & 1.0121\\
+\hspace{1em}Humidity & 1.0057\\
+\hspace{1em}Wind & 1.0329\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Interventions}}\\
+\hspace{1em}Interventions (ban of mass gatherings) & 0.9729\\
+\hspace{1em}Interventions (school and kindergarten closures) & 0.9277\\
+\hspace{1em}Interventions (contact restrictions) & 0.8314\\
+\hspace{1em}Interventions (mandatory face masks) & 0.9064\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Demographic}}\\
+\hspace{1em}Age (pop. 65 and older) & 0.9953\\
+\hspace{1em}Age (pop. younger 18) & 1.0120\\
+\hspace{1em}Foreign citizens & 1.0048\\
+\hspace{1em}Foreign citizens (refugees) & 0.9985\\
+\hspace{1em}Gender & 0.9925\\
+\hspace{1em}Nursing homes & 1.0011\\
+\hspace{1em}Population density & 1.0000\\
+\hspace{1em}Socio-economic status & 0.9982\\
 \bottomrule
@@
 
-Within our framework, we saw significant effects for mobility in
-retail/recreational areas and essential shopping (grocery and pharmacy).
-For retail/recreation, an increase of 1 percent point mobility compared
-to the reference period (03 January to 06 February 2020) leads to an
-increase of the daily reported case number by about 0.8 percent.
-Contrarily, a corresponding increase of 1 percent point for the areas of
-grocery/pharmacy leads to a decrease in the reported case number by
-approximately 0.5 percent. Mobility on workplaces showed a small effect
-of 0.3 increase in case numbers for every 1 percent point increase in
-mobility. Other causal effects of mobility were insubstantial and not
-consistent in their direction (99 percent confidence intervals of
-estimates include 1). Fig~\ref{fig:effectsrange} shows the effects of
-mobility on a range of possible values. Thus, we expect an increase of
-daily cases by approximately 23 percent if mobility in retail/recreation
-reaches baseline levels of 0 percent difference to the reference period.
-On the other hand, an increase of mobility for grocery/pharmacy by 10
-percent points compared to the reference period leads to a reduction of
-the infection rate by approximately 7 percent.
+Within our framework, we saw very different effects for mobility variables.
+For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent.
+Similarly, Mmbility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase.
+Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.
+An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.23 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.03 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.97 percent.
+Fig~\ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
+Thus, we expect an increase of daily cases by approximately 7.8 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
+On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 1.8 percent.
 
 ``Awareness'' had two opposite effects on the outcome in our DAG.
-Awareness measured by Google searches for \emph{corona} had a positive
-effect on the number of reported cases. An one percent point increase of
-the state's Google searches (relative to other states and the
-observation period) leads to an increase of approximately 1 percent. For
-example, if a district shows 10 percent points more relative searches
-for \emph{corona} than another one, we expect approximately 11 percent
-more infections for this district after 8 days. \emph{COVID-19 burden}
-(reported number of cases on day of exposure) affected the outcome
-negatively, where every additional daily case in the district leads to a
-0.6 percent decrease in newly reported case numbers. The last plot in
-Fig~\ref{fig:effectsrange} visualizes this relationship: For a local
-outbreak with 20 daily cases as COVID-19 burden, we estimate as total
-causal effect a subsequent reduction of infection rate by 11 percent.
-
-Within our model, we observed a causal effect of temperature and all
-other weather variables. Every increase of 1 degree Celsius in
-temperature leads to a reduction of the daily reported case numbers by
-approximately 0.8 percent. On the other hand, we found an increasing
-effect of rainfall: One millimeter (=1 liter per square meter) more
-rainfall leads to an increase of reported case numbers by approximately
-1.1 percent. We observe effects for humidity and wind as well (higher
-humidity leading to more cases, stronger wind leading to less cases). In
-perspective (Fig~\ref{fig:effectsrange}), with temperature we expect
-an increase by approximately 17.8 percent at a daily average temperature
-of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately
-11.4 percent compared to a day with no precipitation.
-
-In all cases we opted to use the reduced optimal adjustment set over the
-minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly
-above 0.3), except for mobility, where the minimal adjustment set had a
-higher pseudo-\(R^2\). Notably, these sets always include most of our
-socio-demographic variables as confounders as well as the policy
-variables (cf.~Table~\ref{tab:adjsets}, with the exception being
-COVID-19 burden). We also decided for the lag of 8 days based on the
-highest pseudo-\(R^2\) values compared to other lags on the chosen
-adjustment sets. Similarly, negative binomial regression was chosen over
-Poisson regression, because the latter showed overdispersion and overall
-lower pseudo-\(R^2\) values.
+Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases.
+An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.89 percent.
+For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 9.3 percent more infections for this district after 5 days.
+\emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.2 percent decrease in newly reported case numbers.
+The corresponding plot in Fig~\ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.9 percent.
+
+Within our model, we observed effects of temperature and all other weather variables.
+Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent.
+On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent.
+We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
+In perspective (Fig~\ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.
+
+The different intervention variables showed the strongest effects in our analysis, see Table~\ref{tab:causaleffects}.
+While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 7.2 percent and mandatory face masks reduced this by another 9.4 percent.
+The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.
+
+The effects of the different socio-demographic factors are quite small in comparison to the effects described above.
+We see an increasing effect on case numbers by additional nursing homes between districts.
+Districts with a younger population, more foreign citizens, higher population density and a lower average social-economic status showed higher case numbers, too.
+
+For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC.
+We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.
+
+We decided for a lag of 5 days based on cross-validation. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
 
@@
 %\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{llllllll}
+\begin{tabular}[t]{lllllllllllllll}
 \toprule
-\rotatebox{90}{ } & \rotatebox{90}{Mobility} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Searches corona} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind}\\
+\rotatebox{90}{ } & \rotatebox{90}{Mobility} & \rotatebox{90}{Searches corona} & \rotatebox{90}{COVID-19 burden} & \rotatebox{90}{Temperature} & \rotatebox{90}{Rainfall} & \rotatebox{90}{Humidity} & \rotatebox{90}{Wind} & \rotatebox{90}{Interventions} & \rotatebox{90}{Age} & \rotatebox{90}{Foreign citizens} & \rotatebox{90}{Gender} & \rotatebox{90}{Nursing homes} & \rotatebox{90}{Population density} & \rotatebox{90}{Socio-economic status}\\
 \midrule
-Pseudo-$R^2$ & 0.332 & 0.151 & 0.298 & 0.298 & 0.286 & 0.306 & 0.286\\
-Weekday & x & x & x & x & x & x & x\\
-Holiday (report) & x & x & x & x & x & x & x\\
-Holiday (exposure) & x & x & x & x & x & x & x\\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Mobility}}\\
-\hspace{1em}retail and recreation &  &  &  &  &  &  & \\
-\hspace{1em}grocery and pharmacy &  &  &  &  &  &  & \\
-\hspace{1em}parks &  &  &  &  &  &  & \\
-\hspace{1em}workplaces &  &  &  &  &  &  & \\
-\hspace{1em}residential &  &  &  &  &  &  & \\
-\hspace{1em}transit stations &  &  &  &  &  &  & \\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Weather}}\\
-\hspace{1em}Rainfall & x & x & x & x &  & x & x\\
-\hspace{1em}Temperature & x & x & x &  &  & x & \\
-\hspace{1em}Humidity &  & x & x &  &  &  & \\
-\hspace{1em}Wind &  & x & x & x & x & x & \\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Policies}}\\
-\hspace{1em}Ban of mass gatherings & x &  & x & x & x & x & x\\
-\hspace{1em}School and kindergarten closures & x &  & x & x & x & x & x\\
-\hspace{1em}Contact restrictions & x &  & x & x & x & x & x\\
-\hspace{1em}Mandatory face masks &  &  & x & x & x & x & x\\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Socio-demographic}}\\
-\hspace{1em}Age & x & x & x & x & x & x & x\\
-\hspace{1em}Gender & x & x & x & x & x & x & x\\
-\hspace{1em}Population density & x & x & x & x & x & x & x\\
-\hspace{1em}Foreign citizens &  & x & x &  &  &  & \\
-\hspace{1em}Socio-economic status & x & x & x & x & x & x & x\\
-\hspace{1em}Turnout &  & x &  &  &  &  & \\
-\hspace{1em}Right-wing populist party votes &  & x & x & x & x & x & x\\
-\hspace{1em}Nursing homes &  & x & x & x & x & x & x\\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Awareness}}\\
-\hspace{1em}Searches corona & x &  &  & x & x & x & x\\
-\hspace{1em}COVID-19 burden & x &  & x & x & x & x & x\\
+Weekday & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
+Holiday (report) & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
+Holiday (exposure) & x & x & x & x & x & x & x & x & x & x & x & x & x & x\\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Mobility}}\\
+\hspace{1em}Mobility &  &  &  &  &  & x &  &  &  &  &  &  &  & \\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Awareness}}\\
+\hspace{1em}Searches corona & x &  &  & x & x & x & x &  &  &  &  & x &  & \\
+\hspace{1em}COVID-19 burden & x & x &  & x & x & x & x & x & x & x & x & x & x & x\\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Weather}}\\
+\hspace{1em}Temperature & x & x & x &  &  & x &  & x & x & x & x &  & x & x\\
+\hspace{1em}Rainfall & x & x & x & x &  & x & x & x & x & x & x &  & x & x\\
+\hspace{1em}Humidity & x & x & x &  &  &  &  & x & x & x & x &  & x & x\\
+\hspace{1em}Wind & x & x & x & x & x & x &  & x & x & x & x &  & x & x\\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Interventions}}\\
+\hspace{1em}Interventions & x & x &  & x & x & x & x &  & x & x & x & x & x & x\\
+\addlinespace[0.3em]
+\multicolumn{15}{l}{\textbf{Socio-demographic}}\\
+\hspace{1em}Age & x & x & x & x & x & x & x & x &  &  &  & x & x & x\\
+\hspace{1em}Gender & x & x & x & x & x & x & x & x & x & x &  & x & x & x\\
+\hspace{1em}Population density & x & x & x & x & x & x & x & x & x & x & x & x &  & x\\
+\hspace{1em}Foreign citizens &  & x & x &  &  &  &  & x & x &  & x &  & x & x\\
+\hspace{1em}Socio-economic status & x & x & x & x & x & x & x & x &  &  &  & x & x & \\
+\hspace{1em}Turnout &  &  & x &  &  &  &  & x &  &  &  &  &  & \\
+\hspace{1em}Right-wing populist party votes & x & x & x & x & x & x & x & x &  &  &  & x &  & \\
+\hspace{1em}Nursing homes & x & x & x & x & x & x & x & x &  &  &  &  & x & \\
 \bottomrule
@@
 
-Our objective was to identify causal effects for COVID-19 cases. We
-found that weather affects the reported number of infections, especially
-temperature (which has a reducing effect on case numbers) and rainfall
-(which increases case numbers). We saw that reports of high case numbers
-in districts led to a reduction in new infection numbers, which
-indicates risk-averse awareness in the population and/or effective
-public health measures to suppress a local outbreak. The overall effect
-of mobility showed no consistent effect, however, in specific areas
-significant causal effects could be measured: Increasing activity in
-retail and recreational areas increased reported case numbers, while
-increased movement for essential shopping (grocery and pharmacy) led to
-reduced case numbers.
-
-Furthermore, we made a strong case for the use of causal DAGs in
-epidemiology and a pandemic like COVID-19: DAGs allow to choose
-confounders for the analysis in a principled and statistically correct
-way while reducing possible causes for bias. Also, the DAG formalization
-allows for discussion about the underlying causal assumptions.
+Our objective was to identify effects of determining factors for COVID-19 cases within a causal framework.
+We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers).
+We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population and/or effective public health measures to suppress a local outbreak.
+Mobility showed distinct effects: Increasing activity in retail and recreational areas, as well as transit stations and workplaces increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) and in parks or residential areas led to reduced case numbers.
+All interventions considered (ban of mass gatherings, school/kindergarten closures, contact restrictions, mandatory face masks) reduced case numbers considerably.
+Socio-demographic variables had small effects individually, but in conjunction they explained larger case numbers in (urban) areas with younger population, lower socio-economic status, and higher population density.
+
+Furthermore, we made a strong case for the use of causal DAGs in epidemiology and a pandemic like COVID-19: DAGs allow to choose confounders for the analysis in a principled and statistically correct way while reducing possible causes for bias.
+Also, the DAG formalization allows for discussion about the underlying causal assumptions.
 
@@
 
-Most research on determinants affecting case numbers of COVID-19 is
-restricted to single aspects
-\cite{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
-To reliably identify causal drivers, one must adjust for confounders. To
-this end, we used an integrated model with variables from different
-aspects like mobility, awareness, weather, or socio-demographics and
-identified confounders by causal analysis with a directed acyclic graph.
-A causal approach is used in another current COVID-19 analysis
-\cite{gencoglu2020causal}. There, however, they identify the causal
-relationships (reconstruct a DAG), while we estimated causal effects for
-a given hypothesized DAG.
-
-Several studies assessing the impact of public health measures on
-mobility have each observed a downward trend accompanied by a decrease
-in the number of newly reported cases
-\cite{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
-
-Our findings regarding awareness/Google Trends analysis are in good
-agreement with the correlations found by others \cite{effenberger_association_2020,higgins_correlations_2020,yuan_trends_2020}, who conclude that alertness to COVID-19
-rises several days before the highest number of cases are reported. At
-this point it should be noted, that awareness is substantially
-influenced by public media coverage, which should be considered, if
-possible, in future studies \cite{higgins_correlations_2020}. As such,
-awareness is difficult to measure and here the number of Google searches
-for ``corona'' could only be a proxy for this concept.
-
-In addition, in alignment with other recent published studies, our
-results confirm evidence which associated a negative effect of
-temperature on new COVID-19 cases
-\cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
-It is however controversial to other scientific literature describing no
-effects
-\cite{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
-or even converse correlations
-\cite{auler_evidence_2020, xie_association_2020}. The conflicting
-results might be explained by different climates and characteristics of
-the populations under study. While we are confident that our strict
-causal analysis resulted in effect estimates as undistorted as possible,
-there might be unconsidered bias in those other studies. Further
-research needs to be done to elucidate the biological characteristics of
-the novel virus SARS-CoV-2 regarding its ambient temperature survival
-and transmission. Finally, we found a positive causal effect of
-increased precipitation (rainfall) and a raise in COVID-19 cases, which supports
-previous observations \cite{sobral_association_2020}.
+Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects \cite{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
+To reliably identify (causal) drivers, one must adjust for confounders.
+To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph.
+A causal approach is used in another current COVID-19 analysis \cite{gencoglu2020causal}.
+There, however, they identify the causal relationships (reconstruct a DAG), while we estimated effects for a given hypothesized causal DAG.
+
+Several studies assessing the impact of public health measures on mobility have each observed a downward trend accompanied by a decrease in the number of newly reported cases \cite{chang_modeling_2020, cowling2020impact, fowler_effect_2020, kraemer_effect_2020, lai_effect_2020, linka_outbreak_2020}.
+
+Our findings regarding awareness/Google Trends analysis are in good agreement with the correlations found by others \cite{effenberger_association_2020,higgins_correlations_2020,yuan_trends_2020}, who conclude that alertness to COVID-19 rises several days before the highest number of cases are reported.
+At this point it should be noted, that awareness is substantially influenced by public media coverage, which should be considered, if possible, in future studies \cite{higgins_correlations_2020}.
+As such, awareness is difficult to measure and here the number of Google searches for ``corona'' could only be a proxy for this concept.
+
+In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \cite{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
+It is however controversial to other scientific literature describing no effects \cite{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \cite{auler_evidence_2020, xie_association_2020}.
+The conflicting results might be explained by different climates and characteristics of the populations under study.
+While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies.
+Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission.
+Finally, we found a positive effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \cite{sobral_association_2020}.
 
@@
 
-While use of a causal DAG is itself a strong tool to identify causal
-effects (and not just statistical associations), it introduces two
-limitations: causal assumptions within the graph (depicted by edges)
-need to be well justified, and the statistical regression model that
-calculates total causal effects needs to be appropriate for the task at
-hand. We endorse our graph as a basis for discussion on residual
-confounding. We did not try to construct the DAG from the available data
-(cf.~\cite{gencoglu2020causal}). As such, our proposed DAG is not
-entirely consistent with the data and there are conditional dependencies
-between variables that cannot be dissolved by adding edges to the DAG
-(e.g.~between the policies like contact restrictions and mandatory face
-masks). Another way to identify potential problems in the proposed DAG
-is to perform a sensitivity analysis of its structure by inspecting its
-maximal ancestral graph (MAG) or its Markov equivalence class
-represented by a complete partially DAG (CPDAG) and the existence of
-valid adjustment sets for these generalized graphs
-\cite{perkovic2017complete}. For the MAG derived from our DAG, only the
-effects for exposures mobility and searches for corona can be estimated
-with valid adjustment sets, while for the Markov equivalence class all
-exposures but COVID-19 burden lead to valid adjustments sets. A further
-analysis of these implications is out of the scope of this paper.
-
-We observed overdispersion and a substantial increase in model
-performance with a negative binomial regression compared to Poisson
-regression, which is in line with the results on COVID-19 daily case
-counts of others \cite{kraemer_effect_2020,liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
-We did not model case counts with a differential equation model like the
-classic SIR-model \cite{kermack1991contributions} and its successors,
-since these are more suited to prediction, e.g.~\cite{rkimodellierung}, while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders
-more reliably. There are more advanced statistical methods for count
-data, e.g.~zero-inflated models and mixed models. We tested both
-approaches as extensions to the negative binomial regression and
-experienced numerical problems and increased computing time, along with
-an insubstantial increase in model performance. Furthermore, our model
-assumed that all variables have effects proportional to the size of
-their measurements. It is possible that some variables show saturation
-effects or opposite effects for low, medium, or high values. This could
-be modelled with polynomial or other transformations of the variables,
-which we did not employ due to limited temporal and spatial data
-availability. Use of a fixed DAG with effect estimation via regression
-assumes that data was generated by the same underlying process for the
-observation period. By inclusion of the successive mitigation policies
-as binary variables we were able to explain some of the variance caused
-by the changing dynamics of case numbers (similar to
-\cite{Islamm2743}).
-
-We stress the point that our effects were deduced on an aggregate
-(district) level in the absence of available data on an individual
-level. As such, conclusions about effects cannot be transferred on
-individuals without the possibility for an ecological fallacy.
-Furthermore, as we were using administrative data for our analysis, the
-results are susceptible to the Modifiable Area Unit Problem (MAUP)
-\cite{Openshaw1984}. The MAUP postulates that different regional
-aggregations of the units of observation may lead to different results
-and conclusions. Due to limited available data for the different
-variables, there is currently no way to overcome these problems that are
-inherent to all analyses on aggregated data level.
-
-Our observation period was restricted to succession from late winter to
-spring and summer (February to July). Nevertheless, this transition with
-increasing temperature was a natural experiment that allowed clues on
-weather effects.
-
-We could not include data on health care utilization during the pandemic
-into our models due to the lack of available resources. This is planned
-for a later follow up to this paper since we rank health care
-utilization and mobility within health care facilities among the strong
-factors for COVID-19 progression: personnel in hospitals and private
-practices is particularly exposed to infection, while the lack of
-adequate care for other diseases has severe effects on general health of
-the population. At the same time, health care facilities are key for
-testing and surveillance of COVID-19 patients.
-
-While our analysis focused on Germany and its districts, we assume that
-results may be transferred to other countries by adjusting for their
-respective weather conditions, mobility habits, socio-demographic
-characteristics, and other determining factors.
-
-The code and resources for our analysis are available on Github, we
-invite other researchers to replicate our analysis with different
-assumptions using the files provided in the repository\footnote{The
-  repository is located here: https://github.com/zidatalab/causalcovid19}
-of the article.
+While use of a causal DAG is itself a strong tool to identify \emph{causal} effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand.
+We endorse our graph as a basis for discussion on residual confounding.
+We did not try to construct the DAG from the available data (cf.~\cite{gencoglu2020causal}).
+As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the interventions like contact restrictions and mandatory face masks).
+Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \cite{perkovic2017complete}.
+For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets.
+A further analysis of these implications is out of the scope of this paper.
+
+We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \cite{kraemer_effect_2020} and others \cite{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
+We did not model case counts with a differential equation model like the classic SIR-model \cite{kermack1991contributions} and its successors, since these are more suited to prediction e.g.~\cite{rkimodellierung}, while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably.
+There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models.
+We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance.
+Furthermore, our model assumed that all variables have effects proportional to the size of their measurements.
+It is possible that some variables show saturation effects or opposite effects for low, medium, or high values.
+This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.
+Interaction effects of variables and confounding effects or mediating variables are explicitly taken care of by deriving the valid adjustment sets for a given exposure based on the causal DAG.
+Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period.
+By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \cite{Islamm2743}).
+While multicollinearity of variables poses less of a problem for a proper causal graph analysis \cite{schisterman2017collinearitycausal}, we addressed the problem of multicollinearity in our predictors by two approaches: principal component analysis for the highly collinear mobility variables as well as a regularized regression approach (ridge regression).
+The latter (in conjunction with cross-validation) also reduced the problem of overfitting.
+
+We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level.
+As such, conclusions about effects cannot be transferred on individuals without the possibility for an ecological fallacy.
+Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) \cite{Openshaw1984}.
+The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions.
+Due to limited available data for the different variables, there is currently no way to overcome these problems that are inherent to all analyses on aggregated data level.
+
+Our observation period was restricted to succession from late winter to spring and summer (February to July).
+Nevertheless, this transition with increasing temperature was a natural experiment that allowed clues on weather effects.
+
+We could not include data on health care utilization during the pandemic into our models due to the lack of available resources.
+This is planned for a later follow up to this paper since we rank health care utilization and mobility within health care facilities among the strong factors for COVID-19 progression: personnel in hospitals and private practices is particularly exposed to infection, while the lack of adequate care for other diseases has severe effects on general health of the population.
+At the same time, health care facilities are key for testing and surveillance of COVID-19 patients.
+
+While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.
+
+The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository\footnote{The repository is located here: https://github.com/zidatalab/causalcovid19} of the article.
 
@@
 
-In our analysis, the adverse effect of mobility in retail/recreation and
-the favorable effect of mobility in grocery/pharmacy indicate that
-policies like contact restrictions which limit the number of individual
-interactions can lead to reduced infection numbers. This is due to
-retail/recreational areas encompassing mostly places of social
-gatherings like restaurants and bars, malls, sports and music venues,
-among others, while if people are doing more of their essential shopping
-at supermarkets, they will most likely stay at home with less contact to
-other people.
-
-The causal effects of awareness measured via searches for ``corona'' and
-the COVID-19 burden are harder to interpret. We assume that within our
-model, the searches for ``corona'' are an insufficient proxy for
-awareness, while the decreasing effect for future case numbers of high
-daily COVID-19 burden indicates it affects individual risk-behavior and
-entails effective non-pharmaceutical interventions.
-
-Similarly, the effects of temperature and rainfall can be interpreted as
-causal effects for indoor and outdoor activities, such that higher
-temperatures and low rainfall indicate more people spending time outdoor
-while lower temperatures and high rainfall result in indoor activities,
-which lead to more infections. Current research suggests this to be due
-to the prevalent airborne and respiratory droplets and aerosol
-transmission of the SARS-CoV-2 virus \cite{world2020transmission}. In
-this light, we advocate for precautious measures like increased hygiene,
-face masks, and air ventilation for unavoidable indoor activities.
+In our analysis, the adverse effects of mobility in retail/recreation and workplaces and the favorable effect of mobility in grocery/pharmacy and residential areas indicate that interventions like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers.
+This is due to retail/recreational and workplace areas encompassing mostly places of (social) gatherings, while if people are doing more of their essential shopping at supermarkets and stay at home with less contact to other people, they are less likely to come in contact with infected individuals.
+
+The effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret.
+We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.
+
+Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections.
+Current research suggests this to be due to the prevalent airborne and respiratory droplets and aerosol transmission of the SARS-CoV-2 virus \cite{world2020transmission}.
+In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.
+
+Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions.
+To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
 
@@
 
-To the best of our knowledge, this is the most comprehensive analysis of
-causes for COVID-19 infections which integrates different data sources
-(all publicly available). Causal reasoning with a DAG allows us to
-estimate the causal effects more reliably.
-
-Our findings suggest that the causal effects of mobility, awareness, and
-weather need to be taken into account when deciding for
-mitigation and suppression measures, depending on the recent and future
-COVID-19 pandemic development.
+To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available).
+Causal reasoning with a DAG allows us to estimate the possible causal effects more reliably.
+
+Our findings suggest that the infection-driving effects of mobility, awareness, and weather (and to some extent socio-demographic factors) need to be taken into account when deciding for mitigation and suppression interventions, depending on the recent and future COVID-19 pandemic development.
 
@@
 \section*{Acknowledgments}\label{acknowledgments}
-We are thankful for feedback from Thomas Czihal, Johannes Textor, and
-Ralph Brinks, who gave helpful suggestions on an earlier draft of the
-manuscript.
+We are thankful for feedback from Thomas Czihal, Johannes Textor, Ralph Brinks, and an anonymous reviewer who gave helpful suggestions on earlier versions of the manuscript.
 

### 2021-03-21 manuscript/v4_medrxiv_revision/results.Rmd
fix typo

@@
 While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (ban of mass gatherings)")%>%pull("effect estimate")), 1)` percent, the closure of schools/kindergartens reduced infections by an additional `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (school and kindergarten closures)")%>%pull("effect estimate")), 1)` percent and mandatory face masks reduced this by another `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (mandatory face masks)")%>%pull("effect estimate")), 1)` percent.
-The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (contact restrictions)")%>%pull("effect estimate")), 1)`.
+The effect of contact restrictions was the strongest in our observation period, with an reduction of the case rates by `r round(100*abs(1-causaleffects%>%filter(cause=="Interventions (contact restrictions)")%>%pull("effect estimate")), 1)`.
 

### 2021-03-21 manuscript/v5_submission_revision/replyreviewer.Rmd
update reply to reviewer

--- 2021-03-11 manuscript/v5_submission_revision/replyreviewer.Rmd
@@
 ---
-title: "reply to reviewer"
-output: html_document
+title: "Reply to Reviewer"
+output:
+  pdf_document: default
 ---
@@
 
-*This is an interesting manuscript that aims to identify "causal" relationships" for COVID-19 transmission based on a DAG analysis and empirical data from other studies for variable selection.*
+We want to thank the reviewer for his feedback. We heavily revised the manuscript to address every remark of the reviewer. Below we will address every point of the review separately.
 
-Yes, we aim to estimate causal effects based on a DAG that was constructed based on other publications.
+**This is an interesting manuscript that aims to identify "causal" relationships" for COVID-19 transmission based on a DAG analysis and empirical data from other studies for variable selection.**
+
+1.
+Yes, we aim to estimate *causal effects* based on a DAG that was constructed based on other publications.
 Empirical data was extracted and aggregated from different publicly available databases: RKI, Google mobility/Trends, INKAR, DWD.
-Variable selection for the different models was based on the proposed DAG using the well-studied methods from Pearl etc.
+Variable selection for the different models was based on the proposed DAG using the well-studied methods from Pearl, Spirtes, and others.
+That is, rather than selecting variables, we identified valid adjustment sets in a principled way to estimate effects in a causal framework.
 
-*The DAG analysis is a multivariate analysis to select a subset of variables that are correlated from a large number of candidates.
+**The DAG analysis is a multivariate analysis to select a subset of variables that are correlated from a large number of candidates.
 The first select the sets of variables that are associated with the outcome using DAG analysis and then put them into the regression.
 The DAG analysis returns different sets of variables, using different selection criterion ( for example, the most parsimonious set or the most variation explained ).
-So the authors assumedly use R-squared to select the final sets of variables in the regression model.*
+So the authors assumedly use R-squared to select the final sets of variables in the regression model.**
 
-Yes, we identify subsets of variables using DAG analysis.
-We stress the point that we do this identification process separately for every variable of interest where we are interested in the causal effect.
-We always regress on the same outcome Y (Reported COVID-19 cases) using negative-binomial regression.
+2.
+Yes, for every exposure variable separately, we identify subsets of variables to include in the regression models using DAG analysis.
+We stress the point that we do this identification process separately for every variable of interest where we are interested in the total causal effect.
+We always regress on the same outcome Y (Reported COVID-19 cases) using negative binomial regression.
 That is, we select one variable Xi where we are interested in the effect of Xi on Y.
-Then we identify all possible subsets of variables in the DAG that allow to estimate the causal effect of Xi on Y (based on Pearl etc).
-Out of these subsets we choose the "best" one in regards to pseudo-R2 (or AIC).
-We repeat this procedure for several variables Xi.
+Then we identify all possible subsets of variables in the DAG that allow to estimate the causal effect of Xi on Y (based on the do-calculus framework by Pearl et al.).
+Out of these subsets we choose the "best" one in regards to minimal AIC (we changed this from R-squared in the new version of the manuscript).
+We repeat this procedure for all possible exposure variables Xi.
+We want to stress the point, that the set of valid adjustment sets is not found by looking at AIC/R-squared (or the actual data), but rather only the structure of the DAG.
+These adjustment sets have the property to allow reliable estimates of the total causal effect from Xi to Y.
+When given multiple adjustment sets for the same Xi, we now chose the one with the minimal AIC.
 
-*The challenge I have here (and having reviewed this as well with our statistical team) is that the claims of causal relationships are not convincing, even if they identify associative relationships that other studies have also found.
-Specifically: 1) The model does not address the nonlinearity of variables (such as temperature); other studies have found that temperature has a u-shaped effect at low and high temperatures.
+**The challenge I have here (and having reviewed this as well with our statistical team) is that the claims of causal relationships are not convincing, even if they identify associative relationships that other studies have also found.**
+
+3.
+We are aware that the causal DAG framework is still a rather new approach to study relationships between different factors.
+It is fascinating how the causal DAG and do-calculus framework allows to actually discern mere associations from actual (direct or total) causal effects.
+Of course, this is only possible if the DAG is reliable and regression model assumptions are met.
+We are happy to discuss the proposed DAG.
+To reflect the concerns that we actually found *causal* effects, we used more careful wording throughout most of the revised paper, to make clear that effects estimates afre based on the causal graph and are strong indicators for causal relationships.
+Within the causal DAG framework, the terms *total*/*direct causal effect* are used to describe the relationships (that is, arrows) between variables/nodes in the causal DAG.
+
+**Specifically: 1) The model does not address the nonlinearity of variables (such as temperature); other studies have found that temperature has a u-shaped effect at low and high temperatures.
 Humidity effects may also be nonlinear.
-The assumption of non-linearity on continuous variables needs to be considered.*
+The assumption of non-linearity on continuous variables needs to be considered.**
 
-Yes, non-linearity might be a problem for our generalized linear model (negative-binomial regression).
-We added a complete check of model diagnostics that addresses linearity assumptions, model specification, independence of observations, and multi-collinearity checks (see below).
-We did not find significant problems with non-linear relationships (see plot XX supplementary?).
+4.
+Yes, non-linearity might be a problem for our generalized linear model (negative binomial regression).
+We did a complete check of model diagnostics that addresses linearity assumptions, model specification/distribution of residuals, independence of observations, and multi-collinearity checks (see below).
+We did not find significant problems with non-linear relationships.
 
-*2) There are no tests for multi-collinearity that are presented, which presents significant concerns for over-fitting the model.
+The fitted vs residual plot shows general good model adequacy, since the residuals are almost evenly distributed around 0 for all fitted values.
+This means that the negative binomial model is adequate and there is no general non-linear trend in the predictors collectively (otherwise the black line would show a clear trend down/up/strongly curved).
+
+The plot for individual predictors versus residuals does not reveal strong non-linear effects for individual predictors.
+The LOESS curves for COVID-19 burden and Mobility Principal Component 3 wiggle upwards for higher values, but these are very few.
+In the majority, curves are horizontal and don't imply model problems with non-linearity.
+
+The distribution of the quantile residuals is a very good fit to a normal distribution, which indicates that the negative binomial model is a good model for the data.
+(For generalized linear models, QQ-plots of Pearson residuals are not meaningful, since a GLM actually makes no assumption about the distribution of errors.)
+
+We did not model "time" explicitly as a variable.
+The residual plot with residuals ordered by time shows that we can assume independence of observations over time within the model, since the residuals are distributed around 0 iid for every date.
+In reality, observations on subsequent days are not independent, but the data does not show a strong temporal trend in residuals.
+
+Another viewpoint on independence of observations: Since our data comes from 401 different districts, a mixed/hierarchical model might be more appropriate, since we cannot assume that observations within the same district are independent of each other.
+The plot of district id vs residuals doesn't show a strong violation of the independence assumption (we actually ordered the districts by their respective residual mean in the plot above).
+Nevertheless, we tried a mixed model with random intercepts for districts, but the very small increase in model performance was in no relation to the increased computing time by several magnitudes.
+
+Assessing these diagnostic plots, we are confident that our negative binomial regression framework is appropriate and that non-linearity is not an issue with this dataset.
+
+![Diagnostic plot: Fitted vs residuals.](figuresreviewer/modelcorrect_fitted_vs_resids.png)
+
+![Diagnostic plot: predictors vs residuals.](figuresreviewer/modelcorrect_predictors_vs_resids.png)
+
+![Diagnostic plot: distribution of residuals.](figuresreviewer/modelcorrect_hist_qresids.png)
+
+![Diagnostic plots: independence of observations, temporal.](figuresreviewer/independence_obs_resids_vs_time.png)
+
+![Diagnostic plot: independence of observations, districts.](figuresreviewer/independence_id_vs_resid.png)
+\newpage
+
+**2) There are no tests for multi-collinearity that are presented, which presents significant concerns for over-fitting the model.
 Temperature and humidity are an example here.
-So is the mobility data.*
+So is the mobility data.**
 
-This an important remark.
-We added a correlation plot for the continuous variables and assessed variance inflation factors (VIFs) for the full model.
-We observed high VIFs for the mobility variables and the different intervention variables.
+5.
+This an important remark which led to some major revisions on the analysis pipeline and the paper.
+We assessed variance inflation factors (VIFs) for the full model.
+We observed high VIFs for mobility variables and intervention variables.
 We decided to transform the original mobility variables by PCA and used the set of these principal components instead of the original mobility variables.
-Furthermore, we summarised the intervention variables into one "Intervention" variable (compare new DAG) by summing these binary variables.
-Analysing the VIFs on the full model with these transformed variables, we are left with only one VIF>5 (PC Mobi 1=5.1xxx), which is a huge improvement on the original model.
-We want to add that causal graph analysis for regression is more robust to multi-collinearity than traditional regression approaches (cite paper).
+Furthermore, we employed ridge regression as a way to further deal with multicollinearity and overfitting.
+We want to add that causal graph analysis with regression is more robust to multicollinearity than traditional regression approaches (see Schisterman et al 2017), and that in the presence of correlated variables, one might miss important effects, but the effects found are still valid.
 
-*2) Feels like they cherry-pick the variables of interest in the end?
+**2) Feels like they cherry-pick the variables of interest in the end?
 Not sure how they arrived at the final variable list for causal effects.
-Whey were the restrictions not included in that analysis?*
+Whey were the restrictions not included in that analysis?**
 
-*3) The term “causal analysis” is a bit strong for what they have done here.
+6.
+At the time of writing, we were more interested in variables that could be prognostive of future developments (especially weather, since weather is not modifiable).
+In light of this remark, we revised the analysis and paper and now we include all variables from the DAG as possible exposures, and we report and discuss all effect estimates in the manuscript.
+To reflect this, we changed the title of the manuscript, too.
+
+**3) The term “causal analysis” is a bit strong for what they have done here.
 The basis of the work is the proposed DAG (Figure 1), but the diagram was constructed from other association studies.
 Other than the DAG analysis, they did not do anything to ensure the results are “causal relationship”.
-So I am not convinced that the analysis or results are causal.How do you avoid over-fitting the model or including mediating variables in the analysis?*
+So I am not convinced that the analysis or results are causal.How do you avoid over-fitting the model or including mediating variables in the analysis?**
+
+7.
+Please see our remarks 1., 2., and 3.
+that address the general idea of/concerns about causal frameworks with DAGs and do-calculus.
+We use the term "causal" the same way that Pearl (Causality: Models, Reasoning, and Inference), Spirtes et al.
+(Causation, Prediction, and Search) and the growing body of work based on causal analysis with DAGs use it.
+We modified the title of our paper to "Causal graph analysis of [...]" to reflect the fundamental assumption of a causal graph to be able to do causal analysis.
+
+Proper identification of valid variable subsets for causal effect estimation takes care of the mediating variable problem, since this is at the heart of causal DAG analysis.
+
+Overfitting might still be a problem, which we addressed by regularized regression via ridge regression (see above) and updating our analysis, estimates, and the manuscript accordingly.
 
 *5) What about interaction terms? For example, residential mobility and colder weather? Or rain? These relationships are not simply a straightforward multivariable model.*
+
+8.
+DAG analysis explicitly takes care of interaction between variables by identifying valid subsets of variables that need to be included or excluded in the multivariable model.
+We added a remark on interaction to the paper to reflect on this remark.
 
@@
 Why were public health interventions removed?*
+
+9. See our remark 6., we added effect estimates and discussion of effects for all variables to the manuscript.
+
+Final remarks: We sincerely hope that our remarks, explanations, and the major revisions we did on the analysis and manuscript address all concerns.
+We included all possible variables as exposure, did a model diagnostics, and added principal component analysis, ridge regression, and cross-validation to the analysis to make it resilient to multicollinearity and overfitting.
+We hope to provide a meaningful addition to the growing body of research on causal graphs/causal analysis in epidemiology, while reporting important results on the determining factors of the COVID-19 pandemic.

### 2021-03-21 R/3_modelassumptions.R

@@ -98,14 +98,14 @@ lines(density(rnorm(length(qresids))),
 # independence of observations
 
 scatter.smooth(seq(length(devresids)), devresids, col='gray')
-myresid_id_daycount <- bind_cols(devresids=devresids, id_daycount)
+myresid_id_daycount <- bind_cols(devresids=devresids, id_daycount) %>% group_by(id) %>% mutate(meanresid=mean(devresids)) %>% ungroup()
 ggplot(myresid_id_daycount, aes(x=daycount, y=devresids)) +
   geom_point() +
   geom_smooth()
 # ggplot(myresid_id_daycount, aes(x=reorder(as.factor(id), devresids, FUN = median), y=devresids)) +
 #   geom_boxplot()
-ggplot(myresid_id_daycount, aes(x=as.factor(id), y=devresids)) +
-  geom_boxplot(aes(col=as.factor(bl_id)))
+ggplot(myresid_id_daycount, aes(x=reorder(as.factor(id), meanresid), y=devresids)) +
+  geom_point()#aes(col=as.factor(bl_id))
 
 # multicollinearity
 library(corrplot)

### 2021-03-22 secondwave_analysis/R/0_1_rkicases.R
remove secondwave analysis

@@ -1,64 +1,0 @@
-library(tidyverse)
-library(lubridate)
-
-csvurl <- "https://opendata.arcgis.com/datasets/dd4580c810204019a7b8eb3e0b329dd6_0.csv"
-mytemp <- tempfile()
-options(timeout=300)
-download.file(csvurl, mytemp, quiet=TRUE)
-myresult <- read_csv(mytemp)
-localrki <- myresult %>%
-  mutate(Meldedatum= as.character(as_date(Meldedatum)),
-         Refdatum= as.character(as_date(Refdatum)),
-         Datenstand=as.character(as.Date(Datenstand, format="%d.%m.%Y")))
-  
-  rkitimeframe <- localrki %>% summarise(mindate=min(date(Meldedatum)),maxdate=max(date(Meldedatum)))
-  rkidays <- date(rkitimeframe$maxdate)-date(rkitimeframe$mindate)
-  rkidates = date(now())-seq(1,rkidays+1)
-  
-  Kreise <- localrki %>%  mutate(date=date(Meldedatum)) %>%
-    rename(id=IdLandkreis,cases=AnzahlFall,deaths=AnzahlTodesfall,recovered=AnzahlGenesen) %>%
-    dplyr::select(id,date,cases,deaths,recovered) %>% group_by(id,date) %>%
-    summarise(cases=sum(cases,na.rm=T),deaths=sum(deaths,na.rm=T),recovered=sum(recovered,na.rm=T)) %>%
-    ungroup() %>% mutate(id=as.integer(id)*1000) %>%
-    filter(!is.na(id)) %>% arrange(id,date)
-  
-  Kreise <- expand.grid("id"=unique(Kreise$id),"date"=rkidates) %>% as_tibble(.) %>%
-    arrange(id,date) %>%
-    left_join(Kreise,by=c("id","date")) %>%
-    mutate(cases=ifelse(is.na(cases),0,cases),
-           deaths=ifelse(is.na(deaths),0,deaths),
-           recovered=ifelse(is.na(recovered),0,recovered)) %>%
-    mutate(date=as.character(date))
-  
-  Laender <- localrki %>% mutate(date=date(Meldedatum)) %>%
-    rename(id=IdBundesland,cases=AnzahlFall,deaths=AnzahlTodesfall,recovered=AnzahlGenesen) %>%
-    dplyr::select(id,date,cases,deaths,recovered) %>% group_by(id,date) %>%
[…]

### 2021-03-22 secondwave_analysis/R/0_2_mobility.R

@@ -1,41 +1,0 @@
-library(sf)
-library(tidyverse)
-library(geojsonsf)
-library(geojsonio)
-
-mydatapath <- "./data/mobility/"
-
-# google data
-gmobillink <- "https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv"
-download.file(gmobillink, 
-              paste0(mydatapath, "googlemobility_raw.csv"),
-              quiet=TRUE)
-google_german_mobility <- read_csv(paste0(mydatapath, "googlemobility_raw.csv")) %>%
-  filter(country_region_code=="DE") %>%
-  dplyr::select(-country_region_code, -country_region, -sub_region_2, -metro_area,
-                -census_fips_code)
-
-en_rs_map_bl <- c("08", "09", "11", "12",
-                  "04", "02", "06", "03",
-                  "13", "05", "07", "10",
-                  "14", "15","01", "16",
-                  "00")
-# names(en_de_map_bl) <- c("Baden-Württemberg", "Bayern", "Berlin", "Brandenburg",
-#                          "Bremen", "Hamburg", "Hessen", "Niedersachsen",
-#                          "Mecklenburg-Vorpommern", "Nordrhein-Westfalen", "Rheinland-Pfalz", "Saarland",
-#                          "Sachsen", "Sachsen-Anhalt","Schleswig-Holstein", "Thüringen")
-names(en_rs_map_bl) <- c("Baden-Württemberg", "Bavaria", "Berlin", "Brandenburg",
-                         "Bremen", "Hamburg", "Hessen", "Lower Saxony",
-                         "Mecklenburg-Vorpommern", "North Rhine-Westphalia", "Rhineland-Palatinate", "Saarland",
-                         "Saxony", "Saxony-Anhalt","Schleswig-Holstein", "Thuringia",
-                         "Gesamt")
-google_german_mobility <- google_german_mobility %>%
-  mutate(sub_region_1=ifelse(is.na(sub_region_1), "Gesamt", sub_region_1)) %>%
-  mutate(grocery_and_pharmacy_percent_change_from_baseline=replace_na(grocery_and_pharmacy_percent_change_from_baseline, 0)) %>%
-  mutate(parks_percent_change_from_baseline=replace_na(parks_percent_change_from_baseline, 0)) %>%
[…]

### 2021-03-22 secondwave_analysis/R/0_3_awareness.R

@@ -1,92 +1,0 @@
-library(tidyverse)
-library(lubridate)
-
-mydatapath <- "./data/awareness/"
-
-bundeslaender_maps <- tibble(
-  BL_ID=1:16,
-  Bundesland=c(
-    "Schleswig-Holstein",
-    "Hamburg",
-    "Niedersachsen",
-    "Bremen",
-    "Nordrhein-Westfalen",
-    "Hessen",
-    "Rheinland-Pfalz",
-    "Baden-Württemberg",
-    "Bayern",
-    "Saarland",
-    "Berlin",
-    "Brandenburg",
-    "Mecklenburg-Vorpommern",
-    "Sachsen",
-    "Sachsen-Anhalt",
-    "Thüringen"
-  ),
-  Kuerzel=c(
-    "SH",
-    "HH",
-    "NI",
-    "HB",
-    "NW",
-    "HE",
-    "RP",
-    "BW",
-    "BY",
-    "SL",
-    "BE",
-    "BB",
-    "MV",
-    "SN",
-    "ST",
-    "TH"
-  )
-)
-
-trends_alle_bl <- tibble()
-
-for (kuerzel in bundeslaender_maps$Kuerzel) {
-  this_bl <- read_csv(paste0("data/awareness/trends_", kuerzel, ".csv"), 
-                      col_names = c("woche", "searches"),
-                      col_types = cols(woche=col_date(format = "%Y-%m-%d"), 
-                                       searches=col_double()), 
-                      skip = 2) %>%
-    mutate(searches=ifelse(is.na(searches), 0, searches),
-           BL_ID=bundeslaender_maps%>%filter(Kuerzel==kuerzel)%>%pull(BL_ID),
-           Kuerzel=kuerzel,
-           Bundesland=bundeslaender_maps%>%filter(Kuerzel==kuerzel)%>%pull(Bundesland)) %>%
-    drop_na()
-  trends_alle_bl <- bind_rows(trends_alle_bl,
-                              this_bl)
-}
-
-relativ_gesamt_2020_03_15 <- read_csv(paste0(mydatapath, 
-                                             "relativ_gesamt_2020-03-15.csv"))
-
-relativ_bl_2020_03_15 <- left_join(trends_alle_bl %>% 
-                                     filter(woche=="2020-03-15"), 
-                                   relativ_gesamt_2020_03_15,
-                                   by="Bundesland") %>%
-  mutate(baseline20200315=searches*Relativtrend/100) %>%
[…]

### 2021-03-22 secondwave_analysis/R/0_4_inkar.R

@@ -1,26 +1,0 @@
-library(readr)
-header <- read_delim(file = "data/sociodemographic/inkar_sdg_etc.csv",
-                      delim=";",
-                      col_names = FALSE,
-                      n_max = 1)
-inkar <- read_delim("data/sociodemographic/inkar_sdg_etc.csv", 
-                            ";", escape_double = FALSE, col_names = FALSE, 
-                            locale = locale(decimal_mark = ",", grouping_mark = "."), 
-                            trim_ws = TRUE, skip = 2)
-
-colnames(inkar) <- header[1, ]
-
-inkar <- inkar %>%
-  mutate(`Age (pop. younger 18)`=`Einwohner unter 6 Jahre`+`Einwohner von 6 bis unter 18 Jahren`,
-         Frauenanteil=`Geschlechterproportion insgesamt`/(1+`Geschlechterproportion insgesamt`)) %>%
-  dplyr::select(contains("Haushalte mit niedrigem"),
-                contains("Jahre"),
-                contains("Ausländeranteil"), contains("Schutzsuchende"),
-                contains("Wahlbeteiligung"), contains("AfD"),
-                Einwohnerdichte, Frauenanteil,
-                contains("Age"),
-                Kennziffer, Raumeinheit, Aggregat,
-                -contains("Einwohner unter 6"), -contains("Einwohner von 6")
-  )
-
-write_csv(inkar, "./data/inkar/inkar.csv")

### 2021-03-22 secondwave_analysis/R/0_5_weather.R

@@ -1,203 +1,0 @@
-# the download of weather files takes a while! ~ 1 h
-
-library(readr)
-library(tidyverse)
-
-mydatapath <- "./data/weather/"
-
-opendata_dwd_path <- "https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/recent/"
-
-download.file(paste0(opendata_dwd_path, "KL_Tageswerte_Beschreibung_Stationen.txt"),
-              paste0(mydatapath, "weatherstations.txt"))
-
-header <- read_table2(file = paste0(mydatapath, "weatherstations.txt"),
-                   col_names = FALSE,
-                   n_max = 1)
-weatherdata <- read_table(file = paste0(mydatapath, "weatherstations.txt"),
-                    col_names = FALSE,
-                    locale = locale(encoding = "WINDOWS-1252"),
-                    skip = 2)
-colnames(weatherdata) <- header[1, ]
-
-extractedfiles <- c()
-
-best_bisdatum <- last(names(sort(table(weatherdata$bis_datum))))
-
-weatherdata <- weatherdata %>%
-  filter(bis_datum==as.numeric(best_bisdatum) & von_datum <= 20200101 & Stationshoehe<1000)
-
-for (stationid in weatherdata$Stations_id) {
-  tryCatch({
-    download.file(paste0(opendata_dwd_path, "tageswerte_KL_", stationid, "_akt.zip"),
-                  paste0(mydatapath, "temp.zip"),
-                  quiet=TRUE)
-    cat("downloaded data for", stationid, "successfully\n")
-    listoffilesinzip <- unzip(paste0(mydatapath, "temp.zip"),
-                              list=TRUE)
-    extractedfile <- unzip(paste0(mydatapath, "temp.zip"),
-                           files=last(listoffilesinzip$Name),
-                           exdir=paste0(mydatapath, "temp"))
-    extractedfiles <- c(extractedfiles, extractedfile)
-  }, error = function(e) {
-    cat(stationid, "did not work\n")
-  }, warning = function(w) {
-    cat(stationid, "did not work\n")
-  })
-  
-}
-
-write_lines(extractedfiles,
[…]

### 2021-03-22 secondwave_analysis/R/0_z_testlags.R

@@ -1,292 +1,0 @@
-mylags <- 1:20
-
-myaic <- rep(0, 20)
-myglms <- vector("list", 20)
-
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-
-mydatapath <- "secondwave_analysis/data/" 
-
-google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
-  mutate(id=as.integer(RS_first2)) %>%
-  dplyr::select(-RS_first2) %>%
-  pivot_wider(names_from = areatype, values_from = perc_change) %>%
-  rename(`Mobility (retail and recreation)`=retail_and_recreation_percent_change_from_baseline,
-         `Mobility (grocery and pharmacy)`=grocery_and_pharmacy_percent_change_from_baseline,
-         `Mobility (parks)`=parks_percent_change_from_baseline,
-         `Mobility (transit stations)`=transit_stations_percent_change_from_baseline,
-         `Mobility (workplaces)`=workplaces_percent_change_from_baseline,
-         `Mobility (residential)`=residential_percent_change_from_baseline)
-
-weather <- read_csv(paste0(mydatapath, "weather/mittleres_kreiswetter.csv")) %>%
-  mutate(date=ymd(MESS_DATUM), id=as.integer(RS)*1000) %>%
-  dplyr::select(-MESS_DATUM, -RS) %>%
-  mutate(Wind=replace_na(Wind, mean(Wind, na.rm=TRUE)),
-         Feuchtigkeit=replace_na(Feuchtigkeit, mean(Feuchtigkeit, na.rm=TRUE))) %>%
-  rename(`Weather (rainfall)`=Niederschlag,
-         `Weather (wind)`=Wind,
-         `Weather (temperature)`=Temperatur,
-         `Weather (humidity)`=Feuchtigkeit)
-
-awareness <- read_csv(paste0(mydatapath, "awareness/awareness.csv")) %>%
-  rename(`Searches corona`=relativtrend) %>%
-  dplyr::select(-Bundesland)
-
-pflege_destatis <- read_delim(paste0(mydatapath, "sociodemographic/pflege_destatis.csv"), 
-                              ";", escape_double = FALSE, locale = locale(decimal_mark = ",", 
[…]

### 2021-03-22 secondwave_analysis/R/1_generate_data.R

@@ -1,283 +1,0 @@
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-source("secondwave_analysis/R/z_auxiliary_functions_for_1.R")
-
-mylag <- 7
-
-mydatapath <- "secondwave_analysis/data/" 
-
-google_mobility <- read_csv(paste0(mydatapath, "mobility/google_mobilitychange_percentages_per_bundesland_and_areatype.csv")) %>%
-  mutate(id=as.integer(RS_first2)) %>%
-  dplyr::select(-RS_first2) %>%
-  pivot_wider(names_from = areatype, values_from = perc_change) %>%
-  rename(`Mobility (retail and recreation)`=retail_and_recreation_percent_change_from_baseline,
-         `Mobility (grocery and pharmacy)`=grocery_and_pharmacy_percent_change_from_baseline,
-         `Mobility (parks)`=parks_percent_change_from_baseline,
-         `Mobility (transit stations)`=transit_stations_percent_change_from_baseline,
-         `Mobility (workplaces)`=workplaces_percent_change_from_baseline,
-         `Mobility (residential)`=residential_percent_change_from_baseline)
-
-weather <- read_csv(paste0(mydatapath, "weather/mittleres_kreiswetter.csv")) %>%
-  mutate(date=ymd(MESS_DATUM), id=as.integer(RS)*1000) %>%
-  dplyr::select(-MESS_DATUM, -RS) %>%
-  mutate(Wind=replace_na(Wind, mean(Wind, na.rm=TRUE)),
-         Feuchtigkeit=replace_na(Feuchtigkeit, mean(Feuchtigkeit, na.rm=TRUE))) %>%
-  rename(`Weather (rainfall)`=Niederschlag,
-         `Weather (wind)`=Wind,
-         `Weather (temperature)`=Temperatur,
-         `Weather (humidity)`=Feuchtigkeit) %>%
-  dplyr::select(-Luftdruck)
-
-awareness <- read_csv(paste0(mydatapath, "awareness/awareness.csv")) %>%
-  rename(`Searches corona`=relativtrend) %>%
-  dplyr::select(-Bundesland)
-
-pflege_destatis <- read_delim(paste0(mydatapath, "sociodemographic/pflege_destatis.csv"), 
-                              ";", escape_double = FALSE, locale = locale(decimal_mark = ",", 
[…]

### 2021-03-22 secondwave_analysis/R/2_dags_for_covidcases.R

@@ -1,61 +1,0 @@
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-library(dagitty)
-library(ggdag)
-library(ggthemes)
-library(pcalg)
-library(pscl)
-source("R/z_auxiliary_functions_for_2.R")
-
-# save tables to:
-tablepath <- "manuscript/v4_medrxiv_revision/tables/secondwave/"
-
-# load dag
-dagfile <- "data/bigdag4covid19.txt"
-dagtext <- readChar(dagfile, file.info(dagfile)$size)
-dag <- dagitty::dagitty(dagtext)
-
-unobserved <- c("Access to tests", 
-                "Tracing capacity",
-                "Health care utilization",
-                "Exposure to SARS-COV-2",
-                "Herd immunity",
-                "Onset of COVID-19",
-                "Trust in society",
-                "Susceptibility")
-
-# Load Data
-modeldata <- read_csv("secondwave_analysis/data/Modeldata_scaled_pcamobility.csv")
-
-# all variables
-res <- my_causal(dag, modeldata, exposure=NULL, unobserved)
-
-file.create(paste0(tablepath, "t_pseudor2s.csv"))
-file.create(paste0(tablepath, "t_aics.csv"))
-# mobility
-res <- my_causal(dag, modeldata, exposure="Mobility", unobserved=unobserved)
-write_cause(res, exposure="Mobility")
-
-# awareness
-res <- my_causal(dag, modeldata, exposure="COVID-19 burden", unobserved=unobserved)
-write_cause(res, exposure="COVID-19 burden")
-res <- my_causal(dag, modeldata, exposure="Searches corona", unobserved=unobserved)
-write_cause(res, exposure="Searches corona")
-
-# weather
-for (myweather in c("Temperature", "Rainfall", "Humidity", "Wind")) {
-  res <- my_causal(dag, modeldata, exposure=myweather, unobserved=unobserved)
-  write_cause(res, exposure=myweather)
-}
-
-# other variables
-othervariables <- setdiff(names(dag), c("Reported new cases COVID-19", unobserved,
-                                        "Mobility", "COVID-19 burden", "Searches corona",
-                                        "Temperature", "Rainfall", "Humidity", "Wind"))
-for (ov in othervariables) {
-  res <- my_causal(dag, modeldata, exposure=ov, unobserved=unobserved)
-  write_cause(res, exposure=ov)
-}

### 2021-03-22 secondwave_analysis/R/2_z_othergraphs.R

@@ -1,62 +1,0 @@
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-library(dagitty)
-library(ggdag)
-library(ggthemes)
-library(pcalg)
-library(pscl)
-source("R/z_auxiliary_functions_for_2.R")
-
-# save tables to:
-tablepath <- "./manuscript/v2_medrxiv/tables/"
-
-# load dag
-dagfile <- "./data/bigdag4covid19.txt"
-dagtext <- readChar(dagfile, file.info(dagfile)$size)
-dag_origin <- dagitty::dagitty(dagtext)
-
-unobserved <- c("Access to tests", 
-                "Tracing capacity",
-                "Health care utilization",
-                "Herd immunity",
-                "Onset of COVID-19",
-                "Trust in society",
-                "Susceptibility")
-
-# Load Data
-modeldata <- read_csv("./data/Modeldata.csv") %>%
-  dplyr::select(-id, -bl_id, -cases, -deaths, -recovered, -daycount,
-                -`Mobility (mean)`,
-                -`Relaxation of measures`,
-                -contains("iso"), -contains("census")
-  ) %>% 
-  rename(`School and kindergarten closures`=`School/Kita closures`,
-         `Holiday (exposure)`=Holiday,
-         `Foreign citizens`=`Foreign residents`,
-         `Foreign citizens (refugees)`=`Foreign residents (refugees)`,
-         `Rainfall`=`Weather (rainfall)`,
-         `Humidity`=`Weather (humidity)`,
-         `Gender`=`Sex`,
-         `Temperature`=`Weather (temperature)`,
-         `Wind`=`Weather (wind)`) %>%
-  filter(date<="2020-07-08") %>%
-  dplyr::select(-date)
-
-adjsets_list <- vector("list", 3)
-names(adjsets_list) <- c("original", "equivalence", "mag")
-
-exposures <- c("Mobility", "Searches corona", "COVID-19 burden", "Temperature", "Rainfall", "Humidity", "Wind")
-outcome <- "Reported new cases COVID-19"
-
-dag_equiv <- equivalenceClass(dag_origin)
-dag_mag <- toMAG(dag_origin)
-
-# original adjustment sets
-for (expo in exposures) {
[…]

### 2021-03-22 secondwave_analysis/R/3_modelassumptions.R

@@ -1,99 +1,0 @@
-library(lubridate)
-library(tidyverse)
-library(DBI)
-library(broom)
-library(MASS)
-library(dagitty)
-library(ggdag)
-library(ggthemes)
-library(pcalg)
-library(pscl)
-source("secondwave_analysis/R/z_auxiliary_functions_for_2.R")
-
-# Load Data
-modeldata <- bind_cols(read_csv("secondwave_analysis/data/Modeldata_scaled_pcamobility.csv"))
-
-id_daycount <- read_csv("secondwave_analysis/data/Modeldata_raw.csv") %>%
-  dplyr::select(id, daycount, date, bl_id) %>%
-  dplyr::select(-date)
-  
-
-modeldata_cont <- modeldata %>%
-  dplyr::select(-c("Weekday 1Mo", "Weekday 2Di", "Weekday 3Mi", "Weekday 5Fr", "Weekday 6Sa", "Weekday 7So",
-                   "Holiday (exposure)", "Holiday (report)",
-                   "Reported new cases COVID-19"))
-
-exposure=NULL
-adjsets=NULL
-myformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))-`Active cases`+."
-myglm <- glm.nb(as.formula(myformula),
-                  data=modeldata) # %>%dplyr::select(-`Mobility (PC4)`)
-mynullformula <- "`Reported new cases COVID-19` ~1+offset(log(`Active cases`+1))" 
-mynullglm <- glm.nb(as.formula(mynullformula),
-                    data=modeldata)
-pseudor2 <- 1-sum((myglm$fitted.values-myglm$y)^2)/sum((mynullglm$fitted.values-mynullglm$y)^2)# 1-summary(myglm)$deviance/summary(mynullglm)$deviance
-myaic <- AIC(myglm)
-mynullaic <- AIC(mynullglm)
-devresids <- residuals(myglm, type="deviance")
-library(statmod)
-qresids <- qresid(myglm)
-
-# linearity # [-1989]
-scatter.smooth(predict(myglm, type="link"), devresids, col='gray')
-
-myresidagainstcont <- bind_cols(modeldata_cont, devresids=devresids) %>%
-  pivot_longer(cols=`Rainfall`:`Active cases`)
-ggplot(myresidagainstcont, aes(x=value, y=devresids)) +
-  geom_point() +
[…]

### 2021-03-22 secondwave_analysis/R/z_auxiliary_functions_for_1.R

@@ -1,17 +1,0 @@
-# transform relative outdoor humidity and temperature to indoor humidity
-qfun <- function(humidity, temperature, pressure, maxq=8){
-  rh <- humidity/100
-  t <- temperature+273.15
-  P <- pressure*100
-  resq <- min(maxq, 1000*5/8*1/P*rh*611.2*exp(17.62*(t-273.15)/(t-30.03)))
-  return(resq)
-}
-
-qlevel <- function(humidity, temperature, pressure, cutoffq=6){
-  rh <- humidity/100
-  t <- temperature+273.15
-  P <- pressure*100
-  resq <- 1000*5/8*1/P*rh*611.2*exp(17.62*(t-273.15)/(t-30.03))
-  rlevel <- ifelse(resq<cutoffq, 0, 1)
-  return(rlevel)
-}

### 2021-03-22 manuscript/v5_submission_revision/replyreviewer.Rmd
fixes lars

@@
 
+To sum it up, our strategy aims to identify for every exposure variable separately the adjustment set with the best statistical properties among all valid adjustment sets.
+
 **The challenge I have here (and having reviewed this as well with our statistical team) is that the claims of causal relationships are not convincing, even if they identify associative relationships that other studies have also found.**
@@
 We are happy to discuss the proposed DAG.
-To reflect the concerns that we actually found *causal* effects, we used more careful wording throughout most of the revised paper, to make clear that effects estimates afre based on the causal graph and are strong indicators for causal relationships.
+To reflect the concerns that we actually found *causal* effects, we used more careful wording throughout most of the revised paper, to make clear that effects estimates are based on the causal graph and are strong indicators for causal relationships.
 Within the causal DAG framework, the terms *total*/*direct causal effect* are used to describe the relationships (that is, arrows) between variables/nodes in the causal DAG.
@@
 We want to add that causal graph analysis with regression is more robust to multicollinearity than traditional regression approaches (see Schisterman et al 2017), and that in the presence of correlated variables, one might miss important effects, but the effects found are still valid.
+To find the most appropriate hyper-parameter for the ridge regression, we employed cross-validation on subsets of the 401 districts.
+We used cross-validation on the similar problem of identifying the most appropriate lag parameter, which was subsequently changed from 8 to 5 days in the manuscript.
 

### 2021-03-22 manuscript/v4_medrxiv_revision/results.Rmd
fixes lars

@@
 
-Within our framework, we saw very different effects for mobility variables.
+Within our framework, we saw very different effects for individual mobility variables.
 For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 2)` percent.
-Similarly, Mmbility on workplaces showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (transit stations)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase.
+Similarly, mobility on workplaces showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (transit stations)")%>%pull("effect estimate")), 2)` increase in case numbers for every 1 percent point increase.
 Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.

### 2021-03-22 manuscript/v5_submission_revision/manuscript_tex4submission.tex
update manuscript for submission
fixes lars

@@
 
-For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
+For different official and political interventions on a daily basis and the district level we used one-hot encoded daily variables, i.e.~ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
 
@@
 
-\subsection{Main results}\label{main-results}
+\subsection*{Main results}\label{main-results}
 
@@
 
-Within our framework, we saw very different effects for mobility variables.
+Within our framework, we saw very different effects for individual mobility variables.
 For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent.
-Similarly, Mmbility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase.
+Similarly, mobility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase.
 Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases.
@@
 While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 7.2 percent and mandatory face masks reduced this by another 9.4 percent.
-The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.
+The effect of contact restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.
 

### 2021-03-22 manuscript/v5_submission_revision/replyreviewer.Rmd
fixes tobias

@@
 
-We want to thank the reviewer for his feedback. We heavily revised the manuscript to address every remark of the reviewer. Below we will address every point of the review separately.
+We want to thank the reviewer for the feedback. We heavily revised the manuscript to address every remark of the reviewer. Below we will address every point of the review separately.
 
@@
 We did not model "time" explicitly as a variable.
-The residual plot with residuals ordered by time shows that we can assume independence of observations over time within the model, since the residuals are distributed around 0 iid for every date.
+The residual plot with residuals ordered by time shows that we can assume independence of observations over time within the model, since the residuals are independent and identically distributed (iid) around 0 for every date.
 In reality, observations on subsequent days are not independent, but the data does not show a strong temporal trend in residuals.

### 2021-03-22 manuscript/v4_medrxiv_revision/results.Rmd
update all fixes. final manuscript for submission plos

@@
 ```{r effectsrange, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Relative causal effects of exposures"}
-knitr::include_graphics(rep('figures/f_exposure_vs_effect.eps'))
+knitr::include_graphics(rep('figures/f_exposure_vs_effect_klein.eps'))
 ```
@@
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 2)` percent.
-We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
+We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases).
 In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-20)), 1)` percent at a daily average temperature of $0^{\circ}C$ compared to a day with $20^{\circ}C$.
@@
 
-For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", where the minimal adjustment set had a lower AIC.
+For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable "nursing homes", for which the minimal adjustment set had the lowest AIC.
 We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.

### 2021-03-22 manuscript/v5_submission_revision/manuscript_tex4submission.tex
update all fixes. final manuscript for submission plos

@@
 or were modelled as a construct consisting of several variables.
-For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
+For our causal graph analysis, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
 
@@
 Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
-Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section Results) which means that we assumed that their effects on the outcome will be visible after 5 days.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 set'' \cite{henckel2019graphical} as the set of variables which is a
-valid adjustment set while having the lowest AIC.
+valid adjustment set while having the lowest Akaike information criterion (AIC).
 
@@
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent.
-We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
+We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases).
 In perspective (Fig~\ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
@@
 
-For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC.
+For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', for which the minimal adjustment set had the lowest AIC.
 We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.
@@
 
-Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions.
+Furthermore, our analyses strongly support the effectiveness of non-pharmaceutical interventions.
 To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.

### 2021-03-23 manuscript/v5_submission_revision/replyreviewer.Rmd

@@
 To find the most appropriate hyper-parameter for the ridge regression, we employed cross-validation on subsets of the 401 districts.
-We used cross-validation on the similar problem of identifying the most appropriate lag parameter, which was subsequently changed from 8 to 5 days in the manuscript.
+Due to employing ridge regression for the effect estimates, we cannot report meaningful confidence intervals/p-values.
+Furthermore, we used cross-validation for identifying the most appropriate lag parameter, which was subsequently changed from 8 to 5 days in the manuscript.
 

### 2021-03-23 manuscript/v4_medrxiv_revision/causal_model.Rmd

@@
 Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.", fig.height=5}
-knitr::include_graphics(rep('figures/f_full_covid19_dag.eps'))
+knitr::include_graphics(rep('figures/f_full_covid19_dag_klein.eps'))
 ```

### 2021-04-19 manuscript/v6_medrxiv_revision_2/data_methods.Rmd
second revision

--- 2021-03-20 manuscript/v4_medrxiv_revision/data_methods.Rmd
@@
 Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables.
-For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
+For our causal graph analysis, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).
 
@@
 
-For different official and political interventions on a day and district level we used one-hot encoded daily variables, i.e.
+For different official and political interventions on a daily basis and the district level we used one-hot encoded daily variables, i.e.
 ban of mass gatherings, school and kindergarten closures and their gradual reopening, contact restrictions, and mandatory face masks for shopping and public transport.
@@
 Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
-Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 5 days.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect.
-Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest AIC.
+Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest Akaike information criterion (AIC).
 

### 2021-04-19 manuscript/v6_medrxiv_revision_2/manuscript.tex
second revision

--- 2021-03-21 manuscript/v4_medrxiv_revision/manuscript.tex
@@
 \centering
-\includegraphics{figures/f_full_covid19_dag.eps}
+\includegraphics{figures/f_full_covid19_dag_klein.eps}
 \caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19 cases on the district level.
@@
 Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables.
-Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 5 days.
+Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.
 
@@
 However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect.
-Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest AIC.
+Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest Akaike information criterion (AIC).
 
@@
 \centering
-\includegraphics{figures/f_exposure_vs_effect.eps}
+\includegraphics{figures/f_exposure_vs_effect_klein.eps}
 \caption{\label{fig:effectsrange}Relative causal effects of exposures}
@@
 On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent.
-We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases).
+We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases).
 In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).

### 2021-04-19 manuscript/v6_medrxiv_revision_2/figures/f_exposure_vs_effect.R

@@ -1,0 +1,106 @@
+library("tidyverse")
+library("cowplot")
+
+figurepath <- "manuscript/v4_medrxiv_revision/figures/"
+
+modeldata_raw <- read_csv("data/Modeldata_raw.csv")
+
+varnames <- c("Mobility (retail and recreation)",
+              "Mobility (grocery and pharmacy)",
+              "Mobility (parks)",
+              "Mobility (transit stations)",
+              "Mobility (workplaces)",
+              "Mobility (residential)",
+              "Searches corona", "COVID-19 burden",
+              "Temperature", "Rainfall", "Humidity", "Wind",
+              # "Interventions (ban of mass gatherings)",
+              # "Interventions (school and kindergarten closures)",
+              # "Interventions (contact restrictions)",
+              # "Interventions (mandatory face masks)",
+              "Age (pop. 65 and older)", "Age (pop. younger 18)", 
+              "Foreign citizens", "Foreign citizens (refugees)",
+              "Gender", "Nursing homes",
+              "Population density", "Socio-economic status") 
+varnames <- factor(varnames, levels=varnames, ordered=TRUE)
+vardata <- tibble(varnames=varnames, means=0, effects=0, rangelow=0, rangehigh=0)
+
+ranges <- apply(modeldata_raw[, varnames], 2, quantile, probs=c(0.025, 0.975))
+
+vardata <- vardata %>%
+  mutate(means=sapply(varnames, function(v) mean(modeldata_raw %>% pull(v) )),
+         rangelow=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[1]),
+         rangehigh=sapply(varnames, function(v) range(modeldata_raw %>% pull(v) )[2]))
+vardata$myrangelow <- ranges[1, varnames] # c(-100, -25, -0, -100, -50, -10, -10, 0, 0, 0, 0, 0)
+vardata$myrangehigh <- ranges[2, varnames] # c(+50, +25, +100, +50, +0, +20, 30, 20, 100, 15, 100, 100)
+vardata[1:6, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Mobility_RedOptAdjSet.csv"))$estimates
+vardata[7, 3] <- read_csv(paste0(figurepath, "../tables/t_effects_Searches corona_RedOptAdjSet.csv"))$estimates
[…]

### 2021-04-19 manuscript/v6_medrxiv_revision_2/discussion.Rmd
add paragraphic social determinants

--- 2021-03-21 manuscript/v4_medrxiv_revision/discussion.Rmd
@@
 Finally, we found a positive effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
+
+A recent review on COVID-19 based on evidence from the US and UK concludes that low socioeconomic status groups are being hit harder by the pandemic [@Wachtler2020Socioeconomic].
+Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions.
+For Germany, a higher case fatality of COVID-19 cases in districts with higher socioeconomic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@rki2021sozialeunterschiede].
 

### 2021-04-19 manuscript/v6_medrxiv_revision_2/discussion.Rmd
update discussion social determinants

--- 2021-03-21 manuscript/v4_medrxiv_revision/discussion.Rmd
@@
 Finally, we found a positive effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].
+
+A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic [@Wachtler2020Socioeconomic].
+Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions.
+For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@rki2021sozialeunterschiede].
+Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.
 
@@
 
-Furthermore, our analyses support the effectiveness of non-pharmaceutical interventions.
-To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
+Furthermore, our analyses strongly support the effectiveness of non-pharmaceutical interventions.
+To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks and more severe outcomes of infection cases.
 

### 2021-04-19 manuscript/v7_submission_revision_2/replyreviewer.Rmd
update reply reviewer (rebuttal)

---
title: "Reply to Reviewer #2 / Journal Requirements"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We want to thank the reviewer for the feedback. We revised the manuscript to address the remark of the reviewer regarding social determinants of COVID-19 infections:

**I find the topic of the paper very important and the methodology very interesting. The author has used a number of variables to analyse possible determinants of COVID-19 infections. While the paper acknowledges that there are social factors which affect the spread of the COVID-19 this is not properly discussed in the paper. I strongly suggest the author to refer to the literature on Social Determinants of Health when discussing social variables. For example see the recent paper by Galanis and Hanieh in Social Science and Medicine on Incorporating Social Determinants of Health into Modelling of COVID-19 and references therein.**

We extended the Discussion section of our manuscript to address this rightful remark. More specifically, we included references to reflect on the situation of social determinants on COVID-19 infections in Germany, most prominently the socio-economic status of households.

Furthermore we reply to the Journal Requirements regarding references within the manuscript:

**Please review your reference list to ensure that it is complete and correct. If you have cited papers that have been retracted, please include the rationale for doing so in the manuscript text, or remove these references and replace them with relevant current references. Any changes to the reference list should be mentioned in the rebuttal letter that accompanies your revised manuscript. If you need to cite a retracted article, indicate the article’s retracted status in the References list and also include a citation and full reference for the retraction notice.**

We checked all references, as of 19 April 2021 none were retracted since the first version of the manuscript. We updated some references that moved from a pre-print status to being published after peer-review. Furthermore, we cleaned up the reference list from faulty ISSN.

### 2021-04-26 manuscript/v7_submission_revision_2/replyreviewer.Rmd
updates reply reviwer and social determinants

@@
 We extended the Discussion section of our manuscript to address this rightful remark.
-More specifically, we included references to reflect on the situation of social determinants on COVID-19 infections in Germany, most prominently the socio-economic status of households.
+More specifically, we included 3 additional references to reflect on the situation of social determinants on COVID-19 infections especially in Germany, most prominently the socio-economic status of households.
 

### 2021-04-26 manuscript/v6_medrxiv_revision_2/discussion.Rmd
updates reply reviwer and social determinants

@@
 Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions.
-For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@rki2021sozialeunterschiede].
+For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic [@DAE218459].
 Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.
@@
 
+Social determinants of health are important factors to consider in an epidemiological framework of a pandemic disease like COVID-19.
+To account for this problem, we included several socio-economic confounders that were available on a district level in Germany.
+While our analysis is not an exhaustive analysis of the effects of social determinants on COVID-19 infections, we emphasize the necessity of their inclusion and our results add to the growing body of evidence that these factors interact with each other and cluster especially among people or within areas of underprivileged conditions, with detrimental effects on population health [@solar2010conceptual].
+
 While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.

### 2021-04-26 manuscript/v6_medrxiv_revision_2/manuscript.tex
updates reply reviwer and social determinants

@@
 
+A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic \citep{Wachtler2020Socioeconomic}.
+Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions.
+For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic \citep{DAE218459}.
+Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.
+
 \hypertarget{limitations-and-strengths}{%
@@
 
+Social determinants of health are important factors to consider in an epidemiological framework of a pandemic disease like COVID-19.
+To account for this problem, we included several socio-economic confounders that were available on a district level in Germany.
+While our analysis is not an exhaustive analysis of the effects of social determinants on COVID-19 infections, we emphasize the necessity of their inclusion and our results add to the growing body of evidence that these factors interact with each other and cluster especially among people or within areas of underprivileged conditions, with detrimental effects on population health \citep{solar2010conceptual}.
+
 While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.
@@
 Furthermore, our analyses strongly support the effectiveness of non-pharmaceutical interventions.
-To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks.
+To a lesser extent, the adverse effects of some socio-demographic factors might help to identify areas that are at higher risk of local COVID-19 outbreaks and more severe outcomes of infection cases.
 

### 2021-04-29 manuscript/v6_medrxiv_revision_2/manuscript.Rmd
fix references and update diff tex/diff pdf

--- 2021-03-20 manuscript/v4_medrxiv_revision/manuscript.Rmd
@@
 # latexdiff("../v2_medrxiv/manuscript.tex", "manuscript.tex", "latexdiffrRmd")
-# system("latexdiff ../v2_medrxiv/manuscript.tex manuscript.tex > diffRmd.tex")
+# system("latexdiff ../v5_submission_revision/manuscript_tex4submission.bbl manuscript_tex4submission.bbl > diff.bbl")
+# system("latexdiff ../v5_submission_revision/manuscript_tex4submission.tex manuscript_tex4submission.tex > diff.tex")
 # tinytex::pdflatex("diffRmd.tex")

### 2021-05-17 manuscript/v8_submission_finalization/assignment_files/manuscript_tex4submission.tex
assignment files

--- 2021-03-22 manuscript/v5_submission_revision/manuscript_tex4submission.tex
@@
 For all exposures, our analysis pipeline opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', for which the minimal adjustment set had the lowest AIC.
+For an overview of all final adjustments sets, see Table~\ref{tab:adjsets}.
 We found that there were no valid adjustment sets for the non-identifiable variables turnout and right-wing populist party votes.
@@
 
+A recent review on COVID-19 based on evidence from the US and UK concludes that low socio-economic status groups are being hit harder by the pandemic \cite{Wachtler2020Socioeconomic}.
+Albeit specific pathways remain unclear, many studies found associations with poverty or its correlates such as poor and potentially overcrowded housing conditions.
+For Germany, a higher case fatality of COVID-19 cases in districts with higher socio-economic deprivation has also been reported just recently, which was especially pronounced in the second wave of the pandemic \cite{DAE218459}.
+Similarly, our analysis identified a decreasing effect on COVID-19 case numbers within districts with a higher socio-economic status during the first wave.
+
 \subsection*{Limitations and strengths}\label{limitations-and-strengths}
@@
 
+Social determinants of health are important factors to consider in an epidemiological framework of a pandemic disease like COVID-19.
+To account for this problem, we included several socio-economic confounders that were available on a district level in Germany.
+While our analysis is not an exhaustive analysis of the effects of social determinants on COVID-19 infections, we emphasize the necessity of their inclusion and our results add to the growing body of evidence that these factors interact with each other and cluster especially among people or within areas of underprivileged conditions, with detrimental effects on population health \cite{solar2010conceptual}.
+
 While our analysis focused on Germany and its districts, we assume that results may be transferred to other countries by adjusting for their respective weather conditions, mobility habits, socio-demographic characteristics, and other determining factors.

### 2021-05-17 manuscript/v8_submission_finalization/assignment_files/manuscript_tex4submission.tex
final adjustments for manuscript

@@
 Edgar Steiger\textsuperscript{1*},
-Tobias Mußgnug\textsuperscript{1},
+Tobias Mussgnug\textsuperscript{1},
 Lars Eric Kroll\textsuperscript{1}%,
@@
 We integrated these findings where possible into the construction of our
-DAG, which can be seen in Fig~\ref{fig:dag-covid-19}.
+DAG, which can be seen in Fig~\ref{fig1}.
 
@@
 Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
-\label{fig:dag-covid-19}
+\label{fig1}
 \end{figure}
@@
 outcome (\(K\cdot T=56{\ }140\) observations). The set of possible
-predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig:dag-covid-19}). Due to
+predictors was derived from our causal DAG (see Table~\ref{tab:model-vars} and Fig~\ref{fig1}). Due to
 modelling and data limitations, some of the predictors were unobserved
@@
 
-We analyzed the DAG from Fig~\ref{fig:dag-covid-19} with the R Software
+We analyzed the DAG from Fig~\ref{fig1} with the R Software
 \cite{rsoftware} and the R packages \texttt{dagitty} (formal
@@
 
-In the observational period, the number of daily reported COVID-19 cases increased till the end of March/beginning of April and continually decreased afterwards till the beginning of June 2020 with a slight increase and decrease afterwards (Fig~\ref{fig:cases-log}A).
-On the other hand, the (log-)ratio of reported cases over active cases decreased steeply till the mid of April and increased steadily afterwards with a slight decrease close to the end of the observation period (Fig~\ref{fig:cases-log}B).
+In the observational period, the number of daily reported COVID-19 cases increased till the end of March/beginning of April and continually decreased afterwards till the beginning of June 2020 with a slight increase and decrease afterwards (Fig~\ref{fig2}A).
+On the other hand, the (log-)ratio of reported cases over active cases decreased steeply till the mid of April and increased steadily afterwards with a slight decrease close to the end of the observation period (Fig~\ref{fig2}B).
 Both figures examplify a considerable variation among the districts (light blue points are individual district's data).
@@
 \caption{{\bf Temporal and district level variation of outcome (log-scale).}}
-\label{fig:cases-log}
+\label{fig2}
 \end{figure}
 
-In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Fig~\ref{fig:timeseries}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,B}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,D}=\) 0.8.
+In Germany, we observed a rebound in mobility after the initial political measures, reductions in incident cases were associated with a diminishing public interest in COVID-19, and temperatures were overall increasing (cf.~Fig~\ref{fig3}); with correlations between temporal progression and mobility in retail and recreation \(r_{A,B}=\) 0.02, awareness (``Searches corona'') \(r_{A,C}=\) -0.3, and temperature \(r_{A,D}=\) 0.8.
 
@@
 determinants.}}
-\label{fig:timeseries}
+\label{fig3}
 \end{figure}
@@
 The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate.
-To put these estimates into perspective, Fig~\ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).
+To put these estimates into perspective, Fig~\ref{fig4} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).
 
@@
 \caption{{\bf Relative causal effects of exposures.}}
-\label{fig:effectsrange}
+\label{fig4}
 \end{figure}
@@
 An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.23 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.03 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.97 percent.
-Fig~\ref{fig:effectsrange} shows the effects of mobility on a range of possible values.
+Fig~\ref{fig4} shows the effects of mobility on a range of possible values.
 Thus, we expect an increase of daily cases by approximately 7.8 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period.
@@
 \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.2 percent decrease in newly reported case numbers.
-The corresponding plot in Fig~\ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.9 percent.
+The corresponding plot in Fig~\ref{fig4} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 3.9 percent.
 
@@
 We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases).
-In perspective (Fig~\ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
+In perspective (Fig~\ref{fig4}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\).
 For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.
@@
 
-The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository\footnote{The repository is located here: https://github.com/zidatalab/causalcovid19} of the article.
+The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository of the article (\url{https://github.com/zidatalab/causalcovid19}).
 
@@
 
-\bibliography{mybibfile}
+%\bibliography{mybibfile}
+
+\begin{thebibliography}{10}
+
+\bibitem{who2020report}
+{WHO Team}. Report of the {WHO-China} Joint Mission on Coronavirus Disease 2019
+  ({COVID-19}); 2020, accessed 2020-06-25.
+\newblock Available from:
+  \url{https://www.who.int/publications-detail/report-of-the-who-china-joint-mission-on-coronavirus-disease-2019-(covid-19)}.
+
+\bibitem{Chinazzi395}
+Chinazzi M, Davis JT, Ajelli M, Gioannini C, Litvinova M, Merler S, et~al.
+\newblock The effect of travel restrictions on the spread of the 2019 novel
+  coronavirus ({COVID-19}) outbreak.
+\newblock Science. 2020;368(6489):395--400.
+\newblock doi:{10.1126/science.aba9757}.
+
+\bibitem{guan2020clinical}
+Guan W, Ni Z, Hu Y, Liang W, Ou C, He J, et~al.
+\newblock Clinical Characteristics of Coronavirus Disease 2019 in {China}.
+\newblock New England Journal of Medicine. 2020;doi:{10.1056/NEJMoa2002032}.
+
+\bibitem{higgins_correlations_2020}
+Higgins TS, Wu AW, Sharma D, Illing EA, Rubel K, Ting JY.
+\newblock Correlations of {Online} {Search} {Engine} {Trends} {With}
+  {Coronavirus} {Disease} ({COVID}-19) {Incidence}: {Infodemiology} {Study}.
+\newblock JMIR Public Health and Surveillance. 2020;6(2):e19702.
+\newblock doi:{10.2196/19702}.
+
+\bibitem{li_retrospective_2020}
+Li C, Chen LJ, Chen X, Zhang M, Pang CP, Chen H.
+\newblock Retrospective analysis of the possibility of predicting the
+  {COVID}-19 outbreak from {Internet} searches and social media data, {China},
+  2020.
+\newblock Euro Surveillance. 2020;25(10).
+\newblock doi:{10.2807/1560-7917.ES.2020.25.10.2000199}.
+
+\bibitem{yuan_trends_2020}
+Yuan X, Xu J, Hussain S, Wang H, Gao N, Zhang L.
+\newblock Trends and {Prediction} in {Daily} {New} {Cases} and {Deaths} of
+  {COVID}-19 in the {United} {States}: {An} {Internet} {Search}-{Interest}
+  {Based} {Model}.
+\newblock Exploratory research and hypothesis in medicine. 2020;5(2):1--6.
+\newblock doi:{10.14218/ERHM.2020.00023}.
+
+\bibitem{bannister-tyrrell_preliminary_2020}
+Bannister-Tyrrell M, Meyer A, Faverjon C, Cameron A.
+\newblock Preliminary evidence that higher temperatures are associated with
+  lower incidence of {COVID}-19, for cases reported globally up to 29th
+  {February} 2020.
+\newblock medRxiv. 2020;doi:{10.1101/2020.03.18.20036731}.
+
+\bibitem{demongeot_temperature_2020}
+Demongeot J, Flet-Berliac Y, Seligmann H.
+\newblock Temperature {Decreases} {Spread} {Parameters} of the {New} {Covid}-19
+  {Case} {Dynamics}.
+\newblock Biology. 2020;9(5).
+\newblock doi:{10.3390/biology9050094}.
+
+\bibitem{liu_impact_2020}
+Liu J, Zhou J, Yao J, Zhang X, Li L, Xu X, et~al.
+\newblock Impact of meteorological factors on the {COVID}-19 transmission: {A}
+  multi-city study in {China}.
+\newblock Science of the Total Environment. 2020;726:138513.
+\newblock doi:{10.1016/j.scitotenv.2020.138513}.
+
+\bibitem{Greenland1999}
+Greenland S, Robins JM, Pearl J.
+\newblock Confounding and {Collapsibility} in {Causal} {Inference}.
+\newblock Statistical Science. 1999;14(1):29--46.
+
+\bibitem{schipf_directed_2011}
+Schipf S, Knüppel S, Hardt J, Stang A.
+\newblock Directed {Acyclic} {Graphs} ({DAGs}) – {Die} {Anwendung} kausaler
+  {Graphen} in der {Epidemiologie}.
+\newblock Gesundheitswesen. 2011;73(12):888--892.
+\newblock doi:{10.1055/s-0031-1291192}.
+
+\bibitem{textor_robust_2017}
+Textor J, van~der Zander B, Gilthorpe MS, Liśkiewicz M, Ellison GT.
+\newblock Robust causal inference using directed acyclic graphs: the {R}
+  package ‘dagitty’.
+\newblock International Journal of Epidemiology. 2017;45(6):1887--1894.
+\newblock doi:{10.1093/ije/dyw341}.
+
+\bibitem{jhucovid19db2020}
+{Center for Systems Science and Engineering (CSSE)}. {COVID-19 Data Repository
+  by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins
+  University}; 2020.
+\newblock Available from: \url{https://github.com/CSSEGISandData/COVID-19}.
+
+\bibitem{Pearl_2014}
+Pearl J, Bareinboim E.
+\newblock External Validity: From Do-Calculus to Transportability Across
+  Populations.
+\newblock Statistical Science. 2014;29(4):579–595.
+\newblock doi:{10.1214/14-sts486}.
+
+\bibitem{chang_modeling_2020}
+Chang MC, Kahn R, Li YA, Lee CS, Buckee CO, Chang HH.
+\newblock Variation in human mobility and its impact on the risk of future
+  {COVID}-19 outbreaks in {Taiwan}.
+\newblock BMC Public Health. 2021;21(1):226.
+\newblock doi:{10.1186/s12889-021-10260-7}.
+
+\bibitem{fowler_effect_2020}
+Fowler JH, Hill SJ, Obradovich N, Levin R.
+\newblock {The Effect of Stay-at-Home Orders on COVID-19 Cases and Fatalities
+  in the United States}.
+\newblock medRxiv. 2020;doi:{10.1101/2020.04.13.20063628}.
+
+\bibitem{kraemer_effect_2020}
+Kraemer MUG, Yang CH, Gutierrez B, Wu CH, Klein B, Pigott DM, et~al.
+\newblock The effect of human mobility and control measures on the {COVID}-19
+  epidemic in {China}.
+\newblock Science (New York, NY). 2020;368(6490):493--497.
+\newblock doi:{10.1126/science.abb4218}.
+
+\bibitem{lasry_timing_2020}
+Lasry A, Kidder D, Hast M, Poovey J, Sunshine G, Winglee K, et~al.
+\newblock Timing of Community Mitigation and Changes in Reported {COVID-19} and
+  Community Mobility - Four {U.S.} Metropolitan Areas, {February 26-April 1,
+  2020}.
+\newblock MMWR Morbidity and mortality weekly report. 2020;69(15):451—457.
+\newblock doi:{10.15585/mmwr.mm6915e2}.
+
+\bibitem{linka_outbreak_2020}
+Linka K, Peirlinck M, Sahli~Costabal F, Kuhl E.
+\newblock Outbreak dynamics of {COVID}-19 in {Europe} and the effect of travel
+  restrictions.
+\newblock Computer Methods in Biomechanics and Biomedical Engineering. 2020; p.
+  1--8.
+\newblock doi:{10.1080/10255842.2020.1759560}.
+
+\bibitem{mazzoli_effects_2020}
+Mazzoli M, Mateo D, Hernando A, Meloni S, Ramasco JJ.
+\newblock Effects of mobility and multi-seeding on the propagation of the
+  {COVID}-19 in {Spain}.
+\newblock medRxiv. 2020;doi:{10.1101/2020.05.09.20096339}.
+
+\bibitem{xiong_data-driven_2020}
+{Xiong} C, {Hu} S, {Yang} M, {Younes} HN, {Luo} W, {Ghader} S, et~al.
+\newblock {Data-Driven Modeling Reveals the Impact of Stay-at-Home Orders on
+  Human Mobility during the COVID-19 Pandemic in the U.S}.
+\newblock arXiv e-prints. 2020; p. arXiv:2005.00667.
+
+\bibitem{juni_impact_2020}
+Jüni P, Rothenbühler M, Bobos P, Thorpe KE, da~Costa BR, Fisman DN, et~al.
+\newblock Impact of climate and public health interventions on the {COVID}-19
+  pandemic: {A} prospective cohort study.
+\newblock Canadian Medical Association Journal. 2020;doi:{10.1503/cmaj.200920}.
+
+\bibitem{lai_effect_2020}
+Lai S, Ruktanonchai NW, Zhou L, Prosper O, Luo W, Floyd JR, et~al.
+\newblock Effect of non-pharmaceutical interventions to contain {COVID-19} in
+  {China}.
+\newblock Nature. 2020;doi:{10.1038/s41586-020-2293-x}.
+
+\bibitem{google_trends}
+{Google LLC}. {Google Trends, search term {``}corona{''}}; 2020, accessed
+  2020-06-25.
+\newblock Available from: \url{https://www.google.com/trends}.
+
+\bibitem{ayyoubzadeh_predicting_2020}
+Ayyoubzadeh SM, Ayyoubzadeh SM, Zahedi H, Ahmadi M, R~Niakan~Kalhori S.
+\newblock Predicting {COVID}-19 {Incidence} {Through} {Analysis} of {Google}
+  {Trends} {Data} in {Iran}: {Data} {Mining} and {Deep} {Learning} {Pilot}
+  {Study}.
+\newblock JMIR Public Health and Surveillance. 2020;6(2):e18828.
+\newblock doi:{10.2196/18828}.
+
+\bibitem{effenberger_association_2020}
+Effenberger M, Kronbichler A, Shin JI, Mayer G, Tilg H, Perco P.
+\newblock Association of the {COVID}-19 pandemic with {Internet} {Search}
+  {Volumes}: {A} {Google} {Trends}({TM}) {Analysis}.
+\newblock International Journal of Infectious Diseases. 2020;95:192--197.
+\newblock doi:{10.1016/j.ijid.2020.04.033}.
+
+\bibitem{lin_google_2020}
+Lin YH, Liu CH, Chiu YC.
+\newblock Google searches for the keywords of ``wash hands'' predict the speed
+  of national spread of {COVID}-19 outbreak among 21 countries.
+\newblock Brain, Behavior, and Immunity. 2020;doi:{10.1016/j.bbi.2020.04.020}.
+
+\bibitem{mavragani_tracking_2020}
+Mavragani A.
+\newblock Tracking {COVID}-19 in {Europe}: {Infodemiology} {Approach}.
+\newblock JMIR Public Health and Surveillance. 2020;6(2):e18941.
+\newblock doi:{10.2196/18941}.
+
+\bibitem{walker_use_2020}
+Walker A, Hopkins C, Surda P.
+\newblock Use of {Google Trends} to investigate loss-of-smell-related searches
+  during the {COVID-19} outbreak.
+\newblock International Forum of Allergy \& Rhinology. 2020;10(7):839--847.
+\newblock doi:{10.1002/alr.22580}.
+
+\bibitem{zhou_effects_2020}
+Zhou WK, Wang AL, Xia F, Xiao YN, Tang SY.
+\newblock Effects of media reporting on mitigating spread of {COVID}-19 in the
+  early phase of the outbreak.
+\newblock Mathematical Biosciences and Engineering. 2020;17(3):2693--2707.
+\newblock doi:{10.3934/mbe.2020147}.
+
+\bibitem{qi_covid-19_2020}
+Qi H, Xiao S, Shi R, Ward MP, Chen Y, Tu W, et~al.
+\newblock {COVID}-19 transmission in {Mainland} {China} is associated with
+  temperature and humidity: {A} time-series analysis.
+\newblock Science of the Total Environment. 2020;728:138778.
+\newblock doi:{10.1016/j.scitotenv.2020.138778}.
+
+\bibitem{shi_impact_2020}
+Shi P, Dong Y, Yan H, Zhao C, Li X, Liu W, et~al.
+\newblock Impact of temperature on the dynamics of the {COVID}-19 outbreak in
+  {China}.
+\newblock Science of the Total Environment. 2020;728:138890.
+\newblock doi:{10.1016/j.scitotenv.2020.138890}.
+
+\bibitem{sobral_association_2020}
+Sobral MFF, Duarte GB, da~Penha~Sobral AIG, Marinho MLM, de~Souza~Melo A.
+\newblock Association between climate variables and global transmission {of}
+  {SARS}-{CoV}-2.
+\newblock Science of the Total Environment. 2020;729:138997.
+\newblock doi:{10.1016/j.scitotenv.2020.138997}.
+
+\bibitem{tosepu_correlation_2020}
+Tosepu R, Gunawan J, Effendy DS, Ahmad LOAI, Lestari H, Bahar H, et~al.
+\newblock Correlation between weather and {Covid}-19 pandemic in {Jakarta},
+  {Indonesia}.
+\newblock Science of the Total Environment. 2020;725:138436.
+\newblock doi:{10.1016/j.scitotenv.2020.138436}.
+
+\bibitem{Wang2020temperature}
+Wang M, Jiang A, Gong L, Luo L, Guo W, Li C, et~al.
+\newblock Temperature significant change {COVID-19} Transmission in 429 cities.
+\newblock medRxiv. 2020;doi:{10.1101/2020.02.22.20025791}.
+
+\bibitem{wu_effects_2020}
+Wu Y, Jing W, Liu J, Ma Q, Yuan J, Wang Y, et~al.
+\newblock Effects of temperature and humidity on the daily new cases and new
+  deaths of {COVID-19} in 166 countries.
+\newblock Science of the Total Environment. 2020;729:139051.
+\newblock doi:{https://doi.org/10.1016/j.scitotenv.2020.139051}.
+
+\bibitem{auler_evidence_2020}
+Auler AC, Cássaro FAM, da~Silva VO, Pires LF.
+\newblock Evidence that high temperatures and intermediate relative humidity
+  might favor the spread of {COVID}-19 in tropical climate: {A} case study for
+  the most affected {Brazilian} cities.
+\newblock Science of the Total Environment. 2020;729:139090.
+\newblock doi:{10.1016/j.scitotenv.2020.139090}.
+
+\bibitem{xie_association_2020}
+Xie J, Zhu Y.
+\newblock Association between ambient temperature and {COVID}-19 infection in
+  122 cities from {China}.
+\newblock Science of the Total Environment. 2020;724:138201.
+\newblock doi:{10.1016/j.scitotenv.2020.138201}.
+
+\bibitem{briz-redon_spatio-temporal_2020}
+Briz-Red{\'o}n {\'A}, Serrano-Aroca {\'A}.
+\newblock A spatio-temporal analysis for exploring the effect of temperature on
+  {COVID}-19 early evolution in {Spain}.
+\newblock Science of the Total Environment. 2020;728:138811.
+\newblock doi:{10.1016/j.scitotenv.2020.138811}.
+
+\bibitem{iqbal_nexus_2020}
+Iqbal N, Fareed Z, Shahzad F, He X, Shahzad U, Lina M.
+\newblock The nexus between {COVID}-19, temperature and exchange rate in
+  {Wuhan} city: {New} findings from partial and multiple wavelet coherence.
+\newblock Science of the Total Environment. 2020;729:138916.
+\newblock doi:{10.1016/j.scitotenv.2020.138916}.
+
+\bibitem{jahangiri_sensitivity_2020}
+Jahangiri M, Jahangiri M, Najafgholipour M.
+\newblock The sensitivity and specificity analyses of ambient temperature and
+  population size on the transmission rate of the novel coronavirus
+  ({COVID}-19) in different provinces of {Iran}.
+\newblock Science of the Total Environment. 2020;728:138872.
+\newblock doi:{10.1016/j.scitotenv.2020.138872}.
+
+\bibitem{yao_no_2020}
+Yao Y, Pan J, Liu Z, Meng X, Wang W, Kan H, et~al.
+\newblock No association of {COVID}-19 transmission with temperature or {UV}
+  radiation in {Chinese} cities.
+\newblock The European Respiratory Journal. 2020;55(5).
+\newblock doi:{10.1183/13993003.00517-2020}.
+
+\bibitem{de2020risk}
+de~Lusignan S, Dorward J, Correa A, Jones N, Akinyemi O, Amirthalingam G,
+  et~al.
+\newblock Risk factors for {SARS}-{CoV}-2 among patients in the {Oxford}
+  {Royal} {College} of {General} {Practitioners} {Research} and {Surveillance}
+  {Centre} primary care network: a cross-sectional study.
+\newblock The Lancet Infectious Diseases;doi:{10.1016/S1473-3099(20)30371-6}.
+
+\bibitem{Dragano2020.06.17.20133918}
+Wahrendorf M, Rupprecht CJ, Dortmann O, Scheider M, Dragano N.
+\newblock Erhöhtes {Risiko} eines {COVID}-19-bedingten
+  {Krankenhausaufenthaltes} für {Arbeitslose}: {Eine} {Analyse} von
+  {Krankenkassendaten} von 1,28 {Mio}. {Versicherten} in {Deutschland}.
+\newblock Bundesgesundheitsblatt - Gesundheitsforschung - Gesundheitsschutz.
+  2021;64(3):314--321.
+\newblock doi:{10.1007/s00103-021-03280-6}.
+
+\bibitem{dohle_wingen_schreiber_2020}
+Dohle S, Wingen T, Schreiber M.
+\newblock Acceptance and Adoption of Protective Measures During the {COVID}-19
+  Pandemic: The Role of Trust in Politics and Trust in Science.
+\newblock Social Psychological Bulletin. 2020;15(4):1--23.
+\newblock doi:{10.32872/spb.4315}.
+
+\bibitem{engle_staying_2020}
+Engle S, Stromme J, Zhou A.
+\newblock Staying at home: mobility effects of {COVID-19}.
+\newblock Available at SSRN. 2020;doi:{10.2139/ssrn.3565703}.
+
+\bibitem{cowling2020impact}
+Cowling BJ, Ali ST, Ng TWY, Tsang TK, Li JCM, Fong MW, et~al.
+\newblock Impact assessment of non-pharmaceutical interventions against
+  coronavirus disease 2019 and influenza in {Hong} {Kong}: an observational
+  study.
+\newblock The Lancet Public Health. 2020;5(5):e279--e288.
+\newblock doi:{10.1016/S2468-2667(20)30090-6}.
+
+\bibitem{casenumbers_rki}
+{Robert Koch-Institut (RKI)}. {Fallzahlen in Deutschland (COVID-19)}; 2020,
+  accessed 2020-07-12.
+\newblock Available from:
+  \url{https://www.rki.de/DE/Content/InfAZ/N/Neuartiges_Coronavirus/Fallzahlen.html}.
+
+\bibitem{google_mobility}
+{Google LLC}. Google {COVID-19} Community Mobility Reports; 2020, accessed
+  2020-06-25.
+\newblock Available from: \url{https://www.google.com/covid19/mobility/}.
+
+\bibitem{dwd_weather}
+{Deutscher Wetterdienst (DWD) Climate Data Center (CDC)}. Recent daily station
+  observations (temperature, pressure, precipitation,sunshine duration, etc.)
+  for {Germany}, quality control not completed yet, version recent; 2020,
+  accessed 2020-07-12.
+\newblock Available from:
+  \url{https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/recent/}.
+
+\bibitem{inkar}
+{Bundesinstitut für Bau-, Stadt- und Raumforschung (BBSR)}. {INKAR –
+  Indikatoren und Karten zur Raum- und Stadtentwicklung}; 2020, accessed
+  2020-06-25.
+\newblock Available from: \url{https://www.inkar.de/}.
+
+\bibitem{mitze2020face}
+Mitze T, Kosfeld R, Rode J, Wälde K.
+\newblock Face Masks Considerably Reduce {COVID-19} Cases in {Germany}.
+\newblock Proceedings of the National Academy of Sciences.
+  2020;117(51):32293--32301.
+\newblock doi:{10.1073/pnas.2015954117}.
+
+\bibitem{spirtes2000causation}
+Spirtes P, Glymour CN, Scheines R, Heckerman D.
+\newblock Causation, prediction, and search.
+\newblock MIT Press; 2000.
+
+\bibitem{pearl2009causality}
+Pearl J.
+\newblock Causality.
+\newblock Cambridge: Cambridge University Press; 2009.
+\newblock Available from:
+  \url{https://www.cambridge.org/core/books/causality/B0046844FAE10CBF274D4ACBDAEB5F5B}.
+
+\bibitem{greenland1999causal}
+Greenland S, Pearl J, Robins JM.
+\newblock Causal {Diagrams} for {Epidemiologic} {Research}.
+\newblock Epidemiology. 1999;10(1):37--48.
+
+\bibitem{henckel2019graphical}
+{Henckel} L, {Perkovi{\'c}} E, {Maathuis} MH.
+\newblock {Graphical Criteria for Efficient Total Effect Estimation via
+  Adjustment in Causal Linear Models}.
+\newblock arXiv e-prints. 2020; p. arXiv:1907.02435.
+
+\bibitem{rsoftware}
+{R Core Team}. R: A Language and Environment for Statistical Computing; 2019.
+\newblock Available from: \url{https://www.R-project.org/}.
+
+\bibitem{pcalg}
+Kalisch M, Mächler M, Colombo D, Maathuis MH, Bühlmann P.
+\newblock Causal {Inference} {Using} {Graphical} {Models} with the {R}
+  {Package} pcalg.
+\newblock Journal of Statistical Software. 2012;47(11):1--26.
+\newblock doi:{10.18637/jss.v047.i11}.
+
+\bibitem{maxlikelihood}
+Hilbe JM, Greene WH.
+\newblock 4 - {Count} {Response} {Regression} {Models}.
+\newblock In: Rao CR, Miller JP, Rao DC, editors. Essential {Statistical}
+  {Methods} for {Medical} {Statistics}. Boston: North-Holland; 2011. p.
+  104--145.
+\newblock Available from:
+  \url{http://www.sciencedirect.com/science/article/pii/B9780444537379500074}.
+
+\bibitem{mass}
+Venables WN, Ripley BD.
+\newblock Modern Applied Statistics with S.
+\newblock 4th ed. New York: Springer; 2002.
+\newblock Available from: \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\bibitem{ridge1970}
+Hoerl AE, Kennard RW.
+\newblock Ridge Regression: Biased Estimation for Nonorthogonal Problems.
+\newblock Technometrics. 1970;12(1):55--67.
+\newblock doi:{10.1080/00401706.1970.10488634}.
+
+\bibitem{glmnet2010}
+Friedman J, Hastie T, Tibshirani R.
+\newblock Regularization Paths for Generalized Linear Models via Coordinate
+  Descent.
+\newblock Journal of Statistical Software. 2010;33(1):1--22.
+
+\bibitem{perkovi2015complete}
+{Perkovi{\'c}} E, {Textor} J, {Kalisch} M, {Maathuis} MH.
+\newblock {A Complete Generalized Adjustment Criterion}.
+\newblock arXiv e-prints. 2015; p. arXiv:1507.01524.
+
+\bibitem{gencoglu2020causal}
+Gencoglu O, Gruber M.
+\newblock Causal Modeling of {Twitter} Activity During {COVID-19}.
+\newblock Computation. 2020;8(4).
+\newblock doi:{10.3390/computation8040085}.
+
+\bibitem{Wachtler2020Socioeconomic}
+Wachtler B, Michalski N, Nowossadeck E, Diercke M, Wahrendorf M,
+  Santos-Hövener C, et~al.
+\newblock Socioeconomic inequalities and COVID-19 – A review of the current
+  international literature.
+\newblock Journal of Health Monitoring. 2020;(S7):3--17.
+\newblock doi:{10.25646/7059}.
+
+\bibitem{DAE218459}
+Hoebel J, Michalski N, Wachtler B, Diercke M, Neuhauser H, Wieler LH, et~al.
+\newblock {Sozioökonomische Unterschiede im Infektionsrisiko während der
+  zweiten SARS-CoV-2-Welle in Deutschland}.
+\newblock Dtsch Arztebl International. 2021;118(15):269--270.
+\newblock doi:{10.3238/arztebl.m2021.0188}.
+
+\bibitem{perkovic2017complete}
+Perkovi{\'c} E, Textor J, Kalisch M, Maathuis MH.
+\newblock Complete graphical characterization and construction of adjustment
+  sets in {Markov} equivalence classes of ancestral graphs.
+\newblock The Journal of Machine Learning Research. 2017;18(1):8132--8193.
+
+\bibitem{Islamm2743}
+Islam N, Sharp SJ, Chowell G, Shabnam S, Kawachi I, Lacey B, et~al.
+\newblock Physical distancing interventions and incidence of coronavirus
+  disease 2019: natural experiment in 149 countries.
+\newblock BMJ. 2020;370.
+\newblock doi:{10.1136/bmj.m2743}.
+
+\bibitem{kermack1991contributions}
+Kermack WO, McKendrick AG.
+\newblock Contributions to the mathematical theory of epidemics--I. 1927.
+\newblock Bulletin of mathematical biology. 1991;53(1-2):33—55.
+\newblock doi:{10.1007/bf02464423}.
+
+\bibitem{rkimodellierung}
+{an der Heiden} M, Buchholz U.
+\newblock {Modellierung von Beispielszenarien der SARS-CoV-2-Epidemie 2020 in
+  Deutschland}. 2020;doi:{10.25646/6571.2}.
+
+\bibitem{schisterman2017collinearitycausal}
+Schisterman EF, Perkins NJ, Mumford SL, Ahrens KA, Mitchell EM.
+\newblock Collinearity and Causal Diagrams: A Lesson on the Importance of Model
+  Specification.
+\newblock Epidemiology. 2017;28(1).
+
+\bibitem{Openshaw1984}
+Openshaw S.
+\newblock Ecological {Fallacies} and the {Analysis} of {Areal} {Census} {Data}.
+\newblock Environment and Planning A: Economy and Space. 1984;16(1):17--31.
+\newblock doi:{10.1068/a160017}.
+
+\bibitem{solar2010conceptual}
+Solar O, Irwin A.
+\newblock A conceptual framework for action on the social determinants of
+  health.
+\newblock WHO Document Production Services; 2010.
+\newblock Available from: \url{https://drum.lib.umd.edu/handle/1903/23135}.
+
+\bibitem{world2020transmission}
+{World Health Organization}, et~al.
+\newblock {Transmission of SARS-CoV-2: implications for infection prevention
+  precautions: Scientific Brief, 09 July 2020}.
+\newblock World Health Organization; 2020.
+
+\end{thebibliography}
 

### 2021-05-31 manuscript/v6_medrxiv_revision_2/data_methods.Rmd
fix/update medrxiv tex and pdf

@@
 
-We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, [@casenumbers_rki], acquired on 12 July 2020) and aggregated them by date.
+We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, [@casenumbers_rki] and aggregated them by date.
 The number of daily active cases for day $d$ was derived by subtracting the total number of reported cases on day $d$ and day $d-14$ (14 days as a conservative estimate for the infectious period, which corresponds here to the required quarantine time in Germany).
@@
 
-We analyzed the DAG from Section \@ref(causal-model) with the R Software [@rsoftware] and the R packages `dagitty` (formal representation of the graph and minimal adjustment sets [@textor_robust_2017]) and `pcalg` (for finding an optimal adjustment set [@pcalg]).
+We analyzed the DAG from Figure \@ref(fig:dag-covid-19)) with the R Software [@rsoftware] and the R packages `dagitty` (formal representation of the graph and minimal adjustment sets [@textor_robust_2017]) and `pcalg` (for finding an optimal adjustment set [@pcalg]).
 For the defined exposures and the outcome "Reported new cases of COVID-19", we computed the minimal and optimal adjustment sets.

### 2021-05-31 manuscript/v6_medrxiv_revision_2/discussion.Rmd
fix/update medrxiv tex and pdf

@@
 
-The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository^[The repository is located here: https://github.com/zidatalab/causalcovid19] of the article.
+The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository (https://github.com/zidatalab/causalcovid19]) of the article.
 

### 2021-05-31 manuscript/v6_medrxiv_revision_2/manuscript.Rmd
fix/update medrxiv tex and pdf

@@
     affiliation: Zi
-  - name: Tobias Mußgnug
+  - name: Tobias Mussgnug
     email: tmussgnug@zi.de

### 2021-05-31 manuscript/v6_medrxiv_revision_2/manuscript.tex
fix/update medrxiv tex and pdf

@@
    \ead{esteiger@zi.de} 
-    \author[Zi]{Tobias Mußgnug}
+    \author[Zi]{Tobias Mussgnug}
    \ead{tmussgnug@zi.de} 
@@
 
-We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, \citep{casenumbers_rki}, acquired on 12 July 2020) and aggregated them by date.
+We downloaded German daily case numbers on district level reported by Robert Koch Institute (RKI, \citep{casenumbers_rki} and aggregated them by date.
 The number of daily active cases for day \(d\) was derived by subtracting the total number of reported cases on day \(d\) and day \(d-14\) (14 days as a conservative estimate for the infectious period, which corresponds here to the required quarantine time in Germany).
@@
 
-We analyzed the DAG from Section \ref{causal-model} with the R Software \citep{rsoftware} and the R packages \texttt{dagitty} (formal representation of the graph and minimal adjustment sets \citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal adjustment set \citep{pcalg}).
+We analyzed the DAG from Figure \ref{fig:dag-covid-19}) with the R Software \citep{rsoftware} and the R packages \texttt{dagitty} (formal representation of the graph and minimal adjustment sets \citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal adjustment set \citep{pcalg}).
 For the defined exposures and the outcome ``Reported new cases of COVID-19'', we computed the minimal and optimal adjustment sets.
@@
 
-The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository\footnote{The repository is located here: https://github.com/zidatalab/causalcovid19} of the article.
+The code and resources for our analysis are available on Github, we invite other researchers to replicate our analysis with different assumptions using the files provided in the repository (https://github.com/zidatalab/causalcovid19{]}) of the article.
 
